# The result {#sec-result}

```{r}
#| label: result-numbers
#| include: false

inc_win <- p6_res(p6_test_results, "S1/K0", "win")
mkt_win <- p6_res(p6_test_results, "S4/K0", "win")
inc_plc <- p6_res(p6_test_results, "S1/K0", "place")
mkt_plc <- p6_res(p6_test_results, "S4/K0", "place")
inc_ewc <- p6_res(p6_test_results, "S1/K0", "eachway_corrected")
mkt_ewc <- p6_res(p6_test_results, "S4/K0", "eachway_corrected")
inc_ew5 <- p6_res(p6_test_results, "S1/K0", "eachway")
mkt_ew5 <- p6_res(p6_test_results, "S4/K0", "eachway")

c_win <- p6_con(p6_test_contrasts, "win", "incumbent")
c_plc <- p6_con(p6_test_contrasts, "place", "incumbent")
c_ew5 <- p6_con(p6_test_contrasts, "eachway", "incumbent")
c_ewc <- p6_con(p6_test_contrasts, "eachway_corrected", "incumbent")

val_win <- p6_valinc(p6_val_incumbent, "win")
```

**Owen's rule survived the search.** Nine selection rules were scored against
it at a flat stake on `r p6_comma(n_val_races)` held-out races, and three
staking rules against a flat stake on whatever selection won. For all three
bet types, at both stages, the mechanical declaration rule returned the
incumbent: no alternative selection and no alternative stake had a higher
out-of-sample return.

That is a different answer from the one a search on in-sample predictions
gives, and the difference is the point of the design (@sec-searchset).

Because the declaration is the incumbent, the test split carries two distinct
rules rather than four: S1/K0, which is simultaneously the declared
selection, the declared stake and paper 5's published rule, and S4/K0, the
market-only control. Two of the four named contrasts are therefore
identically zero by construction, and are reported that way rather than
dropped.

```{r}
#| label: tbl-headline
#| tbl-cap: "The test split, scored once. Each-way appears under both paper 5's flat terms and the corrected industry terms; see @sec-terms."
p6_results_table(p6_test_results) |> kable(align = "llrrrrrrrl")
```

## Against a rule with no model in it

The remaining question is whether the model earns its place in the rule at
all. S4 backs the market favourite in every race at a flat stake and reads no
model output. On the `r p6_comma(c_win$n_races)` races both arms bet:

| Market | Incumbent | Market favourite | Difference | 90% interval |
|---|---|---|---|---|
| win | `r p6_fmt_pct(c_win$roi_a)` | `r p6_fmt_pct(c_win$roi_b)` | `r p6_fmt_pts(c_win$diff_point)` | `r p6_fmt_pts_ci(c_win$ci_lo, c_win$ci_hi)` |
| place | `r p6_fmt_pct(c_plc$roi_a)` | `r p6_fmt_pct(c_plc$roi_b)` | `r p6_fmt_pts(c_plc$diff_point)` | `r p6_fmt_pts_ci(c_plc$ci_lo, c_plc$ci_hi)` |
| each-way, paper 5 terms | `r p6_fmt_pct(c_ew5$roi_a)` | `r p6_fmt_pct(c_ew5$roi_b)` | `r p6_fmt_pts(c_ew5$diff_point)` | `r p6_fmt_pts_ci(c_ew5$ci_lo, c_ew5$ci_hi)` |
| each-way, corrected terms | `r p6_fmt_pct(c_ewc$roi_a)` | `r p6_fmt_pct(c_ewc$roi_b)` | `r p6_fmt_pts(c_ewc$diff_point)` | `r p6_fmt_pts_ci(c_ewc$ci_lo, c_ewc$ci_hi)` |

Three of the four contain zero. **On the win market the model rule and the
model-free rule cannot be told apart**: `r p6_fmt_pts(c_win$diff_point)`
points on an interval `r p6_fmt_pts_ci(c_win$ci_lo, c_win$ci_hi)` wide enough
to hold a large edge in either direction. The point estimate favours the
favourite; the interval will not carry it.

The place market is the exception and goes the other way. The incumbent
returns `r p6_fmt_pct(c_plc$roi_a)` against the favourite's
`r p6_fmt_pct(c_plc$roi_b)`, a difference of
`r p6_fmt_pts(c_plc$diff_point)` points on
`r p6_fmt_pts_ci(c_plc$ci_lo, c_plc$ci_hi)` — the one interval in the paper
that excludes zero. It is a smaller claim than it looks: backing favourites
to place is expensive because a place market built from the starting-price
book prices short horses hardest, so the contrast is as much about the place
market's shape as about the model (@sec-control).

## Two things that are structural, not modelling

**The whole of the incumbent's win-market loss is the over-round.** Priced at
a zero-margin book — the same starting prices, renormalised within race to
sum to one — its win ROI moves from `r p6_fmt_pct(inc_win$roi)` to
`r p6_fmt_pct(inc_win$roi_fair)`, a margin of
`r p6_fmt_pct(inc_win$roi_fair - inc_win$roi)`. The favourite rule moves from
`r p6_fmt_pct(mkt_win$roi)` to `r p6_fmt_pct(mkt_win$roi_fair)`. Both are
close to break-even or better against a fair book and lose against a real
one. Closing that gap is a question about the price, not about the rule
(@sec-breakeven).

**The validation slice could not have seen a small edge.** The incumbent
returned `r p6_fmt_pct(val_win$roi)` on the win market there, on
`r p6_comma(val_win$n_bets)` bets, with a 90% interval of
`r p6_fmt_pct_ci(val_win$ci_lo, val_win$ci_hi)`; on test it returned
`r p6_fmt_pct(inc_win$roi)`, interval
`r p6_fmt_pct_ci(inc_win$ci_lo, inc_win$ci_hi)`. Both figures are out of
sample and the two intervals overlap heavily, so the swing of about twenty
points between them is what sampling noise looks like at this scale, not a
finding. A search set that wide cannot resolve differences of a few points,
and the declaration it produced should be read as "nothing separated itself
from the incumbent", not as "the incumbent is best" (@sec-limitations).

The rest of the paper is how those numbers were arrived at.
