# The search, on the validation slice {#sec-search}

```{r}
#| label: search-numbers
#| include: false

cov <- p6_price_coverage_val
v_win <- p6_valinc(p6_val_incumbent, "win")
v_plc <- p6_valinc(p6_val_incumbent, "place")
v_ewc <- p6_valinc(p6_val_incumbent, "eachway_corrected")
s1_a_win <- p6_grid(p6_stage_a_grid, "S1", "K0", "win")
s4_a_win <- p6_grid(p6_stage_a_grid, "S4", "K0", "win")
n_struck <- p6_gate_demonstration |>
  dplyr::filter(bet == "win", !eligible) |> nrow()
n_demo <- p6_gate_demonstration |> dplyr::filter(bet == "win") |> nrow()
```

`r p6_comma(p6_search_set_check$n_rows_frame)` runner-rows over
`r p6_comma(n_val_races)` races.
`r p6_search_set_check$n_rows_dropped` rows carry no usable starting price.
Overlap with the test split: `r p6_search_set_check$n_test_overlap` races.
Overlap with the `r p6_comma(n_fit_races)` fitting races the model was
trained on: `r p6_search_set_check$n_fit_overlap`. All three are assertions
on the pipeline graph, not observations.

## Prices

```{r}
#| label: tbl-coverage
#| tbl-cap: "Price coverage on the validation slice, by year. `SP` is the industry starting price from the results archive; `forecast` is the racecard forecast price from the overnight feed, counted only where the feed row was written strictly before the meeting date."
cov |>
  dplyr::transmute(
    Year = year, Rows = p6_comma(n_rows), Races = p6_comma(n_races),
    `SP rows` = p6_fmt_pct(pct_rows_sp, 1),
    `SP races complete` = p6_fmt_pct(pct_races_sp, 1),
    `Forecast rows` = p6_fmt_pct(pct_rows_fp, 1),
    `Forecast races complete` = p6_fmt_pct(pct_races_fp, 1)
  ) |>
  kable(align = "lrrrrrr")
```

The validation slice runs from December 2010, so it sits entirely inside the
period where both price sources exist — unlike the training period as a
whole, where the racecard forecast is absent before March 2008 and patchy for
a year after, as paper 4's Appendix A records. Every rule in this paper is
priced at the starting price; the forecast column is here because it bounds
what a future paper could select on, not because anything here uses it.

```{r}
#| label: tbl-overround
#| tbl-cap: "The starting-price book by field size, validation slice. The margin is what a flat bettor pays before any skill."
p6_overround_val |>
  dplyr::filter(n_races >= 10) |>
  dplyr::transmute(
    `Field size` = field_size, Races = p6_comma(n_races),
    `Mean book` = sprintf("%.4f", mean_book),
    `Median book` = sprintf("%.4f", median_book),
    `Mean margin` = sprintf("%.2f%%", mean_margin_pct)
  ) |>
  kable(align = "rrrrr")
```

The margin rises with field size throughout. Any rule that shifts betting
towards smaller fields reduces the margin it pays without picking better,
which is why every ROI in this paper is reported at a fair book alongside the
real one.

## Each-way terms {#sec-terms}

Paper 5 settles every each-way bet at one fifth the odds on a top-three
finish, whatever the field size. The real industry terms for a handicap are a
ladder: win-only at four runners, a quarter the odds on the first two from
five to seven, a fifth on the first three from eight to eleven, a quarter on
the first three from twelve to fifteen, and a quarter on the first four from
sixteen.

`r p6_fmt_pct(p6_eachway_terms_decision$share_4_to_7_val, 1)` of validation
races and
`r p6_fmt_pct(p6_eachway_terms_decision$share_4_to_7_test, 1)` of test races
have four to seven runners, where paper 5's terms are most obviously wrong.
Correcting only that band would leave the twelve-and-up band wrong as well,
so the whole ladder is applied. Paper 5's terms differ from it on
`r p6_fmt_pct(p6_eachway_terms_decision$share_mispriced_val, 1)` of validation
races.

Every each-way number in this paper is therefore reported twice: under paper
5's flat terms, so the comparison with paper 5 is like for like, and under
the corrected terms, which is what an each-way bet actually settles on. The
comparator is recomputed under each. The place market is left at paper 5's
top-three Harville construction throughout — the same criticism applies to
it, and correcting it would mean redefining the synthetic market rather than
repricing a bet, which is a different change from the one this paper makes.

## Where Owen's thresholds sit {#sec-thresholds}

```{r}
#| label: tbl-deciles
#| tbl-cap: "Deciles of the two quantities S1 filters on, over the model's top-rated horse in each race, on both splits."
p6_positions$deciles |>
  tidyr::pivot_wider(names_from = split,
                     values_from = c(p_mod, ratio)) |>
  dplyr::transmute(
    Decile = sprintf("%.0f%%", 100 * decile),
    `P_mod, validation` = sprintf("%.4f", p_mod_validation),
    `P_mod, test` = sprintf("%.4f", p_mod_test),
    `Ratio, validation` = sprintf("%.4f", ratio_validation),
    `Ratio, test` = sprintf("%.4f", ratio_test)
  ) |>
  kable(align = "lrrrr")
```

```{r}
#| label: tbl-s1-position
#| tbl-cap: "Owen's two absolute cuts, located in each split's own distribution."
p6_positions$s1_position |>
  dplyr::transmute(
    Split = split, Quantity = quantity, Threshold = s1_threshold,
    `Quantile it sits at` = p6_fmt_pct(quantile_of_top_rated, 1),
    `Top-rated horses passing` = p6_fmt_pct(share_passing, 1)
  ) |>
  kable(align = "llrrr")
```

The probability floor of 0.15 is close to inert on both splits: it passes
`r p6_fmt_pct(p6_positions$s1_position$share_passing[1], 1)` of top-rated
horses on validation and
`r p6_fmt_pct(p6_positions$s1_position$share_passing[3], 1)` on test. The
ratio threshold does the work, passing
`r p6_fmt_pct(p6_positions$s1_position$share_passing[2], 1)` and
`r p6_fmt_pct(p6_positions$s1_position$share_passing[4], 1)`. Owen's rule is,
on this data, a ratio filter with a probability floor attached that does
almost nothing — and because that filter is absolute rather than a quantile,
it bets a visibly different fraction of the two splits. Every candidate rule
in the search is parameterised the other way for exactly that reason.

## Stage A — selection, at a flat stake

```{r}
#| label: tbl-stage-a-win
#| tbl-cap: "Stage A, win market: nine selection rules at a flat stake on the validation slice, best first."
p6_grid_table(p6_stage_a_grid, "win") |> kable(align = "lrrrrrrrll")
```

```{r}
#| label: tbl-stage-a-place
#| tbl-cap: "Stage A, place market."
p6_grid_table(p6_stage_a_grid, "place") |> kable(align = "lrrrrrrrll")
```

```{r}
#| label: tbl-stage-a-ew
#| tbl-cap: "Stage A, each-way at the corrected terms."
p6_grid_table(p6_stage_a_grid, "eachway_corrected") |>
  kable(align = "lrrrrrrrll")
```

**S1 is first on all three.** Its validation win ROI,
`r p6_fmt_pct(s1_a_win$roi)`, is the only positive figure in the win table,
and no other selection rule comes within
`r sprintf("%.0f", 100 * (s1_a_win$roi - max(p6_stage_a_grid$roi[p6_stage_a_grid$bet == "win" & p6_stage_a_grid$selection != "S1"])))`
points of it. The narrower two of the three middle-band rules — S2b and S2c,
built to test paper 4's implication that the ratio filter selects for the
model's own errors — occupy the bottom two places in all three tables; the
widest, S2a, is middling. Inverting the ratio filter made things worse, not
better, so on this evidence the high-ratio filter is picking something real.
The intervals in @sec-limitations are wide enough that this is a direction
rather than a measurement.

@tbl-stage-a-win also shows the control in its unrestricted form. S4/K0 bets
every race and returns `r p6_fmt_pct(s4_a_win$roi)` on validation, against
S1/K0's `r p6_fmt_pct(s1_a_win$roi)`. Those two are not directly comparable —
S1 bets `r p6_comma(s1_a_win$n_bets)` of the `r p6_comma(n_val_races)` races
and S4 bets all of them — which is what the paired contrast on common races
is for; @sec-test does it on the split that counts.

## Stage B — staking, on the declared selection

```{r}
#| label: tbl-stage-b
#| tbl-cap: "Stage B: the three staking rules on the selection stage A declared, validation slice."
p6_stage_b_grid |>
  dplyr::filter(bet %in% c("win", "place", "eachway_corrected")) |>
  dplyr::arrange(match(bet, c("win", "place", "eachway_corrected")),
                 staking) |>
  dplyr::transmute(
    Bet = P6_BET_LABEL[bet], Arm = paste0(selection, "/", staking),
    Bets = p6_comma(n_bets), ROI = p6_fmt_pct(roi),
    `ROI, fair book` = p6_fmt_pct(roi_fair),
    `Mean stake` = sprintf("%.2f", mean_stake),
    `SD unit return` = sprintf("%.2f", sd_unit_return),
    `Max drawdown` = sprintf("%.0f", max_drawdown),
    Eligible = dplyr::if_else(eligible, "yes", "no")
  ) |>
  kable(align = "llrrrrrrl")
```

Flat staking wins every comparison. Both alternatives lower the return on
every market, and they do it while raising the dispersion of the unit return
— K1 stakes more than one unit on average and K2 much less, but neither buys
anything back. The reading is not that variable staking is a bad idea in
general; it is that staking proportional to a disagreement is only worth
doing if the disagreement carries information, and paper 4 already found that
against the settled price it does not.

Note that under S1 the staking rules do not shed any bets: every horse
clearing a ratio above 1.3 has a positive probability edge and, given the
over-round, a positive Kelly fraction too, so all three arms bet the same
`r p6_comma(s1_a_win$n_bets)` races. That is a property of S1, not of the
staking rules, and @sec-gate shows what happens under selections that do not
have it.

## The eligibility gate {#sec-gate}

The gate struck nothing from either stage's pool. That is not because it is
inert: stage A runs at a flat stake, where every selected race is staked by
construction, and stage B runs on S1, where the ratio filter guarantees a
positive edge. Both pools are immune to the failure mode by construction.

To show the gate has teeth, it is applied to the full nine-by-three
cross-product on the validation slice. This declares nothing —
the declaration is passed the stage grids, never this one — and exists only
to make the gate's behaviour visible.

```{r}
#| label: tbl-gate-demo
#| tbl-cap: "The eligibility gate applied to all 27 combinations on the validation slice, win market. Diagnostic only: nothing here declares anything."
p6_gate_demonstration |>
  dplyr::filter(bet == "win") |>
  dplyr::arrange(dplyr::desc(roi)) |>
  dplyr::transmute(
    Arm = paste0(selection, "/", staking),
    Selected = p6_comma(n_races_selected), Bets = p6_comma(n_bets),
    `Staked share` = p6_fmt_pct(staked_share, 0),
    `Projected test bets` = sprintf("%.0f", projected_test_bets),
    ROI = p6_fmt_pct(roi),
    Eligible = dplyr::if_else(eligible, "yes", "no"),
    `Struck because` = dplyr::coalesce(struck_because, "—")
  ) |>
  kable(align = "lrrrrrll")
```

`r n_struck` of the `r n_demo` combinations fail the gate, and the top of the
table shows why it matters. S3a/K2 returns
`r p6_fmt_pct(p6_grid(p6_gate_demonstration, "S3a", "K2", "win")$roi)` on the
validation slice — the best figure in the table — from
`r p6_grid(p6_gate_demonstration, "S3a", "K2", "win")$n_bets` bets across
`r p6_comma(p6_grid(p6_gate_demonstration, "S3a", "K2", "win")$n_races_selected)`
selected races, a
`r p6_fmt_pct(p6_grid(p6_gate_demonstration, "S3a", "K2", "win")$staked_share, 0)`
staked share. An unguarded maximum takes it. The gate strikes it before the
maximum is taken, on both conditions.

## What was declared

```{r}
#| label: tbl-declaration
#| tbl-cap: "The declaration, both stages. Frozen and committed before any test target existed."
p6_declaration |>
  dplyr::transmute(
    Stage = stage, Bet = P6_BET_LABEL[bet],
    Declared = paste0(selection, "/", staking),
    `Top eligible candidate` = paste0(cand_selection, "/", cand_staking),
    `Candidate ROI` = p6_fmt_pct(cand_roi),
    `Comparator ROI` = p6_fmt_pct(comparator_roi),
    `Eligible candidates` = n_eligible,
    Note = note
  ) |>
  kable(align = "lllrrrrl")
```

In every case the top eligible candidate *is* the comparator, so steps 2 and
3 of the declaration rule never engage: there is no difference to bootstrap.
The incumbent is declared for all three bet types at both stages.
