# The result {#sec-result}

```{r}
#| label: example-setup
#| include: false

# The worked example, selected by a rule rather than by eye: the first
# runner, in race and runner order, on the validation slice whose career
# consists of exactly eight prior runs.
ex_key <- p5_sequences$key |>
  dplyr::mutate(row = dplyr::row_number(),
                seq_len = p5_sequences$seq_len,
                career = p5_sequences$career_runs_prior)

ex <- ex_key |>
  dplyr::filter(race_id %in% p5_slice$val_race_ids,
                seq_len == 8L, career == 8L) |>
  dplyr::arrange(race_id, runner_id) |>
  dplyr::slice(1) |>
  dplyr::left_join(
    dplyr::select(p5_runners_interactions, race_id, runner_id, name,
                  pos_lag1_zero, pos_lag1_nonzero, pos_lag2_zero,
                  pos_lag2_nonzero, has_wins, going_runs_prior,
                  going_sr_shrunk, going_sr_delta, days_LTO_log),
    by = c("race_id", "runner_id")
  )

ex_seq <- p5_sequences$x[ex$row, seq_len(ex$seq_len), , drop = FALSE]
dimnames(ex_seq) <- list(NULL, NULL, p5_sequences$features)

ex_tbl <- tibble::tibble(
  `Run` = paste0(seq_len(ex$seq_len), if_else(seq_len(ex$seq_len) == 1L,
                                              " (last time)", "")),
  `Finish` = as.integer(ex_seq[1, , "finish_pos"]),
  `Beaten (l)` = sprintf("%.2f", ex_seq[1, , "beaten_dist"]),
  `Field` = as.integer(ex_seq[1, , "field_size"]),
  `Class` = as.integer(ex_seq[1, , "class"]),
  `Dist (f)` = as.integer(ex_seq[1, , "distance_furlongs"]),
  `Going (ord.)` = as.integer(ex_seq[1, , "going_ordinal"]),
  `Days since previous` = as.integer(ex_seq[1, , "days_since_prev"])
)
ex_name <- ex$name
```

The models in this series score a horse on 26 features. Most of them
describe the horse and the race rather than its recent form: the official
rating, the weight carried, age, sex, headgear, the draw, the course,
today's going, the days since the horse last ran, and strike rates for the
trainer, the jockey and the sire. Eight of the 26 describe the horse's own
run history, and those eight are the only ones this paper changes.

Four of the eight are position lags: last time out's finishing position,
recorded only if the horse finished in the first four and zero otherwise,
with a companion indicator for the zero case, and the same pair for the run
before that. The other four are an ever-won flag and three going-affinity
terms.

This paper replaces those eight with the runs themselves. A gated recurrent
encoder reads up to the twenty most recent prior runs, cross-surface, with
eight values describing each run: the **raw** finishing position, how far
the horse was beaten and whether that reading exists, the going, the class,
the field size, the race distance, and the gap to the run before. The other
eighteen features stay in both models, as do the races, the likelihood and
the scorer.

Take `r ex_name`, running at
`r format(p5_frame$race_date[p5_frame$race_id == ex$race_id & p5_frame$runner_id == ex$runner_id][1], "%d %B %Y")`
with eight prior runs behind it. Its record, most recent first, is what the
encoder reads:

```{r}
#| label: tbl-example-form
#| tbl-cap: !expr glue::glue("{ex_name}'s eight prior runs, as the encoder receives them. Going is on the series' ordinal scale; the days column is the gap back to the run before that one, so the earliest run in the career has no reading and carries the unavailable code −1.")
kable(ex_tbl, align = "lrrrrrrr")
```

Paper 3 sees this record as four features:
`pos_lag1_zero` = `r ex$pos_lag1_zero`,
`pos_lag1_nonzero` = `r ex$pos_lag1_nonzero`,
`pos_lag2_zero` = `r ex$pos_lag2_zero`,
`pos_lag2_nonzero` = `r ex$pos_lag2_nonzero` — plus `has_wins` =
`r ex$has_wins` and a going block of
(`r ex$going_runs_prior`, `r sprintf("%.2f", ex$going_sr_shrunk)`,
`r sprintf("%.2f", ex$going_sr_delta)`).

Last time out the horse finished `r ex_tbl$Finish[1]`th of
`r ex_tbl$Field[1]`, beaten `r ex_tbl[["Beaten (l)"]][1]` lengths. Because
that is outside the first four, the position lag records it as zero and
raises the indicator. A horse that finished last of twenty, beaten forty
lengths, produces exactly the same pair. The run before was a fourth of
`r ex_tbl$Field[2]`, beaten `r ex_tbl[["Beaten (l)"]][2]` lengths, and is
recorded as the single figure 4 — the same value a fourth beaten a
short-head in a six-runner race would produce.

The rest of the table — the six earlier runs, the beaten distances, the
class and distance of each race, the `r ex_tbl[["Days since previous"]][3]`-day
gap before the third-last run — is not in paper 3's feature set at all.

Both models use the same MLP scorer and differ only in how the horse's run
history reaches it: the eight hand-built summary features in one, the
encoder in the other. Both were fixed before any test race was scored. On
`r p5_comma(n_test_races)` held-out test races:

```{r}
#| label: tbl-headline
#| tbl-cap: "Encoder against hand-built features, held-out test split. Paired race-level bootstrap, B = 2000, 90% intervals."
p5_test_contrasts |>
  dplyr::filter(stringr::str_detect(contrast, "rung 3b.*rung 1")) |>
  p5_contrast_table("Encoder", "Hand-built") |>
  kable(align = "lrrrrl")
```

Every interval excludes zero. The same three comparisons against paper 3's
gradient-boosted trees [@chenguestrin2016] are larger still, and the
hand-built model itself beats paper 3 by a smaller margin, so the three
models order strictly on all three measures:

$$\text{paper 3 (trees)} \;<\; \text{hand-built features (MLP)} \;<\;
\text{sequence encoder (MLP)}$$

Nine contrasts, nine intervals excluding zero. That is worth setting beside
what came before it. In papers 1 to 3 no model could be distinguished from
another on this same test split: every paired interval in paper 3 against
paper 2b contained zero, and the only contrast in that paper whose interval
excluded zero was the market beating both models. Three models now separate
cleanly.

Three things did not change. **The market still ranks better than any of
them.** The encoder reaches P1_rank
`r fmt_metric(p5_ct(p5_test_contrasts, "P1_rank", "rung 3b.*rung 1")$point_a)`
and Brier_place
`r fmt_metric(p5_ct(p5_test_contrasts, "Brier_place", "rung 3b.*rung 1")$point_a)`
against the market baseline of roughly `r sprintf("%.5f", market_p1)` and
`r sprintf("%.4f", market_brier)` that papers 2b and 3 published. The gap
narrows on both and closes on neither. The encoder and the market were not
compared directly here; these are published figures set side by side.

**No betting return improves in a way that can be distinguished from
nothing.** The single-bet-per-race backtest gives the encoder a win ROI of
`r fmt_pct(p5_bt(p5_test_backtests, "rung3b", "win")$roi)` against paper 3's
`r fmt_pct(p5_bt(p5_test_backtests, "paper3", "win")$roi)` — the best in the
series — but the difference is
`r fmt_pts(p5_roi(p5_roi_contrasts, "rung 3b.*paper 3", "win")$diff_point)`
points with a 90% interval of
`r fmt_pts_ci(p5_roi(p5_roi_contrasts, "rung 3b.*paper 3", "win")$ci_lo, p5_roi(p5_roi_contrasts, "rung 3b.*paper 3", "win")$ci_hi)`.
Eight of the nine ROI intervals contain zero. Every model loses money.

**And the gain is not about long careers.** Truncating the encoder to the
five most recent runs keeps
`r sprintf("%.1f%%", p5_seqlen(p5_seq_length_table, 5)$share_of_full_gain)`
of its advantage, so what it is doing is reading the recent runs properly
rather than reading further back.
