# The result {#sec-result}

```{r}
#| label: example-setup
#| include: false

# The worked example, selected by a rule rather than by eye: the first
# runner, in race and runner order, on the validation slice whose career
# consists of exactly eight prior runs.
ex_key <- p5_sequences$key |>
  dplyr::mutate(row = dplyr::row_number(),
                seq_len = p5_sequences$seq_len,
                career = p5_sequences$career_runs_prior)

ex <- ex_key |>
  dplyr::filter(race_id %in% p5_slice$val_race_ids,
                seq_len == 8L, career == 8L) |>
  dplyr::arrange(race_id, runner_id) |>
  dplyr::slice(1) |>
  dplyr::left_join(
    dplyr::select(p5_runners_interactions, race_id, runner_id, name,
                  pos_lag1_zero, pos_lag1_nonzero, pos_lag2_zero,
                  pos_lag2_nonzero, has_wins, going_runs_prior,
                  going_sr_shrunk, going_sr_delta, days_LTO_log),
    by = c("race_id", "runner_id")
  )

ex_seq <- p5_sequences$x[ex$row, seq_len(ex$seq_len), , drop = FALSE]
dimnames(ex_seq) <- list(NULL, NULL, p5_sequences$features)

ex_tbl <- tibble::tibble(
  `Run` = paste0(seq_len(ex$seq_len), if_else(seq_len(ex$seq_len) == 1L,
                                              " (last time)", "")),
  `Finish` = as.integer(ex_seq[1, , "finish_pos"]),
  `Beaten (l)` = sprintf("%.2f", ex_seq[1, , "beaten_dist"]),
  `Field` = as.integer(ex_seq[1, , "field_size"]),
  `Class` = as.integer(ex_seq[1, , "class"]),
  `Dist (f)` = as.integer(ex_seq[1, , "distance_furlongs"]),
  `Going (ord.)` = as.integer(ex_seq[1, , "going_ordinal"]),
  `Days since previous` = as.integer(ex_seq[1, , "days_since_prev"])
)
ex_name <- ex$name
```

## What changed

Every model in this series has been given a horse's recent form as a few
hand-built numbers. The largest of them is the position lag block: last time
out's finishing position, recorded only if the horse finished in the first
four and zero otherwise, with a companion indicator for the zero case, and
the same pair for the run before that. Four numbers, two runs. Alongside
them sit an ever-won flag and three going-affinity terms.

This paper removes those eight numbers and gives the model the runs
themselves. A gated recurrent encoder [@cho2014] reads up to the twenty most
recent prior runs, cross-surface, eight channels a run: the **raw**
finishing position, how far the horse was beaten, the going, the class, the
field size, the race distance, and the gap to the run before. Everything
downstream is unchanged — the same races, the same likelihood, the same
scorer, the same seed.

## What one horse looks like to each model

Take `r ex_name`, running at
`r format(p5_frame$race_date[p5_frame$race_id == ex$race_id & p5_frame$runner_id == ex$runner_id][1], "%d %B %Y")`
with eight prior runs behind it. Its record, most recent first, is what the
encoder reads:

```{r}
#| label: tbl-example-form
#| tbl-cap: !expr glue::glue("{ex_name}'s eight prior runs, as the encoder receives them. Going is on the series' ordinal scale; the days column is the gap back to the run before that one, so the earliest run in the career has no reading and carries the unavailable code −1.")
kable(ex_tbl, align = "lrrrrrrr")
```

Paper 3 sees this record as four numbers:
`pos_lag1_zero` = `r ex$pos_lag1_zero`,
`pos_lag1_nonzero` = `r ex$pos_lag1_nonzero`,
`pos_lag2_zero` = `r ex$pos_lag2_zero`,
`pos_lag2_nonzero` = `r ex$pos_lag2_nonzero` — plus `has_wins` =
`r ex$has_wins` and a going block of
(`r ex$going_runs_prior`, `r sprintf("%.2f", ex$going_sr_shrunk)`,
`r sprintf("%.2f", ex$going_sr_delta)`).

Last time out the horse finished `r ex_tbl$Finish[1]`th of
`r ex_tbl$Field[1]`, beaten `r ex_tbl[["Beaten (l)"]][1]` lengths. Because
that is outside the first four, the position lag records it as zero and
raises the indicator. A horse that finished last of twenty, beaten forty
lengths, produces exactly the same two numbers. The run before was a fourth
of `r ex_tbl$Field[2]`, beaten `r ex_tbl[["Beaten (l)"]][2]` lengths, and is
recorded as the single figure 4 — the same value a fourth beaten a
short-head in a six-runner race would produce.

The rest of the table — the six earlier runs, the beaten distances, the
class and distance of each race, the `r ex_tbl[["Days since previous"]][3]`-day
gap before the third-last run — is not in paper 3's feature set at all.

## What happened

On `r p5_comma(n_test_races)` held-out test races, with nothing tuned
between validation and test:

```{r}
#| label: tbl-headline
#| tbl-cap: "Encoder against hand-built summaries, held-out test split. Paired race-level bootstrap, B = 2000, seed 42, 90% intervals."
p5_test_contrasts |>
  dplyr::filter(stringr::str_detect(contrast, "rung 3b.*rung 1")) |>
  p5_contrast_table("Encoder", "Summaries") |>
  kable(align = "lrrrrl")
```

Every interval excludes zero. The same three comparisons against paper 3's
gradient-boosted trees [@chenguestrin2016] are larger still, and the
hand-built arm itself beats paper 3 by a smaller margin, so the three models
order strictly on all three measures:

$$\text{paper 3 (trees)} \;<\; \text{hand-built summaries (MLP)} \;<\; \text{sequence encoder}$$

Nine contrasts, nine intervals excluding zero. That is worth setting beside
what came before it. In papers 1 to 3 no model could be distinguished from
another on this same test split: every paired interval in paper 3 against
paper 2b contained zero, and the only contrast in that paper whose interval
excluded zero was the market beating both models. Three models now separate
cleanly.

## What did not change

**The market still ranks better than any of them.** The encoder reaches
P1_rank
`r fmt_metric(p5_ct(p5_test_contrasts, "P1_rank", "rung 3b.*rung 1")$point_a)`
and Brier_place
`r fmt_metric(p5_ct(p5_test_contrasts, "Brier_place", "rung 3b.*rung 1")$point_a)`
against the market baseline of roughly `r sprintf("%.5f", market_p1)` and
`r sprintf("%.4f", market_brier)` that papers 2b and 3 published. The gap
narrows on both and closes on neither. No market contrast was run in this
paper.

**No betting return improves in a way that can be distinguished from
nothing.** The single-bet-per-race backtest gives the encoder a win ROI of
`r fmt_pct(p5_bt(p5_test_backtests, "rung3b", "win")$roi)` against paper 3's
`r fmt_pct(p5_bt(p5_test_backtests, "paper3", "win")$roi)` — the best in the
series — but the difference is
`r fmt_pts(p5_roi(p5_roi_contrasts, "rung 3b.*paper 3", "win")$diff_point)`
points with a 90% interval of
`r fmt_pts_ci(p5_roi(p5_roi_contrasts, "rung 3b.*paper 3", "win")$ci_lo, p5_roi(p5_roi_contrasts, "rung 3b.*paper 3", "win")$ci_hi)`.
Eight of the nine ROI intervals contain zero. Every arm loses money.

**And the gain is not about long careers.** Truncating the encoder to the
five most recent runs keeps
`r sprintf("%.1f%%", p5_seqlen(p5_seq_length_table, 5)$share_of_full_gain)`
of its advantage. Reading twenty runs instead of five is worth the remaining
`r sprintf("%.1f%%", 100 - p5_seqlen(p5_seq_length_table, 5)$share_of_full_gain)`.
What the encoder is doing is reading the recent runs properly, not reading
further back.
