# Data and scope {#sec-data}

## The race universe is paper 3's, unchanged

Nothing about which races are modelled changes in this paper. The scope is
the series' own: UK All-Weather Flat handicaps in classes 2 to 5 at Kempton,
Lingfield, Southwell and Wolverhampton, from 1 January 2006 to 14 October
2015, non-maiden, four to sixteen declared runners, complete cases on the
model's features.

The rows are taken from the matrices paper 3 was fitted and scored on rather
than re-derived from the filter rules, and row identity is asserted as a
sorted set on `(race_id, runner_id)` rather than checked by counting races:

```{r}
#| label: tbl-identity
#| tbl-cap: "Row identity against paper 3's frame."
p5_sequence_identity |>
  dplyr::rename(Check = check, Value = value) |>
  kable(align = "lr")
```

The train/test split is the series' frozen chronological 70/30 at
2012-12-30, giving `r p5_comma(n_train_races)` training races
(`r p5_comma(n_train_rows)` runner-rows) and `r p5_comma(n_test_races)` test
races (`r p5_comma(n_test_rows)` runner-rows). No within-race information
crosses the boundary.

## A validation slice, so the test split is scored once

The models in this paper have hyperparameters — layer widths, learning
rates, a hidden size, a dropout rate, an epoch count — and every one of them
had to be chosen. Papers 2a and 2b had no such choices; paper 3 made them by
cross-validation on the training split.

This paper carves a **validation slice** out of the training split by the
same rule the series uses one level up: the date at which 70% of the
training period's races have been run, taken as a calendar date so it stays
reproducible if the race universe changes. That date is
`r format(p5_slice$boundary, "%d %B %Y")`, giving
`r p5_comma(n_fit_races)` fitting races
(`r format(p5_slice$fit_date_range[1], "%b %Y")` to
`r format(p5_slice$fit_date_range[2], "%b %Y")`) and
`r p5_comma(n_val_races)` validation races
(`r format(p5_slice$val_date_range[1], "%b %Y")` to
`r format(p5_slice$val_date_range[2], "%b %Y")`).

Every choice in this paper — features, architecture, hyperparameters, epoch
count, which model goes forward — was made on that slice. The test split was
scored once, at the end, after all of them were fixed.

## The sequences

For each runner-race, the encoder receives the horse's most recent prior
runs, ordered most recent first, taken from the full cross-surface career
history the series already extracts. The temporal boundary is the series'
own: strictly before the meeting date of the race being predicted.

Eight values describe each run:

| value | what it carries |
|---|---|
| `finish_pos` | finishing position, **raw**, not capped at four |
| `beaten_dist` | cumulative lengths behind the winner |
| `beaten_missing` | indicator, distinguishing a real zero from an absent reading |
| `going_ordinal` | the going that day, on the series' ordinal scale |
| `class` | race class on the 2006 scheme |
| `field_size` | runners in that race |
| `distance_furlongs` | race distance |
| `days_since_prev` | days from the run before it to that run |

One convention matters for reading the tables below. **Unavailable readings
take −1, not zero**, because zero is a legitimate value for several of these
— a winner is zero lengths behind the winner.

### How much history there is

Training split only, `r p5_comma(p5_sequence_summary$n_obs)` runner-rows:

```{r}
#| label: tbl-seqlen
#| tbl-cap: "Prior-run counts on the training split. The capped row is what the encoder reads; the uncapped row is the horse's true career length, which enters the model separately as a single feature."
dplyr::bind_rows(p5_sequence_summary$seq_len,
                 p5_sequence_summary$career) |>
  dplyr::rename(Measure = measure, Min = min, Q1 = q1, Median = median,
                Q3 = q3, P90 = p90, Max = max, Mean = mean) |>
  kable(align = "lrrrrrrr")
```

`r sprintf("%.1f%%", p5_sequence_summary$truncated$share_pct)` of training
observations have careers longer than the twenty-run cap, and
`r p5_comma(p5_sequence_summary$empty$n)`
(`r sprintf("%.1f%%", p5_sequence_summary$empty$share_pct)`) have no prior
run at all. Because truncation discards a horse's oldest runs, the
**untruncated** career length is carried alongside the sequence as a single
feature, `career_runs_prior`, so nothing about career length is lost at the
cap. A sequence of length zero is exactly "no prior run", which is the fact
the discarded indicators encoded.

### The values as the model reads them

Each of the eight values was checked against its own scale, and out-of-range
readings were handled in one of two ways. Training split, real slots only:

```{r}
#| label: tbl-run-values
#| tbl-cap: "The eight values describing each prior run, training split, over real (unpadded) slots."
p5_sequence_summary$channels |>
  dplyr::transmute(
    Value = channel,
    `Real slots` = p5_comma(n_real_slots),
    Min = min, Median = median, Mean = mean, Max = max,
    `% zero` = pct_zero
  ) |>
  kable(align = "lrrrrrr")
```

A pre-2006 class code of 42 is not a class at all, so it is marked
unavailable and takes −1, as a finishing position of 99 does. A field size of
118 is on the right scale but implausible, so it is capped at 40 — as days
since previous is capped at 1,000 and beaten distance at
`r p5_beaten_audit$bound` lengths. @sec-appx-b gives the counts.

## One feature the sequences do not carry

The eight features the encoder replaces all summarise the horse's history. A
ninth that does the same, `days_LTO_log` — the gap from the horse's most
recent run to today's race — is **kept**.

The sequences record the gap between one prior run and the one before it,
not the gap from the last run to today's race, because today's race is not
in the sequence. The encoder therefore keeps `days_LTO_log` as a separate
feature, as it keeps `career_runs_prior`. @sec-ladder reports what it is
worth: on the validation slice, restoring it moved the encoder from
`r fmt_loss(loss_rung3)` to `r fmt_loss(loss_rung3b)` per race, about
`r sprintf("%.0f%%", 100 * (loss_rung3 - loss_rung3b) / gain_encoder)` of
the encoder's whole advantage over the hand-built model.
