# Data and scope {#sec-data}

## The race universe is paper 3's, unchanged

Nothing about which races are modelled changes in this paper. The scope is
the series' own: UK All-Weather Flat handicaps in classes 2 to 5 at Kempton,
Lingfield, Southwell and Wolverhampton, from 1 January 2006 to 14 October
2015, non-maiden, four to sixteen declared runners, complete cases on the
model's features.

The rows are taken from the matrices paper 3 was fitted and scored on rather
than re-derived from the filter rules, and row identity is asserted as a
sorted set on `(race_id, runner_id)` rather than checked by counting races:

```{r}
#| label: tbl-identity
#| tbl-cap: "Row identity against paper 3's frame."
p5_sequence_identity |>
  dplyr::rename(Check = check, Value = value) |>
  kable(align = "lr")
```

The train/test split is the series' frozen chronological 70/30 at
2012-12-30, giving `r p5_comma(n_train_races)` training races
(`r p5_comma(n_train_rows)` runner-rows) and `r p5_comma(n_test_races)` test
races (`r p5_comma(n_test_rows)` runner-rows). No within-race information
crosses the boundary.

## A validation slice, so the test split is scored once

The models in this paper have hyperparameters — layer widths, learning
rates, a hidden size, a dropout rate, an epoch count — and every one of them
had to be chosen. Papers 2a and 2b had no such choices; paper 3 made them by
cross-validation on the training split.

This paper carves a **validation slice** out of the training split by the
same rule the series uses one level up: the date at which 70% of the
training period's races have been run, taken as a calendar date so it stays
reproducible if the race universe changes. That date is
`r format(p5_slice$boundary, "%d %B %Y")`, giving
`r p5_comma(n_fit_races)` fitting races
(`r format(p5_slice$fit_date_range[1], "%b %Y")` to
`r format(p5_slice$fit_date_range[2], "%b %Y")`) and
`r p5_comma(n_val_races)` validation races
(`r format(p5_slice$val_date_range[1], "%b %Y")` to
`r format(p5_slice$val_date_range[2], "%b %Y")`).

Every choice in this paper — features, architecture, hyperparameters, epoch
count, which arm goes forward — was made on that slice. The test split was
scored once, at the end, after all of them were fixed.

## The sequences

For each runner-race, the encoder receives the horse's most recent prior
runs, ordered most recent first, taken from the full cross-surface career
history the series already extracts. The temporal boundary is the series'
own: strictly before the meeting date of the race being predicted.

Eight channels a run:

| channel | what it carries |
|---|---|
| `finish_pos` | finishing position, **raw**, not capped at four |
| `beaten_dist` | cumulative lengths behind the winner |
| `beaten_missing` | indicator, distinguishing a real zero from an absent reading |
| `going_ordinal` | the going that day, on the series' ordinal scale |
| `class` | race class on the 2006 scheme |
| `field_size` | runners in that race |
| `distance_furlongs` | race distance |
| `days_since_prev` | days from the run before it to that run |

Two conventions are load-bearing. **Unavailable readings take −1, not
zero**, because zero is a legitimate value for several channels — a winner
is zero lengths behind the winner — and is also the padding value. And
**padding is masked by an explicit length** rather than being relied upon to
be inert; @sec-appx-b gives the gate that asserts this.

### How much history there is

Training split only, `r p5_comma(p5_sequence_summary$n_obs)` runner-rows:

```{r}
#| label: tbl-seqlen
#| tbl-cap: "Prior-run counts on the training split. The capped row is what the encoder reads; the uncapped row is the horse's true career length, which enters the model separately as a scalar."
dplyr::bind_rows(p5_sequence_summary$seq_len,
                 p5_sequence_summary$career) |>
  dplyr::rename(Measure = measure, Min = min, Q1 = q1, Median = median,
                Q3 = q3, P90 = p90, Max = max, Mean = mean) |>
  kable(align = "lrrrrrrr")
```

`r sprintf("%.1f%%", p5_sequence_summary$truncated$share_pct)` of training
observations have careers longer than the twenty-run cap, and
`r p5_comma(p5_sequence_summary$empty$n)`
(`r sprintf("%.1f%%", p5_sequence_summary$empty$share_pct)`) have no prior
run at all. Because truncation discards a horse's oldest runs, the
**untruncated** career length is carried alongside the sequence as a scalar,
`career_runs_prior`, so nothing about career length is lost at the cap. A
sequence of length zero is exactly "no prior run", which is the fact the
discarded indicators encoded.

### The channels as the model reads them

Every channel is bounded, and the bounds were set by inspecting the
career-run universe rather than assumed. Training split, real slots only:

```{r}
#| label: tbl-channels
#| tbl-cap: "Sequence channels on the training split, over real (unpadded) slots."
p5_sequence_summary$channels |>
  dplyr::transmute(
    Channel = channel,
    `Real slots` = p5_comma(n_real_slots),
    Min = min, Median = median, Mean = mean, Max = max,
    `% zero` = pct_zero
  ) |>
  kable(align = "lrrrrrr")
```

Two kinds of cleaning are applied, and the distinction is the point.
Readings that are **not on the channel's own scale** — a pre-2006 class code
of 42, a finishing position of 99 — are read as unavailable, because a class
of 42 is not "class 42". Readings that **are** on the channel's scale but
whose exact magnitude stops being informative are winsorised, which keeps
the ordering: field size at 40, days since previous at 1,000, beaten
distance at `r p5_beaten_audit$bound` lengths. @sec-appx-b gives the counts
behind each.

## One feature the sequences do not carry

Eight hand-built features are removed in this paper. A ninth, `days_LTO_log`
— the gap from the horse's most recent run to today's race — is **kept**,
and the reason is a property of what a sequence representation carries
rather than a correction to it.

The `days_since_prev` channel is the gap between a historical run and the
one before it. Read across the sequence it gives the rhythm of a career:
whether the horse has been running every fortnight or once a season. It does
not give the distance from the most recent run to the race being predicted,
because that interval ends at a race the sequence does not contain. A
sequence of prior runs is a record of what happened; the layoff a horse
carries into today is a fact about today.

So the encoder arm keeps `days_LTO_log` as a scalar term, in the same way it
keeps `career_runs_prior`. Both are facts about the horse's history that the
sequence, as constructed, cannot express. @sec-ladder reports what the term
is worth: on the validation slice, restoring it moved the encoder from
`r fmt_loss(loss_rung3)` to `r fmt_loss(loss_rung3b)` per race, about
`r sprintf("%.0f%%", 100 * (loss_rung3 - loss_rung3b) / gain_encoder)` of
the encoder's whole advantage over the hand-built arm.
