# Method {#sec-method}

## The objective is paper 2b's and paper 3's

The likelihood does not change in this paper. Each runner $j$ in race $r$
receives a score $z_{rj}$, and the probability of the observed finishing
order down to depth $k$ is the Plackett–Luce form
[@plackett1975; @luce1959] that paper 2b's exploded conditional logit fits
and paper 3's custom objective optimises:

$$
\Pr(\text{top-}k \text{ order}) \;=\; \prod_{s=1}^{S}
\frac{\exp(z_{r,(s)})}{\sum_{j \in R_{r,s}} \exp(z_{rj})},
\qquad S = \min(k,\, J_r - 1),
$$

where $(s)$ is the horse that finished $s$-th and $R_{r,s}$ is the set still
unplaced at stage $s$. Depth $k = 3$ throughout, as in papers 2b and 3. At
$k = 1$ this is McFadden's conditional logit [@mcfadden1974], which is
paper 1's model.

What changes across the series is only the function producing $z$: a linear
predictor in papers 1, 2a and 2b, a gradient-boosted tree ensemble in paper
3, and a neural scorer here. The loss is implemented directly in `{torch}`
and verified against paper 3's analytic gradient, so this paper descends the
same surface paper 3 did; @sec-appx-a gives the verification.

Weights are shared across horses. The scorer maps one runner's features to
one number, and runners are coupled only through the per-race denominator,
so variable field sizes need nothing beyond masking.

## A ladder, one change at a rung

The paper is built as a ladder. Each rung changes exactly one thing about
how a horse's history reaches the scorer and is compared with the rung
below it, so that a difference in ranking can be attributed to that change
and not to a bundle of them.

| rung | the one change | comparator |
|---|---|---|
| 1 | linear scorer → multi-layer perceptron, identical features | paper 2b's linear predictor |
| 2 | three strike rates → learned entity embeddings | rung 1 |
| 3 | eight history summaries → sequence encoder | rung 1 |

This discipline is the paper's method as much as the models are. Rung 1 took
four attempts, and three were spent on arms that differed from their
comparator in two respects at once — a feature encoding and a function
class, then a function class and a feature count, then a function class and
a selection budget. None of those three could test what the rung existed to
test. The rule that came out of it is that the term-by-term difference
between two arms is stated before anything is fitted and asserted in code,
not left to prose. Every rung below carries such an assertion, and the
tables in @sec-ladder and @sec-results are those assertions rendered.

## Selection, and what it was allowed to see

Nine configurations per arm, declared in full before any of them was scored,
selected on validation Plackett–Luce loss and nothing else. Selection never
saw a ranking metric, a betting return, or a test row.

The neural arms overfit this data quickly: across three rungs the best epoch
was reached within about three passes over the training data and the
validation loss then rose monotonically. Epoch selection is therefore doing
real work, and it is done the same way for every arm — the parameters from
the best validation epoch are the ones kept.

Rungs 1 and 2 ran nine configurations for 200 epochs each. Rung 3 ran nine
for 60 and then refitted the selected configuration alone at 200, which
reproduced the shorter run's trace exactly and did not improve on it: the
optimum was at the same epoch either way. That is what makes the shorter
search budget safe to use, and it is checked rather than assumed.

## The two arms that reach the test split

```{r}
#| label: tbl-arm-difference
#| tbl-cap: "The two test-split arms, term by term. Asserted in the pipeline before either was scored."
p5_test_arm_difference |>
  dplyr::transmute(
    Respect = respect,
    `Hand-built (rung 1)` = rung_1,
    `Encoder (rung 3b)` = rung_3b,
    Differs = dplyr::if_else(differs, "yes", "")
  ) |>
  kable(align = "llll")
```

Both arms are refitted on the **full** training split — the validation slice
folded back in — at the configuration the slice selected, then score the
test split once. Both train for the same number of epochs, the epoch each
selected on validation, and there is no per-epoch evaluation at that stage:
with no held-out set left, scoring the test split each epoch would put it
inside the selection loop.

One consequence is worth stating plainly. The full training split is about
43% larger than the fitting partition used for selection, so an epoch is 43%
more gradient steps than it was during selection. Both arms take that change
identically, so the contrast between them is unaffected by it.

## The backend is a frozen parameter

Every model in this paper is fitted on the CPU `libtorch` build in double
precision. Results do not reproduce across backends, so the backend is
treated as a parameter of the whole paper rather than a runtime detail: it
was decided once, before the ladder was built, and not revisited.

The lockfile that pins the R packages cannot capture which C++ runtime is
installed, so the backend is captured inside every fitted model and asserted
against the pipeline's own record. A report cannot claim a backend a fit did
not use. @sec-appx-c gives the mechanism and the reasoning behind choosing
CPU on this hardware.

```{r}
#| label: tbl-backend
#| tbl-cap: "The backend, captured at fit time and asserted for both test-split arms."
p5_test_backend_check |>
  dplyr::transmute(Device = device, `CUDA available` = cuda_is_available,
                   torch = torch_version, libtorch = libtorch_version,
                   Threads = torch_threads, R = r_version,
                   `Fits checked` = n_runs_checked) |>
  kable(align = "lllrrlr")
```
