# The ladder, on the validation slice {#sec-ladder}

Everything in this section is measured on the `r p5_comma(n_val_races)`
validation races. No test row is involved.

## Rung 1: the control that was meant to tie, and did not

Rung 1 exists to establish a baseline, not to find anything. The plan was to
give a multi-layer perceptron exactly the features paper 2b's linear
predictor has, on the same likelihood and the same races, and confirm the
two tie — so that any later gain from a sequence encoder could be attributed
to the representation rather than to having switched to a neural network.

They do not tie.

```{r}
#| label: tbl-rung1
#| tbl-cap: "MLP against a linear scorer on identical 26 features and an equal selection budget, validation slice. Paired race-level bootstrap, B = 2000, 90% intervals."
p5_mlp_vs_linear_v4 |>
  p5_contrast_table("MLP", "Linear") |>
  kable(align = "lrrrrl")
```

Both intervals exclude zero. The two models carry identical features, the
same loss, the same slice and an equal budget of nine configurations at 200
epochs each.

@tbl-budget-decomp separates two changes. Paper 2b's linear model has 17
features. Giving the same linear model nine more — the four going-affinity
terms, the draw position and the four course indicators paper 2b lacks —
improved it by `r fmt_gain(gain_ninefeature)` per race. Switching that same
26-feature model from a linear scorer to an MLP improved it by a further
`r fmt_gain(gain_nonlinear)`, about
`r sprintf("%.0f", gain_nonlinear / gain_ninefeature)` times as much.

```{r}
#| label: tbl-budget-decomp
#| tbl-cap: "Three fits on the validation slice: paper 2b's linear model on 17 features, the same linear model on 26, and the MLP on those same 26."
p5_budget_decomposition |>
  dplyr::transmute(
    Model = arm, Features = terms, Budget = budget,
    `Val PL loss` = purrr::map_chr(val_pl_loss, fmt_loss),
    `Gain on previous row` = dplyr::if_else(
      is.na(gain_vs_prev), "—", purrr::map_chr(gain_vs_prev, fmt_gain))
  ) |>
  kable(align = "lrlrr")
```

**This sits against paper 3's conclusion, and the paper should say so
directly.** Paper 3 replaced a linear predictor with a gradient-boosted tree
ensemble on the same features and objective, found every paired interval
containing zero, and read that as the function class not being the binding
constraint. On the evidence in front of it that reading was sound, but the
evidence was one function class. A different nonlinear model, on essentially
the same information and the same objective, does better than the linear
predictor here. What paper 3 established is that trees did not help; what it
did not establish is that nonlinearity did not.

Two differences between the comparisons are worth keeping in view: paper 3's
contrast was scored on the test split while this one is on a validation
slice, and paper 3's trees route missing values natively where this model
imputes them. Neither changes the direction of the finding.

## Rung 2: learned representations for trainer, jockey and sire

An embedding is a short list of numbers the model learns for each trainer,
jockey and sire, used in place of a single strike rate. Because the model
fits those numbers itself, they can carry more than a win rate can — the
supplementary note accompanying this paper gives the construction.

Rung 2 makes the swap. The three strike rates `trainerSR`, `jockeySR` and
`sireSR` come out, leaving 23 features, and a four-number representation of
the trainer, of the jockey and of the sire goes in. Those representations
are learned only from the fitting rows, and an entity with fewer than twenty
runs there is pooled into a shared bucket rather than given its own.

```{r}
#| label: tbl-rung2
#| tbl-cap: "Learned representations against the strike rates they replaced, validation slice."
p5_rung2_vs_rung1 |>
  p5_contrast_table("Representations", "Strike rates") |>
  kable(align = "lrrrrl")
```

Both intervals exclude zero, against the representations. The swap costs
`r fmt_gain(cost_embeddings)` per race on validation loss, and the
separation is complete: the best of the nine configurations with
representations, `r fmt_loss(min(p5_rung2_scores$best_val_pl_loss))`, is
worse than the worst of the nine with strike rates,
`r fmt_loss(max(p5_mlp_scores_v4$best_val_pl_loss))`. The reading was fixed
before the fit and it fired as *worse — report and stop, do not remediate*.
No remediation was attempted.

### The mechanism

A strike rate is arithmetic: the entity's cumulative wins over runs at the
latest meeting date before this race, computed over the whole 1.72M-row
archive. It updates after every race the trainer has, and it is defined for
trainers the model has never been fitted on.

A learned representation is neither. It is one fixed vector per entity,
learned only from the
`r p5_comma(nrow(p5_arm_data$fit$key))` in-scope fitting rows, and it exists
only for entities that appear there often enough to be worth one.

```{r}
#| label: tbl-vocab
#| tbl-cap: "Which entities got their own representation, frozen on the fitting partition."
p5_entity_indices$summary |>
  dplyr::transmute(
    Entity = entity,
    `Distinct in fitting` = p5_comma(n_distinct_fit),
    `Given a representation` = p5_comma(vocab_size),
    `% fitting rows pooled` = pct_fit_bucket,
    `% validation rows pooled` = pct_val_bucket,
    `% validation rows unseen` = pct_val_unseen
  ) |>
  kable(align = "lrrrrr")
```

Under a chronological split that matters. About half the validation rows
carry at least one entity in the shared bucket, against about a fifth of
fitting rows, and on those rows the model reads a generic vector where the
strike rate it replaced was a real number — `trainerSR`, `jockeySR` and
`sireSR` are missing on no row of either partition.

Stated generally: **a learned representation loses to a hand-built feature
when the hand-built feature is computed from a wider information set than
the model can be fitted on.**

## Rung 3: the sequence encoder

Rung 3 removes the history summaries and adds the encoder. Nine
configurations were declared, varying the four settings that belong to the
encoder itself: the hidden size, the number of recurrent layers, the dropout
rate and the learning rate, spread around a centre of hidden size 32, one
layer, dropout 0.1 and a learning rate of 1e-3. The downstream scorer is
held at rung 1's selected architecture throughout, so the only architecture
being searched is the encoder's.

@tbl-rung3b-grid gives the nine. Three things to take from it: dropout is
the setting that matters, and its ordering is monotone — the highest setting
is best and zero is second worst; capacity past a hidden size of 32 buys
nothing; and the spread from best to worst configuration,
`r fmt_gain(max(p5_rung3b_scores_60$best_val_pl_loss) - min(p5_rung3b_scores_60$best_val_pl_loss))`,
is small against the `r fmt_gain(gain_encoder)` the encoder gains over the
hand-built model.

```{r}
#| label: tbl-rung3b-grid
#| tbl-cap: "The nine encoder configurations, validation slice, 60 epochs each. Selection on validation PL loss and nothing else."
p5_rung3b_scores_60 |>
  dplyr::transmute(
    Config = config, Hidden = hidden, Layers = layers, Dropout = dropout,
    `Learning rate` = formatC(lr, format = "e", digits = 0),
    `Best val PL loss` = purrr::map_chr(best_val_pl_loss, fmt_loss),
    `Best epoch` = best_epoch,
    `Loss at epoch 60` = purrr::map_chr(final_epoch_val_pl_loss, fmt_loss)
  ) |>
  kable(align = "rrrrrrrr")
```

Refitting the selected configuration at 200 epochs reproduced its 60-epoch
trace exactly — maximum absolute difference
`r p5_rung3b_budget_check$max_abs_trace_diff_first_60` — and did not improve
on it. The optimum is at epoch
`r p5_rung3b_budget_check$best_epoch_at_200` either way.

The encoder keeps `days_LTO_log`, for the reason given in @sec-data. On the
validation slice that term is worth `r fmt_gain(loss_rung3 - loss_rung3b)`
per race:

```{r}
#| label: tbl-rung3b
#| tbl-cap: "The layoff term in the encoder model, validation slice."
p5_rung3b_vs_rung3 |>
  p5_contrast_table("Layoff kept", "Layoff dropped") |>
  kable(align = "lrrrrl")
```

## The ladder in one table

```{r}
#| label: tbl-ladder
#| tbl-cap: "Validation Plackett–Luce loss per race. Positive gain is better than rung 1."
p5_rung3b_decomposition |>
  dplyr::transmute(
    Model = arm,
    `Val PL loss` = purrr::map_chr(val_pl_loss, fmt_loss),
    `Gain over rung 1` = purrr::map_chr(gain_over_rung1, \(x)
      if (abs(x) < 1e-12) "—" else fmt_gain(x))
  ) |>
  kable(align = "lrr")
```

The encoder is worth `r fmt_gain(gain_encoder)` per race over the hand-built
model — about `r sprintf("%.1f", gain_encoder / gain_nonlinear)` times the
nonlinearity and `r sprintf("%.0f", gain_encoder / gain_ninefeature)` times
the nine extra features — on **seven fewer** hand-built features. Which
model went to the test split was decided on this column and recorded in the
pipeline, not chosen by hand.
