# The ladder, on the validation slice {#sec-ladder}

Everything in this section is measured on the `r p5_comma(n_val_races)`
validation races. No test row is involved.

## Rung 1: the control that was meant to tie, and did not

Rung 1 exists to establish a baseline, not to find anything. The plan was to
give a multi-layer perceptron exactly the features paper 2b's linear
predictor has, on the same likelihood and the same races, and confirm the
two tie — so that any later gain from a sequence encoder could be attributed
to the representation rather than to having switched to a neural network.

They do not tie.

```{r}
#| label: tbl-rung1
#| tbl-cap: "MLP against a linear scorer on identical 26 terms and an equal selection budget, validation slice. Paired race-level bootstrap, B = 2000, seed 42, 90% intervals."
p5_mlp_vs_linear_v4 |>
  p5_contrast_table("MLP", "Linear") |>
  kable(align = "lrrrrl")
```

Both intervals exclude zero. The two arms carry identical features, the same
loss, the same slice, the same seed, and — after the third attempt at this
rung — an equal budget of nine configurations at 200 epochs each. On
validation Plackett–Luce loss:

```{r}
#| label: tbl-budget-decomp
#| tbl-cap: "What each change is worth, per race, on the validation slice."
p5_budget_decomposition |>
  dplyr::transmute(
    Arm = arm, Terms = terms, Budget = budget,
    `Val PL loss` = purrr::map_chr(val_pl_loss, fmt_loss),
    `Gain on previous row` = dplyr::if_else(
      is.na(gain_vs_prev), "—", purrr::map_chr(gain_vs_prev, fmt_gain))
  ) |>
  kable(align = "lrlrr")
```

The nonlinearity is worth `r fmt_gain(gain_nonlinear)` per race; the nine
features paper 2b lacks are worth `r fmt_gain(gain_ninefeature)`. The
function class buys about
`r sprintf("%.1f", gain_nonlinear / gain_ninefeature)` times what those
features buy.

**This sits against paper 3's conclusion, and the paper should say so
directly.** Paper 3 replaced a linear predictor with a gradient-boosted tree
ensemble on the same features and objective, found every paired interval
containing zero, and read that as the function class not being the binding
constraint. On the evidence in front of it that reading was sound. But the
evidence was one function class. A different nonlinear class, on
essentially the same information and the same objective, does better than
the linear predictor here — so the conclusion, as stated, was drawn more
broadly than the experiment supported. What paper 3 established is that
*trees* did not help; what it did not establish is that nonlinearity did not.

Two differences between the comparisons are worth keeping in view rather
than using to explain the tension away: paper 3's contrast was scored on the
test split while this one is on a validation slice carved out of training,
and paper 3's trees route missing values natively where this arm imputes
them. Neither changes the direction of the finding.

## Rung 2: embeddings for trainer, jockey and sire, and a clean null

Rung 2 removes the three strike rates — `trainerSR`, `jockeySR`, `sireSR` —
and puts learned embeddings on `trainer_id`, `jockey_id` and `sire_id` in
their place: 26 dense terms become 23, plus three four-dimensional vectors
the model fits for itself. Vocabularies are frozen on the fitting partition,
with entities below twenty runs pooled into a shared bucket.

```{r}
#| label: tbl-rung2
#| tbl-cap: "Entity embeddings against the strike rates they replaced, validation slice."
p5_rung2_vs_rung1 |>
  p5_contrast_table("Embeddings", "Strike rates") |>
  kable(align = "lrrrrl")
```

Both intervals exclude zero, both against the embeddings. On validation loss
the swap costs `r fmt_gain(cost_embeddings)` per race, about
`r sprintf("%.1f", cost_embeddings / gain_nonlinear)` times what the
nonlinearity in rung 1 bought. The separation is complete across the search:
the best of the nine embedding configurations,
`r fmt_loss(min(p5_rung2_scores$best_val_pl_loss))`, is worse than the worst
of the nine strike-rate configurations,
`r fmt_loss(max(p5_mlp_scores_v4$best_val_pl_loss))`. There is no
configuration-selection story available.

The reading was fixed before the fit and it fired as *worse — report and
stop, do not remediate*. No remediation was attempted.

### The mechanism is the finding

A strike rate here is not a parameter. It is arithmetic: the entity's
cumulative wins over runs at the latest meeting date strictly before this
race, computed over the whole 1.72M-row historical archive. It therefore
varies within an entity over time, and it is defined for entities the model
has never been fitted on.

An embedding is the opposite on both counts. It is one static vector per
entity, learned only from the `r p5_comma(nrow(p5_arm_data$fit$key))`
in-scope fitting rows, and it exists only for entities that appear in them
often enough to be worth a vector.

Under a chronological split the second point bites hard, because the
validation slice is later than the fitting partition and contains people and
sires the earlier period did not:

```{r}
#| label: tbl-vocab
#| tbl-cap: "Entity vocabularies, frozen on the fitting partition."
p5_entity_indices$summary |>
  dplyr::transmute(
    Entity = entity,
    `Distinct in fitting` = p5_comma(n_distinct_fit),
    `Vocabulary` = p5_comma(vocab_size),
    `% fitting rows bucketed` = pct_fit_bucket,
    `% validation rows bucketed` = pct_val_bucket,
    `% validation rows unseen` = pct_val_unseen
  ) |>
  kable(align = "lrrrrr")
```

Roughly half the validation rows carry at least one of the three entities in
the shared rare bucket, against about a fifth of fitting rows. On those rows
the model reads a generic vector where the strike rate it replaced was a
real number: `trainerSR`, `jockeySR` and `sireSR` are missing on **no** row
of either partition.

Stated generally, and this is the transferable part: **a learned
representation loses to a hand-built feature when the hand-built feature is
computed from a wider information set than the model can be fitted on.** The
strike rate summarises an archive; the embedding can only summarise the
sample. No amount of tuning inside rung 2 addresses that, which is why the
rung stops rather than remediating.

## Rung 3: the sequence encoder

Rung 3 removes the history summaries and adds the encoder. Nine
configurations over hidden size, layers, dropout and learning rate — the
downstream scorer held at rung 1's selected architecture, so the only
architecture being searched is the encoder's own.

The first version of this rung dropped nine summaries including
`days_LTO_log`, which as @sec-data explains the sequences do not carry.
Restoring it as a scalar is one change from that arm, and it is worth
having:

```{r}
#| label: tbl-rung3b
#| tbl-cap: "Restoring the layoff term to the encoder arm, validation slice."
p5_rung3b_vs_rung3 |>
  p5_contrast_table("Layoff kept", "Layoff dropped") |>
  kable(align = "lrrrrl")
```

Both intervals exclude zero. The single term is worth
`r fmt_gain(loss_rung3 - loss_rung3b)` per race, about
`r sprintf("%.1f", (loss_rung3 - loss_rung3b) / gain_nonlinear)` times what
the entire nonlinearity bought in rung 1. Eight of the nine configurations
beat the arm without it. The encoder arm carried forward is therefore the
one with the layoff term: eight summaries removed, not nine.

```{r}
#| label: tbl-rung3b-grid
#| tbl-cap: "The nine encoder configurations, validation slice, 60 epochs each. Selection on validation PL loss and nothing else."
p5_rung3b_scores_60 |>
  dplyr::transmute(
    Config = config, Hidden = hidden, Layers = layers, Dropout = dropout,
    `Learning rate` = formatC(lr, format = "e", digits = 0),
    `Best val PL loss` = purrr::map_chr(best_val_pl_loss, fmt_loss),
    `Best epoch` = best_epoch,
    `Loss at epoch 60` = purrr::map_chr(final_epoch_val_pl_loss, fmt_loss)
  ) |>
  kable(align = "rrrrrrrr")
```

Dropout is the axis that matters, and its ordering is monotone: the highest
setting is best and zero is second worst. Capacity past a hidden size of 32
buys nothing, and the smallest hidden size is the worst configuration in the
grid. The spread from best to worst is
`r fmt_gain(max(p5_rung3b_scores_60$best_val_pl_loss) - min(p5_rung3b_scores_60$best_val_pl_loss))`,
small against the `r fmt_gain(gain_encoder)` the arm gains over rung 1.

Refitting the selected configuration at 200 epochs reproduced its 60-epoch
trace exactly — maximum absolute difference
`r p5_rung3b_budget_check$max_abs_trace_diff_first_60` — and did not improve
on it. The optimum is at epoch
`r p5_rung3b_budget_check$best_epoch_at_200` either way.

## The ladder in one table

```{r}
#| label: tbl-ladder
#| tbl-cap: "Validation Plackett–Luce loss per race. Positive gain is better than rung 1."
p5_rung3b_decomposition |>
  dplyr::transmute(
    Arm = arm,
    `Val PL loss` = purrr::map_chr(val_pl_loss, fmt_loss),
    `Gain over rung 1` = purrr::map_chr(gain_over_rung1, \(x)
      if (abs(x) < 1e-12) "—" else fmt_gain(x))
  ) |>
  kable(align = "lrr")
```

The encoder is worth `r fmt_gain(gain_encoder)` per race over the hand-built
arm — about `r sprintf("%.1f", gain_encoder / gain_nonlinear)` times the
nonlinearity and `r sprintf("%.0f", gain_encoder / gain_ninefeature)` times
the nine extra features — on **seven fewer** hand-built terms. Which arm
went to the test split was decided on this column and recorded in the
pipeline, not chosen by hand.
