# The test split {#sec-results}

The test split was scored once, after every choice in @sec-ladder was fixed.
Both models were refitted on the full training split, scored on
`r p5_comma(n_test_races)` held-out races, and passed through paper 3's own
evaluation code — the same prediction builder, the same ranking metrics, the
same backtest — with nothing differing between the models but the score
vector. All three models are compared on an identical race universe.

## Ranking

```{r}
#| label: tbl-test-enc-hand
#| tbl-cap: "Sequence encoder against the MLP on hand-built features, held-out test split. Paired race-level bootstrap over the common races, B = 2000, 90% intervals."
p5_test_contrasts |>
  dplyr::filter(stringr::str_detect(contrast, "rung 3b.*rung 1")) |>
  p5_contrast_table("Encoder", "Hand-built") |>
  kable(align = "lrrrrl")
```

```{r}
#| label: tbl-test-enc-p3
#| tbl-cap: "Sequence encoder against paper 3's gradient-boosted trees, same races and same method."
p5_test_contrasts |>
  dplyr::filter(stringr::str_detect(contrast, "rung 3b.*paper 3")) |>
  p5_contrast_table("Encoder", "Paper 3") |>
  kable(align = "lrrrrl")
```

```{r}
#| label: tbl-test-hand-p3
#| tbl-cap: "The MLP on hand-built features against paper 3's gradient-boosted trees, same races and same method."
p5_test_contrasts |>
  dplyr::filter(stringr::str_detect(contrast, "rung 1.*paper 3")) |>
  p5_contrast_table("Hand-built", "Paper 3") |>
  kable(align = "lrrrrl")
```

Nine intervals, nine exclusions of zero, every one favouring the newer
model. The ordering is strict and identical on all three measures:

$$\text{paper 3 (trees)} \;<\; \text{MLP on hand-built features} \;<\;
\text{sequence encoder}$$

Two things about the sizes. The encoder's margin over paper 3 is larger than
its margin over the hand-built model, which is what the ladder predicts: the
hand-built model is itself ahead of paper 3. And that model's own margin
over paper 3 is roughly a fifth of the encoder's, so most of the distance
between paper 3 and the encoder comes from the representation rather than
from the switch to a neural scorer.

**The effect is larger on test than it was on validation.** The model was
finalised on the validation slice and then scored on the test races without
further adjustment. On the validation slice the encoder beat the hand-built
model by `r fmt_sci(p5_ct(p5_rung3_vs_rung1, "P1_rank")$diff_point)` on
P1_rank before the layoff term was restored; on test the corresponding
figure is
`r fmt_sci(p5_ct(p5_test_contrasts, "P1_rank", "rung 3b.*rung 1")$diff_point)`.
This is the same model measured on a different and larger sample, not an
improvement in it.

### Against the market

The encoder reaches P1_rank
`r fmt_metric(p5_ct(p5_test_contrasts, "P1_rank", "rung 3b.*rung 1")$point_a)`
and Brier_place
`r fmt_metric(p5_ct(p5_test_contrasts, "Brier_place", "rung 3b.*rung 1")$point_a)`.
The over-round-adjusted starting price, on the baseline papers 2b and 3
published, reaches about `r sprintf("%.5f", market_p1)` and
`r sprintf("%.4f", market_brier)`.

The gap narrows on both measures and closes on neither. The encoder and the
market were not compared directly here; these are published figures set side
by side. Paper 4 already established the relevant result directly: against
the settled starting price this series' models add nothing distinguishable,
and their value sits in the interval between an early price and the off.

## Betting

The single-bet-per-race rule is the series' own, and it selected nothing in
this paper: no configuration, epoch, feature or model was chosen on a
return. Among horses clearing the naive filter — model win probability above
0.15 and a model-to-market ratio above 1.3 — the rule backs the one with the
highest model win probability. Win bets settle at starting price; place and
each-way settle from the real-price payout tables.

```{r}
#| label: tbl-backtest
#| tbl-cap: "Single-bet-per-race backtest on the held-out test split."
p5_backtest_table(
  p5_test_backtests,
  c(rung1 = "Hand-built", rung3b = "Encoder", paper3 = "Paper 3")
) |>
  dplyr::arrange(factor(Bet, levels = c("win", "place", "each-way")),
                 factor(Arm, levels = c("Encoder", "Hand-built", "Paper 3"))) |>
  dplyr::rename(Model = Arm) |>
  kable(align = "llrrrrr")
```

No model makes money. The encoder is the best of the three on win
(`r fmt_pct(p5_bt(p5_test_backtests, "rung3b", "win")$roi)`) and on each-way
(`r fmt_pct(p5_bt(p5_test_backtests, "rung3b", "eachway")$roi)`), and its
win return is the best in the series to date.

**It is also worst of the two paper-5 models on place**, at
`r fmt_pct(p5_bt(p5_test_backtests, "rung3b", "place")$roi)` against the
hand-built model's
`r fmt_pct(p5_bt(p5_test_backtests, "rung1", "place")$roi)`, despite having
the better Brier_place. The two measures are not in conflict: Brier_place
scores calibration across every runner in the field, while the bet rule
picks the single highest model win probability that clears a market filter,
so they are computed on different quantities over different subsets. The
result is recorded as it stands.

### Intervals on the returns

Every other contrast in this paper carries an interval, and the returns
should too. The per-race bet units were re-derived from the stored
predictions, and checked to reproduce the published returns and bet counts
exactly before anything was attached to them.

```{r}
#| label: tbl-roi-ci
#| tbl-cap: "Paired race-level bootstrap on single-bet ROI, over the common test races, B = 2000, 90% intervals. Differences in percentage points."
p5_roi_contrasts |>
  dplyr::transmute(
    Comparison = dplyr::recode(
      contrast,
      "rung 3b (encoder) - rung 1 (summaries)" = "Encoder − Hand-built",
      "rung 3b (encoder) - paper 3 (GBT)" = "Encoder − Paper 3",
      "rung 1 (MLP) - paper 3 (GBT)" = "Hand-built − Paper 3"
    ),
    Bet = dplyr::recode(bet, win = "win", place = "place",
                        eachway = "each-way"),
    First = purrr::map_chr(roi_a, \(x) fmt_pct(x, 2)),
    Second = purrr::map_chr(roi_b, \(x) fmt_pct(x, 2)),
    Difference = purrr::map_chr(diff_point, fmt_pts),
    `90% interval` = purrr::map2_chr(ci_lo, ci_hi, fmt_pts_ci),
    `Excludes zero` = dplyr::if_else(excludes_zero, "yes", "")
  ) |>
  kable(align = "llrrrrl")
```

Eight of the nine intervals contain zero, and they are wide — roughly five
to fourteen points either side. **No betting improvement in this paper is
distinguishable from no improvement at all.**

One interval does exclude zero: the hand-built model's place ROI is
`r fmt_pts(p5_roi(p5_roi_contrasts, "rung 1.*paper 3", "place")$diff_point)`
points above paper 3's, interval
`r fmt_pts_ci(p5_roi(p5_roi_contrasts, "rung 1.*paper 3", "place")$ci_lo, p5_roi(p5_roi_contrasts, "rung 1.*paper 3", "place")$ci_hi)`
— the first ROI interval in this series to exclude zero. The paper does not
build on it: it is one of nine intervals with no multiplicity control and
none pre-registered, it does not involve the paper's model, and both models
lose money.
