# The test split {#sec-results}

The test split was scored once, after every choice in @sec-ladder was fixed.
Both arms were refitted on the full training split, scored on
`r p5_comma(n_test_races)` held-out races, and passed through paper 3's own
evaluation code — the same prediction builder, the same ranking metrics, the
same backtest — with nothing differing between the arms but the score
vector. All three arms are compared on an identical race universe.

## Ranking

```{r}
#| label: tbl-test-all
#| tbl-cap: "All three contrasts on the held-out test split. Paired race-level bootstrap over the common races, B = 2000, seed 42, 90% intervals."
tibble::tribble(
  ~grep,                 ~a,          ~b,
  "rung 3b.*rung 1",     "Encoder",   "Summaries",
  "rung 3b.*paper 3",    "Encoder",   "Paper 3",
  "rung 1.*paper 3",     "Summaries", "Paper 3"
) |>
  purrr::pmap(function(grep, a, b) {
    p5_test_contrasts |>
      dplyr::filter(stringr::str_detect(contrast, grep)) |>
      p5_contrast_table(a, b) |>
      dplyr::mutate(Contrast = paste(a, "−", b), .before = 1) |>
      dplyr::rename(A = 3, B = 4)
  }) |>
  purrr::list_rbind() |>
  kable(align = "llrrrrl")
```

Nine intervals, nine exclusions of zero, every one favouring the newer arm.
The ordering is strict and identical on all three measures:

$$\text{paper 3} \;<\; \text{hand-built summaries} \;<\; \text{sequence encoder}$$

Two things about the sizes. The encoder's margin over paper 3 is larger than
its margin over the hand-built arm, which is what the ladder predicts: the
hand-built arm is itself ahead of paper 3. And the hand-built arm's own
margin over paper 3 is roughly a fifth of the encoder's, so most of the
distance between paper 3 and the encoder comes from the representation
rather than from the switch to a neural scorer.

**The effect is larger on test than it was on validation**, and nothing was
tuned between the two. On the validation slice the encoder beat the
hand-built arm by
`r fmt_sci(p5_ct(p5_rung3_vs_rung1, "P1_rank")$diff_point)` on P1_rank
before the layoff term was restored; on test the corresponding figure is
`r fmt_sci(p5_ct(p5_test_contrasts, "P1_rank", "rung 3b.*rung 1")$diff_point)`.
This is the same model measured on a different and larger sample, not an
improvement in it.

### Against the market

The encoder reaches P1_rank
`r fmt_metric(p5_ct(p5_test_contrasts, "P1_rank", "rung 3b.*rung 1")$point_a)`
and Brier_place
`r fmt_metric(p5_ct(p5_test_contrasts, "Brier_place", "rung 3b.*rung 1")$point_a)`.
The over-round-adjusted starting price, on the baseline papers 2b and 3
published, reaches about `r sprintf("%.5f", market_p1)` and
`r sprintf("%.4f", market_brier)`.

The gap narrows on both measures and closes on neither. No market contrast
was run in this paper, so nothing stronger than that comparison of published
figures is claimed. Paper 4 already established the relevant result
directly: against the settled starting price this series' models add nothing
distinguishable, and their value sits in the interval between an early price
and the off.

## Betting

The single-bet-per-race rule is the series' own, and it selected nothing in
this paper: no configuration, epoch, feature or arm was chosen on a return.
Among horses clearing the naive filter — model win probability above 0.15
and a model-to-market ratio above 1.3 — the rule backs the one with the
highest model win probability. Win bets settle at starting price; place and
each-way settle from the real-price payout tables.

```{r}
#| label: tbl-backtest
#| tbl-cap: "Single-bet-per-race backtest on the held-out test split."
p5_backtest_table(
  p5_test_backtests,
  c(rung1 = "Summaries", rung3b = "Encoder", paper3 = "Paper 3")
) |>
  dplyr::arrange(factor(Bet, levels = c("win", "place", "each-way")),
                 factor(Arm, levels = c("Encoder", "Summaries", "Paper 3"))) |>
  kable(align = "llrrrrr")
```

No arm makes money. The encoder is the best of the three on win
(`r fmt_pct(p5_bt(p5_test_backtests, "rung3b", "win")$roi)`) and on each-way
(`r fmt_pct(p5_bt(p5_test_backtests, "rung3b", "eachway")$roi)`), and its
win return is the best in the series to date.

**It is also worst of the two paper-5 arms on place**, at
`r fmt_pct(p5_bt(p5_test_backtests, "rung3b", "place")$roi)` against the
hand-built arm's
`r fmt_pct(p5_bt(p5_test_backtests, "rung1", "place")$roi)`, despite having
the better Brier_place. The two measures are not in conflict: Brier_place
scores calibration across every runner in the field, while the bet rule
picks the single highest model win probability that clears a market filter,
so they are computed on different quantities over different subsets. The
result is recorded as it stands.

### Intervals on the returns

Every other contrast in this paper carries an interval, and the returns
should too — a ten-point difference reported alone invites a reading its
precision does not support. The per-race bet units were not stored by the
backtest, only the summaries, so they are re-derived from the stored
predictions through the same chain the backtest used and checked to
reproduce the published returns and bet counts exactly before anything is
attached to them. No model was refitted and no test row rescored.

```{r}
#| label: tbl-roi-ci
#| tbl-cap: "Paired race-level bootstrap on single-bet ROI, over the common test races, B = 2000, seed 42, 90% intervals. Differences in percentage points."
p5_roi_contrasts |>
  dplyr::transmute(
    Contrast = dplyr::recode(
      contrast,
      "rung 3b (encoder) - rung 1 (summaries)" = "Encoder − Summaries",
      "rung 3b (encoder) - paper 3 (GBT)" = "Encoder − Paper 3",
      "rung 1 (MLP) - paper 3 (GBT)" = "Summaries − Paper 3"
    ),
    Bet = dplyr::recode(bet, win = "win", place = "place",
                        eachway = "each-way"),
    A = purrr::map_chr(roi_a, \(x) fmt_pct(x, 2)),
    B = purrr::map_chr(roi_b, \(x) fmt_pct(x, 2)),
    Difference = purrr::map_chr(diff_point, fmt_pts),
    `90% interval` = purrr::map2_chr(ci_lo, ci_hi, fmt_pts_ci),
    `Excludes zero` = dplyr::if_else(excludes_zero, "yes", "")
  ) |>
  kable(align = "llrrrrl")
```

Eight of the nine intervals contain zero, and they are wide — roughly five
to fourteen points either side. The headline betting comparison is among
them: the encoder's
`r fmt_pts(p5_roi(p5_roi_contrasts, "rung 3b.*paper 3", "win")$diff_point)`
points of win ROI over paper 3 carries an interval of
`r fmt_pts_ci(p5_roi(p5_roi_contrasts, "rung 3b.*paper 3", "win")$ci_lo, p5_roi(p5_roi_contrasts, "rung 3b.*paper 3", "win")$ci_hi)`.
A difference of that size on about
`r p5_comma(p5_bt(p5_test_backtests, "rung3b", "win")$n_bets)` single bets,
where the payoff is a roughly sevenfold return landing on about
`r sprintf("%.0f%%", 100 * p5_bt(p5_test_backtests, "rung3b", "win")$n_wins / p5_bt(p5_test_backtests, "rung3b", "win")$n_bets)`
of them, does not resolve at this sample size. **No betting improvement in
this paper is distinguishable from no improvement at all.**

One interval does exclude zero, and it is recorded here rather than left in
the table to be found:

> The hand-built arm's place ROI is
> `r fmt_pts(p5_roi(p5_roi_contrasts, "rung 1.*paper 3", "place")$diff_point)`
> points above paper 3's, interval
> `r fmt_pts_ci(p5_roi(p5_roi_contrasts, "rung 1.*paper 3", "place")$ci_lo, p5_roi(p5_roi_contrasts, "rung 1.*paper 3", "place")$ci_hi)`.
> That is the first ROI interval in this series to exclude zero.

Four things belong beside it, and together they are why the paper does not
build on it. It is one of nine intervals with no multiplicity control, and
none was pre-registered; nine 90% intervals under a global null would
produce at least one exclusion more often than not. It does not involve the
paper's model — the encoder's own place contrast against paper 3 is
`r fmt_pts(p5_roi(p5_roi_contrasts, "rung 3b.*paper 3", "place")$diff_point)`
points with an interval of
`r fmt_pts_ci(p5_roi(p5_roi_contrasts, "rung 3b.*paper 3", "place")$ci_lo, p5_roi(p5_roi_contrasts, "rung 3b.*paper 3", "place")$ci_hi)`,
containing zero. Both arms lose money, so it is a smaller loss rather than a
profit. And the test split is now closed, so the honest route to promoting
it would be a pre-registered test on data this paper has not used.
