# What the encoder is doing {#sec-discussion}

The result says the representation matters. It does not say what the encoder
found. Two diagnostics narrow that, both on the validation slice, both
descriptive. **No interval is attached to either and neither is a tested
claim.** They describe the question rather than settle it.

## How much history the gain needs

The selected configuration was refitted with the sequence truncated to the
five, ten and fifteen most recent runs — same grid position, same seed, same
slice, only the number of runs the encoder can read changing.

```{r}
#| label: tbl-seqlen-sens
#| tbl-cap: "Sequence length against validation Plackett–Luce loss. The gain column is against the hand-built arm."
p5_seq_length_table |>
  dplyr::transmute(
    `Prior runs` = max_len,
    `Mean length` = mean_len_fit,
    `% at the cap` = pct_at_cap_fit,
    `Best val PL loss` = purrr::map_chr(best_val_pl_loss, fmt_loss),
    `Gain over rung 1` = purrr::map_chr(gain_over_rung1, fmt_gain),
    `Share of the full gain` = sprintf("%.1f%%", share_of_full_gain)
  ) |>
  kable(align = "rrrrrr")
```

Five runs keep
`r sprintf("%.1f%%", p5_seqlen(p5_seq_length_table, 5)$share_of_full_gain)`
of the encoder's advantage. Going from five to twenty — quadrupling the
history available — buys the remaining
`r sprintf("%.1f%%", 100 - p5_seqlen(p5_seq_length_table, 5)$share_of_full_gain)`.
The direction is consistent and monotone, and the size is small.

Paper 3 sees two prior runs. On this evidence the encoder's advantage is
about **reading the recent runs properly** rather than about depth of
history or structure across a long span. What "properly" means is visible in
@tbl-example-form: the raw finishing position instead of a top-four code,
with the beaten distance, the going, the class, the field size and the
distance of each race attached to it. A ninth of fourteen beaten six lengths
and a twentieth of twenty beaten forty are the same number to paper 3's
feature set and different rows to the encoder.

## Whether it is re-encoding what it replaced

If the encoder were simply computing the discarded summaries for itself, its
output would reconstruct them. Each removed feature was regressed on the
encoder's `r p5_encoder_diag$hidden_dim` output dimensions, on validation
rows.

```{r}
#| label: tbl-recovery
#| tbl-cap: "Linear recovery of the removed features from the encoder's output. Reference rows are terms the encoder arm kept, and were never required to encode."
dplyr::bind_rows(
  p5_encoder_diag$recovery |>
    dplyr::transmute(Feature = feature, `R²` = fmt_r2(r_squared),
                     n = p5_comma(n), Role = "removed"),
  p5_encoder_diag$reference |>
    dplyr::transmute(Feature = feature, `R²` = fmt_r2(r_squared),
                     n = p5_comma(n), Role = "reference (kept)")
) |>
  kable(align = "lrrl")
```

The reference rows set the floor: a
`r p5_encoder_diag$hidden_dim`-dimension summary of a horse's history
correlates with most things about that horse, and terms the encoder never
had to represent sit at
`r fmt_r2(min(p5_encoder_diag$reference$r_squared))`–`r fmt_r2(max(p5_encoder_diag$reference$r_squared))`.

Against that floor, only one removed feature is substantially recovered:
whether the horse finished outside the first four last time, at
`r fmt_r2(p5_rec(p5_encoder_diag$recovery, "pos_lag1_zero"))`. Its
two-runs-back counterpart follows at
`r fmt_r2(p5_rec(p5_encoder_diag$recovery, "pos_lag2_zero"))`. Everything
else lies between
`r fmt_r2(min(p5_encoder_diag$recovery$r_squared))` and
`r fmt_r2(p5_rec(p5_encoder_diag$recovery, "pos_lag1_nonzero"))`, and three
of the eight sit at or below the reference floor. The going-affinity term
paper 3 built for this purpose, `going_sr_delta`, is at
`r fmt_r2(p5_rec(p5_encoder_diag$recovery, "going_sr_delta"))`.

The striking row is `pos_lag1_nonzero` at
`r fmt_r2(p5_rec(p5_encoder_diag$recovery, "pos_lag1_nonzero"))`. Last
time's finishing position is channel one of the sequence's most recent slot,
and a linear read of the encoder's output recovers under a third of its
variance.

**The caveat governs the whole table: these are linear $R^2$.** A low value
means the feature is not *linearly* recoverable from those dimensions, not
that the information is absent — a recurrent state is free to hold it in a
form no regression will find. What the table supports is the narrow claim
that the encoder is not a linear re-encoding of the summaries it replaced.
It does not establish what else the encoder is representing, and no stronger
reading should be taken from it.

# Rungs that were not run {#sec-declined}

Two further rungs were specified in advance and deliberately not run. Both
were dropped for reasons that would not have been resolved by running them.

**Embeddings alongside the strike rates rather than in place of them.** The
obvious response to rung 2 is to keep both, so the rare-bucket problem stops
being fatal: an entity outside the vocabulary would still carry its strike
rate. The difficulty is what remains for the embedding to do. It can only
contribute where an entity appears often enough in the fitting partition to
be worth a vector, and those are exactly the entities whose strike rate is
computed from the most runs and is therefore already reliable. The rung
would test whether an embedding adds anything on the rows where the
hand-built feature is strongest, which is not the question rung 2 raised.
The response that does address it is pre-training the embeddings on the full
archive rather than on the in-scope rows — which means fitting on races
outside this paper's scope, a change to the series' data scope rather than a
rung on its ladder.

**A Deep Sets encoder over today's field** [@zaheer2017]. Every rung in this
paper encodes one horse's own history. A field encoder would instead read
the other runners in today's race — a permutation-invariant summary of who
else is in it — and give each horse a term for the company it keeps. The
objection is arithmetic. Such a model is fitted at race grain, not
runner-race grain: `r p5_comma(n_train_races)` training races against
`r p5_comma(n_train_rows)` runner-rows, an order of magnitude less data than
every other rung in the paper, and it introduces a new architecture at the
same time. The ladder's rule is one change a rung, and this would be two on a
tenth of the sample.

# Limitations {#sec-limitations}

**The handicapper has already compressed the same history.** `or_relative`
— the horse's official rating relative to its race — stays in both arms, and
an official rating is the handicapper's own summary of the horse's form,
revised after every run, from a wider base of races than this paper models
and with knowledge the feed does not carry. The encoder is therefore
competing for the residual: whatever remains in a horse's run history once a
professional assessment of that history is already in the model. The gain
reported here is a gain on top of that, which makes it more interesting and
also bounds it.

**Going barely varies in scope.** All-Weather surfaces are consistent by
design and the in-scope races are overwhelmingly Standard going, so the
going channel does most of its work on the cross-surface part of a horse's
history — its turf and jumps runs — rather than on the races being
predicted. A going-affinity signal has little room to operate here, which is
consistent with both paper 3's finding and the low recovery of
`going_sr_delta` above, and says nothing about how it would behave on turf.

**These models overfit this data quickly.** Across three rungs the best
validation epoch arrived within about three passes over the training data,
and the loss then rose monotonically — in the worst configuration by more
than two nats per race by epoch 200. Early stopping on a held-out slice is
doing a great deal of the work. That is a statement about how little
`r p5_comma(n_train_races)` training races is for models of this size, and
it is the strongest reason to be cautious about extending the architecture
before extending the data.

**The twenty-run cap is generous and nothing hangs on it.** Paper 2b already
found that a third prior-run lag added nothing, and @tbl-seqlen-sens shows
the encoder's gain is essentially complete by five runs. The cap was set
wide to avoid making the truncation a live parameter, and the sensitivity
analysis confirms it never became one.

**The backend is fixed and the comparisons are internal.** All fitting is
CPU `libtorch` in double precision. Results do not reproduce across
backends, so every number here belongs to that one; @sec-appx-c records what
would be needed to reproduce it.

**One measure moved the wrong way.** The encoder has the better
Brier_place and the worse place ROI of the two paper-5 arms. That is
recorded in @sec-results as it stands rather than resolved.

# Conclusion {#sec-conclusion}

Replacing hand-built summaries of a horse's recent form with a sequence
encoder that reads the runs themselves improves ranking on held-out data,
by margins whose intervals exclude zero on every measure tested, and by more
than the switch from trees to a neural scorer buys. After four papers in
which no model could be distinguished from another on this test split, three
now order strictly.

The ladder that got there produced two results of its own. A multi-layer
perceptron beat a linear scorer on identical features, which paper 3's
conclusion about function class did not anticipate and which that paper's
evidence did not cover. And learned entity embeddings lost to the strike
rates they replaced, cleanly and without remediation, because a hand-built
feature computed from a wider information set than the model can be fitted
on is a hard thing for a learned representation to beat.

What has not changed is the market. It ranks better than the encoder, as it
ranked better than everything before it, and no betting return in this paper
is distinguishable from no improvement at all. The representation of a
horse's history was a real constraint and is now a smaller one. It was not
the constraint that separates these models from the price.
