# What the encoder is doing {#sec-discussion}

The result says the representation matters. It does not say what the encoder
found. Two diagnostics narrow that, both on the validation slice, both
descriptive.

## How much history the gain needs

The selected configuration was refitted with the sequence truncated to the
five, ten and fifteen most recent runs — same seed, same slice, only the
number of runs the encoder can read changing.

```{r}
#| label: tbl-seqlen-sens
#| tbl-cap: "Sequence length against validation Plackett–Luce loss. The gain column is against the MLP on hand-built features."
p5_seq_length_table |>
  dplyr::transmute(
    `Prior runs` = max_len,
    `Mean length` = mean_len_fit,
    `% at the cap` = pct_at_cap_fit,
    `Best val PL loss` = purrr::map_chr(best_val_pl_loss, fmt_loss),
    `Gain over rung 1` = purrr::map_chr(gain_over_rung1, fmt_gain),
    `Share of the full gain` = sprintf("%.1f%%", share_of_full_gain)
  ) |>
  kable(align = "rrrrrr")
```

Five runs keep
`r sprintf("%.1f%%", p5_seqlen(p5_seq_length_table, 5)$share_of_full_gain)`
of the encoder's advantage, and quadrupling the history available buys the
rest. Paper 3 sees two prior runs. The advantage is therefore about reading
the recent runs properly — the raw finishing position instead of a top-four
code, with the beaten distance, going, class and field size attached to it —
rather than about depth of history.

## Whether it is re-encoding what it replaced

If the encoder were simply computing the discarded features for itself, its
output would reconstruct them. Each removed feature was regressed on the
encoder's `r p5_encoder_diag$hidden_dim` output dimensions, on validation
rows.

```{r}
#| label: tbl-recovery
#| tbl-cap: "Linear recovery of the removed features from the encoder's output. Reference rows are features the encoder model kept, and was never required to encode."
dplyr::bind_rows(
  p5_encoder_diag$recovery |>
    dplyr::transmute(Feature = feature, `R²` = fmt_r2(r_squared),
                     n = p5_comma(n), Role = "removed"),
  p5_encoder_diag$reference |>
    dplyr::transmute(Feature = feature, `R²` = fmt_r2(r_squared),
                     n = p5_comma(n), Role = "reference (kept)")
) |>
  kable(align = "lrrl")
```

The reference rows set the floor: features the encoder never had to
represent sit at
`r fmt_r2(min(p5_encoder_diag$reference$r_squared))`–`r fmt_r2(max(p5_encoder_diag$reference$r_squared))`.
Against that floor only one removed feature is substantially recovered —
whether the horse finished outside the first four last time, at
`r fmt_r2(p5_rec(p5_encoder_diag$recovery, "pos_lag1_zero"))`. Last time's
finishing position itself, which is the first of the eight values in the
most recent slot, comes out at
`r fmt_r2(p5_rec(p5_encoder_diag$recovery, "pos_lag1_nonzero"))`.

These are linear $R^2$, so a low value means the feature is not *linearly*
recoverable from those dimensions, not that the information is absent. What
the table supports is the narrow claim that the encoder is not a linear
re-encoding of the features it replaced.

# Conclusion {#sec-conclusion}

Replacing hand-built summaries of a horse's recent form with a sequence
encoder that reads the runs themselves improves ranking on held-out data,
by margins whose intervals exclude zero on every measure tested, and by more
than the switch from trees to a neural scorer buys. After four papers in
which no model could be distinguished from another on this test split, three
now order strictly.

Two other things were found along the way. A multi-layer perceptron beat a
linear scorer on identical features, which paper 3's conclusion about
function class did not anticipate and which that paper's evidence did not
cover. And learned entity representations lost to the strike rates they
replaced, cleanly and without remediation, because a hand-built feature
computed from a wider information set than the model can be fitted on is a
hard thing for a learned representation to beat.

Two limitations bound the result. `or_relative` — the horse's official
rating relative to its race — stays in both models, and an official rating
is the handicapper's own summary of the same history, revised after every
run from a wider base of races than this paper models. The encoder is
competing for what is left after that, which makes the gain more interesting
and also bounds it. And these models overfit quickly: the best epoch arrived
within about three passes over the training data in every rung, which says
how little `r p5_comma(n_train_races)` training races is for models of this
size, and is the strongest reason to be cautious about extending the
architecture before extending the data.

All fitting was done on the CPU `libtorch` build in double precision.
Results do not reproduce across backends, so every number here belongs to
that one; @sec-appx-c records what reproducing it requires.

What has not changed is the market. It ranks better than the encoder, as it
ranked better than everything before it, and no betting return in this paper
is distinguishable from no improvement at all. The representation of a
horse's history was a real constraint and is now a smaller one. It was not
the constraint that separates these models from the price.
