# Appendix B — Sequence construction {#sec-appx-b}

## Schema

For each runner-race the builder produces a fixed-shape array plus two
counts:

| object | shape | what it is |
|---|---|---|
| `x` | `r p5_comma(dim(p5_sequences$x)[1])` × `r dim(p5_sequences$x)[2]` × `r dim(p5_sequences$x)[3]` | prior runs × values per run, most recent first |
| `seq_len` | `r p5_comma(dim(p5_sequences$x)[1])` | real runs in that row, capped |
| `career_runs_prior` | `r p5_comma(dim(p5_sequences$x)[1])` | untruncated career length |

Prior runs come from the series' cached cross-surface career history — every
race code and every surface, not All-Weather alone — under the series'
temporal rule: strictly before the meeting date of the race being predicted.
Declared non-runners are dropped, matching the treatment the going features
already give them: a declared non-runner did not run, so it is not a prior
run.

## The cap, and why nothing is lost at it

The array is capped at `r p5_sequences$max_len` runs. Career length is
counted **before** the cap and carried separately as `career_runs_prior`, so
a horse with fifty prior runs is not presented as one with twenty. On the
training split
`r sprintf("%.1f%%", p5_sequence_summary$truncated$share_pct)` of
observations exceed the cap.

The cap was set wide so that its value would not become another setting
requiring selection. @tbl-seqlen-sens confirms it never did: the encoder's
advantage is essentially complete by five runs.

## Cleaning

Two treatments, chosen by whether the reading is on its own scale.

**Read as unavailable, taking −1**, when the value is not on the scale at
all, so its magnitude would be meaningless:

- `class` outside 1–7. The 2006 restructuring replaced multi-digit codes
  whose meaning is not comparable with the current scheme, and those values
  are entirely pre-2006.
- `finish_pos` beyond 40. Values of 98 and 99 are sentinels, not placings.

**Capped**, when the value is on the scale but its exact magnitude stops
being informative, so the ordering is worth keeping: `field_size` at 40, the
largest legitimate British field; `days_since_prev` at 1,000, past which the
exact gap carries nothing and against which the largest observed value of
37,918 days is plainly not a real layoff; and `beaten_dist` at
`r p5_beaten_audit$bound` lengths.

−1 rather than 0 marks an unavailable reading because zero is legitimate for
several of these values — a winner is zero lengths behind the winner, which
is definitional rather than imputed.

### The beaten-distance cap

`beaten_dist` is the only cap that touches an appreciable share of rows, so
it is documented rather than asserted. Over
`r p5_comma(p5_beaten_audit$n_rows)` career-run rows,
`r p5_comma(p5_beaten_audit$n_missing)`
(`r sprintf("%.3f%%", p5_beaten_audit$pct_missing)`) have no reading at all.

```{r}
#| label: tbl-beaten-thresholds
#| tbl-cap: "Rows above each beaten-distance threshold, whole career-run universe."
p5_beaten_audit$thresholds |>
  kable(align = "rrr")
```

The tail is real data, not a sentinel: the value is cumulative lengths
behind the winner, and the long tail is jumps racing.

```{r}
#| label: tbl-beaten-racetype
#| tbl-cap: "Beaten distance by race type."
p5_beaten_audit$by_race_type |>
  kable(align = "lrrrrr")
```

A tailed-off horse in a hurdle race genuinely finishes hundreds of lengths
adrift, so the reading is capped rather than discarded — the ordering still
means "beaten a very long way", and −1 would misreport a real result as
absent. Beyond `r p5_beaten_audit$bound` lengths the exact figure adds
nothing. The cap also does the most for the scale the encoder sees: before
it, the largest value sits about 23 standard deviations above the mean;
after it, about 6.
