# Appendix B — Sequence construction {#sec-appx-b}

## Schema

For each runner-race the builder produces a fixed-shape array plus two
scalars:

| object | shape | what it is |
|---|---|---|
| `x` | `r p5_comma(dim(p5_sequences$x)[1])` × `r dim(p5_sequences$x)[2]` × `r dim(p5_sequences$x)[3]` | prior runs × channels, most recent first, right-padded |
| `seq_len` | `r p5_comma(dim(p5_sequences$x)[1])` | real runs in that row, capped |
| `career_runs_prior` | `r p5_comma(dim(p5_sequences$x)[1])` | untruncated career length |

Prior runs come from the series' cached cross-surface career history —
every race code and every surface, not All-Weather alone — under the
series' temporal rule: strictly before the meeting date of the race being
predicted. Declared non-runners are dropped, matching the treatment the
going features already give them: a declared non-runner did not run, so it
is not a prior run.

## The cap, and why nothing is lost at it

The array is capped at `r p5_sequences$max_len` runs. Career length is
counted **before** the cap and carried separately as `career_runs_prior`, so
a horse with fifty prior runs is not presented as one with twenty. On the
training split
`r sprintf("%.1f%%", p5_sequence_summary$truncated$share_pct)` of
observations exceed the cap.

The cap was set wide deliberately, so that its value would not become a
parameter requiring selection. @tbl-seqlen-sens confirms it never did: the
encoder's advantage is essentially complete by five runs.

## Channel cleaning

Two treatments, and the distinction between them is the whole of the
reasoning.

**Read as unavailable (−1)** when the value is not on the channel's own
scale, so its magnitude would be meaningless:

- `class` outside 1–7. The 2006 restructuring replaced multi-digit codes
  whose meaning is not comparable with the current scheme, and those values
  are entirely pre-2006. A class of 42 is not "class 42".
- `finish_pos` beyond 40. Values of 98 and 99 are sentinels, not placings.

**Winsorised** when the value *is* on the channel's scale but its exact
magnitude stops being informative, so the ordering is worth keeping:

- `field_size` at 40, the largest legitimate British field.
- `days_since_prev` at 1,000. Past about three years the exact gap carries
  nothing an encoder can use, and the maximum observed is 37,918 days, which
  is not a real layoff.
- `beaten_dist` at `r p5_beaten_audit$bound` lengths.

The unavailable fill is **−1 rather than 0** because zero is a legitimate
reading for several channels — a winner is zero lengths behind the winner,
which is definitional rather than imputed — and because zero is the padding
value. The two must not collide.

### The beaten-distance bound

`beaten_dist` is the only bound that touches an appreciable share of rows,
so it is documented rather than asserted. Over
`r p5_comma(p5_beaten_audit$n_rows)` career-run rows,
`r p5_comma(p5_beaten_audit$n_missing)`
(`r sprintf("%.3f%%", p5_beaten_audit$pct_missing)`) have no reading at all.

```{r}
#| label: tbl-beaten-thresholds
#| tbl-cap: "Rows above each beaten-distance threshold, whole career-run universe."
p5_beaten_audit$thresholds |>
  kable(align = "rrr")
```

The tail is real data, not a sentinel: the column is cumulative lengths
behind the winner, and the long tail is jumps racing.

```{r}
#| label: tbl-beaten-racetype
#| tbl-cap: "Beaten distance by race type."
p5_beaten_audit$by_race_type |>
  kable(align = "lrrrrr")
```

A tailed-off horse in a hurdle race genuinely finishes hundreds of lengths
adrift, so the reading is winsorised rather than discarded — the ordering
still means "beaten a very long way", and −1 would misreport a real result
as absent. Beyond `r p5_beaten_audit$bound` lengths the exact figure adds
nothing: the horse was comprehensively beaten either way. The bound also
does the most for the scale the encoder sees. Before it, the channel maximum
sits about 23 standard deviations above the channel mean; after it, about 6.

## Padding and masking

Sequences are **right-padded**: real runs occupy the leading slots and
padding follows. The encoder's output is the recurrent state at slot
`seq_len`, so the recurrence never consumes a padded slot.

Left-padding would not do, and the reason is not obvious enough to leave
unstated. A gated recurrent unit fed $k$ zero vectors from a zero initial
state does not remain at zero, because the update and reset gates carry
biases. The state entering the first real run would then depend on how much
padding preceded it, and two horses with identical records but different
career lengths would receive different summaries.

Runs are reversed into chronological order before the encoder reads them, so
the last thing it sees is the most recent run.

A row with no prior run at all has no last real slot; it receives the zero
vector, which is what a zero-length sequence should give and is the fact the
discarded "no prior run" indicator carried.

### The masking gate

`scripts/verify_p5_gru_mask.R` asserts the property directly rather than
relying on the argument above, on a randomly initialised encoder in
evaluation mode, across three seeds. It is also on the pipeline graph, so
it runs whenever the encoder code changes.

```{r}
#| label: tbl-mask-gate
#| tbl-cap: "The masking gate. Differences are between encoder outputs that must be identical."
p5_gru_mask_check |>
  dplyr::transmute(Check = check,
                   `Max abs difference` = formatC(max_abs_diff,
                                                  format = "e", digits = 2),
                   Tolerance = formatC(tol, format = "e", digits = 0),
                   Passed = dplyr::if_else(passed, "yes", "NO")) |>
  kable(align = "lrrl")
```

The third check is the one that earns its place. An off-by-one in the index
that selects the last real slot would summarise the wrong run while leaving
both padding-invariance checks intact — the encoder would be perfectly
consistent about reading the wrong thing. Feeding each row's real runs with
no padding at all is what pins the index.
