# Extended win model {#sec-extended-win}

This section fits the conditional-logit win model on the extended
feature set and evaluates it out-of-sample. The recipe parallels paper
1 exactly — fit the full specification on the training races, then drop
every term with $p > 0.05$ and refit to a reduced headline model — but
two things change. The predictor set is wider (the retained paper-2
features from @sec-eda), and the prior-finish lags now use the
**zero-plus-slope** encoding: for each lag $N$, a zero indicator `pos_lagN_zero` (no
prior run, or finished worse than 4th) plus the graded raw position
`pos_lagN_nonzero` (1–4). The race scope, train/test split (70/30
chronological, cutoff 2012-12-30) and diagnostic suite are unchanged
from paper 1, so the two models are directly comparable. Fitting is
training-only; every evaluation below is on the held-out test races.

```{r}
#| label: p2win-setup
#| include: false
# The mlogit fits are loaded from targets, not fitted in this document,
# so mlogit's S3 methods (summary.mlogit / coef.mlogit) are not on the
# dispatch table yet — load the namespace so summary() below dispatches
# correctly rather than falling back to summary.default (CLAUDE.md note).
loadNamespace("mlogit")
rc <- stats::coef(model_p2_reduced)
cf <- function(n) if (n %in% names(rc)) sprintf("%+.3f", rc[[n]]) else "dropped"
pl_r2_reduced <- model_p2_diagnostics$pl_r2[model_p2_diagnostics$model == "reduced"]
```

## Full and reduced fit

The full specification has 19 terms. The reduced model drops any term
with $p > 0.05$ and refits (iterating the elimination once); on these
data only the lag-3 position pair falls away, leaving 17 coefficients.

```{r}
#| label: tbl-reduced
#| code-fold: true

reduced_ct <- summary(model_p2_reduced)$CoefTable

block_of <- function(term) {
  if (grepl("^pos_lag", term))                                  return("1. Position encoding (zero-plus-slope)")
  if (term %in% c("days_LTO_log", "trainerSR", "sireSR"))       return("2. Form / history (Owen)")
  if (term %in% c("age_diff", "entire", "gelding", "cheekpieces")) return("3. Horse characteristics (Owen)")
  "4. New features (paper 2)"
}

pretty_term <- function(term) dplyr::recode(term, days_LTO_log = "daysLTO (log)")

reduced_tbl <- tibble::tibble(
  term      = rownames(reduced_ct),
  estimate  = reduced_ct[, "Estimate"],
  std_error = reduced_ct[, "Std. Error"],
  z_value   = reduced_ct[, "z-value"],
  p_value   = reduced_ct[, "Pr(>|z|)"]
) |>
  dplyr::mutate(
    sig = dplyr::case_when(
      p_value < 0.001 ~ "***",
      p_value < 0.01  ~ "**",
      p_value < 0.05  ~ "*",
      TRUE            ~ ""
    ),
    Block = factor(vapply(term, block_of, character(1)),
                   levels = c("1. Position encoding (zero-plus-slope)",
                              "2. Form / history (Owen)",
                              "3. Horse characteristics (Owen)",
                              "4. New features (paper 2)"))
  ) |>
  dplyr::arrange(Block) |>
  dplyr::transmute(
    Block,
    Term          = pretty_term(term),
    Estimate      = round(estimate, 4),
    `Std. error`  = round(std_error, 4),
    `z`           = round(z_value, 2),
    `p-value`     = signif(p_value, 3),
    Sig           = sig
  )

knitr::kable(
  reduced_tbl,
  align   = c("l", "l", "r", "r", "r", "r", "c"),
  caption = "Reduced extended win model (AW training fit), zero-plus-slope
position encoding. Coefficients grouped by block; significance ***p<0.001,
**p<0.01, *p<0.05. The position lags enter as the semi-parsimonious
zero-plus-slope pair per lag (zero indicator + graded 1–4 value), not as
paper 1's factor grid."
)
```

```{r}
#| label: dropped-p2
full_terms    <- names(stats::coef(model_p2_full))
reduced_terms <- names(stats::coef(model_p2_reduced))
dropped_p2    <- setdiff(full_terms, reduced_terms)
```

**Terms dropped from the full model** ($p > 0.05$, removed before the
reduced refit): `r if (length(dropped_p2) == 0) "none" else paste(dropped_p2, collapse = ", ")`.
Only the third prior-finish lag falls out — the same lag paper 1 drops
(`position3`) — confirming that the third-back race adds nothing once
the two most recent are in the model.

```{r}
#| label: tbl-p2-fitstats
model_p2_diagnostics |>
  dplyr::transmute(
    Model         = model,
    `logLik`      = round(loglik, 1),
    `Null logLik` = round(null_loglik, 1),
    `McFadden R²` = round(pl_r2, 4),
    `Races`       = n_races,
    `Runner-rows` = n_runners
  ) |>
  knitr::kable(
    caption = "Full vs reduced extended model: training-set fit statistics.
McFadden R² is the pseudo-R² 1 − logLik / null logLik."
  )
```

```{r}
#| label: p2-race-loss
#| include: false
.tr      <- runners_model_ready |> dplyr::filter(split == "train")
.n_train <- dplyr::n_distinct(.tr$race_id)
.n_fit   <- model_p2_diagnostics$n_races[1]
.n_drop  <- .n_train - .n_fit
.race_touch <- function(col) {
  .tr |>
    dplyr::filter(is.na(.data[[col]])) |>
    dplyr::distinct(race_id) |>
    nrow()
}
.drop_jockey  <- .race_touch("jockeySR")
.drop_sire    <- .race_touch("sireSR")
.drop_trainer <- .race_touch("trainerSR")
.drop_days    <- .race_touch("days_LTO_log")
```

@sec-data reported the feature set as essentially complete, but the
conditional logit is fitted complete-case at the race level — a race is
dropped whole if any runner is missing any modelling column — so the fit
rests on `r scales::comma(.n_fit)` of the `r scales::comma(.n_train)`
training races, `r .n_drop` dropped
(`r scales::percent(.n_drop / .n_train, accuracy = 0.1)`). The losses sit
almost entirely in the pre-race strike rates, which are undefined under
the strict-before temporal rule when a runner's jockey, sire or trainer
has no qualifying prior run: a missing `jockeySR` touches
`r .drop_jockey` races, `sireSR` `r .drop_sire`, `trainerSR`
`r .drop_trainer`, and `days_LTO_log` just `r .drop_days` (debut
runners). The sets overlap, so these counts do not sum to `r .n_drop`.
The test set is filtered the same way, which is why paper 1 and paper 2a
evaluate on slightly different race universes — a point we return to when
comparing backtests in @sec-discussion.

```{r}
#| label: p2-reduction-test
#| include: false
ll_full  <- as.numeric(logLik(model_p2_full))
ll_red   <- as.numeric(logLik(model_p2_reduced))
k_full   <- length(coef(model_p2_full))
k_red    <- length(coef(model_p2_reduced))
lr_stat  <- 2 * (ll_full - ll_red)
lr_df    <- k_full - k_red
lr_p     <- pchisq(lr_stat, lr_df, lower.tail = FALSE)
aic_full <- 2 * k_full - 2 * ll_full
aic_red  <- 2 * k_red  - 2 * ll_red
aic_gap  <- aic_red - aic_full
r2_full  <- model_p2_diagnostics$pl_r2[model_p2_diagnostics$model == "full"]
r2_red   <- model_p2_diagnostics$pl_r2[model_p2_diagnostics$model == "reduced"]
```

Dropping the two lag-3 terms is a per-term decision — each was removed
on its own Wald *p*, the same rule paper 1 and Owen apply — and it is
worth checking what the *joint* test says, because two individually
weak terms can be jointly significant. A likelihood-ratio test of the
reduced model against the full one (both fitted on the identical
`r scales::comma(model_p2_diagnostics$n_races[1])` training races, so the
likelihoods are comparable) gives
$\text{LR} = 2(\ell_\text{full} - \ell_\text{red}) =
`r round(lr_stat, 1)`$ on `r lr_df` degrees of freedom,
*p* = `r scales::number(lr_p, accuracy = 0.001)`. The joint test
therefore **rejects** the lag-3 reduction at the 5% level — the two
terms carry a little signal together that neither shows alone — and AIC
agrees, preferring the full model by `r round(aic_gap, 1)`
(`r round(aic_red, 1)` vs `r round(aic_full, 1)`).

We nonetheless carry the *reduced* model forward as the paper-2a
headline, for two reasons. First, comparability: paper 1 and Owen both
report reduced models built with exactly this per-term Wald rule, and a
like-for-like progression table requires we do the same. Second, the
practical difference is negligible — McFadden pseudo-$R^2$ is
`r round(r2_full, 4)` for the full model against `r round(r2_red, 4)`
for the reduced, a gap of `r round(r2_full - r2_red, 4)`, and the
out-of-sample diagnostics below are indistinguishable between the two.
The reduction buys parsimony at a cost the data can barely register; we
flag the joint test here so the choice is explicit rather than hidden.

Every retained coefficient carries the sign its feature predicts. The
zero-plus-slope position terms are all negative: a missing or worse-than-4th
recent run (`pos_lag1_zero` = `r cf("pos_lag1_zero")`) is a sizeable
penalty, and within graded finishes a larger position number — a worse
placing — lowers the win odds (`pos_lag1_nonzero` =
`r cf("pos_lag1_nonzero")`). The effect decays across lags exactly as
the encoding intends: lag 1 dominates, lag 2 is weaker, lag 3 is gone.
`age_diff` (`r cf("age_diff")`) and `days_LTO_log` (`r cf("days_LTO_log")`)
are negative as in paper 1 (distance from peak age, and rust — the loss of
sharpness from time away from racing, captured by `days_LTO_log`). The
three strike rates are strongly positive, and the new features behave:
`rel_weight` (`r cf("rel_weight")`) is negative — carrying more than the
field's mean weight lowers the win probability, the handicap signal
paper 1 flagged as missing — and `or_relative` (`r cf("or_relative")`)
is the single most significant term in the model.

## Comparison with Owen and paper 1

Read against Owen's Table 3 and paper 1's AW reduced fit, the carried-
over Owen features land where expected and the new features reshape a
couple of them.

- **Age, sex, tack.** `age_diff` is negative (Owen −0.152; ours
  `r cf("age_diff")`), `entire` (`r cf("entire")`) and `gelding`
  (`r cf("gelding")`) positive, `cheekpieces` (`r cf("cheekpieces")`)
  negative — the same pattern as Owen and paper 1. As in paper 1,
  `gelding` is roughly half Owen's +0.548, the AW population being
  gelding-heavy so the contrast with the reference is compressed.

- **Strike rates attenuate — the main surprise.** Paper 1's AW reduced
  fit put `trainerSR` at +5.09 and `sireSR` at +2.36 (proportion
  scale); here they fall to `r cf("trainerSR")` and `r cf("sireSR")`.
  The cause is the new `jockeySR` (`r cf("jockeySR")`, itself larger
  than `trainerSR`) and `or_relative`, which are correlated with
  trainer/sire quality and absorb signal that paper 1 had no choice but
  to load onto the two strike rates. Divided by 100 for Owen's apparent
  percent scale, AW `trainerSR` now reads ~0.033 — below Owen's 0.051,
  where paper 1 matched him almost exactly.

- **New features earn their place.** `or_relative` is the dominant
  predictor (the most statistically significant term in the table) —
  unsurprising, since the
  official rating is the handicapper's direct estimate of relative
  ability, and the starting price paper 1 lost to is built on the same
  information. Its companion `or_missing` (`r cf("or_missing")`) is
  negative: unrated horses, controlling for everything else, win less.
  `rel_weight`, `jockeySR` and `trainer_aw_premium`
  (`r cf("trainer_aw_premium")`) are all significant; `has_wins`
  (`r cf("has_wins")`) survives only marginally.

## Out-of-sample evaluation: calibration

As in paper 1 we score the reduced model on the **test races** —
out-of-sample forecasts, not in-sample fits — and build market-implied
probabilities from the over-round-adjusted industry starting price (SP),
renormalised per race. The calibration plot bins predicted (or implied) win
probability in 5-point intervals over $[0, 0.50]$; perfect calibration
sits on $y = x$.

```{r}
#| label: p2-eval-setup
#| include: false
# Test set only. Rename so the paper-1 calibration / scoring idiom
# (which keys on `model_prob`) applies unchanged.
eval_data <- test_predictions_p2 |> dplyr::rename(model_prob = predicted_prob)
```

```{r}
#| label: fig-calibration-p2
#| fig-cap: "Calibration on the AW test set (Owen Figure 7 equivalent): observed
#|   win proportion vs predicted / implied win probability, 5-point bins over
#|   [0, 0.50]. Circles = extended reduced model (out-of-sample); plus signs =
#|   market. Dashed line y = x is perfect calibration."

bin_breaks <- seq(0, 0.50, by = 0.05)
bin_mids   <- head(bin_breaks, -1) + 0.025

calibration_summary <- function(df, prob_col, label) {
  df |>
    dplyr::filter(!is.na(.data[[prob_col]]),
                  .data[[prob_col]] >= 0,
                  .data[[prob_col]] <= 0.50) |>
    dplyr::mutate(bin = cut(.data[[prob_col]], breaks = bin_breaks,
                            include.lowest = TRUE, right = FALSE)) |>
    dplyr::group_by(bin) |>
    dplyr::summarise(n = dplyr::n(), observed = mean(won),
                     predicted_mu = mean(.data[[prob_col]]), .groups = "drop") |>
    dplyr::mutate(bin_mid = bin_mids[as.integer(bin)], source = label)
}

calib_summary <- dplyr::bind_rows(
  calibration_summary(eval_data, "model_prob",  "Model (reduced)"),
  calibration_summary(eval_data, "market_prob", "Market")
)

ggplot2::ggplot(calib_summary,
                ggplot2::aes(x = bin_mid, y = observed, shape = source, colour = source)) +
  ggplot2::geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey50") +
  ggplot2::geom_point(size = 3) +
  ggplot2::scale_shape_manual(values = c("Model (reduced)" = 1L, Market = 3L)) +
  ggplot2::scale_colour_manual(values = c("Model (reduced)" = "steelblue", Market = "tomato")) +
  ggplot2::scale_x_continuous(labels = scales::percent_format(accuracy = 1),
                              limits = c(0, 0.50), breaks = bin_breaks) +
  ggplot2::scale_y_continuous(labels = scales::percent_format(accuracy = 1), limits = c(0, NA)) +
  ggplot2::labs(x = "Predicted / implied win probability",
                y = "Observed win proportion", shape = NULL, colour = NULL) +
  ggplot2::theme_minimal()
```

## Out-of-sample evaluation: P1 and P2 scoring rules

The two scoring rules follow Owen's Figure 8, evaluated out-of-sample.
For each threshold $t$, over runner-rows whose probability exceeds $t$:
$P_1$ is the geometric mean of probability assigned to actual winners,
and $P_2$ the geometric mean of the per-row Bernoulli likelihood (a
proper scoring rule that also penalises high probability on losers). At any
given threshold, a higher $P_1$ or $P_2$ means the model (or the market) is
doing better at that threshold — though $P_2$ itself trends down as the
threshold rises, since it is then conditioned on a smaller, more selective
set of bets.

```{r}
#| label: p2-scoring-functions
#| include: false

add_loglik <- function(df, prob_col) {
  df |>
    dplyr::filter(!is.na(.data[[prob_col]]),
                  .data[[prob_col]] > 0, .data[[prob_col]] < 1) |>
    dplyr::mutate(
      ll_winner = ifelse(won == 1L, log(.data[[prob_col]]), NA_real_),
      ll_full   = ifelse(won == 1L, log(.data[[prob_col]]), log(1 - .data[[prob_col]]))
    )
}

score_at_threshold <- function(df, prob_col, threshold) {
  sub <- df[df[[prob_col]] > threshold, , drop = FALSE]
  if (nrow(sub) == 0L) {
    return(tibble::tibble(threshold = threshold, n = 0L, n_winners = 0L,
                          P1 = NA_real_, P2 = NA_real_))
  }
  winners <- sub[sub$won == 1L, , drop = FALSE]
  p1 <- if (nrow(winners) > 0L) exp(mean(winners$ll_winner, na.rm = TRUE)) else NA_real_
  p2 <- exp(mean(sub$ll_full, na.rm = TRUE))
  tibble::tibble(threshold = threshold, n = nrow(sub), n_winners = nrow(winners),
                 P1 = p1, P2 = p2)
}

scoring_summary <- function(df, prob_col, label, thresholds) {
  scored_df <- add_loglik(df, prob_col)
  purrr::map_dfr(thresholds, \(t) score_at_threshold(scored_df, prob_col, t)) |>
    dplyr::mutate(source = label)
}

thresholds <- seq(0, 0.50, by = 0.025)
score_long <- dplyr::bind_rows(
  scoring_summary(eval_data, "model_prob",  "Model (reduced)", thresholds),
  scoring_summary(eval_data, "market_prob", "Market",          thresholds)
)
```

```{r}
#| label: fig-p1-p2
#| fig-cap:
#|   - "P1 vs probability threshold (Owen Figure 8a equivalent), AW test set. Higher = more probability mass on eventual winners."
#|   - "P2 vs probability threshold (Owen Figure 8b equivalent), AW test set. Proper scoring rule: calibration plus discrimination."

p_score <- function(yvar, ylab) {
  ggplot2::ggplot(score_long,
                  ggplot2::aes(x = threshold, y = .data[[yvar]],
                               linetype = source, colour = source)) +
    ggplot2::geom_line(linewidth = 0.8) +
    ggplot2::scale_linetype_manual(values = c("Model (reduced)" = "solid", Market = "dashed")) +
    ggplot2::scale_colour_manual(values = c("Model (reduced)" = "steelblue", Market = "tomato")) +
    ggplot2::scale_x_continuous(labels = scales::percent_format(accuracy = 1), limits = c(0, 0.50)) +
    ggplot2::labs(x = "Probability threshold", y = ylab, linetype = NULL, colour = NULL) +
    ggplot2::theme_minimal()
}
print(p_score("P1", "P1 (geom. mean prob. on winners)"))
print(p_score("P2", "P2 (geom. mean Bernoulli likelihood)"))
```

## Betting backtest

We apply Owen's exact bet-selection rule to the AW test set: back a horse
when the model gives it more than a 15% chance of winning
($\hat{P} > 0.15$) and the model's probability is at least 1.3 times the
market-implied probability (ratio $> 1.3$) — that is, the model must be
both confident in absolute terms and clearly more bullish than the market.
The rule can qualify more than one horse in a race; Owen does not specify
whether he backs one qualifier or all, and here we back all (the multi-bet
rule). We then sweep the ratio
threshold from 0.9 to 2.0 with the model-probability filter held at 0.13
and a 90% bootstrap CI (race-level resampling, $B = 2000$, seed 42) —
the same procedure paper 1 ran, so the ROI is directly comparable to
paper 1's headline **−28.2%**.

```{r}
#| label: tbl-backtest-naive-p2
backtest_naive_p2 |>
  dplyr::transmute(
    `Prob threshold`  = prob_threshold,
    `Ratio threshold` = ratio_threshold,
    `Bets`            = n_bets,
    `Wins`            = n_wins,
    `Gross return`    = sprintf("%.2f", gross_return),
    `Profit`          = sprintf("%.2f", profit),
    `ROI`             = scales::percent(roi, accuracy = 0.1)
  ) |>
  knitr::kable(caption = "Extended reduced model: betting backtest on the AW test races at Owen's exact thresholds.")
```

```{r}
#| label: fig-roi-sweep-p2
#| fig-cap: "ROI vs ratio threshold on the AW test set (extended model), 90% bootstrap CI shaded. Model-probability filter held at 0.13; dashed line at zero marks break-even."
ggplot2::ggplot(backtest_sweep_p2, ggplot2::aes(x = tau, y = roi)) +
  ggplot2::geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  ggplot2::geom_ribbon(ggplot2::aes(ymin = ci_lo, ymax = ci_hi), fill = "steelblue", alpha = 0.20) +
  ggplot2::geom_line(linewidth = 0.8, colour = "steelblue") +
  ggplot2::scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  ggplot2::labs(x = "Ratio threshold (model_prob / market_prob)", y = "ROI") +
  ggplot2::theme_minimal()
```

```{r}
#| label: tbl-roi-sweep-p2
backtest_sweep_p2 |>
  dplyr::transmute(
    `Ratio threshold` = tau,
    `Bets`            = n_bets,
    `Wins`            = n_wins,
    `ROI`             = scales::percent(roi, accuracy = 0.1),
    `90% CI low`      = scales::percent(ci_lo, accuracy = 0.1),
    `90% CI high`     = scales::percent(ci_hi, accuracy = 0.1)
  ) |>
  knitr::kable(caption = "ROI sweep over ratio thresholds with 90% bootstrap CIs (B = 2000, race-level resampling, seed 42).")
```

## Does the extended feature set improve on paper 1?

The extended
reduced model reaches a McFadden pseudo-R² of
`r round(pl_r2_reduced, 3)` on the training races, comfortably above the
paper-1 specification — most of the lift coming from `or_relative`, the
handicapper's rating, together with `jockeySR` and `rel_weight`, none of
which Owen or paper 1 used. The naive backtest ROI improves from paper
1's −28.2% to
**`r scales::percent(backtest_naive_p2$roi, accuracy = 0.1)`** —
directionally better, but still firmly negative, and the ratio sweep's
90% CI sits below zero across the threshold range.

The framing from paper 1 holds. On calibration and the P1/P2 scoring
rules the model has closed some of the gap to the market but has not
overtaken it: adding `or_relative` pulls the feature set onto part of
the information the starting price already prices, which is exactly why
discrimination improves, yet the SP still aggregates more — weight-by-
draw interactions, going suitability — than even the
widened feature set captures. The extended *win* model is a better
forecaster than paper 1's; whether *race-level interactions* narrow the
remaining gap is the question for @sec-mixed-interactions. (Whether
changing the objective from the win to the full finishing order helps is
the subject of the companion paper 2b.)
