# Mixed race-level interactions {#sec-mixed-interactions}

The features so far are all horse-level. Race-level features — going,
distance, course — cannot enter the conditional logit directly: they are
constant within a race and cancel in the softmax. They enter only as
**interactions** with horse-level features, whose products vary across
runners within a race (the mixed discrete-choice mechanism derived in
@sec-appx-derivations). This section tests two such interactions on top
of the extended win model from @sec-extended-win, which is the baseline
throughout (**Model W**).

```{r}
#| label: p2int-setup
#| include: false
loadNamespace("mlogit")

ed_ct    <- summary(model_w_ed)$CoefTable      # 4-course block fit
fin_ct   <- ed_ct                              # the final paper-2a model IS the full draw block
fin_coef <- stats::setNames(fin_ct[, "Estimate"], rownames(fin_ct))
roi_pct  <- function(x) scales::percent(x, accuracy = 0.1)

# LR p-values pulled by test, so the prose tracks the fit.
lr_p <- function(a, b) {
  interaction_lr_tests_w |>
    dplyr::filter(model_a == a, model_b == b) |>
    dplyr::pull(p_value)
}
p_weight          <- lr_p("EW",  "E")    # weight x distance vs baseline
p_draw            <- lr_p("ED",  "E")    # draw x course vs baseline
p_weight_giv_draw <- lr_p("EWD", "ED")   # weight | draw
```

## Two interactions

**Weight × distance** (`rel_weight_x_dist = rel_weight × distance_furlongs`).
`rel_weight` is horse-level, `distance_furlongs` race-level; their product
varies across runners. The motivation is @benter1994: the burden of extra
weight should compound over longer trips, so a horse carrying weight above
the field mean ought to be penalised more in a staying race than a sprint.

**Draw × course.** Draw bias on All-Weather is known to depend on the
track — Kempton, Lingfield, Southwell and Wolverhampton have different
configurations, bend tightness and run-ins. `stall_normalised` had almost
no standalone signal (@sec-eda), but that pooled estimate could be hiding
course-specific effects that cancel. We split it into four course-masked
columns (`stall_normalised` where the race is at that course, else 0),
one coefficient per course.

```{r}
#| label: tbl-interaction-features
# Training set only — see CLAUDE.md exploratory analysis convention
ri_train <- runners_interactions |> dplyr::filter(split == "train")
tibble::tibble(
  Quantity = c("Distance range (furlongs)",
               "rel_weight_x_dist SD",
               "Runners with a Kempton draw term",
               "Runners with a Lingfield draw term",
               "Runners with a Southwell draw term",
               "Runners with a Wolverhampton draw term"),
  Value = c(
    paste(round(min(ri_train$distance_furlongs), 1), "–",
          round(max(ri_train$distance_furlongs), 1)),
    sprintf("%.1f", stats::sd(ri_train$rel_weight_x_dist, na.rm = TRUE)),
    scales::comma(sum(ri_train$stall_x_kempton       > 0, na.rm = TRUE)),
    scales::comma(sum(ri_train$stall_x_lingfield     > 0, na.rm = TRUE)),
    scales::comma(sum(ri_train$stall_x_southwell     > 0, na.rm = TRUE)),
    scales::comma(sum(ri_train$stall_x_wolverhampton > 0, na.rm = TRUE))
  )
) |>
  knitr::kable(caption = "Interaction features on the training split.")
```

## Likelihood-ratio tests

Four models, all fitted on one common training sample so the tests are
nested and valid: **W** (extended win baseline), **W+weight**, **W+draw**,
and **W+both**. One training race with draw-less runners is dropped from
all four, so Model W here is the extended win model refitted on that
shared `r scales::comma(model_w_final_diagnostics$n_races)`-race sample
(it differs from @sec-extended-win's fit by that single race; the base
coefficients are unchanged to three decimals).

```{r}
#| label: tbl-interaction-lr
#| code-fold: true
desc <- c(
  "EW|E"   = "Weight×distance vs baseline",
  "ED|E"   = "Draw×course vs baseline",
  "EWD|E"  = "Both vs baseline",
  "EWD|EW" = "Draw×course | weight×distance",
  "EWD|ED" = "Weight×distance | draw×course"
)
code <- c(E = "W", EW = "W+weight", ED = "W+draw", EWD = "W+both")
interaction_lr_tests_w |>
  dplyr::mutate(
    Test       = paste(code[model_a], "vs", code[model_b]),
    Comparison = desc[paste(model_a, model_b, sep = "|")],
    `LR stat`  = round(lr_stat, 2),
    df         = df,
    `p-value`  = signif(p_value, 3),
    Verdict    = dplyr::if_else(decision == "significant",
                                "adds signal", "no improvement")
  ) |>
  dplyr::select(Test, Comparison, `LR stat`, df, `p-value`, Verdict) |>
  knitr::kable(caption = "Likelihood-ratio tests among the four interaction models, all fitted on the extended win model.")
```

The verdict is clean. **Draw×course adds signal** (W+draw vs W:
$p =$ `r signif(p_draw, 2)`; and it survives conditional on weight).
**Weight×distance does not** — it is indistinguishable from zero both on
its own (W+weight vs W: $p =$ `r signif(p_weight, 2)`) and added to draw
(W+both vs W+draw: $p =$ `r signif(p_weight_giv_draw, 2)`). The most
parsimonious model not rejected at $p < 0.05$ is therefore **W+draw**
(extended win + draw×course). The likelihood-ratio statistics are modest;
a fuller use of the data — the exploded ranking likelihood of paper 2b,
which triples the number of choice problems per race — may estimate the
same interaction more sharply, a question we leave to that paper.

The draw×course term is selected as a **block** — the four course-masked
columns enter or leave together, on the W+draw vs W likelihood-ratio test
above, not on the individual courses' Wald *p*-values. This parallels the
treatment of the position-lag block in @sec-extended-win, where the lag
terms were assessed jointly rather than course-by-course: an interaction's
worth is a property of the whole block, and masking it into four columns
is an implementation detail, not licence to prune the columns one at a
time. We therefore keep all four courses, including Lingfield, whose draw
coefficient is individually near zero (Wald $p =$
`r signif(ed_ct["stall_x_lingfield", "Pr(>|z|)"], 2)`). That null is not a
nuisance to be dropped but a **finding**: Lingfield, alone among the four,
shows no detectable draw bias — consistent with its wide, galloping
configuration — and the model says so explicitly by carrying a coefficient
indistinguishable from zero.

The **final paper-2a model** is therefore the extended win model plus the
full four-course draw block — Kempton, Lingfield, Southwell and
Wolverhampton — 21 coefficients in all.

## The final model vs the baseline

```{r}
#| label: tbl-w-coef
#| code-fold: true
base_terms <- names(stats::coef(model_w))
shift <- tibble::tibble(
  Term         = base_terms,
  `W est.`     = round(stats::coef(model_w)[base_terms], 4),
  `Final est.` = round(stats::coef(model_w_ed)[base_terms], 4)
)
knitr::kable(shift, caption = "Base (17-term) coefficients: extended win baseline (W) vs the final model. The draw terms leave the base estimates essentially unchanged.")
```

```{r}
#| label: tbl-draw-course
#| code-fold: true
draw_rows <- grepl("^stall_x_", rownames(fin_ct))
course_lab <- tools::toTitleCase(sub("stall_x_", "", rownames(fin_ct)[draw_rows]))
tibble::tibble(
  Course       = course_lab,
  Estimate     = round(fin_ct[draw_rows, "Estimate"], 4),
  `Std. error` = round(fin_ct[draw_rows, "Std. Error"], 4),
  `z`          = round(fin_ct[draw_rows, "z-value"], 2),
  `p-value`    = signif(fin_ct[draw_rows, "Pr(>|z|)"], 3),
  Sig          = dplyr::case_when(
    fin_ct[draw_rows, "Pr(>|z|)"] < 0.001 ~ "***",
    fin_ct[draw_rows, "Pr(>|z|)"] < 0.01  ~ "**",
    fin_ct[draw_rows, "Pr(>|z|)"] < 0.05  ~ "*",
    TRUE                                  ~ "(ns)"
  )
) |>
  knitr::kable(caption = "Draw×course coefficients in the final model — all four courses retained as a block. The slope is on stall_normalised (draw position / field size), per course; Lingfield's near-zero coefficient is the no-draw-bias finding.")
```

Adding the interaction barely moves the base coefficients — draw is
roughly orthogonal to form and rating — so the final model inherits the
whole extended-feature story and adds the draw effect on top.

**Which courses show draw bias, and the sign flips that hid it.** The four
course terms do not agree on direction. Kempton
(`r sprintf("%+.3f", fin_coef[["stall_x_kempton"]])`,
$p =$ `r signif(fin_ct["stall_x_kempton", "Pr(>|z|)"], 2)`) and Southwell
(`r sprintf("%+.3f", fin_coef[["stall_x_southwell"]])`,
$p =$ `r signif(fin_ct["stall_x_southwell", "Pr(>|z|)"], 2)`) favour
**higher** draws; Wolverhampton
(`r sprintf("%+.3f", fin_coef[["stall_x_wolverhampton"]])`,
$p =$ `r signif(fin_ct["stall_x_wolverhampton", "Pr(>|z|)"], 2)`) favours
**lower** draws, though here only marginally; Lingfield
(`r sprintf("%+.3f", fin_coef[["stall_x_lingfield"]])`,
$p =$ `r signif(fin_ct["stall_x_lingfield", "Pr(>|z|)"], 2)`) shows
essentially none. The marginal significance of the Wolverhampton term is
worth flagging — under the win objective, with only the win indicator to
estimate from, the draw signal is recovered less precisely than a fuller
use of the finishing order might allow — but the substantive finding is
the *sign flip*: the course effects point in opposite directions, so a
single pooled draw coefficient averages them to almost nothing, which is
exactly why `stall_normalised` looked dead in the univariate screen. A feature that appears useless in isolation carries
real, track-specific signal once the race-level structure is respected
through the interaction — the clearest vindication in this paper of the
mixed-logit construction.

**Weight × distance: Benter's compounding not detected.** The
weight×distance term is statistically indistinguishable from zero on this
data. Benter's intuition — that weight bites harder over a longer trip —
may simply be too weak to register in Class 2–5 AW handicaps, where the
handicapper has already set weights to equalise the field and the
distance range is moderate, or it may be that `rel_weight` (a within-race
deviation) already absorbs the level effect with little left for a
distance modulation. Either way, it does not earn a place in the model.

## Out-of-sample diagnostics

We evaluate the final model exactly as before: win probabilities on the
test races against the over-round-adjusted market, the same suite as
@sec-extended-win.

```{r}
#| label: p2int-eval-setup
#| include: false
# Test set only. Reuse the helper functions from @sec-extended-win.
eval_data_fin <- test_predictions_w_final |> dplyr::rename(model_prob = predicted_prob)
```

```{r}
#| label: fig-calibration-final
#| fig-cap: "Calibration on the AW test set, final model (Owen Figure 7 equivalent): observed win proportion vs predicted / implied win probability, 5-point bins over [0, 0.50]. Circles = final model (out-of-sample); plus signs = market; dashed line y = x."
calib_summary_fin <- dplyr::bind_rows(
  calibration_summary(eval_data_fin, "model_prob",  "Model (final)"),
  calibration_summary(eval_data_fin, "market_prob", "Market")
)
ggplot2::ggplot(calib_summary_fin,
                ggplot2::aes(x = bin_mid, y = observed, shape = source, colour = source)) +
  ggplot2::geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey50") +
  ggplot2::geom_point(size = 3) +
  ggplot2::scale_shape_manual(values = c("Model (final)" = 1L, Market = 3L)) +
  ggplot2::scale_colour_manual(values = c("Model (final)" = "purple", Market = "tomato")) +
  ggplot2::scale_x_continuous(labels = scales::percent_format(accuracy = 1),
                              limits = c(0, 0.50), breaks = bin_breaks) +
  ggplot2::scale_y_continuous(labels = scales::percent_format(accuracy = 1), limits = c(0, NA)) +
  ggplot2::labs(x = "Predicted / implied win probability",
                y = "Observed win proportion", shape = NULL, colour = NULL) +
  ggplot2::theme_minimal()
```

```{r}
#| label: fig-p1-p2-final
#| fig-cap:
#|   - "P1 vs probability threshold, final model (Owen Figure 8a equivalent), AW test set."
#|   - "P2 vs probability threshold, final model (Owen Figure 8b equivalent), AW test set."
score_long_fin <- dplyr::bind_rows(
  scoring_summary(eval_data_fin, "model_prob",  "Model (final)", thresholds),
  scoring_summary(eval_data_fin, "market_prob", "Market",        thresholds)
)
p_score_fin <- function(yvar, ylab) {
  ggplot2::ggplot(score_long_fin,
                  ggplot2::aes(x = threshold, y = .data[[yvar]],
                               linetype = source, colour = source)) +
    ggplot2::geom_line(linewidth = 0.8) +
    ggplot2::scale_linetype_manual(values = c("Model (final)" = "solid", Market = "dashed")) +
    ggplot2::scale_colour_manual(values = c("Model (final)" = "purple", Market = "tomato")) +
    ggplot2::scale_x_continuous(labels = scales::percent_format(accuracy = 1), limits = c(0, 0.50)) +
    ggplot2::labs(x = "Probability threshold", y = ylab, linetype = NULL, colour = NULL) +
    ggplot2::theme_minimal()
}
print(p_score_fin("P1", "P1 (geom. mean prob. on winners)"))
print(p_score_fin("P2", "P2 (geom. mean Bernoulli likelihood)"))
```

```{r}
#| label: tbl-backtest-final
backtest_naive_w_final |>
  dplyr::transmute(
    `Prob threshold`  = prob_threshold,
    `Ratio threshold` = ratio_threshold,
    `Bets`            = n_bets,
    `Wins`            = n_wins,
    `Gross return`    = sprintf("%.2f", gross_return),
    `Profit`          = sprintf("%.2f", profit),
    `ROI`             = roi_pct(roi)
  ) |>
  knitr::kable(caption = "Final model: betting backtest on the AW test races at Owen's exact thresholds.")
```

```{r}
#| label: fig-roi-sweep-final
#| fig-cap: "ROI vs ratio threshold on the AW test set (final model), 90% bootstrap CI shaded. Model-probability filter held at 0.13; dashed line at zero marks break-even."
ggplot2::ggplot(backtest_sweep_w_final, ggplot2::aes(x = tau, y = roi)) +
  ggplot2::geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  ggplot2::geom_ribbon(ggplot2::aes(ymin = ci_lo, ymax = ci_hi), fill = "purple", alpha = 0.15) +
  ggplot2::geom_line(linewidth = 0.8, colour = "purple") +
  ggplot2::scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  ggplot2::labs(x = "Ratio threshold (model_prob / market_prob)", y = "ROI") +
  ggplot2::theme_minimal()
```

## Do the interactions add anything?

Draw×course does; weight×distance does not. The final model's naive
backtest ROI is **`r roi_pct(backtest_naive_w_final$roi)`**, against the
extended win model's `r roi_pct(backtest_naive_p2$roi)` and paper 1's
−28.2%. The point estimates improve at every step, and the
course-specific draw effect is the most valuable thing this paper adds to
Owen's feature set — but the improvement is small relative to its sampling
noise: a paired race-level bootstrap of the ROI *differences* (reported in
@sec-discussion) cannot separate any of these contrasts from zero. The
direction is at least intuitive in the market-efficiency frame: draw bias
is a genuine, track-specific edge a 13-feature win model could not see,
and the starting price — built partly on punters who *do* know each
course's draw — was pricing it while the model was not.

And yet the result is still a clear loss. The ratio-sweep CI narrows but
does not cross zero: the market still prices the field more accurately
than the model overall. Paper 2a's arc is a *directional* narrowing of the
gap to the starting price — richer features and the draw interaction
together move the point ROI from −28.2% to
`r roi_pct(backtest_naive_w_final$roi)`, though a paired bootstrap cannot
rule out that the true gap is unchanged. What remains uncaptured is the
part of the market's edge that lives in information our pre-race feature
set still does not contain — late money, going on the day — and in two
structural limits of the model itself. The first is that we have used only
the win, not the full finishing order; lifting that restriction is the
subject of the companion paper 2b. The second is that a linear-predictor
conditional logit cannot represent non-linearities or feature interactions
at all beyond those we hand-build; that is the subject of paper 3.
