# Mixed race-level interactions {#sec-mixed-interactions}

The features so far are all horse-level. Race-level features — going,
distance, course — cannot enter the conditional logit directly: they are
constant within a race and cancel in the softmax. They enter only as
**interactions** with horse-level features, whose products vary across
runners within a race (the mixed discrete-choice mechanism derived in
@sec-appx-derivations). This section tests two such interactions on top
of the extended win model from @sec-extended-win, which is the baseline
throughout.

```{r}
#| label: p2int-setup
#| include: false
loadNamespace("mlogit")

ed_ct    <- summary(model_w_ed)$CoefTable      # full 4-course block (selection stage)
fin_ct   <- summary(model_w_final)$CoefTable   # final model: the 2 retained courses
fin_coef <- stats::setNames(fin_ct[, "Estimate"], rownames(fin_ct))
roi_pct  <- function(x) scales::percent(x, accuracy = 0.1)

# LR p-values pulled by test, so the prose tracks the fit.
lr_p <- function(a, b) {
  interaction_lr_tests_w |>
    dplyr::filter(model_a == a, model_b == b) |>
    dplyr::pull(p_value)
}
p_weight          <- lr_p("EW",  "E")    # weight x distance vs baseline
p_draw            <- lr_p("ED",  "E")    # draw x course vs baseline
p_weight_giv_draw <- lr_p("EWD", "ED")   # weight | draw
```

## Two interactions

**Weight × distance** (`rel_weight_x_dist = rel_weight × distance_furlongs`).
`rel_weight` is horse-level, `distance_furlongs` race-level; their product
varies across runners. The motivation is @benter1994: the burden of extra
weight should compound over longer trips, so a horse carrying weight above
the field mean ought to be penalised more in a staying race than a sprint.

**Draw × course.** Draw bias on All-Weather is known to depend on the
track — Kempton, Lingfield, Southwell and Wolverhampton have different
configurations, bend tightness and run-ins. `stall_normalised` had almost
no standalone signal (@sec-eda), but that pooled estimate could be hiding
course-specific effects that cancel. We split it into four course-masked
columns (`stall_normalised` where the race is at that course, else 0),
one coefficient per course.

```{r}
#| label: tbl-interaction-features
# Training set only — see CLAUDE.md exploratory analysis convention
ri_train <- runners_interactions |> dplyr::filter(split == "train")
tibble::tibble(
  Quantity = c("Distance range (furlongs)",
               "rel_weight_x_dist SD",
               "Runners with a Kempton draw term",
               "Runners with a Lingfield draw term",
               "Runners with a Southwell draw term",
               "Runners with a Wolverhampton draw term"),
  Value = c(
    paste(round(min(ri_train$distance_furlongs), 1), "–",
          round(max(ri_train$distance_furlongs), 1)),
    sprintf("%.1f", stats::sd(ri_train$rel_weight_x_dist, na.rm = TRUE)),
    scales::comma(sum(ri_train$stall_x_kempton       > 0, na.rm = TRUE)),
    scales::comma(sum(ri_train$stall_x_lingfield     > 0, na.rm = TRUE)),
    scales::comma(sum(ri_train$stall_x_southwell     > 0, na.rm = TRUE)),
    scales::comma(sum(ri_train$stall_x_wolverhampton > 0, na.rm = TRUE))
  )
) |>
  knitr::kable(caption = "Interaction features on the training split.")
```

## Likelihood-ratio tests

Four models, all fitted on one common training sample so the tests are
nested and valid: **W** (extended win baseline), **W+weight**, **W+draw**,
and **W+both** (the **W** prefix denoting the extended win baseline of
@sec-extended-win, and the suffixes its two candidate interactions). One
training race with draw-less runners is dropped from all four, so the
baseline here is the extended win model refitted on that
shared `r scales::comma(model_w_final_diagnostics$n_races)`-race sample
(it differs from @sec-extended-win's fit by that single race; the base
coefficients are unchanged to three decimals).

```{r}
#| label: tbl-interaction-lr
#| code-fold: true
desc <- c(
  "EW|E"   = "Weight×distance vs baseline",
  "ED|E"   = "Draw×course vs baseline",
  "EWD|E"  = "Both vs baseline",
  "EWD|EW" = "Draw×course | weight×distance",
  "EWD|ED" = "Weight×distance | draw×course"
)
code <- c(E = "W", EW = "W+weight", ED = "W+draw", EWD = "W+both")
interaction_lr_tests_w |>
  dplyr::mutate(
    Test       = paste(code[model_a], "vs", code[model_b]),
    Comparison = desc[paste(model_a, model_b, sep = "|")],
    `LR stat`  = round(lr_stat, 2),
    df         = df,
    `p-value`  = signif(p_value, 3),
    Verdict    = dplyr::if_else(decision == "significant",
                                "adds signal", "no improvement")
  ) |>
  dplyr::select(Test, Comparison, `LR stat`, df, `p-value`, Verdict) |>
  knitr::kable(caption = "Likelihood-ratio tests among the four interaction models, all fitted on the extended win model.")
```

The verdict is clean. **Draw×course adds signal** (W+draw vs W:
$p =$ `r signif(p_draw, 2)`; and it survives conditional on weight).
**Weight×distance does not** — it is indistinguishable from zero both on
its own (W+weight vs W: $p =$ `r signif(p_weight, 2)`) and added to draw
(W+both vs W+draw: $p =$ `r signif(p_weight_giv_draw, 2)`). The most
parsimonious model not rejected at $p < 0.05$ is therefore **W+draw**
(extended win + draw×course). The likelihood-ratio statistics are modest;
a fuller use of the data — the exploded ranking likelihood of paper 2b,
which triples the number of choice problems per race — may estimate the
same interaction more sharply, a question we leave to that paper.

The draw×course term enters as a **block**: the four course-masked columns
are admitted together on the W+draw vs W likelihood-ratio test above, which
establishes that course-specific draw bias carries signal the pooled term
could not. Admitting the block is one question; which individual course
slopes to *retain* is another, and we settle it with the same per-term
Wald rule paper 1 and Owen use, and that reduces the position-lag block in
@sec-extended-win — a course slope individually non-significant at
$p < 0.05$ is dropped. Of the four slopes in the fitted block, Kempton
(`r sprintf("%+.3f", ed_ct["stall_x_kempton", "Estimate"])`,
$p =$ `r signif(ed_ct["stall_x_kempton", "Pr(>|z|)"], 2)`) and Southwell
(`r sprintf("%+.3f", ed_ct["stall_x_southwell", "Estimate"])`,
$p =$ `r signif(ed_ct["stall_x_southwell", "Pr(>|z|)"], 2)`) clear the
threshold; Lingfield
(`r sprintf("%+.3f", ed_ct["stall_x_lingfield", "Estimate"])`,
$p =$ `r signif(ed_ct["stall_x_lingfield", "Pr(>|z|)"], 2)`) and
Wolverhampton
(`r sprintf("%+.3f", ed_ct["stall_x_wolverhampton", "Estimate"])`,
$p =$ `r signif(ed_ct["stall_x_wolverhampton", "Pr(>|z|)"], 2)`) do not,
and are dropped.

```{r}
#| label: tbl-draw-reduction
#| code-fold: true
final_reduction_lr_w |>
  dplyr::mutate(
    Drop      = c("Lingfield", "Wolverhampton"),
    `LR stat` = round(lr_stat, 2),
    df        = df,
    `p-value` = signif(p_value, 3),
    Verdict   = dplyr::if_else(decision == "significant",
                               "retain term", "drop term")
  ) |>
  dplyr::select(Drop, `LR stat`, df, `p-value`, Verdict) |>
  knitr::kable(caption = "Within-block reduction of the draw×course term: two sequential one-degree-of-freedom likelihood-ratio tests, each dropping one individually non-significant course slope from the model above it, all on the common training sample. Neither drop is rejected at p < 0.05, so the final model retains only Kempton and Southwell.")
```

The reduction is confirmed by the two sequential reduction tests in
@tbl-draw-reduction — each a one-degree-of-freedom likelihood-ratio test,
which drops a single coefficient and checks whether the simpler model loses
significant fit. Dropping Lingfield from the four-course block costs nothing
($p =$ `r signif(final_reduction_lr_w$p_value[1], 2)`), and dropping
Wolverhampton from the remaining three is likewise not rejected at the 5%
level ($p =$ `r signif(final_reduction_lr_w$p_value[2], 2)`, though it is
the closer call). The same inclusion rule — individual Wald $p < 0.05$ —
was applied identically to both courses: this is a mechanical rule
decision, not a judgement call, even though Wolverhampton's $p$ sits close
to the line. The **final paper-2a model** is therefore the extended win
model plus two course-specific draw slopes — Kempton and Southwell —
`r length(stats::coef(model_w_final))` coefficients in all.

## The final model vs the baseline

```{r}
#| label: tbl-w-coef
#| code-fold: true
base_terms <- names(stats::coef(model_w))
shift <- tibble::tibble(
  Term         = base_terms,
  `W est.`     = round(stats::coef(model_w)[base_terms], 4),
  `Final est.` = round(stats::coef(model_w_final)[base_terms], 4)
)
knitr::kable(shift, caption = "Base (17-term) coefficients: extended win baseline (W) vs the final model. The draw terms leave the base estimates essentially unchanged.")
```

```{r}
#| label: tbl-draw-course
#| code-fold: true
draw_rows  <- grepl("^stall_x_", rownames(ed_ct))
course_lab <- tools::toTitleCase(sub("stall_x_", "", rownames(ed_ct)[draw_rows]))
retained   <- rownames(ed_ct)[draw_rows] %in% names(stats::coef(model_w_final))
tibble::tibble(
  Course       = course_lab,
  Estimate     = round(ed_ct[draw_rows, "Estimate"], 4),
  `Std. error` = round(ed_ct[draw_rows, "Std. Error"], 4),
  `z`          = round(ed_ct[draw_rows, "z-value"], 2),
  `p-value`    = signif(ed_ct[draw_rows, "Pr(>|z|)"], 3),
  Sig          = dplyr::case_when(
    ed_ct[draw_rows, "Pr(>|z|)"] < 0.001 ~ "***",
    ed_ct[draw_rows, "Pr(>|z|)"] < 0.01  ~ "**",
    ed_ct[draw_rows, "Pr(>|z|)"] < 0.05  ~ "*",
    TRUE                                 ~ "(ns)"
  ),
  Retained     = dplyr::if_else(retained, "yes", "dropped")
) |>
  knitr::kable(caption = "Draw×course coefficients in the full four-course block (the selection-stage fit). The slope is on stall_normalised (draw position / field size), per course. Kempton and Southwell are individually significant and retained; Lingfield and Wolverhampton are non-significant at p < 0.05 and dropped, leaving the two-course final model.")
```

Adding the interaction barely moves the base coefficients — draw is
roughly orthogonal to form and rating — so the final model inherits the
whole extended-feature story and adds the draw effect on top.

**Which courses carry draw bias, and the sign flips that hid it.** Within
the fitted block the four course slopes disagree on both direction and
strength. Kempton
(`r sprintf("%+.3f", ed_ct["stall_x_kempton", "Estimate"])`,
$p =$ `r signif(ed_ct["stall_x_kempton", "Pr(>|z|)"], 2)`) and Southwell
(`r sprintf("%+.3f", ed_ct["stall_x_southwell", "Estimate"])`,
$p =$ `r signif(ed_ct["stall_x_southwell", "Pr(>|z|)"], 2)`) show a clear,
significant advantage to **higher** draws and are the two courses the final
model keeps. Wolverhampton
(`r sprintf("%+.3f", ed_ct["stall_x_wolverhampton", "Estimate"])`,
$p =$ `r signif(ed_ct["stall_x_wolverhampton", "Pr(>|z|)"], 2)`) leans the
other way, towards **lower** draws, but only marginally; Lingfield
(`r sprintf("%+.3f", ed_ct["stall_x_lingfield", "Estimate"])`,
$p =$ `r signif(ed_ct["stall_x_lingfield", "Pr(>|z|)"], 2)`) shows
essentially none. Under the win objective — with only the win indicator to
estimate from — neither clears the 5% bar, so both drop out. The
substantive finding survives the reduction: the per-course effects point in
*opposite directions*, so a single pooled draw coefficient averages them to
almost nothing, which is exactly why `stall_normalised` looked dead in the
univariate screen. A feature that appears useless in isolation carries
real, track-specific signal once the race-level structure is respected
through the interaction — the clearest vindication in this paper of the
mixed-logit construction.

**Weight × distance: Benter's compounding not detected.** The
weight×distance term is statistically indistinguishable from zero on this
data. Benter's intuition — that weight bites harder over a longer trip —
may simply be too weak to register in Class 2–5 AW handicaps, where the
handicapper has already set weights to equalise the field and the
distance range is moderate, or it may be that `rel_weight` (a within-race
deviation) already absorbs the level effect with little left for a
distance modulation. Either way, it does not earn a place in the model.

## Out-of-sample diagnostics

We evaluate the final model exactly as before: win probabilities on the
test races against the over-round-adjusted market, the same suite as
@sec-extended-win.

```{r}
#| label: p2int-eval-setup
#| include: false
# Test set only. Reuse the helper functions from @sec-extended-win.
eval_data_fin <- test_predictions_w_final |> dplyr::rename(model_prob = predicted_prob)
```

```{r}
#| label: fig-calibration-final
#| fig-cap: "Calibration on the AW test set, final model (Owen Figure 7 equivalent): observed win proportion vs predicted / implied win probability, 5-point bins over [0, 0.50]. Circles = final model (out-of-sample); plus signs = market; dashed line y = x."
calib_summary_fin <- dplyr::bind_rows(
  calibration_summary(eval_data_fin, "model_prob",  "Model (final)"),
  calibration_summary(eval_data_fin, "market_prob", "Market")
)
ggplot2::ggplot(calib_summary_fin,
                ggplot2::aes(x = bin_mid, y = observed, shape = source, colour = source)) +
  ggplot2::geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey50") +
  ggplot2::geom_point(size = 3) +
  ggplot2::scale_shape_manual(values = c("Model (final)" = 1L, Market = 3L)) +
  ggplot2::scale_colour_manual(values = c("Model (final)" = "purple", Market = "tomato")) +
  ggplot2::scale_x_continuous(labels = scales::percent_format(accuracy = 1),
                              limits = c(0, 0.50), breaks = bin_breaks) +
  ggplot2::scale_y_continuous(labels = scales::percent_format(accuracy = 1), limits = c(0, NA)) +
  ggplot2::labs(x = "Predicted / implied win probability",
                y = "Observed win proportion", shape = NULL, colour = NULL) +
  ggplot2::theme_minimal()
```

```{r}
#| label: fig-p1-p2-final
#| fig-cap:
#|   - "P1 vs probability threshold, final model (Owen Figure 8a equivalent), AW test set."
#|   - "P2 vs probability threshold, final model (Owen Figure 8b equivalent), AW test set."
score_long_fin <- dplyr::bind_rows(
  scoring_summary(eval_data_fin, "model_prob",  "Model (final)", thresholds),
  scoring_summary(eval_data_fin, "market_prob", "Market",        thresholds)
)
p_score_fin <- function(yvar, ylab) {
  ggplot2::ggplot(score_long_fin,
                  ggplot2::aes(x = threshold, y = .data[[yvar]],
                               linetype = source, colour = source)) +
    ggplot2::geom_line(linewidth = 0.8) +
    ggplot2::scale_linetype_manual(values = c("Model (final)" = "solid", Market = "dashed")) +
    ggplot2::scale_colour_manual(values = c("Model (final)" = "purple", Market = "tomato")) +
    ggplot2::scale_x_continuous(labels = scales::percent_format(accuracy = 1), limits = c(0, 0.50)) +
    ggplot2::labs(x = "Probability threshold", y = ylab, linetype = NULL, colour = NULL) +
    ggplot2::theme_minimal()
}
print(p_score_fin("P1", "P1 (geom. mean prob. on winners)"))
print(p_score_fin("P2", "P2 (geom. mean Bernoulli likelihood)"))
```

```{r}
#| label: tbl-backtest-final
backtest_naive_w_final |>
  dplyr::transmute(
    `Prob threshold`  = prob_threshold,
    `Ratio threshold` = ratio_threshold,
    `Bets`            = n_bets,
    `Wins`            = n_wins,
    `Gross return`    = sprintf("%.2f", gross_return),
    `Profit`          = sprintf("%.2f", profit),
    `ROI`             = roi_pct(roi)
  ) |>
  knitr::kable(caption = "Final model: betting backtest on the AW test races at Owen's exact thresholds.")
```

```{r}
#| label: fig-roi-sweep-final
#| fig-cap: "ROI vs ratio threshold on the AW test set (final model), 90% bootstrap CI shaded. Model-probability filter held at 0.13; dashed line at zero marks break-even."
ggplot2::ggplot(backtest_sweep_w_final, ggplot2::aes(x = tau, y = roi)) +
  ggplot2::geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  ggplot2::geom_ribbon(ggplot2::aes(ymin = ci_lo, ymax = ci_hi), fill = "purple", alpha = 0.15) +
  ggplot2::geom_line(linewidth = 0.8, colour = "purple") +
  ggplot2::scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  ggplot2::labs(x = "Ratio threshold (model_prob / market_prob)", y = "ROI") +
  ggplot2::theme_minimal()
```

## Do the interactions add anything?

The final model's naive
backtest ROI is **`r roi_pct(backtest_naive_w_final$roi)`**, against the
extended win model's `r roi_pct(backtest_naive_p2$roi)` and paper 1's
−28.2%. The point estimates improve at every step; the extended feature
set — chiefly the official rating — accounts for most of that gain, with
the course-specific draw effect adding a smaller further increment. But the
improvement is small relative to its sampling noise: a paired race-level
bootstrap of the ROI *differences* (reported in @sec-discussion) cannot
separate any of these contrasts from zero. The
direction is at least intuitive in the market-efficiency frame: draw bias
is a genuine, track-specific edge a 13-feature win model could not see,
and the starting price — built partly on punters who *do* know each
course's draw — was pricing it while the model was not.

And yet the result is still a clear loss. The ratio-sweep CI narrows but
does not cross zero: the market still prices the field more accurately
than the model overall. Paper 2a's arc is a *directional* narrowing of the
gap to the starting price — richer features and the draw interaction
together move the point ROI from −28.2% to
`r roi_pct(backtest_naive_w_final$roi)`, though a paired bootstrap cannot
rule out that the true gap is unchanged. What remains uncaptured is the
part of the market's edge that lives in information our pre-race feature
set still does not contain — late money, going on the day — together with a
genuine structural limit of the model itself: a linear-predictor
conditional logit cannot represent non-linearities or feature interactions
beyond those we hand-build, which is the subject of paper 3. This paper also
deliberately scopes to the win outcome rather than the full finishing
order; using the finishing order is the subject of the companion paper 2b —
a change of objective, not a shortfall of this one.
