# Discussion {#sec-discussion}

```{r}
#| label: discussion-setup
#| include: false
loadNamespace("mlogit")

# Paper-1 reduced model McFadden R² from its own fitted object — padded
# probability slots are zero, real ones positive.
paper1_plr2 <- {
  pf  <- fitted_final$probabilities
  nrr <- rowSums(pf > 0)
  1 - as.numeric(stats::logLik(fitted_final)) / (-sum(log(nrr)))
}

fin_ct <- summary(model_w_ed)$CoefTable   # 4-course block = the final model
roi_pct <- function(x) scales::percent(x, accuracy = 0.1)

# Mean bookmaker over-round across the test races the backtest bets on:
# raw 1/SP sums to > 1 per race; the excess is the bookmaker margin.
overround_mean <- test_predictions_w_final |>
  dplyr::filter(!is.na(starting_price_decimal), starting_price_decimal > 1) |>
  dplyr::group_by(race_id) |>
  dplyr::summarise(book = sum(1 / starting_price_decimal), .groups = "drop") |>
  dplyr::summarise(m = mean(book) - 1) |>
  dplyr::pull(m)
```

## Model progression

Paper 2a makes two additions to paper 1's model, and each earns its
place. @tbl-progression lines up the three fits on a common footing: the
McFadden pseudo-$R^2$ on the training races, and the naive betting ROI
and race count on the held-out test set.

```{r}
#| label: tbl-progression
#| code-fold: true
tibble::tibble(
  Model = c("Paper 1 (replication)",
            "Extended win model",
            "Final (extended win + draw×course)"),
  `McFadden R²` = c(
    paper1_plr2,
    model_p2_diagnostics$pl_r2[model_p2_diagnostics$model == "reduced"],
    model_w_final_diagnostics$pl_r2
  ),
  `Naive ROI` = c(backtest_naive$roi, backtest_naive_p2$roi,
                  backtest_naive_w_final$roi),
  `Test races` = c(
    dplyr::n_distinct(test_predictions$race_id),
    dplyr::n_distinct(test_predictions_p2$race_id),
    dplyr::n_distinct(test_predictions_w_final$race_id)
  )
) |>
  dplyr::mutate(
    `McFadden R²` = round(`McFadden R²`, 4),
    `Naive ROI`   = roi_pct(`Naive ROI`),
    `Test races`  = scales::comma(`Test races`)
  ) |>
  knitr::kable(caption = "Paper-2a model progression: McFadden R² on training, naive ROI and test-race count on the held-out set. The three models evaluate on slightly different test universes — paper 1 retains races that paper 2a's extra features drop to NA (@sec-extended-win) — so the ROIs are not struck on an identical race set; the paired bootstrap below removes that confound by intersecting each pair's common races.")
```

The picture is one of modest, point-estimate improvement. The extended
feature set lifts discrimination above paper 1, the single largest
contributors being `or_relative` — the handicapper's own assessment of
ability, the strongest new predictor in the model — and, for the
backtest, the course-specific draw effect. The naive ROI climbs at each
step, from −28.2% in paper 1 to
`r roi_pct(backtest_naive_w_final$roi)` in the final model. Whether that
climb is real or sampling noise is the question the next table settles.

Because the three models are scored on overlapping but non-identical test
races, and because each model's ROI is itself a noisy statistic, the
marginal bootstrap intervals of @sec-mixed-interactions cannot tell us
whether the *differences* between models are real. @tbl-roi-diff answers
that directly: for each pair it restricts to the common races, resamples
races (paired, so both models see the same resample), and reports the 90%
interval of the ROI difference.

```{r}
#| label: tbl-roi-diff
#| code-fold: true
roi_difference_bootstrap |>
  dplyr::transmute(
    Contrast      = contrast,
    `ROI diff.`   = roi_pct(diff_point),
    `90% CI`      = paste0("[", roi_pct(ci_lo), ", ", roi_pct(ci_hi), "]"),
    `Common races` = scales::comma(n_common)
  ) |>
  knitr::kable(caption = "Paired race-level bootstrap of the ROI difference between models (B = 2000, seed 42), each restricted to the pair's common test races. A 90% interval spanning zero means the two models' ROIs are not statistically distinguishable on these data.")
```

The verdict is sobering. Every point difference is positive — the
ordering paper 1 < extended < final holds — but every 90% interval spans
zero. The narrowing of the gap to the market is **directionally
consistent but not statistically distinguishable**: on this test set we
cannot reject the hypothesis that the extended features and the
draw×course interaction leave betting ROI unchanged. The improvement is
real in the coefficients and the McFadden $R^2$, where the training
sample is large; it is not resolvable in the much noisier, smaller-sample
betting return. We report the point gains plainly and decline to call
them established.

## Why the market still wins

For all that, the final model still loses. Its naive ROI is
`r roi_pct(backtest_naive_w_final$roi)`, and the ratio-threshold sweep in
@sec-mixed-interactions has a 90% bootstrap interval that does not cross
zero at any threshold. The reading is the one paper 1 reached, only on a
narrower margin: the starting price aggregates information our pre-race
feature set cannot observe. `or_relative` encodes the handicapper's
*historical* assessment of a horse, but not its current readiness —
paddock condition, work on the morning, stable confidence — nor the late
market moves of informed money that the SP absorbs in the minutes before
the off. The favourite–longshot bias documented in paper 1 persists here,
and a model built only on form, ratings and draw cannot price the part of
the field's chances that lives in same-day, private information.

A second qualification concerns the benchmark. The ROI is measured at the
**industry starting price**, which carries a substantial bookmaker
over-round — about `r scales::percent(overround_mean, accuracy = 0.1)` on
these test races — and is therefore the harshest price a bettor could
take. We remove that over-round only where we compare the model to the
market's *probabilities* (the calibration plot and the P1/P2 rules
renormalise 1/SP to sum to one); it is necessarily kept in the ROI,
because a winning bet returns the actual SP, margin and all. The same
selections struck on a betting exchange — where Betfair SP charges only
commission, of order 2–5%, rather than that margin — would return
materially more, though whether enough to reach break-even cannot be
settled without exchange prices, which this dataset does not carry. This
is the benchmark that matters for the live-trading direction noted under
limitations below.

## Draw bias by course

The draw×course result is the clearest illustration in the paper of why
the mixed discrete-choice construction matters. The four course terms
disagree on direction: Kempton
(`r sprintf("%+.3f", fin_ct["stall_x_kempton", "Estimate"])`,
$p =$ `r signif(fin_ct["stall_x_kempton", "Pr(>|z|)"], 2)`) and Southwell
(`r sprintf("%+.3f", fin_ct["stall_x_southwell", "Estimate"])`,
$p =$ `r signif(fin_ct["stall_x_southwell", "Pr(>|z|)"], 2)`) favour
higher draws, while Wolverhampton
(`r sprintf("%+.3f", fin_ct["stall_x_wolverhampton", "Estimate"])`,
$p =$ `r signif(fin_ct["stall_x_wolverhampton", "Pr(>|z|)"], 2)`, only
marginally) favours lower draws; Lingfield
(`r sprintf("%+.3f", fin_ct["stall_x_lingfield", "Estimate"])`,
$p =$ `r signif(fin_ct["stall_x_lingfield", "Pr(>|z|)"], 2)`) shows
essentially no bias — a substantive finding in its own right, kept in the
block rather than pruned, recording that the track has no draw advantage
to price. Because the non-null course effects point in opposite
directions, a single pooled draw coefficient averages them to almost
nothing — which is exactly why `stall_normalised` looked useless in the
univariate screen. A feature that appears dead in isolation carries
strong, track-specific signal once the race-level structure is respected
through the interaction. This is the mixed discrete-choice mechanism
(@sec-appx-derivations) doing real work, not decoration. Under the win
objective the draw terms are estimated relatively imprecisely —
Wolverhampton only marginally significant — because only the win indicator
informs them; whether the ranking likelihood of paper 2b, which draws on
more of each race, recovers them more sharply is a question we leave to
that paper.

## Limitations

- **No live-data test.** The model is fitted and evaluated on the same
  2006–2015 window as paper 1; nothing here tests it on current racing,
  and the AW programme has changed since (new tracks, fixture inflation).
- **Industry-SP benchmark.** Profitability is evaluated at the bookmaker
  starting price, with its full over-round; a betting exchange (Betfair
  SP, commission only) is a more favourable — and, for the live-trading
  direction, more relevant — benchmark. The Smartform data carries only
  industry SP (the morning forecast price and Tote dividends aside), so a
  Betfair-SP backtest is left to future work.
- **Going deferred.** A horse-specific going-affinity feature
  (`going_ordinal × horse_going_sr`) would require a going-filtered
  career-history sub-query, and as a race-level feature it only enters
  the conditional logit through an interaction. It is the natural next
  feature for paper 3, where a tree-based model can take going as a raw
  predictor without the softmax-cancellation constraint that forces the
  interaction here.
- **`or_relative` missing-not-at-random.** The zero-imputation with a
  missing indicator is defensible, but unrated horses remain a distinct
  population rather than average ones; the fitted `or_missing` coefficient
  is negative and significant
  (`r sprintf("%+.3f", fin_ct["or_missing", "Estimate"])`,
  $p < 0.001$), confirming that they underperform even the imputed zero
  and that the indicator is carrying a real effect rather than a nuisance.

## Where the series goes next

This paper has taken the win objective as fixed and asked how far a richer
feature set can go; the answer is "a little way". Two companion papers
take the two remaining levers. **Paper 2b** holds the feature set fixed
and changes the *objective*: it fits the exploded conditional logit (the
Plackett–Luce ranking model) on the same features and evaluates it on
ranking metrics against a discounted-Harville place-market baseline,
asking whether predicting the finishing order, rather than just the win,
recovers more of the market's edge. **Paper 3** changes the *model class*
while holding everything else fixed: it replaces the conditional-logit
linear predictor with a tree-based ensemble fitted under a ranking loss,
keeping the same data scope, the same chronological train/test split, and
the same diagnostic suite. The point of holding all of that constant is to
isolate one thing — how much of the remaining gap to the market is
recoverable from model-class flexibility alone, non-linearities,
automatic interactions, and raw features such as going — as distinct from
the feature-set improvements paper 2a has banked.
