# Data {#sec-data}

```{r}
#| label: p4-data-setup
#| include: false

cov_split <- p4_coverage$by_split
cov_year  <- p4_coverage$by_year
or_src    <- p4_overround$by_source
or_field  <- p4_overround$by_field_size
comp_fit  <- p4_compression$fit_summary

arch_or <- or_src |> dplyr::filter(source == "Archived forecast")
sp_or   <- or_src |> dplyr::filter(source == "Starting price")

feed_test  <- cov_split$feed_pct[cov_split$split == "test"]
feed_train <- cov_split$feed_pct[cov_split$split == "train"]

acct <- p4_common$accounting

fs_lo <- min(or_field$field_size)
fs_hi <- max(or_field$field_size)
at    <- function(v, fs) v[or_field$field_size == fs]
grad_span  <- function(v) at(v, fs_hi) - at(v, fs_lo)
grad_ratio <- grad_span(or_field$median_archive) / grad_span(or_field$median_sp)
```

The race universe is unchanged from papers 2b and 3: Class 2–5
All-Weather Flat handicaps at Kempton, Lingfield, Southwell and
Wolverhampton, 2006-01-01 to 2015-10-14, non-maiden, four to sixteen
starters, with the same chronological 70/30 train/test split at
2012-12-30. Paper 1 sets out the scope decisions. Only the price data is
new, and only that is covered here. Everything distributional below is
training-split; coverage is reported on both splits.

## Coverage

A price is usable if present and strictly greater than 1. The pre-race
forecast additionally requires its feed row to have been written before
the meeting date.

```{r}
#| label: tbl-coverage-year
cov_year |>
  dplyr::transmute(
    Year = year,
    Races = comma(races),
    `Runner rows` = comma(runner_rows),
    `Archived forecast` = scales::percent(archive_pct, accuracy = 0.1),
    `Pre-race forecast` = scales::percent(feed_pct, accuracy = 0.1),
    `Starting price` = scales::percent(sp_pct, accuracy = 0.1)
  ) |>
  knitr::kable(caption = "Usable-price coverage by year, over all in-scope runner rows.")
```

The two forecast prices fail in opposite places. The archived value is
complete from 2008 and thin only in 2006, at
`r scales::percent(cov_year$archive_pct[cov_year$year == 2006], accuracy = 0.1)`.
The pre-race value does not exist before March 2008, when Smartform's
daily feed begins, so training-split coverage is
`r scales::percent(feed_train, accuracy = 0.1)`. On the test split, where
this paper fits, it is `r scales::percent(feed_test, accuracy = 0.1)`.
The pre-race value is therefore not a drop-in replacement for the
archived one elsewhere in the series.

Races are dropped whole, so race-level completeness governs the fit.

```{r}
#| label: tbl-intersection
acct |>
  dplyr::transmute(
    `Price set` = price_set,
    `Races complete` = comma(races_complete),
    `Of` = comma(races_offered),
    `%` = scales::percent(pct_complete, accuracy = 0.01),
    `Lost` = races_lost
  ) |>
  knitr::kable(caption = "Race-level completeness over the test universe, and the intersection every arm is fitted on.")
```

The intersection costs one race, lost to the pre-race forecast. Every arm
is fitted on the remaining `r comma(n_common)`.

## Overround and shape

```{r}
#| label: tbl-overround
or_src |>
  dplyr::transmute(
    Source = source,
    Races = comma(races),
    `25th pct` = sprintf("%.3f", p25),
    Median = sprintf("%.3f", median),
    `75th pct` = sprintf("%.3f", p75),
    IQR = sprintf("%.3f", iqr)
  ) |>
  knitr::kable(caption = sprintf("Race overround (sum of 1/price over the field) by price source. Training split, and the %s races carrying a complete set of all three prices — the pre-race forecast does not exist before March 2008, so putting the three on their own larger sets would not compare like with like.", comma(or_src$races[1])))
```

Median margins are similar — `r sprintf("%.3f", arch_or$median)` for the
archived forecast against `r sprintf("%.3f", sp_or$median)` for the
starting price — but the forecast prices are more dispersed, with an
interquartile range of `r sprintf("%.3f", arch_or$iqr)` against
`r sprintf("%.3f", sp_or$iqr)`. The margin also widens with field size,
from `r sprintf("%.3f", at(or_field$median_archive, fs_lo))` at
`r fs_lo` runners to
`r sprintf("%.3f", at(or_field$median_archive, fs_hi))` at `r fs_hi`, the
largest field with enough races in this set — a gradient `r scales::percent(grad_ratio - 1, accuracy = 1)` steeper than
the starting price's.

Regressing the within-race centred log archived-forecast probability on
the within-race centred log starting-price probability gives a slope of
`r sprintf("%.3f", comp_fit$slope)`
(95% CI `r sprintf("[%.3f, %.3f]", comp_fit$conf_low, comp_fit$conf_high)`,
`r comma(comp_fit$n_runners)` runners in `r comma(comp_fit$n_races)`
training races). Below 1 means the forecast price separates the field
less than the starting price does: it spreads runners over a narrower
range of probabilities, making favourites too long and longshots too
short.
