# Appendix A: Forecast-price provenance in Smartform {#sec-appx-provenance .appendix}

```{r}
#| label: p4-provenance-setup
#| include: false

pf <- p4_provenance_facts

amd_arch <- p4_amendment2$coefficients |>
  dplyr::filter(stringr::str_detect(term, "mkt_C"))
amd_feed <- p4_amendment2$coefficients |>
  dplyr::filter(stringr::str_detect(term, "mkt_A"))
amd_lr_arch <- p4_amendment2$lr |>
  dplyr::filter(stringr::str_detect(contrast, "feed only"))

corr_ac <- p4_market_correlations |>
  dplyr::filter(x == "log_p_mkt_A", y == "log_p_mkt_C")

feed_alone <- p4_amendment2$feed_only
arch_alone <- p4_amendment2$archive_only
```

This appendix is for anyone else reusing Smartform. It records a
divergence between two columns the documentation describes identically.

Smartform holds a forecast price in two places. `daily_runners` carries
the forthcoming day's racecards, available from the night before racing;
`historic_runners` is the results archive, written the following day with
the previous day's results. On this machine the daily tables update at
00:30 and the historic tables at 05:30. Arm A of this paper takes the
daily value, arm B the starting price from the archive, and arm C the
archive's own forecast price.

The Smartform manual [@smartform2011] documents both forecast columns
identically, as the forecast price from the racecard. It gives no reason
for the two to differ, and it predates the end of this paper's data
window by four years.

They do differ. On the `r comma(pf$n_compared)` training-split runner
rows carrying both values and a starting price, the two disagree on
`r scales::percent(pf$pct_disagreeing, accuracy = 0.1)` — typically by
one or two steps of the bookmakers' fractional ladder, 10/1 against 12/1
rather than anything wholesale. Where they disagree the archived value
sits closer to the starting price: mean absolute log distance
`r sprintf("%.3f", pf$mean_dist_arch)` against
`r sprintf("%.3f", pf$mean_dist_feed)`, with
`r scales::percent(pf$pct_archive_closer_to_sp, accuracy = 0.1)` of the
`r comma(pf$n_disagreeing)` disagreeing rows moving toward it.

Fitted alone on the test universe, the archived value is the better
predictor of the two.

```{r}
#| label: tbl-provenance-alone
dplyr::bind_rows(feed_alone, arch_alone) |>
  dplyr::transmute(
    Column = dplyr::if_else(stringr::str_detect(price_source, "feed"),
                            "Pre-race (daily)", "Archived (historic)"),
    `$\\beta_{\\text{mkt}}$` = fmt_b(estimate),
    SE = fmt_b(std_error),
    `$z$` = fmt_z(z),
    `Log-likelihood` = sprintf("%.1f", logLik),
    `McFadden $R^2$` = fmt_r2(mcfadden_r2)
  ) |>
  knitr::kable(caption = "Each forecast-price column fitted alone as the market term, same races.")
```

Carrying both in one conditional logit, the archived value adds to the
pre-race one: a likelihood-ratio statistic of
`r sprintf("%.1f", amd_lr_arch$lr_stat)` on `r amd_lr_arch$df` degree of
freedom against the pre-race-only model. The two columns correlate within
race at `r sprintf("%.3f", corr_ac$corr_within_race)`, so that fit is
identified off little residual variation and its individual coefficients
are unstable — the pre-race coefficient turns negative
(`r fmt_b(amd_feed$estimate)`) while the same column fitted alone is
strongly positive (`r fmt_b(feed_alone$estimate)`). Nothing here rests on
that sign.

Two explanations remain live, and they bias $\beta_{\text{mod}}$ in
opposite directions.

* **A later racecard edition.** Forecast prices are revised between
  editions, the daily feed holds an earlier one and the archive a later
  one. Arm C is then contaminated by information that accrued after arm A
  was available.
* **The same price with transcription noise removed.** The daily value
  carries error, the archive is the corrected version, and the two differ
  by noise rather than timing. Arm A then carries an errors-in-variables
  attenuation that inflates $\beta_{\text{mod}}$.

The design does not separate them. Both predict exactly what is observed:
a later price knows more, and a de-noised price also predicts better and
sits closer to the starting price once the noisy one is conditioned on.
The manual, documenting both as the same racecard price, gives no support
for the first over the second. We do not claim to have resolved it, and
@sec-discussion carries the consequence — that the arm A to arm C gap is
an upper bound on the attenuation rather than an estimate of it.

For reuse: take the forecast price from `daily_runners` wherever a
verifiably pre-race value is needed, expecting it to be absent before
March 2008 and patchy for a year after (@sec-data). The `historic_runners`
column is written after the race and does not reproduce it.
