# Return on investment {#sec-roi}

```{r}
#| label: roi-key
#| include: false
c_2b_2a <- p6m_row(p6m_contrasts, "contrast", "2b - 2a")
c_3_2b  <- p6m_row(p6m_contrasts, "contrast", "3 - 2b")
c_5_3   <- p6m_row(p6m_contrasts, "contrast", "5 - 3")
c_5_hb  <- p6m_row(p6m_contrasts, "contrast", "5 - hand-built")
first_excl <- p6m_contrasts |> dplyr::filter(excludes_zero)
h1 <- p6m_row(p6m_race_sets, "race_set", "validation slice, first half")
h2 <- p6m_row(p6m_race_sets, "race_set", "validation slice, second half")
ts <- p6m_row(p6m_race_sets, "race_set", "test split")
s_2a <- p6m_row(p6m_series, "model", "2a")
s_5  <- p6m_row(p6m_series, "model", "5")
```

Return on investment has been reported throughout this series, alongside the
ranking measures, and it cannot carry the model-to-model comparison on its
own. Each figure is an unbiased estimate of what that rule returned on the
races it was measured on, and nothing was miscomputed. The intervals around
the differences between them are simply wider than the differences the series
was trying to detect, so a sequence of point estimates across papers cannot
establish that the models improved.

Two demonstrations follow. The first is the published record: the four
consecutive-model differences and their paired bootstraps. The second holds
the model and the rule fixed and changes only the races.

Everything here runs on stored predictions and refits nothing, so these
comparisons are out of sample as comparisons. The models being compared are
not: they were selected with knowledge of these test races across five
papers, where the feature set, the objective, the function class, the encoder
architecture and the rung ladder were each chosen after seeing an earlier
paper's test results.

## The published record

Single-bet win return, one bet per race under Owen's naive rule [@owen2019] at
the series-standard thresholds, settled at the real starting price:

```{r}
#| label: tbl-series
p6m_tbl(
  p6m_series |>
    dplyr::transmute(
      Model = model, Bets = p6m_n(n_bets), Wins = p6m_n(n_wins),
      ROI = p6m_pct(roi, 2, TRUE),
      `ROI less top winner` = p6m_pct(roi_drop_top1, 2, TRUE)),
  caption = "Single-bet win return across the series, on the test split."
)
```

Paper 1 is absent because it has no single-bet figure: it reported only the
bet-every-qualifier rule, at
`r p6m_pct(p6m_backtests_main$naive_1$roi, 2, TRUE)` over
`r p6m_n(p6m_backtests_main$naive_1$n_bets)` bets.

The point estimates improve by
`r sprintf("%.1f", 100 * (s_5$roi - s_2a$roi))` points from paper 2a to paper
5, which reads as steady improvement. The paired race-level bootstraps of the
consecutive differences do not support that reading.

The four differences below were recomputed for this paper on one bet rule,
because the versions the series published do not share one: papers 2b and 3
compared return on the bet-every-qualifier rule while paper 5 compared it on
the single-bet rule, and the two rules move the earlier point differences by
about three points. The recomputation uses the series' own paired bootstrap
and reproduces paper 5's two published figures exactly, on the point
difference and both interval ends, so all four rows are the output of one
procedure.

```{r}
#| label: tbl-published
p6m_tbl(
  p6m_published |>
    dplyr::transmute(
      Contrast = label, Basis = basis,
      Difference = p6m_pp(diff_point),
      `90% interval` = p6m_ci_pp(ci_lo, ci_hi)),
  caption = "The four contrasts as the series published them, with the bet rule each was computed on."
)
```

```{r}
#| label: tbl-contrasts
p6m_tbl(
  p6m_contrasts |>
    dplyr::transmute(
      Contrast = contrast,
      Difference = p6m_pp(diff_point),
      `90% interval` = p6m_ci_pp(ci_lo, ci_hi),
      Races = p6m_n(n_races),
      `Excludes zero` = p6m_yn(excludes_zero)),
  caption = "The same four contrasts on the single-bet rule. Paired race-level bootstrap, B = 2000, seed 42."
)
```

`r first_excl$contrast[1]` is the only one of the four whose interval excludes
zero, and it is the first step in the sequence rather than the largest
difference in it. The three later contrasts all contain zero: their intervals
are `r sprintf("%.0f", 100 * (c_3_2b$ci_hi - c_3_2b$ci_lo))`,
`r sprintf("%.0f", 100 * (c_5_3$ci_hi - c_5_3$ci_lo))` and
`r sprintf("%.0f", 100 * (c_5_hb$ci_hi - c_5_hb$ci_lo))` points wide against
point differences of `r p6m_pp(c_3_2b$diff_point, 1)`,
`r p6m_pp(c_5_3$diff_point, 1)` and `r p6m_pp(c_5_hb$diff_point, 1)`. Every
comparison after the first is consistent with an improvement, no change, or a
decline.

## One model, one rule, three sets of races

The second demonstration holds the model and the rule fixed and changes only
the races. Paper 5's encoder, Owen's rule, one bet per race, nothing refitted
and nothing selected.

```{r}
#| label: tbl-race-sets
p6m_tbl(
  p6m_race_sets |>
    dplyr::transmute(
      `Race set` = race_set, Fit = fit,
      Races = ifelse(is.na(n_races), "---", p6m_n(n_races)),
      Bets = p6m_n(n_bets), Wins = p6m_n(n_wins),
      ROI = p6m_pct(roi, 1, TRUE)),
  caption = "One model and one rule on three sets of races."
)
```

The first two rows are the 1,505 races carved out of the training split at
2010-12-15 for paper 5's model selection, split chronologically at
`r unique(p6m_val_halves$cut_date)` into `r p6m_n(h1$n_races)` and
`r p6m_n(h2$n_races)` races. The same model under the same rule returns
`r p6m_pct(h1$roi, 1, TRUE)` on one and `r p6m_pct(h2$roi, 1, TRUE)` on the
other, a spread of `r sprintf("%.0f", 100 * (h1$roi - h2$roi))` points across
two adjacent sets of races of almost equal size. On the test split it returns
`r p6m_pct(ts$roi, 1, TRUE)`.

Those three figures are not identical in provenance: the validation slice is
scored by the encoder's fitting-partition fit and the test split by the
full-training-split refit, which are two fits of one architecture. The spread
between the two halves needs no such qualification, since both are scored by
the same fit.
