# ROI cannot compare one model against the next {#sec-roi}

```{r}
#| label: roi-key
#| include: false
c_2b_2a <- p6m_row(p6m_contrasts, "contrast", "2b - 2a")
c_3_2b  <- p6m_row(p6m_contrasts, "contrast", "3 - 2b")
c_5_3   <- p6m_row(p6m_contrasts, "contrast", "5 - 3")
c_5_hb  <- p6m_row(p6m_contrasts, "contrast", "5 - hand-built")
n_excl  <- sum(p6m_contrasts$excludes_zero)
roi_row <- p6m_row(p6m_resolution, "measure", "single-bet win ROI")
rank_rows <- p6m_resolution |> dplyr::filter(measure != "single-bet win ROI")
h1 <- p6m_row(p6m_race_sets, "race_set", "validation slice, first half")
h2 <- p6m_row(p6m_race_sets, "race_set", "validation slice, second half")
ts <- p6m_row(p6m_race_sets, "race_set", "test split")
p3_frame <- p6m_row(p6m_race_identity, "frame", "3")
```

Every paper in this series except paper 4 has reported a return on investment,
and the series has used the sequence of those figures to judge whether each
model improved on the one before. It cannot support that use. Of the four
consecutive-model contrasts, only
`r c("none", "one", "two", "three", "all four")[n_excl + 1]` excludes zero,
and it is the earliest.

The figures themselves are sound. Each is an unbiased estimate of what that
rule returned on those races, and nothing was miscomputed. The problem is that
the intervals around the differences are wider than the differences the series
was trying to detect.

## The published record

The series' single-bet win ROI, one bet per race under Owen's naive rule
[@owen2019] at the series-standard thresholds, settled at the real starting
price:

```{r}
#| label: tbl-series
p6m_tbl(
  p6m_series |>
    dplyr::transmute(
      Model = model, Bets = p6m_n(n_bets), Wins = p6m_n(n_wins),
      ROI = p6m_pct(roi, 2, TRUE),
      `ROI less top winner` = p6m_pct(roi_drop_top1, 2, TRUE)),
  caption = "Single-bet win ROI across the series, on the test split."
)
```

Paper 1 is absent because it has no single-bet figure: it reported only the
bet-every-qualifier rule, at
`r p6m_pct(p6m_backtests_main$naive_1$roi, 2, TRUE)` over
`r p6m_n(p6m_backtests_main$naive_1$n_bets)` bets, and no single-bet
equivalent was computed at the time.

The point estimates improve by
`r sprintf("%.1f", 100 * (p6m_row(p6m_series, "model", "5")$roi - p6m_row(p6m_series, "model", "2a")$roi))`
points from paper 2a to paper 5, which reads as steady improvement. The paired race-level bootstraps of the consecutive
differences do not support that reading.

```{r}
#| label: tbl-contrasts
p6m_tbl(
  p6m_contrasts |>
    dplyr::transmute(
      Contrast = contrast,
      Difference = p6m_pp(diff_point),
      `90% interval` = p6m_ci_pp(ci_lo, ci_hi),
      Races = p6m_n(n_races),
      `Excludes zero` = p6m_yn(excludes_zero)),
  caption = "Consecutive-model differences in single-bet win ROI. Paired race-level bootstrap, B = 2000, seed 42."
)
```

Only `r c_2b_2a$contrast` excludes zero, at `r p6m_pp(c_2b_2a$diff_point)`
with interval `r p6m_ci_pp(c_2b_2a$ci_lo, c_2b_2a$ci_hi)`. The three later
contrasts have intervals `r sprintf("%.0f", 100 * (c_3_2b$ci_hi - c_3_2b$ci_lo))`,
`r sprintf("%.0f", 100 * (c_5_3$ci_hi - c_5_3$ci_lo))` and
`r sprintf("%.0f", 100 * (c_5_hb$ci_hi - c_5_hb$ci_lo))` points wide against
point differences of `r p6m_pp(c_3_2b$diff_point, 1)`,
`r p6m_pp(c_5_3$diff_point, 1)` and `r p6m_pp(c_5_hb$diff_point, 1)`.

### A note on the basis of these four figures

The four contrasts above were recomputed for this paper. The versions the
series published do not share a bet rule: papers 2b and 3 compared ROI on the
bet-every-qualifier rule, while paper 5 compared it on the single-bet rule.

```{r}
#| label: tbl-published
p6m_tbl(
  p6m_published |>
    dplyr::transmute(
      Contrast = label, Basis = basis,
      Difference = p6m_pp(diff_point),
      `90% interval` = p6m_ci_pp(ci_lo, ci_hi),
      `Excludes zero` = p6m_yn(excludes_zero)),
  caption = "The same four contrasts as published, with the bet rule each was computed on."
)
```

Switching the two earlier contrasts to the single-bet rule moves them by
`r sprintf("%.1f", 100 * abs(p6m_basis_check$published[1] - p6m_basis_check$single_bet[1]))`
and
`r sprintf("%.1f", 100 * abs(p6m_basis_check$published[2] - p6m_basis_check$single_bet[2]))`
points, which is why they were recomputed rather than quoted. The
recomputation uses the series' own paired bootstrap and reproduces paper 5's
two figures exactly, so all four rows are the output of one procedure:

```{r}
#| label: tbl-recompute-check
p6m_tbl(
  p6m_recompute_check |>
    dplyr::transmute(
      Contrast = contrast,
      Recomputed = p6m_pp(new_diff, 4),
      Published = p6m_pp(pub_diff, 4),
      `Max absolute difference` = p6m_sig(pmax(d_diff, d_lo, d_hi), 2)),
  caption = "The two contrasts that appear in both tables. Agreement is exact, on the point difference and both interval ends."
)
```

The conclusion is unchanged by the change of basis: one of four contrasts
excludes zero on either rule, and it is the earliest one either way.

## Three measures that resolve, one that does not

Papers 2b and 3 are the clearest case. They tie on all three ranking measures
the series reports, and their single-bet win ROI point estimates differ by
`r sprintf("%.1f", 100 * abs(roi_row$diff_point))` points.

```{r}
#| label: tbl-resolution
p6m_tbl(
  p6m_resolution |>
    dplyr::transmute(
      Measure = measure,
      `Paper 2b` = p6m_sig(point_2b, 4),
      `Paper 3` = p6m_sig(point_3, 4),
      Difference = p6m_sig(diff_point, 3),
      `90% interval` = sprintf("[%s, %s]", p6m_sig(ci_lo, 3), p6m_sig(ci_hi, 3))),
  caption = "Papers 2b and 3 on four measures. Every interval contains zero."
)
```

```{r}
#| label: tbl-resolution-basis
p6m_tbl(
  p6m_resolution |>
    dplyr::transmute(
      Measure = measure,
      `Computed over` = n_basis,
      Width = p6m_sig(ci_width, 3),
      `Width / value` = sprintf("%.1f%%", 100 * width_over_value)),
  caption = "What each measure is computed over, and its interval width as a share of the value it measures (paper 2b's, from Table 5)."
)
```

Every interval contains zero, so on all four measures the two models tie. What
differs is how much that tells us, and the measure of that is how wide each
interval is against the size of the quantity it measures. On the three ranking
measures the interval is
`r sprintf("%.1f%%", 100 * min(rank_rows$width_over_value))` to
`r sprintf("%.1f%%", 100 * max(rank_rows$width_over_value))` of the value
being measured, so the tie is a real answer: any difference between the two
models is at most a few per cent of the quantity. The ROI interval is
`r sprintf("%.0f%%", 100 * roi_row$width_over_value)` of the value being
measured and spans `r p6m_ci_pp(roi_row$ci_lo, roi_row$ci_hi, 1)`, which is
consistent with paper 3 being much worse than paper 2b, slightly better, or
the same. That is not a tie, it is an absence of information.

The reason is the number of observations each measure gets. Brier_place is
computed over `r p6m_n(p3_frame$n_rows)` runner-rows and the other two
ranking measures over `r p6m_n(p3_frame$n_races)` races. The ROI is computed
over the `r p6m_n(p6m_backtests_main$single_2b$n_bets)` and
`r p6m_n(p6m_backtests_main$single_3$n_bets)` races where Owen's rule finds a
bet, and within those almost all of the variance sits in whether a small
number of long-priced bets landed.

## One model, one rule, three sets of races

The second demonstration holds the model and the rule fixed and changes only
the races. Paper 5's encoder, Owen's rule, one bet per race, nothing refitted
and nothing selected.

```{r}
#| label: tbl-race-sets
p6m_tbl(
  p6m_race_sets |>
    dplyr::transmute(
      `Race set` = race_set, Fit = fit,
      Races = ifelse(is.na(n_races), "---", p6m_n(n_races)),
      Bets = p6m_n(n_bets), Wins = p6m_n(n_wins),
      ROI = p6m_pct(roi, 1, TRUE)),
  caption = "One model and one rule on three sets of races."
)
```

The first two rows are the 1,505 races carved out of the training split at
2010-12-15 for paper 5's model selection, split chronologically at
`r unique(p6m_val_halves$cut_date)` into `r p6m_n(h1$n_races)` and
`r p6m_n(h2$n_races)` races. The same model under the same rule returns
`r p6m_pct(h1$roi, 1, TRUE)` on one and `r p6m_pct(h2$roi, 1, TRUE)` on the
other, a spread of
`r sprintf("%.0f", 100 * (h1$roi - h2$roi))` points across two adjacent sets
of races of almost equal size. On the test split it returns
`r p6m_pct(ts$roi, 1, TRUE)`.

Those three figures are not identical in provenance: the validation slice is
scored by the encoder's fitting-partition fit and the test split by the
full-training-split refit, which are two fits of one architecture. The spread
between the two validation halves needs no such qualification, since both are
scored by the same fit.

## What this establishes

The ROI figures in papers 1 to 5 are unbiased estimates of what each rule
returned on the races it was measured on. The intervals around the differences
between them are wider than the differences the series was trying to detect,
and a single figure moves by
`r sprintf("%.0f", 100 * (h1$roi - h2$roi))` points between two adjacent sets
of races when the model and the rule are held fixed. A sequence of ROI point
estimates across papers therefore cannot establish that the models improved.

The measures that do resolve are the ones computed over every race or every
runner-row rather than over the subset of races a betting rule selects.
Section 2 takes the scoring rules the series has for that job and reports what
they say.
