# Owen's scoring rules {#sec-p2}

```{r}
#| label: p2-key
#| include: false
sc <- function(rule, col) {
  out <- p6m_scores |> dplyr::filter(rule == !!rule, column == !!col)
  stopifnot(nrow(out) == 1L); out
}
bench <- function(rule, model, benchmark) {
  out <- p6m_scores_bench |>
    dplyr::filter(rule == !!rule, model == !!model, benchmark == !!benchmark)
  stopifnot(nrow(out) == 1L); out
}
pair <- function(rule, contrast) {
  out <- p6m_scores_pairs |>
    dplyr::filter(rule == !!rule, contrast == !!contrast)
  stopifnot(nrow(out) == 1L); out
}
n_bench <- nrow(p6m_scores_bench)
n_bench_excl <- sum(p6m_scores_bench$excludes_zero)
bot <- p6m_bins |> dplyr::filter(bin == "0-2%")
bin_enc <- p6m_bins |> dplyr::filter(column == "p_5")
bin_sp <- p6m_bins |> dplyr::filter(column == "p_sp")
SHORT <- c("Paper 2b logit" = "2b", "Paper 3 GBT" = "3",
           "Paper 5 encoder" = "5", "Starting price" = "SP",
           "Forecast price" = "FP")
KEY <- "Columns: 2b = paper 2b logit, 3 = paper 3 GBT, 5 = paper 5 encoder, SP = starting price, FP = forecast price."
```

Paper 1 reports two scoring rules over the runner-rows whose probability
exceeds a threshold $t$: $P_1(t)$, the geometric mean of the probability
assigned to winners, and $P_2(t)$, the geometric mean of the per-row Bernoulli
likelihood. Both are kept here, with two changes.

## What each rule scores

Every source in this paper is a distribution over the race. The model
probabilities come from a softmax or a conditional logit over the race's
runners, and both price columns are renormalised within race, so each sums to
one over the field.

Under that constraint the two rules score two different outcomes. **P1 at
$t = 0$ is the exponentiated mean, over races, of the log probability assigned
to the horse that won** — there is exactly one winner per race, so the mean
over winning rows is a mean over races. That is the log score for the
race-level categorical outcome. **P2 sums a Bernoulli log-likelihood over
every runner-row and divides by the row count**, so it scores each runner's
marginal win probability as an independent binary event and does not use the
one-winner constraint. Each is proper for the outcome it scores.

The series' own headline ranking measure is built the same way as P1:
`score_p1_rank` is the exponentiated mean log probability of the observed
top-three finishing order, which is P1's construction at depth three.

## Two changes

**Disjoint bins instead of nested thresholds.** Subsetting rows on the
source's own probability before scoring affects both rules equally: under
$P_1(t)$ and $P_2(t)$ the subset at $t = 0.15$ still contains every row above
0.15, so consecutive thresholds are nearly the same rows and no band can be
read on its own. Disjoint bins are what address that. The bins below 0.10 are
two, two, two and four points wide, finer than any threshold the series has
used, because that is where the models' long-priced statements sit.

**Two benchmarks.** The over-round-adjusted settled starting price, as the
series has always used, and the overnight racecard forecast price from
`daily_runners`, renormalised per race the same way. Paper 4 found the model
ties the first and beats the second, so one benchmark alone cannot say whether
a model's probabilities are good.

Every source is scored on the same rows. Paper 4's price panel supplies both
benchmark probabilities and papers 2b and 3's stored predictions; paper 5's
encoder is joined on from its own stored test predictions.

```{r}
#| label: tbl-p2-cover
p6m_tbl(
  p6m_p2_cover |>
    dplyr::transmute(
      Rows = p6m_n(rows_test), Races = p6m_n(races_test),
      `Rows scored` = p6m_n(rows_panel),
      `Races scored` = p6m_n(races_panel)),
  caption = "The test split against the rows both benchmarks cover."
)
```

Requiring a usable pre-race forecast price costs
`r p6m_n(p6m_p2_cover$rows_dropped)` rows in
`r p6m_n(p6m_p2_cover$races_dropped)` race, so both benchmarks are scored on
identical rows.

## The three models under both benchmarks

```{r}
#| label: tbl-scores
p6m_scores |>
  dplyr::mutate(score_txt = sprintf("%.5f", score),
                ci_txt = sprintf("[%.5f, %.5f]", ci_lo, ci_hi)) |>
  dplyr::select(source, rule, score_txt, ci_txt) |>
  tidyr::pivot_wider(names_from = rule,
                     values_from = c(score_txt, ci_txt)) |>
  dplyr::arrange(dplyr::desc(as.numeric(score_txt_P2))) |>
  dplyr::transmute(Source = source, P1 = score_txt_P1,
                   `P1 interval` = ci_txt_P1, P2 = score_txt_P2,
                   `P2 interval` = ci_txt_P2) |>
  p6m_tbl(caption = "Both rules, every source, over every scored runner-row. Higher is better. Race-level bootstrap, B = 2000, seed 42.")
```

The two rules give the same ordering,
`r p6m_rule_agreement$all_sources_order`, so nothing below turns on which is
read.

```{r}
#| label: tbl-scores-bench
p6m_tbl(
  p6m_scores_bench |>
    dplyr::transmute(
      Rule = rule, Model = model, Benchmark = benchmark,
      Difference = sprintf("%+.5f", diff_point),
      `90% interval` = sprintf("[%+.5f, %+.5f]", ci_lo, ci_hi)),
  caption = "Each model against each benchmark, on both rules, paired by race. Every interval excludes zero."
)
```

All `r n_bench` contrasts exclude zero: every model loses to the starting
price and beats the forecast price on both rules. That is paper 4's finding in
a different measure, and it is why one benchmark is not enough. Against the
settled price alone the models look worse than the market; against the
overnight price alone they look better; both are true.

## The three models against each other

```{r}
#| label: tbl-scores-pairs
p6m_tbl(
  p6m_scores_pairs |>
    dplyr::transmute(
      Rule = rule, Contrast = contrast,
      Difference = sprintf("%+.5f", diff_point),
      `90% interval` = sprintf("[%+.5f, %+.5f]", ci_lo, ci_hi),
      `Excludes zero` = p6m_yn(excludes_zero)),
  caption = "The three models against each other, on both rules, paired by race."
)
```

On both rules the encoder separates from the two earlier models and papers 2b
and 3 do not separate from each other. @sec-roi could do neither: the
single-bet return difference between the encoder and paper 3 was
`r p6m_pp(p6m_row(p6m_contrasts, "contrast", "5 - 3")$diff_point, 1)` with an
interval `r sprintf("%.0f", 100 * (p6m_row(p6m_contrasts, "contrast", "5 - 3")$ci_hi - p6m_row(p6m_contrasts, "contrast", "5 - 3")$ci_lo))`
points wide. Both rules are computed over all
`r p6m_n(sc("P2", "p_5")$n_rows)` runner-rows; the return difference was
computed over the races Owen's rule selects, and within those over the
handful that won.

## By bin

The bins carry calibration and no score. A score computed inside a bin is
largely set by the bin's win rate, and P1 reads only the winning rows, so four
of the fifty source-by-bin cells hold no winner at all. What a source states
against what happens is the reading that survives, and it is the only one the
bins are used for here.

```{r}
#| label: tbl-bins-calib
p6m_bins |>
  dplyr::mutate(src = unname(SHORT[source]),
                v = sprintf("%+.1f", 100 * calib_gap)) |>
  dplyr::select(bin, src, v) |>
  tidyr::pivot_wider(names_from = src, values_from = v) |>
  dplyr::rename(Bin = bin) |>
  p6m_tbl(caption = paste("Observed win rate less mean stated probability, in percentage points. A negative gap means the source states more than it wins.", KEY))
```

Every source states more than it wins at the long-priced end and less than it
wins at the short-priced end, and the models overshoot further than the prices
do. The encoder's 30-35% bin wins
`r sprintf("%+.1f", 100 * p6m_row(bin_enc, "bin", "30-35%")$calib_gap)` points
more often than it says, against
`r sprintf("%+.1f", 100 * p6m_row(bin_sp, "bin", "30-35%")$calib_gap)` for the
starting price, which is within
`r sprintf("%.1f", 100 * max(abs(bin_sp$calib_gap[bin_sp$bin != "35%+"])))`
points of its stated probability in every bin but its top one. The nested form
could not show this, because it pooled everything above each threshold and the
upper bands were swamped by the rows below them.

The bottom bin is thin and its gap is close to minus the stated probability by
construction: it holds `r p6m_n(min(bot$n))` to `r p6m_n(max(bot$n))` rows
with at most `r max(bot$wins)` winners depending on the source.
