# P1 out, P2 refined {#sec-p2}

```{r}
#| label: p2-key
#| include: false
src_sp <- p6m_row(p6m_p2_all, "column", "p_sp")
src_fc <- p6m_row(p6m_p2_all, "column", "p_fc")
src_5  <- p6m_row(p6m_p2_all, "column", "p_5")
src_2b <- p6m_row(p6m_p2_all, "column", "p_2b")
src_3  <- p6m_row(p6m_p2_all, "column", "p_3")
d_5_sp <- p6m_p2_diffs |>
  dplyr::filter(model == "Paper 5 encoder", benchmark == "Starting price")
d_5_fc <- p6m_p2_diffs |>
  dplyr::filter(model == "Paper 5 encoder", benchmark == "Forecast price")
pair_5_3 <- p6m_row(p6m_p2_pairs, "contrast", "Paper 5 encoder - Paper 3 GBT")
pair_2b_3 <- p6m_row(p6m_p2_pairs, "contrast", "Paper 2b logit - Paper 3 GBT")
bin_enc <- p6m_p2_bins |> dplyr::filter(column == "p_5")
bin_sp <- p6m_p2_bins |> dplyr::filter(column == "p_sp")
thin_rows <- p6m_p2_bins |> dplyr::filter(thin)

# Short column labels for the pivoted bin tables. A header long enough to wrap
# shifts every value in its column up a row in the PDF, so these stay to two
# or three characters and the caption carries the key.
SHORT <- c("Paper 2b logit" = "2b", "Paper 3 GBT" = "3",
           "Paper 5 encoder" = "5", "Starting price" = "SP",
           "Forecast price" = "FP")
KEY <- "Columns: 2b = paper 2b logit, 3 = paper 3 GBT, 5 = paper 5 encoder, SP = starting price, FP = forecast price."
pivot_bins <- function(d, value_col) {
  d |>
    dplyr::mutate(src = unname(SHORT[source])) |>
    dplyr::select(bin, src, dplyr::all_of(value_col)) |>
    tidyr::pivot_wider(names_from = src,
                       values_from = dplyr::all_of(value_col)) |>
    dplyr::rename(Bin = bin)
}
```

Owen's paper 1 reports two scoring rules over the runner-rows whose
probability exceeds a threshold $t$:

- $P_1(t)$, the geometric mean of the probability assigned to winners;
- $P_2(t)$, the geometric mean of the per-row Bernoulli likelihood.

**P1 is dropped.** It reads only the winning rows, so a source can raise it by
inflating every probability it publishes and nothing in the rule penalises
that. P2 reads the losing rows too and is the proper rule of the pair.

P2 is kept with three changes.

**Disjoint bins instead of nested thresholds.** Under $P_2(t)$ the subset at
$t = 0.15$ still contains every row above 0.15, so no band can be read on its
own and consecutive thresholds are nearly the same rows. Here each row falls
in exactly one bin.

**Finer resolution below 0.10.** A 0.10 threshold pools every price longer
than 9/1, which is where the models' long-priced statements sit. The bins
below 0.10 are two, two, two and four points wide. The bottom bin is kept at
0 to 2% rather than merged upward: it is thin, and the thinness is itself
reportable, since the models almost never state a probability that low. Its
one consequence is that `r nrow(thin_rows)` bin of the
`r nrow(p6m_p2_bins)` has too few rows to support an interval.

**Two benchmarks.** P2 is reported against the over-round-adjusted settled
starting price, as the series has always done, and against the overnight
racecard forecast price from `daily_runners`, renormalised per race the same
way. Paper 4 found the model ties the first and beats the second, so one
benchmark alone cannot say whether a model's probabilities are good.

Every source is scored on the same rows. Paper 4's price panel supplies both
benchmark probabilities and papers 2b and 3's stored predictions; paper 5's
encoder is joined on from its own stored test predictions.

```{r}
#| label: tbl-p2-cover
p6m_tbl(
  p6m_p2_cover |>
    dplyr::transmute(
      Rows = p6m_n(rows_test), Races = p6m_n(races_test),
      `Rows scored` = p6m_n(rows_panel),
      `Races scored` = p6m_n(races_panel)),
  caption = "The test split against the rows both benchmarks cover."
)
```

Requiring a usable pre-race forecast price costs
`r p6m_n(p6m_p2_cover$rows_dropped)` rows in
`r p6m_n(p6m_p2_cover$races_dropped)` race, so the forecast price covers the
test split almost completely and both benchmarks are scored on identical rows.

## The three models under both benchmarks

```{r}
#| label: tbl-p2-overall
p6m_tbl(
  p6m_p2_all |>
    dplyr::transmute(
      Source = source, P2 = sprintf("%.5f", p2),
      `90% interval` = sprintf("[%.5f, %.5f]", ci_lo, ci_hi)),
  caption = "P2 over every scored runner-row. Higher is better. Race-level bootstrap, B = 2000, seed 42."
)
```

```{r}
#| label: tbl-p2-diffs
p6m_tbl(
  p6m_p2_diffs |>
    dplyr::transmute(
      Model = model, Benchmark = benchmark,
      Difference = sprintf("%+.5f", diff_point),
      `90% interval` = sprintf("[%+.5f, %+.5f]", ci_lo, ci_hi),
      `Excludes zero` = p6m_yn(excludes_zero)),
  caption = "Each model against each benchmark on P2, paired by race."
)
```

The model ordering is the same under both benchmarks, because both are scored
on the same rows and so the models' own P2 cannot change between them. That is
worth stating because it was the thing to check: a benchmark available on
fewer races would have moved the ordering through the row set rather than
through the models.

What changes completely is the verdict. Every model loses to the starting
price and every model beats the forecast price, and all six intervals exclude
zero.

```{r}
#| label: tbl-p2-order
p6m_tbl(
  p6m_p2_order |>
    dplyr::transmute(
      Benchmark = benchmark,
      Above = n_models_above,
      `Above, excl. zero` = n_above_excluding_zero,
      `Below, excl. zero` = n_below_excluding_zero),
  caption = "How many of the three models beat each benchmark on P2."
)
```

The ordering under both benchmarks is
`r unique(p6m_p2_order$"model ordering")`.

The encoder loses `r sprintf("%.5f", abs(d_5_sp$diff_point))` to the starting
price and beats the forecast price by
`r sprintf("%.5f", d_5_fc$diff_point)`. This is paper 4's result in a
different measure: the model knows less than the settled price and more than
the overnight one.

## What P2 resolves that ROI did not

```{r}
#| label: tbl-p2-pairs
p6m_tbl(
  p6m_p2_pairs |>
    dplyr::transmute(
      Contrast = contrast, Difference = sprintf("%+.5f", diff_point),
      `90% interval` = sprintf("[%+.5f, %+.5f]", ci_lo, ci_hi),
      `Excludes zero` = p6m_yn(excludes_zero)),
  caption = "The three models against each other on P2, paired by race."
)
```

P2 separates the encoder from both earlier models, with both intervals
excluding zero, and reports papers 2b and 3 as a tie. @sec-roi could do
neither: the single-bet ROI difference between the encoder and paper 3 was
`r p6m_pp(p6m_row(p6m_contrasts, "contrast", "5 - 3")$diff_point, 1)` with an
interval `r sprintf("%.0f", 100 * (p6m_row(p6m_contrasts, "contrast", "5 - 3")$ci_hi - p6m_row(p6m_contrasts, "contrast", "5 - 3")$ci_lo))`
points wide, and every contrast after the first contained zero.

The reason is the one @sec-roi ends on. P2 is computed over all
`r p6m_n(src_5$n_rows)` scored runner-rows; the ROI difference was computed
over the races where Owen's rule finds a bet, and within those over the
handful that won.

## By bin

```{r}
#| label: tbl-p2-bins
p6m_p2_bins |>
  dplyr::mutate(v = sprintf("%.4f", p2)) |>
  pivot_bins("v") |>
  p6m_tbl(caption = paste("P2 by disjoint bin. Bins are cut on the scored source's own probability, so a bin holds different rows for different sources.", KEY))
```

```{r}
#| label: tbl-p2-bin-counts
p6m_p2_bins |>
  dplyr::mutate(v = sprintf("%s (%s)", p6m_n(n), p6m_n(wins))) |>
  pivot_bins("v") |>
  p6m_tbl(caption = paste(sprintf("Rows in each bin, winners in brackets. The one bin too thin to support an interval is %s's bottom bin: %s rows, %s winners.", thin_rows$source[1], p6m_n(thin_rows$n[1]), p6m_n(thin_rows$wins[1])), KEY))
```

P2 is not comparable across bins. The best attainable value depends on the win
rate in the bin, so a bin with few winners scores close to 1 for any source,
which is why the counts sit beside the scores. The comparison that is valid
within a bin is between what a source states and what happens.

```{r}
#| label: tbl-p2-calib
p6m_p2_bins |>
  dplyr::mutate(v = sprintf("%+.1f", 100 * calib_gap)) |>
  pivot_bins("v") |>
  p6m_tbl(caption = paste("Observed win rate less mean stated probability, in percentage points. A negative gap means the source states more than it wins.", KEY))
```

The bins show where the models lose to the starting price, which the nested
form could not: it pooled everything above each threshold, so the upper bands
were swamped by the rows below them.

The encoder states more than it wins at the long-priced end and less than it
wins at the short-priced end. In its 4-6% bin it states
`r p6m_pct(p6m_row(bin_enc, "bin", "4-6%")$mean_p, 2)` and wins
`r p6m_pct(p6m_row(bin_enc, "bin", "4-6%")$win_rate, 2)`; in its 30-35% bin
it states `r p6m_pct(p6m_row(bin_enc, "bin", "30-35%")$mean_p, 2)` and wins
`r p6m_pct(p6m_row(bin_enc, "bin", "30-35%")$win_rate, 2)`, a gap of
`r sprintf("%+.1f", 100 * p6m_row(bin_enc, "bin", "30-35%")$calib_gap)`
points. The starting price is within
`r sprintf("%.1f", 100 * max(abs(p6m_row(bin_sp, "bin", "4-6%")$calib_gap), abs(p6m_row(bin_sp, "bin", "25-30%")$calib_gap)))`
points of its stated probability in every bin from 4-6% through 30-35%, and
misses only in its top bin, where it states
`r p6m_pct(p6m_row(bin_sp, "bin", "35%+")$mean_p, 2)` and wins
`r p6m_pct(p6m_row(bin_sp, "bin", "35%+")$win_rate, 2)`.
