# How much of the model's disagreement the market accepts {#sec-beta}

```{r}
#| label: beta-key
#| include: false
b_slope <- p6m_row(p6m_beta, "term", "slope")
b_int <- p6m_row(p6m_beta, "term", "intercept")
p4_sp <- p6m_p4_cited |> dplyr::filter(benchmark == "settled starting price")
p4_fc <- p6m_p4_cited |> dplyr::filter(benchmark == "overnight forecast price")
p4_fc_a <- p4_fc |> dplyr::filter(sample == "test-A")
p4_fc_b <- p4_fc |> dplyr::filter(sample == "test-B")
```

**This section is speculation. Nothing in the paper rests on it.**

Paper 4 fitted a two-stage conditional logit with a market probability and a
model probability, both coefficients free, and found the model adds nothing
distinguishable against the settled starting price and a great deal against
the overnight racecard forecast price.

```{r}
#| label: tbl-p4-cited
p6m_tbl(
  p6m_p4_cited |>
    dplyr::transmute(
      Benchmark = benchmark, Sample = sample, Model = model,
      Coefficient = sprintf("%.3f", estimate),
      `95% interval` = sprintf("[%+.3f, %+.3f]", conf_low, conf_high)),
  caption = "Paper 4's model-probability coefficients, read from its own stored fits."
)
```

That result is taken as given here and not re-argued. What it implies is that
the model's information, such as it is, sits between the overnight price and
the off. The quantity that matters is then how much of the model's
disagreement with the overnight price the market subsequently accepts.

## The regression

On the test split, on the encoder's stored predictions, with $p_{\text{close}}$
the over-round-adjusted settled starting price, $p_{\text{fcst}}$ the
renormalised overnight forecast price and $\hat p$ the encoder's probability:

$$
\operatorname{logit}(p_{\text{close}}) - \operatorname{logit}(p_{\text{fcst}})
  = \alpha + \beta\left[\operatorname{logit}(\hat p) -
    \operatorname{logit}(p_{\text{fcst}})\right] + \varepsilon .
$$

Standard errors are a **race-level bootstrap**, B = 2000, seed 42, which is
the resampling unit and replicate count used for every other interval in this
paper. The alternative was race-clustered analytic errors; the bootstrap is
used for consistency with the rest of the series.

The rows are the same `r p6m_n(b_slope$n_rows)` runner-rows in
`r p6m_n(b_slope$n_races)` races that @sec-p2 scores, because the regression
needs both prices and the forecast-price requirement bites in the same place.

```{r}
#| label: tbl-beta
p6m_tbl(
  p6m_beta |>
    dplyr::transmute(
      Term = term, Estimate = sprintf("%.4f", estimate),
      `Std. error` = sprintf("%.4f", se),
      `90% interval` = sprintf("[%+.4f, %+.4f]", ci_lo, ci_hi)),
  caption = "The pooled fit. Standard errors and intervals from a race-level bootstrap, B = 2000, seed 42."
)
```

The pooled coefficient is `r sprintf("%.3f", b_slope$estimate)` with a
standard error of `r sprintf("%.3f", b_slope$se)` and a 90% interval of
`r sprintf("[%+.3f, %+.3f]", b_slope$ci_lo, b_slope$ci_hi)`.

What that says is narrow. Where the encoder disagrees with the published
forecast, the market's own later move away from that forecast goes about
`r sprintf("%.0f%%", 100 * b_slope$estimate)` of the way in the same
direction, so the encoder anticipates most of a move the market was going to
make anyway. It is not a return, not an edge, and not a price improvement. It
is agreement between two opinions about a third, and the third is a published
forecast rather than a price anyone bet into.

The intercept is `r sprintf("%.3f", b_int$estimate)`. Both price columns are
renormalised to sum to one within each race, so no difference in over-round
between the two books survives into the fit; what remains is that the settled
book is more spread out than the forecast book, putting
`r p6m_pct(p6m_row(p6m_spread, "source", "settled starting price")$mean_favourite_prob, 1)`
on its favourite against
`r p6m_pct(p6m_row(p6m_spread, "source", "overnight forecast price")$mean_favourite_prob, 1)`,
so the typical runner's closing logit sits below its forecast logit even where
the model and the forecast agree.

Only the pooled coefficient is reported. There is no breakdown by price band,
field size or anything else, because that would be a search and @sec-roi is
the reason searches on this data do not work.

## Two caveats

**The forecast price is a compiler's published opinion, not a price money has
traded at.** `daily_runners.forecast_price_decimal` is written the evening
before racing by someone whose job is to publish an estimate. An early
exchange price aggregates people who have committed funds, and it will
already contain much of what a compiler's forecast misses. The coefficient
above is therefore an upper bound on anything tradeable: whatever fraction of
the model's disagreement with a compiler the market later accepts, the
fraction it would accept from an early traded price is smaller.

**A coefficient near zero would close the idea off. A positive coefficient
does not establish it.** This result is the second of those. At
`r sprintf("%.3f", b_slope$estimate)`, with an interval well clear of zero,
the market does move in the encoder's direction rather than ignoring it, so
the idea is not closed off. That is all it shows. It does not show that the
movement can be captured, that it is large enough to survive costs, or that
any of it survives against a price that has been traded rather than
published.
