# Data and features {#sec-data}

```{r}
#| label: p3-data-setup
#| include: false

n_races_ranking_2b <- dplyr::n_distinct(ranking_eval_runners_2b$race_id)
n_races_winonly_2b <- dplyr::n_distinct(test_predictions_2b$race_id)
n_races_p3_test    <- dplyr::n_distinct(gbt_test_data$key$race_id)
n_races_p3_train   <- dplyr::n_distinct(gbt_train_data$key$race_id)

feature_names <- gbt_train_data$feature_names
n_features    <- length(feature_names)

# Race-level going_ordinal distribution over the training races (constant
# within a race, so de-duplicate on race_id first).
going_ordinal_by_race <- tibble::tibble(
  race_id       = gbt_train_data$key$race_id,
  going_ordinal = gbt_train_data$X[, "going_ordinal"]
) |>
  dplyr::distinct(race_id, going_ordinal)
pct_standard_train <- 100 * mean(going_ordinal_by_race$going_ordinal == 4)

# going_sr_delta: NA share by split, and exact-zero share among defined
# values (train).
going_na_share <- function(X, label) {
  v <- X[, "going_sr_delta"]
  tibble::tibble(split = label, n = length(v), n_na = sum(is.na(v)),
                 pct_na = 100 * sum(is.na(v)) / length(v))
}
going_na_table <- dplyr::bind_rows(
  going_na_share(gbt_train_data$X, "train"),
  going_na_share(gbt_test_data$X,  "test")
)
delta_train <- gbt_train_data$X[, "going_sr_delta"]
delta_train_defined <- delta_train[!is.na(delta_train)]
n_delta_defined  <- length(delta_train_defined)
n_delta_zero     <- sum(delta_train_defined == 0)
pct_delta_zero   <- 100 * n_delta_zero / n_delta_defined

# Train-split-only decile table of going_sr_delta (general model-ready
# population), with mean field size joined in.
field_size <- qualifying_runners |>
  dplyr::group_by(race_id) |>
  dplyr::summarise(field_size = dplyr::n(), .groups = "drop")

train_decile <- runners_model_ready |>
  dplyr::filter(split == "train", !is.na(going_sr_delta)) |>
  dplyr::left_join(field_size, by = "race_id") |>
  dplyr::mutate(decile = dplyr::ntile(going_sr_delta, 10))

decile_table <- train_decile |>
  dplyr::group_by(decile) |>
  dplyr::summarise(
    n               = dplyr::n(),
    delta_min       = min(going_sr_delta),
    delta_max       = max(going_sr_delta),
    win_rate        = mean(won),
    mean_field_size = mean(field_size),
    .groups = "drop"
  )

field_size_range <- range(decile_table$mean_field_size)
```

The data, the race-selection rules, and the feature set before the additions
below are identical to papers 2a and 2b; we refer the reader there (paper 2a,
§2; paper 2b, @sec-data) rather than repeat them. The chronological
race-level train/test split at 2012-12-30 and the complete-case rule
(a race is dropped if any modelling feature is undefined for any of its
runners) both carry over unchanged.

## The test universe

A win backtest needs only to know who won, so paper 2b's largest test
universe keeps every race with a usable win probability. The ranking
construction instead needs a clean top three — finishing positions one,
two and three each recorded exactly once — so it works from a smaller
universe. Paper 3 shares that second requirement, so it uses the
`r format(n_races_p3_test, big.mark = ",")`-race universe throughout, and
@sec-results reports every comparison against papers 2a and 2b on both
universes.

## Feature set

@tbl-feature-set lists the `r n_features` features the model is fitted on:
paper 2b's own final specification, plus the four going-affinity features
introduced in this paper.

```{r}
#| label: tbl-feature-set
tibble::tribble(
  ~Feature,               ~Description,                                                 ~Type,   ~Source,
  "pos_lag1_nonzero",     "Most recent prior finishing position (1st-4th; 0 if none)",   "horse", "paper 1",
  "pos_lag2_nonzero",     "Second most recent prior finishing position",                 "horse", "paper 1",
  "age_diff",             "Symmetric distance from peak age (3 on AW)",                  "horse", "paper 1",
  "days_LTO_log",         "Log days since last run",                                     "horse", "paper 1",
  "trainerSR",            "Cumulative pre-race trainer strike rate",                     "horse", "paper 1",
  "sireSR",               "Cumulative pre-race sire strike rate",                        "horse", "paper 1",
  "jockeySR",             "Cumulative pre-race jockey strike rate",                      "horse", "paper 2a",
  "entire / gelding",     "Sex indicators",                                              "horse", "paper 1",
  "cheekpieces",          "Tack indicator",                                              "horse", "paper 1",
  "rel_weight",           "Weight carried minus race-mean weight",                       "horse", "paper 2a",
  "or_relative",          "Official rating minus race-mean rating (0 where undefined)",  "horse", "paper 2a",
  "or_missing",           "Indicator: official rating was undefined",                    "horse", "paper 2a",
  "trainer_aw_premium",   "Trainer AW strike rate minus overall strike rate",            "horse", "paper 2a",
  "has_wins",             "Indicator: any prior career win",                             "horse", "paper 2a",
  "stall_normalised",     "Starting stall divided by field size",                        "horse", "paper 2a",
  "course (4 indicators)","Which of the four AW courses",                                "race",  "paper 2a",
  "going_runs_prior",     "Count of prior runs on today's kind of going",                "horse", "paper 3",
  "going_sr_shrunk",      "Win rate on today's kind of going, shrunk toward career rate","horse", "paper 3",
  "going_sr_delta",       "going_sr_shrunk minus overall career win rate",               "horse", "paper 3",
  "going_ordinal",        "Today's going, on a common fast-to-slow scale",               "race",  "paper 3"
) |>
  knitr::kable(caption = sprintf("Paper-3 feature set: %d features (course and the sex indicators are each one row here but several terms).", n_features))
```

Two substitutions replace terms that exist only to work around what a
conditional logit needs and a tree does not.

- **Position lags.** Paper 2b encodes each prior finishing position as a
  pair of terms — an indicator for "no prior run" and the raw position
  where one exists — because a linear model needs the split to avoid
  reading a real finish on the same numeric scale as "no run." A tree
  splits on the raw position directly, so the indicator is dropped: four
  terms become two.
- **Course and draw.** Paper 2b's course effect only appears crossed with
  the horse-level draw, because a conditional logit cannot use a
  race-level covariate on its own — it cancels in the softmax. Here the
  draw and the four course indicators are entered separately, and a tree
  is free to split on either, or both together, at any depth, for any
  course — not only the ones the linear reduction happened to keep. Three
  terms become five.

Two further cases keep paper 2b's own form. `or_relative` and its
companion flag `or_missing` stay as built: the imputation is a considered
missing-not-at-random strategy in its own right, not merely an encoding
workaround. `days_LTO_log` also carries over unchanged, since a tree only
asks whether a value falls above or below a threshold, and taking logs
does not change which runners fall on which side.

## Going affinity

`going_runs_prior`, `going_sr_shrunk`, `going_sr_delta` and `going_ordinal`
are the one feature paper 2 deferred to this paper — going is a race-level
quantity that cancels in a conditional-logit softmax unless interacted
with a horse-level term, and a tree needs no such interaction. They are
built from each horse's full cross-surface career history, strictly before
today's race.

Today's going is `r sprintf("%.1f", pct_standard_train)`% "Standard" across
the `r format(n_races_p3_train, big.mark = ",")` training races
(`going_ordinal` = 4 on the project's 1–7 fast-to-slow scale). Because
"Standard" pools AW Standard with turf Good — the two buckets are defined
within surface, at that surface's own terciles of going, and AW's terciles
both land on Standard — `going_sr_shrunk` and `going_sr_delta` describe,
for nearly every runner, a horse's win rate on Standard-or-Good ground
relative to its own career rate. That is a measure of ground-type form, not
a preference for today's specific going in any finer sense; with going this
close to constant across the training data, there is little room for a
finer sense to be estimated from it.

A horse with very few prior runs on today's kind of going has an
unreliable win rate on it, so the figure used, `going_sr_shrunk`, is a
blend of that rate and the horse's overall career rate — weighted toward
the career rate when the horse has run few times on this going, and
toward the observed rate as those runs accumulate:
$(\text{wins}_{\text{bucket}} + m \cdot \text{career\_sr}) /
(\text{runs}_{\text{bucket}} + m)$, with $m = 5$ a judgement call, not
derived from the data or tuned. `going_sr_delta` is that shrunk rate minus
the career rate, the term that separates going affinity from general
ability. `going_sr_delta` is exactly zero by
construction whenever every one of a horse's prior runs falls in the same
going bucket as today's (`r sprintf("%.1f", pct_delta_zero)`% of the
`r format(n_delta_defined, big.mark = ",")` training runner-rows with a
defined value), and undefined for a horse with no prior run in today's
bucket, including a true debut
(`r sprintf("%.2f", going_na_table$pct_na[going_na_table$split == "train"])`%
of training runner-rows,
`r sprintf("%.2f", going_na_table$pct_na[going_na_table$split == "test"])`%
of test). These values are left missing rather than imputed: the model
learns a default routing direction for a missing value at each split,
rather than being given a stand-in value in its place.

```{r}
#| label: tbl-going-decile
#| code-fold: true
decile_table |>
  dplyr::transmute(
    Decile           = decile,
    `Runner-rows`    = format(n, big.mark = ","),
    `Delta range`    = sprintf("[%.3f, %.3f]", delta_min, delta_max),
    `Win rate`       = scales::percent(win_rate, accuracy = 0.1),
    `Mean field size` = round(mean_field_size, 1)
  ) |>
  knitr::kable(caption = "Training-split win rate by going_sr_delta decile, with mean field size.")
```

The win rate rises from decile 1 to decile 10 without a corresponding trend
in mean field size — the ten deciles' mean field sizes range from
`r sprintf("%.1f", field_size_range[1])` to
`r sprintf("%.1f", field_size_range[2])` runners, which rules out a smaller
or larger field as the mechanical explanation. A race-clustered bootstrap
of the win-rate difference between decile 10 and decile 1 — the same
resample-races-with-replacement design as the ROI-difference bootstrap
used throughout this series, races resampled `r going_decile_bootstrap$n_boot`
times with seed 42, the decile assignment and both win rates recomputed
fresh within each resample rather than treating the two deciles as
independent samples — gives a point estimate of
`r sprintf("%+.3f", going_decile_bootstrap$point_diff)` with a 90% CI of
[`r sprintf("%.3f", going_decile_bootstrap$ci_lo)`, `r sprintf("%.3f", going_decile_bootstrap$ci_hi)`].
