# Data and features {#sec-data}

```{r}
#| label: p3-data-setup
#| include: false

n_races_ranking_2b <- dplyr::n_distinct(ranking_eval_runners_2b$race_id)
n_races_winonly_2b <- dplyr::n_distinct(test_predictions_2b$race_id)
n_races_p3_test    <- dplyr::n_distinct(gbt_test_data$key$race_id)
n_races_p3_train   <- dplyr::n_distinct(gbt_train_data$key$race_id)

feature_names <- gbt_train_data$feature_names
n_features    <- length(feature_names)

# Race-level going_ordinal distribution over the training races (constant
# within a race, so de-duplicate on race_id first).
going_ordinal_by_race <- tibble::tibble(
  race_id       = gbt_train_data$key$race_id,
  going_ordinal = gbt_train_data$X[, "going_ordinal"]
) |>
  dplyr::distinct(race_id, going_ordinal)
pct_standard_train <- 100 * mean(going_ordinal_by_race$going_ordinal == 4)

# going_sr_delta: NA share by split, and exact-zero share among defined
# values (train).
going_na_share <- function(X, label) {
  v <- X[, "going_sr_delta"]
  tibble::tibble(split = label, n = length(v), n_na = sum(is.na(v)),
                 pct_na = 100 * sum(is.na(v)) / length(v))
}
going_na_table <- dplyr::bind_rows(
  going_na_share(gbt_train_data$X, "train"),
  going_na_share(gbt_test_data$X,  "test")
)
delta_train <- gbt_train_data$X[, "going_sr_delta"]
delta_train_defined <- delta_train[!is.na(delta_train)]
n_delta_defined  <- length(delta_train_defined)
n_delta_zero     <- sum(delta_train_defined == 0)
pct_delta_zero   <- 100 * n_delta_zero / n_delta_defined

# Train-split-only decile table of going_sr_delta (general model-ready
# population, matching scripts/verify_going_features.R's own diagnostic),
# with mean field size joined in.
field_size <- qualifying_runners |>
  dplyr::group_by(race_id) |>
  dplyr::summarise(field_size = dplyr::n(), .groups = "drop")

train_decile <- runners_model_ready |>
  dplyr::filter(split == "train", !is.na(going_sr_delta)) |>
  dplyr::left_join(field_size, by = "race_id") |>
  dplyr::mutate(decile = dplyr::ntile(going_sr_delta, 10))

decile_table <- train_decile |>
  dplyr::group_by(decile) |>
  dplyr::summarise(
    n               = dplyr::n(),
    delta_min       = min(going_sr_delta),
    delta_max       = max(going_sr_delta),
    win_rate        = mean(won),
    mean_field_size = mean(field_size),
    .groups = "drop"
  )

field_size_range <- range(decile_table$mean_field_size)
```

The data, the race-selection rules, and the feature set before the additions
below are identical to papers 2a and 2b; we refer the reader there (paper 2a,
§2; paper 2b, @sec-data) rather than repeat them. The chronological
race-level train/test split at 2012-12-30 and the complete-case rule
(a race is dropped if any modelling column is undefined for any of its
runners) both carry over unchanged.

## The test universe

Paper 2b evaluates on two different test universes, because its two tasks
need different things from a race. A win backtest only needs to know who
won, so `test_predictions_2b` keeps every test race with a usable win
probability — `r format(n_races_winonly_2b, big.mark = ",")` races. The
ranking task needs a clean top three (positions 1, 2 and 3 each recorded
exactly once, so the field can be exploded into nested choice sets), which
`r format(n_races_winonly_2b - n_races_ranking_2b, big.mark = ",")` fewer
races satisfy; `ranking_eval_runners_2b` accordingly covers
`r format(n_races_ranking_2b, big.mark = ",")` races.

Paper 3's feature matrix is built by the same exploded construction paper
2b's ranking model uses (`prepare_exploded_data()`), because the training
objective is the same depth-3 Plackett–Luce likelihood. Its test build
therefore uses the clean-top-three universe throughout —
`r format(n_races_p3_test, big.mark = ",")` races, matching
`ranking_eval_runners_2b` exactly (checked at build time; see
`R/gbt_results.R::check_gbt_race_universe()`) — for every table in
@sec-results, including the win-picking and betting comparisons against
papers 2a and 2b's own, larger win-only universes. Those comparisons are
reported on both the published universe and this restricted one, so a
reader can see whether the restriction changes anything. The training
side uses the same construction: `r format(n_races_p3_train, big.mark = ",")`
races, the same set `model_2b_exploded_draw_final` was fitted and evaluated
on.

## Feature set

The feature set is paper 2b's own final specification —
`r length(stats::coef(model_2b_exploded_draw_final))` terms — plus the four
going-affinity columns below, giving `r n_features` columns in all:
`r paste(feature_names, collapse = ", ")`.

Two substitutions were made where paper 2b's specification exists only to
work around what a conditional logit needs and a tree does not:

- **Position lags.** Paper 2b encodes each prior finishing position as a
  pair of terms, `pos_lagN_zero` (an indicator for "no prior run") and
  `pos_lagN_nonzero` (the raw position where one exists) — a linear model
  needs the split so a real finish of, say, 2nd is not read on the same
  numeric scale as "no run." A tree splits on `pos_lagN_nonzero == 0`
  directly, so the indicator is dropped and only the raw value kept: four
  columns become two (`pos_lag1_nonzero`, `pos_lag2_nonzero`).
- **Course and draw.** Paper 2b's course effect only appears at all because
  it is crossed with the horse-level draw (`stall_x_kempton`,
  `stall_x_southwell`, `stall_x_wolverhampton`) — a conditional logit cannot
  use a race-level covariate on its own; it cancels in the softmax. Here
  `stall_normalised` and one-hot columns for all four courses
  (`course_Kempton`, `course_Lingfield`, `course_Southwell`,
  `course_Wolverhampton`) are entered separately, and a tree is free to
  split on course, on draw, or on both together, at any depth, for any
  course — not only the three the linear reduction happened to keep. Three
  columns become five.

Two further cases keep paper 2b's own form rather than being substituted.
`or_relative` (the official-rating gap to the race average) and its
companion flag `or_missing` stay as 2b built them: the zero-imputation plus
flag is a considered missing-not-at-random strategy in its own right (an
unrated horse is a distinct population, not a value to be guessed), not
merely an encoding workaround, so there is no clear case for replacing it
with XGBoost's native missing-value handling here. `days_LTO_log` also
carries over unchanged; a tree's splits are invariant to a monotonic
transform of a single feature; keeping the log form costs nothing and
keeps this paper's training pseudo-$R^2$
(`r sprintf("%.5f", model_3_gbt_meta$train_pl_r2)`, in-sample) on the same
footing as paper 2b's own
(`r sprintf("%.5f", paper2b_train_pl_r2)`, likewise in-sample).

## Going affinity

`going_runs_prior`, `going_sr_shrunk`, `going_sr_delta` and `going_ordinal`
are the one feature paper 2 deferred to this paper (going is a race-level
quantity that cancels in a conditional-logit softmax unless interacted with
a horse-level term; a tree needs no such interaction). They are built from
each horse's full cross-surface career history, strictly before today's
race, in `R/build_going_features.R`.

Today's going is `r sprintf("%.1f", pct_standard_train)`% "Standard" across
the `r format(n_races_p3_train, big.mark = ",")` training races
(`going_ordinal` = 4 on the project's 1–7 fast-to-slow scale). Because
"Standard" pools AW Standard with turf Good — the two buckets are defined
within surface, at that surface's own terciles of going, and AW's terciles
both land on Standard — `going_sr_shrunk` and `going_sr_delta` describe,
for nearly every runner, a horse's win rate on Standard-or-Good ground
relative to its own career rate. That is a measure of ground-type form, not
a preference for today's specific going in any finer sense; with going this
close to constant across the training data, there is little room for a
finer sense to be estimated from it.

`going_sr_shrunk` is the horse's win rate in today's going bucket, shrunk
toward its own overall career win rate `career_sr` with prior weight
$m = 5$: $(\text{wins}_{\text{bucket}} + m \cdot \text{career\_sr}) /
(\text{runs}_{\text{bucket}} + m)$. $m = 5$ is a judgement call, not derived
from the data, and is not tuned. `going_sr_delta` is
`going_sr_shrunk` minus `career_sr` — the term that separates going
affinity from general ability, since the raw bucket rate alone would partly
proxy what `or_relative` already carries. `going_sr_delta` is exactly zero
by construction whenever every one of a horse's prior runs falls in the
same going bucket as today's: the bucket win rate and the career win rate
are then identical, so the shrinkage formula returns `career_sr` unchanged.
That accounts for `r sprintf("%.1f", pct_delta_zero)`% of the
`r format(n_delta_defined, big.mark = ",")` training runner-rows with a
defined `going_sr_delta`.

`going_sr_delta` is undefined (`NA`) for a horse with no prior run in
today's bucket — including a true debut, where all three horse-level
columns are `NA` — in
`r sprintf("%.2f", going_na_table$pct_na[going_na_table$split == "train"])`%
of training runner-rows and
`r sprintf("%.2f", going_na_table$pct_na[going_na_table$split == "test"])`%
of test runner-rows. These values are left missing rather than imputed:
XGBoost learns a default split direction for each tree node from the
training data and routes every missing value that way, so the fact that a
horse has no relevant record is available to the model as information in
its own right, rather than being masked by a stand-in value.

```{r}
#| label: tbl-going-decile
#| code-fold: true
decile_table |>
  dplyr::transmute(
    Decile           = decile,
    `Runner-rows`    = format(n, big.mark = ","),
    `Delta range`    = sprintf("[%.3f, %.3f]", delta_min, delta_max),
    `Win rate`       = scales::percent(win_rate, accuracy = 0.1),
    `Mean field size` = round(mean_field_size, 1)
  ) |>
  knitr::kable(caption = "Training-split win rate by going_sr_delta decile, with mean field size.")
```

The win rate rises from decile 1 to decile 10 without a corresponding trend
in mean field size — the ten deciles' mean field sizes range from
`r sprintf("%.1f", field_size_range[1])` to
`r sprintf("%.1f", field_size_range[2])` runners, which rules out a smaller
or larger field as the mechanical explanation. A race-clustered bootstrap
of the win-rate difference between decile 10 and decile 1 —
`R/gbt_results.R::bootstrap_decile_difference()`, the same
resample-races-with-replacement design as the ROI-difference bootstrap
used throughout this series, races resampled `r going_decile_bootstrap$n_boot`
times with seed 42, the decile assignment and both win rates recomputed
fresh within each resample rather than treating the two deciles as
independent samples — gives a point estimate of
`r sprintf("%+.3f", going_decile_bootstrap$point_diff)` with a 90% CI of
[`r sprintf("%.3f", going_decile_bootstrap$ci_lo)`, `r sprintf("%.3f", going_decile_bootstrap$ci_hi)`].
