# Model fitted {#sec-model}

```{r}
#| label: p3-model-setup
#| include: false

all_results <- gbt_tuning_results$all_results
sel         <- gbt_selected_config

n_grid_points   <- nrow(all_results)
total_cv_secs   <- sum(all_results$elapsed_secs)
total_cv_hours  <- total_cv_secs / 3600

# Grid dimensions, read off the actual grid rather than restated by hand.
grid_dims <- list(
  max_depth        = sort(unique(all_results$max_depth)),
  eta              = sort(unique(all_results$eta)),
  min_child_weight = sort(unique(all_results$min_child_weight)),
  subsample        = sort(unique(all_results$subsample)),
  colsample_bytree = sort(unique(all_results$colsample_bytree))
)

# Tie window: points within 0.001 of the grid maximum.
best_val         <- max(all_results$fold_mean_pl_r2)
tie_contenders   <- all_results[all_results$fold_mean_pl_r2 >= best_val - 0.001, ]
n_tie_contenders <- nrow(tie_contenders)

# Paired comparison against the grid maximum: the SAME 5 folds run through
# every grid point, so a paired difference/SE is the correct yardstick
# (not the marginal per-point fold sd).
fold_cols  <- paste0("fold_r2_", 1:5)
max_idx    <- which.max(all_results$fold_mean_pl_r2)
max_row    <- all_results[max_idx, ]
max_folds  <- as.numeric(max_row[fold_cols])
paired_mat <- t(apply(all_results[fold_cols], 1, function(r) as.numeric(r) - max_folds))
paired_mean <- rowMeans(paired_mat)
paired_se   <- apply(paired_mat, 1, function(d) stats::sd(d) / sqrt(length(d)))
# The maximum's own row compares to itself: mean 0, sd 0, se 0 -- defined as
# 0 paired-SE from itself rather than 0/0, so it is counted among the points
# "within 1 paired SE of the maximum" below.
se_multiple <- ifelse(paired_se == 0, 0, abs(paired_mean) / paired_se)
n_within_1se <- sum(se_multiple <= 1)

sel_match <- which(
  all_results$max_depth        == sel$max_depth &
  all_results$eta               == sel$eta &
  all_results$min_child_weight  == sel$min_child_weight &
  all_results$subsample         == sel$subsample &
  all_results$colsample_bytree  == sel$colsample_bytree
)
sel_se_multiple <- se_multiple[sel_match]

# Top 10 by fold_mean_pl_r2: subsample values.
top10 <- all_results[order(-all_results$fold_mean_pl_r2), ][1:10, ]
top10_all_subsample_07 <- all(top10$subsample == 0.7)

sel_is_boundary <- sel$max_depth == max(grid_dims$max_depth) ||
                   sel$eta       == max(grid_dims$eta)

n_144_grid <- length(grid_dims$max_depth) * length(grid_dims$eta) *
              length(grid_dims$min_child_weight) * length(grid_dims$subsample) * 2L

# Tie-break evidence: does the selected model's top-10 gain-importance
# ranking match a refit at the untied grid maximum?
gm <- gbt_gridmax_vs_selected_importance
n_top10_same_feature <- sum(!is.na(gm$gridmax_rank) & gm$gridmax_rank <= 10)
n_top10_same_rank    <- sum(gm$selected_rank == gm$gridmax_rank, na.rm = TRUE)
```

## The objective

The GBT is fitted under the same objective as paper 2b's ranking model:
the depth-$k = 3$ Plackett–Luce likelihood over each race's finishing order
— the derivation is identical, only the function class changes (from a
linear predictor to an additive tree ensemble). `{xgboost}` takes a custom
objective as a gradient and a Hessian, one value of each per training row;
`R/pl_objective.R` supplies both in closed form (@sec-appx-derivations).
The Hessian XGBoost consumes is the diagonal only — the second derivative
of each row's own score, ignoring the cross terms between runners in the
same race that the full Plackett–Luce likelihood actually has (a race's
denominator couples every runner still in contention at each stage). This
is an approximation to the true Hessian, standard practice for a grouped
loss under a boosting library built around one Hessian value per row, and
is the same approximation underlying `{xgboost}`'s own built-in ranking
objectives.

`{xgboost}`'s built-in ranking objectives (`rank:pairwise`, `rank:ndcg`,
LambdaMART-style objectives) are a different objective — they optimise
pairwise or list-wise ranking losses, not the Plackett–Luce likelihood —
and were not used; the custom objective is what keeps this paper's model
and paper 2b's model comparable at the same likelihood.

## Tuning procedure

The tuning grid was fixed in advance of fitting anything, before any
test-set number existed: a full factorial over five hyperparameters,
`r n_grid_points` points in all.

```{r}
#| label: tbl-tuning-grid
knitr::kable(
  tibble::tibble(
    Hyperparameter = c("max_depth", "eta", "min_child_weight", "subsample", "colsample_bytree"),
    `Values tested` = c(
      paste(grid_dims$max_depth, collapse = ", "),
      paste(grid_dims$eta, collapse = ", "),
      paste(grid_dims$min_child_weight, collapse = ", "),
      paste(grid_dims$subsample, collapse = ", "),
      paste(grid_dims$colsample_bytree, collapse = ", ")
    )
  ),
  caption = sprintf("The %d-point tuning grid (full factorial).", n_grid_points)
)
```

Each of the `r n_grid_points` points was fit under race-grouped 5-fold
cross-validation: races, not runner-rows, are assigned to folds, so no
race is split across an analysis/assessment pair. Within each fold, boosting
ran for up to 2,000 rounds with early stopping — the number of rounds is
determined by the fold's own held-out Plackett–Luce pseudo-$R^2$
(`pl_r2`), patience 50 rounds. This per-fold early-stopped round count
decides `nrounds` **during selection only**; a separate decision governs
the eventual refit on all training data, below.

**Fallback ladder.** The full grid also varies `colsample_bytree` over
`{0.7, 1.0}` — `r n_144_grid` points, five-fold, projected at roughly 12.9
hours, not comfortably under this project's working limit of about 12
hours for a single run. The fixed rule was to stop cutting as soon as the
projection came in under that limit. Fixing `colsample_bytree` at `{0.7}`
alone cut the grid to the `r n_grid_points` points actually run, projected
at roughly 6.5 hours, and that was enough: the further rungs (also fixing
`subsample`, dropping `eta = 0.01`, or narrowing `min_child_weight`) were
not needed. The measured cost came in higher than the projection: summed
across all `r n_grid_points` points, cross-validation used
`r sprintf("%.1f", total_cv_hours)` hours — at that rate the un-reduced
`r n_144_grid`-point grid would have taken roughly twice as long,
confirming the cut was necessary rather than a precaution that turned out
unneeded.

**Selection rule.** The winning point is the one with the highest
fold-mean `pl_r2`; among points within 0.001 of that maximum, the rule
prefers the shallowest tree, then the fewest rounds. `r n_tie_contenders`
of the `r n_grid_points` points fell within that 0.001 window, and the
selected point —
max_depth `r sel$max_depth`, eta `r sel$eta`, min_child_weight
`r sel$min_child_weight`, subsample `r sel$subsample`, colsample_bytree
`r sel$colsample_bytree` — was the shallowest and fastest-converging among
them.

Because the same 5 folds run through every grid point, a **paired**
difference and standard error against the grid maximum is the correct
comparison — not the marginal spread of fold scores within each point
taken on its own. On that basis, `r n_within_1se` of `r n_grid_points`
points (the maximum itself included) sit within one paired standard error
of the maximum; the selected point is not one of them; it sits
`r sprintf("%.2f", sel_se_multiple)` paired standard errors from the
maximum. The 0.001 tie window used for selection is therefore wider than
the paired standard errors actually observed near the top of this grid —
a fixed constant chosen before the paired structure of the data was known,
rather than a multiple of it. The rule was fixed in advance and stands;
this note records that, as specified, it selected among points a paired
test would not call tied.

Whether that gap matters was checked with one refit: the grid maximum
(max_depth `r gbt_gridmax_config$max_depth`, eta `r gbt_gridmax_config$eta`,
min_child_weight `r gbt_gridmax_config$min_child_weight`, subsample
`r gbt_gridmax_config$subsample`) refit on all training data at its own
fold-mean rounds (`r gbt_gridmax_meta$nrounds`), the same way the selected
model was. This is one refit; it does not reopen model selection, which
stays the point chosen by the fixed rule above.

```{r}
#| label: tbl-gridmax-importance
#| code-fold: true
gm |>
  dplyr::transmute(
    Feature,
    `Selected: rank` = selected_rank,
    `Selected: gain` = round(selected_gain, 4),
    `Grid max: rank` = gridmax_rank,
    `Grid max: gain` = round(gridmax_gain, 4)
  ) |>
  knitr::kable(caption = "Gain importance, top 10 by the selected model's own ranking, compared against a refit at the untied grid maximum.")
```

`r n_top10_same_feature` of the selected model's top 10 features by gain
also appear in the grid maximum's own top 10, and `r n_top10_same_rank`
of those hold the identical rank in both models. The tie-break's practical
consequences are confined to whatever reordering the table above shows;
it did not admit a materially different model.

Every one of the top ten points by fold-mean `pl_r2`
`r if (top10_all_subsample_07) "has subsample 0.7" else "does not all share one subsample value"`.
The selected configuration is
`r if (sel_is_boundary) "" else "not "`on a boundary of the grid
(`max_depth` `r if (sel$max_depth == max(grid_dims$max_depth)) "is" else "is not"`
the largest tested, `eta`
`r if (sel$eta == max(grid_dims$eta)) "is" else "is not"` the largest
tested).

No grid point diverged: `gbt_tuning_results$n_grid_points_with_divergence`
is `r gbt_tuning_results$n_grid_points_with_divergence` across all
`r n_grid_points` points (`R/gbt_tuning.R`'s divergence guard logs, rather
than silently drops, any configuration whose scores run away during
boosting; none did here).

**The refit.** The selected configuration is refit once on all training
data, no held-out set and no early stopping (there is none left to stop
against): `nrounds` is the **mean**, across the selected point's five
folds, of each fold's own early-stopped round count, rounded —
`r model_3_gbt_meta$nrounds` rounds — rather than the fold median or the
fold maximum, for consistency with fold-mean `pl_r2` already being the
grid's own selection criterion. The resulting fit's training
$\text{pl\_r2}$ is `r sprintf("%.5f", model_3_gbt_meta$train_pl_r2)`,
in-sample.

**Where the compute goes.** The custom objective and its matching
evaluation metric are each evaluated once per boosting round — every round
needs both the gradient/Hessian pair for the next tree and the held-out
`pl_r2` for early stopping — and this dominates the per-round cost;
building the trees themselves, once the gradient and Hessian are in hand,
is comparatively cheap. Summed cross-validation time for the
`r n_grid_points`-point grid was `r sprintf("%.1f", total_cv_hours)` hours.

**A limitation carried by row subsampling.** `subsample` in XGBoost draws
a random fraction of training *rows* before growing each tree — rows here
are individual horses, not races, because XGBoost has no notion of a
race-level grouping to subsample by. Under the Plackett–Luce objective,
rows in the same race are not independent: a race's likelihood couples
every horse still in contention at each stage, so dropping some of a
race's runners at random changes the remaining rows' own gradients and
Hessians, not just the sample size. Every one of the top ten configurations
in this grid uses `subsample = 0.7`, so this row-level, cross-race
subsampling clearly helps; @sec-discussion returns to what it means that
the subsampling unit and the likelihood's own grouping unit do not match.
