# Model fitted {#sec-model}

```{r}
#| label: p3-model-setup
#| include: false

all_results <- gbt_tuning_results$all_results
sel         <- gbt_selected_config

n_grid_points   <- nrow(all_results)
total_cv_secs   <- sum(all_results$elapsed_secs)
total_cv_hours  <- total_cv_secs / 3600

# Grid dimensions, read off the actual grid rather than restated by hand.
grid_dims <- list(
  max_depth        = sort(unique(all_results$max_depth)),
  eta              = sort(unique(all_results$eta)),
  min_child_weight = sort(unique(all_results$min_child_weight)),
  subsample        = sort(unique(all_results$subsample)),
  colsample_bytree = sort(unique(all_results$colsample_bytree))
)

# Top 10 by fold_mean_pl_r2: subsample values (Appendix B).
top10 <- all_results[order(-all_results$fold_mean_pl_r2), ][1:10, ]
top10_all_subsample_07 <- all(top10$subsample == 0.7)

sel_is_boundary <- sel$max_depth == max(grid_dims$max_depth) ||
                   sel$eta       == max(grid_dims$eta)

# Fallback ladder (Appendix B): the un-reduced grid also varies
# colsample_bytree over two values.
n_144_grid <- length(grid_dims$max_depth) * length(grid_dims$eta) *
              length(grid_dims$min_child_weight) * length(grid_dims$subsample) * 2L

# Tie-break evidence: does the selected model's top-10 gain-importance
# ranking match a refit at the untied grid maximum?
gm <- gbt_gridmax_vs_selected_importance
n_top10_same_rank <- sum(gm$selected_rank == gm$gridmax_rank, na.rm = TRUE)

# Paired comparison against the grid maximum (§6): the same 5 folds run
# through every grid point, so a paired difference/SE is the right
# yardstick for how flat the grid is near the top -- not the marginal
# per-point fold sd.
fold_cols   <- paste0("fold_r2_", 1:5)
max_idx     <- which.max(all_results$fold_mean_pl_r2)
max_folds   <- as.numeric(all_results[max_idx, fold_cols])
paired_mat  <- t(apply(all_results[fold_cols], 1, function(r) as.numeric(r) - max_folds))
paired_mean <- rowMeans(paired_mat)
paired_se   <- apply(paired_mat, 1, function(d) stats::sd(d) / sqrt(length(d)))
# The maximum's own row compares to itself: mean 0, sd 0 -- defined as 0
# paired-SE from itself (not 0/0) so it counts among the points "within
# 1 paired SE of the maximum".
se_multiple  <- ifelse(paired_se == 0, 0, abs(paired_mean) / paired_se)
n_within_1se <- sum(se_multiple <= 1)
```

## The objective

The GBT is fitted under the same objective as paper 2b's ranking model:
the depth-$k = 3$ Plackett–Luce likelihood over each race's finishing
order — the derivation is identical, only the function class changes (from
a linear predictor to an additive tree ensemble). This objective is
supplied as a gradient and a Hessian, one value of each per training row
(@sec-appx-derivations). The Hessian is the diagonal only — the second
derivative of each row's own score, dropping the cross terms between
runners in the same race that the full likelihood actually has. This is a
standard approximation for a grouped loss fitted one row-level Hessian at
a time, and the same one the function class's own built-in ranking
objectives make.

Those built-in ranking objectives (pairwise and list-wise losses) are a
different objective from the Plackett–Luce likelihood and were not used;
the custom objective is what keeps this paper's model and paper 2b's
comparable at the same likelihood.

## Tuning procedure

The tuning grid was fixed in advance of fitting anything, before any
test-set number existed: a full factorial over five hyperparameters,
`r n_grid_points` points in all.

```{r}
#| label: tbl-tuning-grid
knitr::kable(
  tibble::tibble(
    Hyperparameter = c("max_depth", "eta", "min_child_weight", "subsample", "colsample_bytree"),
    `Values tested` = c(
      paste(grid_dims$max_depth, collapse = ", "),
      paste(grid_dims$eta, collapse = ", "),
      paste(grid_dims$min_child_weight, collapse = ", "),
      paste(grid_dims$subsample, collapse = ", "),
      paste(grid_dims$colsample_bytree, collapse = ", ")
    )
  ),
  caption = sprintf("The %d-point tuning grid (full factorial).", n_grid_points)
)
```

Each point was fit under race-grouped 5-fold cross-validation — races,
not runner-rows, are assigned to folds, so no race is split across an
analysis/assessment pair — with early stopping determining the number of
boosting rounds from each fold's own held-out score (@sec-appx-software
has the implementation budget behind this grid).

The winning point is the one with the highest fold-mean score; among
points within 0.001 of that maximum, the rule prefers the shallowest
tree, then the fewest rounds. The selected configuration is max_depth
`r sel$max_depth`, eta `r sel$eta`, min_child_weight
`r sel$min_child_weight`, subsample `r sel$subsample`, colsample_bytree
`r sel$colsample_bytree`.

It is refit once on all training data, with no held-out set left to
early-stop against: the number of rounds is the mean, across its five
folds, of each fold's own early-stopped count, rounded —
`r model_3_gbt_meta$nrounds` rounds. The resulting fit's training
pseudo-$R^2$ is `r sprintf("%.5f", model_3_gbt_meta$train_pl_r2)`,
in-sample.

The pre-declared tie-break window was wider than the paired standard
errors observed near the top of the grid, so the rule selected among
points a paired test would not call tied. A refit at the grid maximum,
taken at its own fold-mean rounds, produces materially the same model —
`r n_top10_same_rank` of the top ten features by gain importance hold the
identical rank in both fits.

The selected configuration is `r if (sel_is_boundary) "" else "not "`on a
boundary of the grid (neither `max_depth` nor `eta` is the largest value
tested).

Row subsampling operates at the level of individual horses, not races,
because there is no race-level grouping to subsample by. Under a grouped
likelihood the runners in a race are not independent, so the subsampling
unit and the likelihood's own grouping unit do not match — a limitation
@sec-discussion returns to.
