# Results {#sec-results}

```{r}
#| label: p3-results-setup
#| include: false

roi_pct <- function(x) scales::percent(x, accuracy = 0.1)
pp      <- function(x) sprintf("%+.1f pp", 100 * x)
comma   <- function(x) format(x, big.mark = ",")

ci_at <- function(sweep, tau0 = 1.3) {
  r <- sweep[abs(sweep$tau - tau0) < 1e-9, ]
  sprintf("[%s, %s]", roi_pct(r$ci_lo), roi_pct(r$ci_hi))
}

p1_gbt    <- ranking_metrics_3      |> dplyr::filter(metric == "P1_rank")
p1_2b     <- ranking_metrics_2b     |> dplyr::filter(metric == "P1_rank")
p1_2a     <- ranking_metrics_2a_fresh |> dplyr::filter(metric == "P1_rank")
brp_gbt   <- ranking_metrics_3      |> dplyr::filter(metric == "Brier_place")
brp_2b    <- ranking_metrics_2b     |> dplyr::filter(metric == "Brier_place")
brp_2a    <- ranking_metrics_2a_fresh |> dplyr::filter(metric == "Brier_place")

boot_2b_p1     <- boot_ranking_3_vs_2b |> dplyr::filter(metric == "P1_rank")
boot_2b_brier  <- boot_ranking_3_vs_2b |> dplyr::filter(metric == "Brier_place")
boot_2b_plr2   <- boot_ranking_3_vs_2b |> dplyr::filter(metric == "test_pl_r2")
boot_mkt_p1    <- boot_ranking_3_vs_market |> dplyr::filter(metric == "P1_rank")
boot_mkt_brier <- boot_ranking_3_vs_market |> dplyr::filter(metric == "Brier_place")

course_splits <- gbt_split_counts |> dplyr::filter(grepl("^course_", Feature))

going_within  <- gbt_perm_importance_within_test |> dplyr::filter(feature %in% c("going_sr_shrunk", "going_sr_delta", "going_runs_prior"))
going_ordinal_across <- gbt_perm_importance_across_test |> dplyr::filter(feature == "going_ordinal")
```

The three questions are the same ones papers 2a and 2b ask, in the same
order: how well the model ranks, how well it picks winners, and whether
that translates into betting value. All three are scored on the
`r comma(n_races_p3_test)`-race test universe (@sec-data).

## Q1 — Ranking

```{r}
#| label: tbl-q1-ranking
tibble::tibble(
  Metric  = c("P1_rank (higher is better)", "Brier_place (lower is better)"),
  GBT     = c(signif(p1_gbt$model, 3),  signif(brp_gbt$model, 3)),
  `Paper 2b` = c(signif(p1_2b$model, 3), signif(brp_2b$model, 3)),
  `Paper 2a*` = c(signif(p1_2a$model, 3), signif(brp_2a$model, 3)),
  Market  = c(signif(p1_gbt$market, 3), signif(brp_gbt$market, 3))
) |>
  knitr::kable(caption = sprintf("Ranking metrics on the %s test races: the GBT, paper 2b's exploded model, paper 2a's win model, and the discounted-Harville market. *Paper 2a's figures are computed fresh from its stored test predictions for this comparison and were not part of its published results.", comma(n_races_p3_test)))
```

On the same test races, test-split McFadden pseudo-$R^2$ is
`r sprintf("%.5f", test_pl_r2_3)` for the GBT and
`r sprintf("%.5f", test_pl_r2_2b)` for paper 2b.

A paired race-level bootstrap ($B = 2\,000$, seed 42) of the GBT against
paper 2b gives a 90% interval containing zero on all three metrics:
P1_rank `r sprintf("%+.2e", boot_2b_p1$diff_point)`
[`r sprintf("%.2e", boot_2b_p1$ci_lo)`, `r sprintf("%.2e", boot_2b_p1$ci_hi)`],
Brier_place `r sprintf("%+.4f", boot_2b_brier$diff_point)`
[`r sprintf("%.4f", boot_2b_brier$ci_lo)`, `r sprintf("%.4f", boot_2b_brier$ci_hi)`],
and test pseudo-$R^2$ `r sprintf("%+.4f", boot_2b_plr2$diff_point)`
[`r sprintf("%.4f", boot_2b_plr2$ci_lo)`, `r sprintf("%.4f", boot_2b_plr2$ci_hi)`].
The same bootstrap against the discounted-Harville market excludes zero
on both metrics: P1_rank `r sprintf("%+.2e", boot_mkt_p1$diff_point)`
[`r sprintf("%.2e", boot_mkt_p1$ci_lo)`, `r sprintf("%.2e", boot_mkt_p1$ci_hi)`],
Brier_place `r sprintf("%+.4f", boot_mkt_brier$diff_point)`
[`r sprintf("%.4f", boot_mkt_brier$ci_lo)`, `r sprintf("%.4f", boot_mkt_brier$ci_hi)`].

The tree ensemble and the linear predictor rank equally well on this test
set: neither ranking metric distinguishes them. The market ranks better
than both, on both metrics, and this is the only contrast in the paper
whose interval excludes zero.

That the two models tie is not the same as the two models agreeing
race by race. Within-race-centred scores from the GBT and paper 2b
correlate at $r =$ `r sprintf("%.3f", score_agreement_3_vs_2b$pearson_r)`
(Spearman `r sprintf("%.3f", score_agreement_3_vs_2b$spearman_r)`), and
the two models pick a different top horse in
`r sprintf("%.1f", 100 * (1 - score_agreement_3_vs_2b$share_same_top))`%
of races. They are different models that rank equally well.

## Predictor importance

**Gain importance** is an in-sample accounting of where the fitted loss
went — evidence the model used a feature, not evidence it helps out of
sample.

```{r}
#| label: tbl-gain-importance
#| code-fold: true
gbt_gain_importance_train |>
  dplyr::transmute(Feature, Rank = rank, Gain = round(Gain, 4)) |>
  knitr::kable(caption = "Gain importance, all 24 features, in-sample.")
```

**Permutation importance**, scored on the test split, is the out-of-sample
measure, reported in two variants. Shuffling which horse holds a value
within its own race is the right test for a horse-level feature, but it
is a structural no-op for a feature constant within every race — today's
going, which course the race is at — so those five race-level features
are instead permuted across races.

```{r}
#| label: tbl-perm-within
#| code-fold: true
gbt_perm_importance_within_test |>
  dplyr::transmute(Feature = feature, Rank = rank,
                   `Mean drop` = round(mean_drop, 5), `SD` = round(sd_drop, 5)) |>
  knitr::kable(caption = "Within-race permutation importance, test split, 19 horse-level features, 30 repeats.")
```

```{r}
#| label: tbl-perm-across
#| code-fold: true
gbt_perm_importance_across_test |>
  dplyr::transmute(Feature = feature, Rank = rank,
                   `Mean drop` = round(mean_drop, 6), `SD` = round(sd_drop, 6)) |>
  knitr::kable(caption = "Across-race permutation importance, test split, 5 race-level features, 30 repeats.")
```

The four course indicators are split on between
`r min(course_splits$n_splits)` and `r max(course_splits$n_splits)` times
each in the fitted trees — evidence the model uses them, since within-race
permutation still returns exactly zero for every one.

Permutation also understates a correlated pair of predictors, since
permuting one leaves its partner free to compensate for the lost signal —
`going_sr_shrunk` and `going_sr_delta` correlate at $r =$
`r sprintf("%.2f", cor(gbt_test_data$X[, "going_sr_shrunk"], gbt_test_data$X[, "going_sr_delta"], use = "complete.obs"))`
by construction, so their individual ranks in @tbl-perm-within are a
lower bound on their joint contribution, not a point estimate.

The three horse-level going features sit mid-table, each clearing its
own noise floor: `going_sr_shrunk` ranks
`r going_within$rank[going_within$feature == "going_sr_shrunk"]`,
`going_runs_prior`
`r going_within$rank[going_within$feature == "going_runs_prior"]`, and
`going_sr_delta`
`r going_within$rank[going_within$feature == "going_sr_delta"]` of 19.
`going_ordinal` ranks last of the five race-level features, its mean drop
smaller in magnitude than its own standard deviation — indistinguishable
from noise, as `r sprintf("%.1f", pct_standard_train)`% of training races
sharing one going value predicts. A model carrying these four features
ranks no better, out of sample, than paper 2b's model, which did not have
them.

## Q2 — Win-picking

Owen's naive rule (model win probability above 0.15, model/market ratio
above 1.3), priced at real starting prices.

```{r}
#| label: tbl-q2-naive
tibble::tibble(
  Model      = c("GBT", "Paper 2b", "Paper 2b", "Paper 2a", "Paper 2a"),
  Universe   = c(comma(n_races_p3_test), "published", sprintf("restricted (%s)", comma(n_races_p3_test)),
                 "published", sprintf("restricted (%s)", comma(n_races_p3_test))),
  Bets       = comma(c(backtest_naive_3_win$n_bets, backtest_naive_2b_win$n_bets,
                       backtest_naive_2b_win_restricted_3$n_bets, backtest_naive_w_final$n_bets,
                       backtest_naive_w_final_restricted_3$n_bets)),
  Wins       = comma(c(backtest_naive_3_win$n_wins, backtest_naive_2b_win$n_wins,
                       backtest_naive_2b_win_restricted_3$n_wins, backtest_naive_w_final$n_wins,
                       backtest_naive_w_final_restricted_3$n_wins)),
  ROI        = roi_pct(c(backtest_naive_3_win$roi, backtest_naive_2b_win$roi,
                         backtest_naive_2b_win_restricted_3$roi, backtest_naive_w_final$roi,
                         backtest_naive_w_final_restricted_3$roi))
) |>
  knitr::kable(caption = "Naive win backtest (prob > 0.15, ratio > 1.3), real starting prices. Published columns are each model's own reported universe; restricted recomputes the same stored predictions on the GBT's 2,183-race universe, no refit.")
```

```{r}
#| label: tbl-q2-boot
dplyr::bind_rows(roi_diff_3_vs_2b, roi_diff_3_vs_2a) |>
  dplyr::transmute(
    Contrast     = contrast,
    `ROI diff.`  = pp(diff_point),
    `90% CI`     = paste0("[", pp(ci_lo), ", ", pp(ci_hi), "]"),
    `Common races` = comma(n_common)
  ) |>
  knitr::kable(caption = "Paired race-level ROI-difference bootstrap, GBT against paper 2b and against paper 2a.")
```

Both intervals contain zero. On the restricted, like-for-like universe the
point estimate runs the wrong way for the GBT against both comparators —
`r roi_pct(backtest_naive_3_win$roi)` against paper 2b's
`r roi_pct(backtest_naive_2b_win_restricted_3$roi)` and paper 2a's
`r roi_pct(backtest_naive_w_final_restricted_3$roi)` — and the bootstrap
intervals do not distinguish any of the three.

In the `r disagreement_set_3_vs_2b$n_disagree` of
`r disagreement_set_3_vs_2b$n_total` races where the GBT and paper 2b
pick a different top horse, the GBT's pick wins
`r sprintf("%.1f", 100 * disagreement_set_3_vs_2b$win_rate_a_dis)`% of the
time against paper 2b's
`r sprintf("%.1f", 100 * disagreement_set_3_vs_2b$win_rate_b_dis)`%, at a
mean starting price of
`r sprintf("%.2f", disagreement_set_3_vs_2b$sp_summary$mean_sp[disagreement_set_3_vs_2b$sp_summary$model == "A (disagreement set)"])`
against `r sprintf("%.2f", disagreement_set_3_vs_2b$sp_summary$mean_sp[disagreement_set_3_vs_2b$sp_summary$model == "B (disagreement set)"])`.
A race-level bootstrap of both differences contains zero: the win-rate
difference is `r sprintf("%+.3f", boot_disagreement_3_vs_2b$winrate_diff_point)`
[`r sprintf("%.3f", boot_disagreement_3_vs_2b$winrate_ci[1])`, `r sprintf("%.3f", boot_disagreement_3_vs_2b$winrate_ci[2])`]
and the price difference is
`r sprintf("%+.2f", boot_disagreement_3_vs_2b$sp_diff_point)`
[`r sprintf("%.2f", boot_disagreement_3_vs_2b$sp_ci[1])`, `r sprintf("%.2f", boot_disagreement_3_vs_2b$sp_ci[2])`].

## Q3 — Betting value

Single bet per race — the highest-model-win-probability qualifier under
Owen's naive rule — priced at real starting prices, win, place and
each-way, the same construction as paper 2b.

```{r}
#| label: tbl-q3-single-bet
tibble::tibble(
  `Bet type` = c("Win", "Win", "Win", "Place", "Place", "Each-way", "Each-way"),
  Model      = c("GBT", "Paper 2b", "Paper 2b",
                 "GBT", "Paper 2b",
                 "GBT", "Paper 2b"),
  Universe   = c(comma(n_races_p3_test), "published", sprintf("restricted (%s)", comma(n_races_p3_test)),
                 comma(n_races_p3_test), "published",
                 comma(n_races_p3_test), "published"),
  Bets       = comma(c(backtest_single_win_3$n_bets, backtest_single_win_2b$n_bets, backtest_single_win_2b_restricted_3$n_bets,
                       backtest_single_place_3$n_bets, backtest_single_place_2b$n_bets,
                       backtest_single_eachway_3$n_bets, backtest_single_eachway_2b$n_bets)),
  ROI        = roi_pct(c(backtest_single_win_3$roi, backtest_single_win_2b$roi, backtest_single_win_2b_restricted_3$roi,
                         backtest_single_place_3$roi, backtest_single_place_2b$roi,
                         backtest_single_eachway_3$roi, backtest_single_eachway_2b$roi)),
  `90% CI`   = c(ci_at(sweep_single_win_3), ci_at(backtest_single_win_sweep_2b), "",
                 ci_at(sweep_single_place_3), ci_at(backtest_single_place_sweep_2b),
                 ci_at(sweep_single_eachway_3), ci_at(backtest_single_eachway_sweep_2b))
) |>
  knitr::kable(caption = "Single-bet-per-race backtest, real starting prices. 90% CI from a race-level bootstrap (B = 2000, seed 42) at the naive ratio threshold; blank where no sweep was run on that restricted comparison.")
```

Paper 2a's single-bet win ROI, for reference, is
`r roi_pct(backtest_single_win_2a$roi)` published and
`r roi_pct(backtest_single_win_2a_restricted_3$roi)` on the restricted
universe; paper 2a reports no place or each-way figures.

The paper 3 and paper 2b intervals overlap on all three bet types.

Overall, the change of function class from a linear predictor to a
gradient-boosted tree ensemble does not improve ranking, win-picking, or
betting value against paper 2b, and does not narrow the gap to the
market on any measure tested. @sec-discussion takes up what that means.
