# What changes {#sec-method}

The bet selection and staking rule, and nothing else. The model is paper 5's
rung-3b encoder, held fixed. Paper 6 fits nothing and refits nothing: both
prediction sets are paper 5's own stored targets.

No feature changes, no objective changes, no hyperparameter changes, and the
races are the same races.

## Three things wrong with the incumbent rule

Owen's rule [@owen2019], carried unchanged through papers 1 to 5, backs a
horse when its model win probability exceeds 0.15 and its ratio of model to
market probability exceeds 1.3. Papers 2a onwards back the single highest
model probability among the qualifiers. Three properties of that rule are
worth testing.

**Flat staking discards the size of the disagreement.** A horse the model
makes a 40% chance at a price implying 20% and a horse it makes 16% at a
price implying 15% attract the same one unit. If the model has any
information the first is worth more than the second, and a flat stake throws
that away. Two staking rules are tried against it: one proportional to the
probability difference, one a quarter-Kelly fraction [@kelly1956;
@maclean2011].

**The thresholds are inherited, not derived.** 0.15 and 1.3 come from Owen's
turf dataset. On the search set they sit at the
`r sprintf("%.0f", 100 * p6_positions$s1_position$quantile_of_top_rated[1])`th
and
`r sprintf("%.0f", 100 * p6_positions$s1_position$quantile_of_top_rated[2])`th
percentiles of the distributions they filter (@sec-thresholds). Nothing about
this dataset put them there. Every threshold this paper searches is instead a
**quantile** of the within-split distribution of the quantity filtered on, for
the reason given in @sec-twofits.

**The ratio filter selects for disagreement.** A model-to-market ratio above
1.3 means the model and the settled market disagree, sharply, about a horse.
Paper 4 asked directly what the model adds *given* the price, and found that
against the settled starting price it adds nothing distinguishable — pooled
`b_mod` of 0.068 and 0.039, intervals reaching no higher than 0.204. If that
is right, races where the two disagree are mostly races where the model is
wrong, and a filter that selects them is selecting for the model's own
errors. Three candidate rules invert the filter, keeping the top-rated horse
only when its ratio sits in a **middle** band rather than a high one, and
three more replace the ratio with a price cut.

## The search set {#sec-searchset}

The rules are searched on paper 5's **validation slice**: the
`r p6_comma(n_val_races)` races carved out of the training period at
2010-12-15, scored by the fit trained on the `r p6_comma(n_fit_races)`
fitting races alone. That fit never saw the validation races, so its
probabilities there are out of sample.

The fitting races themselves are not used. The model was fitted on them and
its probabilities there are far sharper than anything it produces out of
sample, which would make a search rank rules by how hard they lean on that
sharpness rather than by how well they bet.

This is a correction to an earlier attempt at this paper, whose search ran on
training-split predictions and did exactly that: it preferred staking rules
that concentrate on the model's most confident calls, and those rules then
staked almost nothing out of sample. That draft is kept, unrendered, at
`papers/06_betting_strategy/SUPERSEDED/` with a note of what went wrong. The
design fact that matters here is the one above — the search set is held out
from the fit that scores it.

## Two fits, and why the thresholds are quantiles {#sec-twofits}

```{r}
#| label: tbl-provenance
#| tbl-cap: "The two fits. Both are paper 5's own stored targets; paper 6 refits neither."
p6_fit_provenance_tbl |>
  dplyr::transmute(
    Fit = fit, Scores = scores,
    `Training races` = p6_comma(n_train_races),
    Config = config, Epochs = epochs,
    `Validation PL loss` = dplyr::if_else(is.na(val_pl_loss), "—",
                                          sprintf("%.6f", val_pl_loss)),
    Target = source
  ) |>
  kable(align = "lllrrrl")
```

The search set is scored by the fitting-partition fit; the test split is
scored by the full-training-split refit, the model paper 5 published. Those
are two fits of one architecture on different amounts of data, and their
score distributions differ.

```{r}
#| label: tbl-splits
#| tbl-cap: "The two splits, over the model's top-rated horse in each race. The distributions are not the same shape, which is what the quantile parameterisation exists to handle."
p6_positions$summary |>
  dplyr::transmute(
    Split = split, Races = p6_comma(n_races),
    `Median P_mod` = sprintf("%.4f", median_p_mod),
    `Median ratio` = sprintf("%.4f", median_ratio),
    `Median SP` = sprintf("%.2f", median_sp),
    `Positive Kelly edge` = p6_fmt_pct(share_positive_kelly_edge, 1)
  ) |>
  kable(align = "lrrrrr")
```

The median top-rated ratio is
`r sprintf("%.3f", p6_pos(p6_positions, "validation")$median_ratio)` on the
validation slice and
`r sprintf("%.3f", p6_pos(p6_positions, "test")$median_ratio)` on test, and
the median starting price of that horse differs too. An absolute cut declared
on one split would therefore select a different fraction of races on the
other, and any difference between two arms would be partly a difference in
how much they bet.

Every candidate threshold in this paper is consequently a **quantile of the
within-split distribution** of the quantity filtered on, taken over the
model's top-rated horse in each race, and recomputed on the test split from
the test split's own distribution. S1 is the exception and keeps its
published absolute form: it is the incumbent being tested, not a candidate
being tuned — which also means S1 does not select the same fraction of races
on both splits, and @sec-thresholds says by how much.

## The ledger

Papers 1 to 5 wired the rule into the backtest: the functions that computed
ROI also decided which horse to back and how much. No other rule could be
priced on the same footing, which is one reason the rule was never examined.

Paper 6 replaces that with one function. Given per-runner model
probabilities, starting prices, finishing positions and field sizes, plus a
selection rule and a staking rule as arguments, it returns one row per bet
with the race, the runner, the stake, the return and the profit. Nothing is
hard-coded.

Settlement is not re-derived. Place and each-way bets settle through paper
2b's own payout functions, called unchanged: the place price is the pure
Harville [@harville1973] top-three probability of the raw starting-price
book, so it carries the real margin, and the each-way place leg pays a
fraction of the starting price on a placed finish. The one addition is a
second each-way payout under corrected place terms (@sec-terms), asserted
equal to paper 2b's when given paper 2b's terms rather than trusted to agree.

The first thing the ledger does is rebuild paper 5's published test backtest.

```{r}
#| label: tbl-gate
#| tbl-cap: "The reproduction gate. Paper 5's encoder rows, rebuilt through the paper-6 ledger under S1/K0 and paper 5's each-way terms."
p6_ledger_gate |>
  dplyr::transmute(
    Bet = dplyr::recode(bet, win = "win", place = "place",
                        eachway = "each-way"),
    `Bets (p6)` = n_bets_p6, `Bets (p5)` = n_bets_p5,
    `Wins (p6)` = n_wins_p6, `Wins (p5)` = n_wins_p5,
    `Profit (p6)` = sprintf("%.2f", profit_p6),
    `Profit (p5)` = sprintf("%.2f", profit_p5),
    `ROI difference` = sprintf("%.1e", d_roi)
  ) |>
  kable(align = "lrrrrrrr")
```

Every column agrees exactly. The gate is a standing script,
`scripts/verify_p6_ledger.R`, and a target on the pipeline graph, so a later
change to the ledger that moved any of these numbers fails rather than
quietly repricing the comparator.

## The candidates

| | Selection |
|---|---|
| S0 | the model's top-rated horse, every race, no filter |
| S1 | Owen's rule: $P_{\text{mod}} > 0.15$ and $P_{\text{mod}}/P_{\text{mkt}} > 1.3$, back the highest $P_{\text{mod}}$ (the incumbent, in its published absolute form) |
| S2a/b/c | the top-rated horse where its ratio falls in a middle band: $[q_{25}, q_{75}]$, $[q_{30}, q_{60}]$, $[q_{40}, q_{70}]$ |
| S3a/b/c | the top-rated horse where its starting price is at or below the $q_{25}$, $q_{50}$, $q_{75}$ cut |
| S4 | the market favourite, every race, no model input |

| | Staking, per leg |
|---|---|
| K0 | flat, one unit |
| K1 | $(P_{\text{mod}} - P_{\text{mkt}})/0.05$, capped at three units |
| K2 | quarter-Kelly at $P_{\text{mod}}$ against the starting price, divided by 0.05, capped at three units |

Both staking constants are round numbers fixed in advance: a five-point
probability edge stakes one unit under K1, and a quarter-Kelly fraction of
five per cent of bankroll stakes one unit under K2. ROI is invariant to that
scale except through the cap, so the constants matter only in where the cap
bites. Both rules read the selected runner's win-side quantities whatever the
bet type being settled, so staking is one rule across the three markets
rather than three.

## Two stages, because they are two questions {#sec-stages}

A single grid over selection and staking confounds them: an arm that changes
both and loses tells you nothing about which change was responsible. So the
declaration runs in two stages, each varying exactly one term.

**Stage A — selection, at a flat stake.** All nine selection rules at K0,
comparator S1/K0.

**Stage B — staking, on the declared selection only.** K0, K1 and K2 on
whatever stage A declared, comparator that selection at K0.

Paper 5's rung-1 lesson — an arm differing from its comparator in two
respects at once cannot test what the contrast exists to test — is enforced
in code rather than by inspection. Before anything is scored,
`p6_assert_one_difference()` checks every pair the paper reads; all
`r nrow(p6_contrast_check)` pass.

## Eligibility, counted in bets placed {#sec-eligibility}

K1 and K2 stake zero where the model's probability does not exceed the price.
That makes them selection rules as well as staking rules, and the effect is
much stronger out of sample than in: the model's top-rated horse carries a
positive Kelly edge in
`r p6_fmt_pct(p6_pos(p6_positions, "validation")$share_positive_kelly_edge, 1)`
of validation races and
`r p6_fmt_pct(p6_pos(p6_positions, "test")$share_positive_kelly_edge, 1)` of
test races. A rule can therefore look strong on a search set while placing
almost no bets on the one that matters.

Two conditions, both applied **before** the argmax rather than after it:

1. projected test bets $\geq$ 300, where projected = validation bets placed
   $\times$ (`r p6_comma(n_test_races)` / `r p6_comma(n_val_races)`);
2. the arm places a positive stake in at least 60% of the races its selection
   rule selects.

A candidate failing either is struck from the grid before the maximum is
taken. @sec-gate reports what it struck.

## The declaration {#sec-declaration}

Within a stage, and with no judgement at any step:

1. Among eligible candidates, take the highest validation ROI at the real
   starting price.
2. Paired race-level bootstrap of its ROI difference against the stage
   comparator on their common races, B = 2000, seed 42.
3. Declare that candidate only if the point difference exceeds one bootstrap
   standard error of the difference. Otherwise declare the comparator.
4. Ties on ROI to four decimal places prefer the lower-numbered selection
   arm, then K0 over K1 over K2.

Both stages, all three bet types, and the validation numbers that produced
them were written to `papers/06_betting_strategy/DECLARED_RULES.md` and
committed before any test target existed. The pipeline re-reads that file
from disk on every build and matches it against the declaration target, so
neither can drift from the other.
