# The test split, scored once {#sec-test}

```{r}
#| label: test-numbers
#| include: false

t_inc_win <- p6_res(p6_test_results, "S1/K0", "win")
t_mkt_win <- p6_res(p6_test_results, "S4/K0", "win")
t_inc_plc <- p6_res(p6_test_results, "S1/K0", "place")
t_mkt_plc <- p6_res(p6_test_results, "S4/K0", "place")
t_inc_ewc <- p6_res(p6_test_results, "S1/K0", "eachway_corrected")
t_mkt_ewc <- p6_res(p6_test_results, "S4/K0", "eachway_corrected")
t_inc_ew5 <- p6_res(p6_test_results, "S1/K0", "eachway")
t_mkt_ew5 <- p6_res(p6_test_results, "S4/K0", "eachway")

k_win <- p6_con(p6_test_contrasts, "win", "incumbent")
k_plc <- p6_con(p6_test_contrasts, "place", "incumbent")
k_ew5 <- p6_con(p6_test_contrasts, "eachway", "incumbent")
k_ewc <- p6_con(p6_test_contrasts, "eachway_corrected", "incumbent")

drop_move <- abs(p6_test_results$roi_drop_top1 - p6_test_results$roi)
```

## The arms

```{r}
#| label: tbl-arms
#| tbl-cap: "The entire test contact. Two distinct rules, because both declaration stages returned the incumbent, so the declared selection, the declared stake and paper 5's rule are all the same arm."
p6_test_arms |>
  dplyr::left_join(dplyr::select(p6_test_selection_sizes, arm,
                                 n_races_selected), by = "arm") |>
  dplyr::transmute(
    Arm = arm, `Races selected` = p6_comma(n_races_selected),
    `Standing for` = roles
  ) |>
  kable(align = "lrl")
```

Nothing else was scored on test.

## Results

```{r}
#| label: tbl-test
#| tbl-cap: "The test split. `ROI, fair book` prices the same bets at starting prices renormalised within race to sum to one; `ROI less top winner` drops the single largest return. Intervals are 90% race-level bootstrap, B = 2000, seed 42."
p6_results_table(p6_test_results) |> kable(align = "llrrrrrrrl")
```

Every arm loses money on every market at real starting prices. The incumbent
returns `r p6_fmt_pct(t_inc_win$roi)` on the win market — unchanged from
paper 5, since it is paper 5's rule on paper 5's predictions — and its
interval, `r p6_fmt_pct_ci(t_inc_win$ci_lo, t_inc_win$ci_hi)`, reaches just
above zero. The market-favourite control returns
`r p6_fmt_pct(t_mkt_win$roi)` on `r p6_comma(t_mkt_win$n_bets)` bets with an
interval that does not.

Dropping the single largest winner moves each ROI by between
`r sprintf("%.1f", 100 * min(drop_move))` and
`r sprintf("%.1f", 100 * max(drop_move))` points, so no row rests on one
result.

## The contrasts

```{r}
#| label: tbl-contrasts
#| tbl-cap: "The four named contrasts, on the races both arms bet. `A` and `B` are each arm's ROI restricted to those races, which is what the difference is a difference of. Degenerate rows are where the two arms are the same rule."
p6_contrast_table(p6_test_contrasts) |> kable(align = "lllrrrllr")
```

**Did selection help?** and **did staking help?** are identically zero: the
declared rule is the incumbent, so both contrasts compare an arm with itself.
That is a result, not a gap in the design — the search was run, it declared
the incumbent, and a search that declares the incumbent has nothing to
contrast against it. The informative statement about those two questions is
in @sec-search, on the validation slice, where the alternatives were actually
scored and lost.

The other two questions are the same question, for the same reason, and are
the paper's central result.

## The control {#sec-control}

S4/K0 backs the market favourite in every test race at a flat stake. It reads
no model output. If a rule containing no model does as well as the model
rule, the model adds nothing to bet selection.

On the `r p6_comma(k_win$n_races)` races the incumbent bets:

- **Win.** `r p6_fmt_pct(k_win$roi_a)` against `r p6_fmt_pct(k_win$roi_b)`, a
  difference of `r p6_fmt_pts(k_win$diff_point)` points,
  `r p6_fmt_pts_ci(k_win$ci_lo, k_win$ci_hi)`. Contains zero. The point
  estimate favours the model-free rule and the interval is far too wide to
  carry it.
- **Place.** `r p6_fmt_pct(k_plc$roi_a)` against
  `r p6_fmt_pct(k_plc$roi_b)`, `r p6_fmt_pts(k_plc$diff_point)` points,
  `r p6_fmt_pts_ci(k_plc$ci_lo, k_plc$ci_hi)`. **Excludes zero**, favouring
  the model.
- **Each-way, paper 5 terms.** `r p6_fmt_pts(k_ew5$diff_point)` points,
  `r p6_fmt_pts_ci(k_ew5$ci_lo, k_ew5$ci_hi)`. Contains zero, essentially
  centred on it.
- **Each-way, corrected terms.** `r p6_fmt_pts(k_ewc$diff_point)` points,
  `r p6_fmt_pts_ci(k_ewc$ci_lo, k_ewc$ci_hi)`. Contains zero, favouring the
  model-free rule.

Note how much the each-way answer moves with the terms. Under paper 5's flat
one-fifth-top-three settlement the two arms are level; under the real
handicap ladder the model-free rule is ahead by
`r sprintf("%.0f", -100 * k_ewc$diff_point)` points. Neither difference is
distinguishable from zero, but the sign flips, and it flips because the
corrected terms take two place-places away in small fields where the
favourite is short and the model's selection is not. That is a settlement
effect, and it is the reason the corrected terms are reported at all.

**The place result is the one interval that excludes zero, and it says less
than it appears to.** The place market here is synthetic: British bookmakers
price a place bet as a fraction of the win odds and the dataset holds no
place dividends, so paper 2b builds one from the Harville top-three
probability of the raw starting-price book (@sec-appx-a). That construction
prices short-priced horses hardest — a favourite's place probability is high,
so its place price is short, and the margin sits on top of a small number.
Backing favourites into it is an expensive strategy for reasons that have
nothing to do with the model. The honest reading is that the incumbent avoids
a bad bet the control cannot avoid, not that the model has found something in
the place market.

## Margin paid, and horses picked {#sec-margin}

A rule can improve its ROI two ways: by picking better horses, or by betting
where the book takes less. The second is not skill, and separating them is
what the fair-book column is for.

```{r}
#| label: tbl-margin
#| tbl-cap: "The margin decomposition. `Margin paid` is the fair-book ROI less the real-SP ROI: the share of stake the over-round takes on the horses that arm actually backs."
p6_margin_table(p6_margin_decomposition) |> kable(align = "llrrrrrr")
```

At a zero-margin book the win-market ordering is unchanged but both arms turn
positive: the incumbent `r p6_fmt_pct(t_inc_win$roi_fair)` and the favourite
rule `r p6_fmt_pct(t_mkt_win$roi_fair)`. A positive fair-book return on
favourites is the favourite–longshot bias, long documented and not a finding
of this paper [@snowberg2010]: favourites win more often than the
renormalised book implies, and longshots less. It is not exploitable here,
because the margin removed to produce that number is the bookmaker's and
there is no counterparty offering it.

The two arms pay almost the same margin —
`r p6_fmt_pct(t_inc_win$roi_fair - t_inc_win$roi)` and
`r p6_fmt_pct(t_mkt_win$roi_fair - t_mkt_win$roi)` on the win market — so the
win-market comparison is not one of them buying a cheaper book. Neither is
the place comparison: `r p6_fmt_pct(t_inc_plc$roi_fair - t_inc_plc$roi)`
against `r p6_fmt_pct(t_mkt_plc$roi_fair - t_mkt_plc$roi)`. Whatever
separates these two rules, it is not the margin they pay. In a paper where
one of the alternatives on offer was a price cut — S3a/b/c, which bet shorter
prices and would have paid a thinner book — that is worth stating: **no arm
that reached the test split improved its ROI by reducing the margin it paid.**

## Figures

```{r}
#| label: fig-cumulative
#| fig-cap: "Cumulative profit in stake units on the test split, in race-date order. The arms stake different totals, so the levels are not comparable; the shape is the point."
#| fig-height: 7
p6_fig_cumulative
```

Both arms lose steadily. Neither result rests on a run of outcomes in one
part of the period.

```{r}
#| label: fig-val-vs-test
#| fig-cap: "The win market. The nine stage-A selection rules on the validation slice as a rug, against the two arms that reached the test split with their 90% intervals. Both axes are out of sample."
#| fig-height: 3.2
p6_fig_val_vs_test
```

@fig-val-vs-test is the honest summary of the exercise: a spread of
out-of-sample validation returns about twenty points wide, a declaration that
landed on the incumbent, and two test intervals wide enough to overlap
almost everything in the rug.
