# The exploded conditional logit and the course×draw search {#sec-exploded-model}

The win model of paper 2a uses only one bit of each race's result: who
won. The finishing order carries more. The **exploded conditional logit**
[@henery1981] uses it by treating the observed order as a sequence of
conditional win problems — the winner is chosen from the full field, the
runner-up from the field minus the winner, the third from the field minus
the first two — and pooling those sub-races into a single conditional-logit
fit, which is exactly maximising the Plackett–Luce likelihood. We truncate
at depth $k = 3$: only the top-3 order is modelled, the tail of the field
being largely noise [@henery1981]. The algebra and the
Plackett–Luce equivalence are derived in @sec-appx-derivations.

The specification is the same 17 terms that survived paper 2a's feature
reduction — the same zero-plus-slope position encoding, the same imputed
`or_relative`/`or_missing` — so any difference at this stage is the
explosion, not the features. The draw×course block is then searched
**independently** of paper 2a (below); we do not assume 2a's surviving
courses carry over.

```{r}
#| label: p2b-model-setup
#| include: false
# Fits loaded from targets (not fitted here) -> load mlogit's namespace so
# summary.mlogit / coef.mlogit dispatch (CLAUDE.md note).
loadNamespace("mlogit")

win_ct  <- summary(model_p2_reduced)$CoefTable          # 2a reduced win model
base_ct <- summary(model_2b_exploded_base)$CoefTable    # exploded, 17-term, no draw
terms   <- rownames(win_ct)

win_coef  <- stats::setNames(win_ct[, "Estimate"], terms)
base_coef <- stats::setNames(base_ct[, "Estimate"], terms)
win_z     <- stats::setNames(win_ct[, "z-value"],  terms)
base_z    <- stats::setNames(base_ct[, "z-value"],  terms)

# Exploded interaction data is a dfidx; access columns as plain vectors
# (its index column breaks dplyr verbs — CLAUDE.md mlogit notes).
exp_races <- dplyr::n_distinct(exploded_interactions_data$race_id)
exp_sets  <- dplyr::n_distinct(exploded_interactions_data$exploded_race_id)
exp_rows  <- nrow(exploded_interactions_data)

# Final draw block coefficients (3-course) and the 4-course block.
full_ct  <- summary(model_2b_exploded_draw_full)$CoefTable
final_ct <- summary(model_2b_exploded_draw_final)$CoefTable
wolf_2a  <- summary(model_w_ed)$CoefTable["stall_x_wolverhampton", ]  # 2a win 4-course block
```

## From win to ranking: the exploded construction

For each race we form $k = 3$ sub-races: choice set $s$ is the horse that
finished $s$th together with every horse not yet ranked — all finishers
placed $s$-or-worse plus any non-finishers, which never get ranked and so
stay in the pool at every depth. The depth-1 sub-race is therefore exactly
the win model's choice set (the full field, winner = 1st), which makes the
two models comparable at depth 1; each sub-race has exactly one winner.

Two race filters apply. Races with any missing value in the modelling
columns are dropped (the same rule as the win model), and races whose top-3
positions are not each recorded exactly once — a dead-heat for 2nd or 3rd,
or a gap — cannot be unambiguously exploded and are dropped too. Finishing
position is the official placing, taking the amended placing where a result
was later revised.

```{r}
#| label: tbl-exploded-size
depth_rows <- as.integer(table(exploded_interactions_data$depth))
tibble::tibble(
  `Sub-race`    = paste0("depth ", seq_along(depth_rows), " (",
                         c("full field", "field − 1st", "field − 1st,2nd"), ")"),
  `Runner-rows` = format(depth_rows, big.mark = ",")
) |>
  knitr::kable(caption = sprintf(
    "Exploded training data (carrying the draw-course columns): %s valid races become %s choice sets (3 per race) over %s runner-rows.",
    format(exp_races, big.mark = ","),
    format(exp_sets,  big.mark = ","),
    format(exp_rows,  big.mark = ",")))
```

The runner-row count falls by one per race at each successive depth: the
same races contribute to all three sub-races, but each deeper sub-race drops
the horses already placed above it (depth 2 removes the winner, depth 3 the
first two), so the choice set shrinks by one runner each time.

## Fitting the ranking model

We carry paper 2a's reduced 17-feature specification — the
feature-by-feature selection was settled there — and re-estimate it under
the ranking objective, entering the four-course draw×course block (Kempton,
Lingfield, Southwell, Wolverhampton) from the start. The draw block is then
reduced by the same per-term $p < 0.05$ rule paper 2a applies to its
win model, run fresh on this fit: the least significant non-significant
course slope is dropped, the model refit, and the step repeated until every
surviving slope clears the threshold.

Re-estimating under the explosion moves the coefficients in two ways: the
estimates shrink in magnitude, and the z-values mostly rise. Both are shown
against paper 2a's win fit on the identical specification.

```{r}
#| label: tbl-base-coef-compare
#| code-fold: true
block_of <- function(term) {
  if (grepl("^pos_lag", term))                                     return("1. Position (zero-plus-slope)")
  if (term %in% c("days_LTO_log", "trainerSR", "sireSR"))          return("2. Form / history")
  if (term %in% c("age_diff", "entire", "gelding", "cheekpieces")) return("3. Horse characteristics")
  "4. New features (introduced in paper 2a)"
}
pretty_term <- function(t) dplyr::recode(t, days_LTO_log = "daysLTO (log)")
tibble::tibble(
  Block           = vapply(terms, block_of, character(1)),
  Term            = pretty_term(terms),
  `Win est.`      = round(win_coef[terms], 4),
  `Win z`         = round(win_z[terms], 2),
  `Exploded est.` = round(base_coef[terms], 4),
  `Exploded z`    = round(base_z[terms], 2)
) |>
  dplyr::arrange(Block) |>
  knitr::kable(
    align   = c("l", "l", "r", "r", "r", "r"),
    caption = "Paper 2a reduced win model (depth-1) vs the exploded fit on the same 17 terms (depth-3 Plackett–Luce). Estimates and z-values side by side.")
```

The estimates shrink — most coefficients are smaller in magnitude under the
explosion, because the win model loads each feature with all of its power to
separate the *winner* from the field, whereas the exploded model must make
the same coefficient explain the 1-vs-rest, 2-vs-rest and 3-vs-rest
contrasts at once. The z-values mostly move the other way, rising, as
depth-3 explosion roughly triples the number of choice problems the same
races supply; the form and position terms, which order the whole field
rather than just crown the winner, change the most.

```{r}
#| label: tbl-draw-reduction-2b
#| code-fold: true
model_2b_draw_reduction_steps |>
  dplyr::transmute(
    Step      = step,
    `Dropped` = dropped,
    `p`       = signif(wald_p, 3),
    `From → to` = paste0(n_before, "-course → ", n_after, "-course"),
    `LR stat` = round(lr_stat, 3),
    df        = lr_df,
    `LR p`    = signif(lr_p, 3)
  ) |>
  knitr::kable(caption = "Per-term reduction of the exploded draw×course block. Each step drops the least significant non-significant course slope and refits; the 1-df likelihood-ratio drop-test is reported alongside.")
```

Only **Lingfield** drops ($p =$
`r signif(model_2b_draw_reduction_steps$wald_p[1], 2)`; the 1-df LR
drop-test agrees, $p =$ `r signif(model_2b_draw_reduction_steps$lr_p[1], 2)`).
The surviving block is **Kempton, Southwell and Wolverhampton** — three
courses, where paper 2a's win model retained only **Kempton and
Southwell**. The difference is Wolverhampton. On the exploded fit
Wolverhampton's draw slope
is `r sprintf("%+.3f", final_ct["stall_x_wolverhampton", "Estimate"])`
($p =$ `r sprintf("%.4f", final_ct["stall_x_wolverhampton", "Pr(>|z|)"])`); in
paper 2a's win model it was `r sprintf("%+.3f", wolf_2a["Estimate"])`
($p \approx 0.066$) — the **same sign and similar
magnitude, estimated far more sharply**. The reading is the narrow one: the
ranking objective estimates the same draw effect paper 2a saw, sharply
enough to retain it, rather than uncovering a new effect the win model
missed. The negative slope (an advantage to lower draws) is the direction
2a already found and is directionally consistent with a tight, left-handed
All-Weather track; it is a single distance-pooled slope, so it is an average
direction across trips, not a trip-specific law.

The final paper-2b model is the 17-term exploded fit plus the three retained
course-draw slopes —
`r length(stats::coef(model_2b_exploded_draw_final))` coefficients in all.

```{r}
#| label: tbl-draw-course-2b
#| code-fold: true
draw_rows  <- grepl("^stall_x_", rownames(full_ct))
course_lab <- tools::toTitleCase(sub("stall_x_", "", rownames(full_ct)[draw_rows]))
retained   <- rownames(full_ct)[draw_rows] %in% names(stats::coef(model_2b_exploded_draw_final))
tibble::tibble(
  Course       = course_lab,
  Estimate     = round(full_ct[draw_rows, "Estimate"], 4),
  `Std. error` = round(full_ct[draw_rows, "Std. Error"], 4),
  `z`          = round(full_ct[draw_rows, "z-value"], 2),
  `p-value`    = signif(full_ct[draw_rows, "Pr(>|z|)"], 3),
  Retained     = dplyr::if_else(retained, "yes", "dropped")
) |>
  knitr::kable(caption = "Draw×course coefficients in the full four-course block (the selection-stage exploded fit). Kempton, Southwell and Wolverhampton are individually significant and retained; Lingfield is not and is dropped, leaving the three-course final model.")
```
