# Evaluation on ranking metrics {#sec-evaluation}

This paper asks two questions, in order. **Q1** is the ranking question the
paper is built around: how well does the exploded model predict the top-3
finishing order, measured against the market? **Q2** is a byproduct:
because the model is fitted on richer (depth-3) ranking supervision rather
than depth-1 win supervision, how does it pick *winners* compared with paper
2a's win-fitted model? The two are answered with different baselines and
reported separately.

## Ranking metrics

We evaluate ranking on two metrics, both computed out-of-sample on the
held-out test races.

**P1_rank** (ranking discrimination): the geometric mean, over test races,
of the Plackett–Luce probability assigned to the observed top-3 finishing
order. Computed as the exponentiated mean log-likelihood per race to avoid
numerical underflow:
$$
\text{P1}_{\text{rank}}
  = \exp\!\left( \frac{1}{n} \sum_{i} \log P(j_1^i, j_2^i, j_3^i) \right),
$$
where $P(j_1, j_2, j_3)$ is the depth-3 Plackett–Luce probability of the
observed order under the model or market. Higher is better. This is the
ranking analogue of Owen's P1.

**Brier_place** (place calibration): the mean squared error between the
predicted place probability and the binary place outcome (top-3 or not),
averaged over all runner-race observations in the test set:
$$
\text{Brier}_{\text{place}}
  = \frac{1}{N} \sum_{i,j} \left( \text{place}_{ij} - p^{\text{place}}_{ij} \right)^2,
$$
where $\text{place}_{ij} = 1$ if horse $j$ finished in the top 3 in race
$i$, and $p^{\text{place}}_{ij}$ is the model- or market-implied
probability of a top-3 finish. Lower is better; a proper scoring rule,
analogous to Owen's P2.

Both market baselines use discounted Harville ($\alpha = 0.80$ for the
second-place conditional, $0.65$ for the third; @sec-data): the place
probabilities for Brier_place and the depth-3 order probability for P1_rank
are both built from the same discounted starting-price win probabilities. The
model side of each metric uses its own ($\alpha = 1$) Plackett–Luce
implication.

```{r}
#| label: p2b-eval-setup
#| include: false
roi_pct <- function(x) scales::percent(x, accuracy = 0.1)
pp      <- function(x) sprintf("%+.1f pp", 100 * x)   # ROI difference in points

p1  <- ranking_metrics_2b |> dplyr::filter(metric == "P1_rank")
brp <- ranking_metrics_2b |> dplyr::filter(metric == "Brier_place")

q1_races <- dplyr::n_distinct(ranking_eval_runners_2b$race_id)
q2_races <- dplyr::n_distinct(model_market_ratio_2b_win$race_id)
rd <- roi_difference_2b_vs_2a
```

## Q1 — Ranking performance against the market

The model is scored on `r format(q1_races, big.mark = ",")` test
races[^nfilter] against the discounted-Harville market baseline.

[^nfilter]: Q1 scores `r format(q1_races, big.mark = ",")` races and Q2
below scores `r format(q2_races, big.mark = ",")`; the difference is that
the ranking metrics require a clean, unambiguous top-3 (positions 1–3 each
recorded exactly once) and a complete market win-probability vector, a
filter the depth-1 win backtest in Q2 does not need.

```{r}
#| label: tbl-ranking-metrics
#| code-fold: true
ranking_metrics_2b |>
  dplyr::transmute(
    Metric    = metric,
    Model     = signif(model, 4),
    Market    = signif(market, 4),
    Better    = c("higher", "lower"),
    Winner    = dplyr::if_else(
      (metric == "P1_rank"     & model > market) |
      (metric == "Brier_place" & model < market), "model", "market")
  ) |>
  knitr::kable(caption = sprintf(
    "Ranking metrics on the %s scorable test races: exploded model vs discounted-Harville market baseline.",
    format(q1_races, big.mark = ",")))
```

The model predicts the top-3 order with clear skill above chance. A
random-guessing baseline scores about 0.003 on P1_rank — roughly the
reciprocal of the number of possible top-3 orderings in a typical field —
and about 0.23 on Brier_place, the score from predicting every horse's place
chance at the base rate. The model scores `r signif(p1$model, 3)` on P1_rank
(higher is better) and `r signif(brp$model, 3)` on Brier_place (lower is
better), well inside the skilled range on both. Measured against the
discounted-Harville market baseline, the market scores higher on each —
`r signif(p1$market, 3)` on P1_rank and `r signif(brp$market, 3)` on
Brier_place. The model has genuine ranking skill; the market, here as in
papers 1 and 2a, sets the higher mark, consistent with the starting price
aggregating information the pre-race feature set does not contain.

## Q2 — Win performance as a byproduct

The same fully-specified three-course exploded model can be judged on the
*win* objective, by reading off its depth-1 win probabilities and running
paper 2a's exact betting backtest. The question is not whether it beats the
market — neither model does — but whether ranking supervision makes it a
better or worse winner-picker than paper 2a's win-fitted model.

```{r}
#| label: tbl-backtest-2b-win
backtest_naive_2b_win |>
  dplyr::transmute(
    `Prob threshold`  = prob_threshold,
    `Ratio threshold` = ratio_threshold,
    `Bets`            = n_bets,
    `Wins`            = n_wins,
    `Gross return`    = sprintf("%.2f", gross_return),
    `Profit`          = sprintf("%.2f", profit),
    `ROI`             = roi_pct(roi)
  ) |>
  knitr::kable(caption = "Exploded (ranking-fitted) model on the depth-1 win backtest, AW test races at Owen's exact thresholds.")
```

The ranking-fitted model returns a naive win ROI of
**`r roi_pct(backtest_naive_2b_win$roi)`**, against paper 2a's win-fitted
headline of **`r roi_pct(backtest_naive_w_final$roi)`** — both struck on the
identical `r format(q2_races, big.mark = ",")`-race test set. Because the
two models are evaluated on the same races we can pair the resampling: a
race-level ROI-difference bootstrap (B = 2000, seed 42) restricted to the
common races gives a point difference of
**`r pp(rd$diff_point)`** in the ranking model's favour, with a 90%
interval of **[`r pp(rd$ci_lo)`, `r pp(rd$ci_hi)`]**.

```{r}
#| label: tbl-roi-diff-2b
#| code-fold: true
roi_difference_2b_vs_2a |>
  dplyr::transmute(
    Contrast       = contrast,
    `ROI diff.`    = pp(diff_point),
    `90% CI`       = paste0("[", pp(ci_lo), ", ", pp(ci_hi), "]"),
    `Common races` = format(n_common, big.mark = ",")
  ) |>
  knitr::kable(caption = "Paired race-level ROI-difference bootstrap (2b ranking-fitted minus 2a win-fitted), restricted to the common test races.")
```

The 90% interval lies above zero, so on these test races the ranking-fitted
model's win ROI is distinguishable from the win-fitted model's. The reading
is that depth-3 ranking supervision yields a somewhat better winner-picker
than depth-1 win supervision. Two things keep it in proportion: the lower
bound of the interval (`r pp(rd$ci_lo)`) sits close to zero, so this is a
distinguishable difference rather than a wide one; and both models return a
negative ROI (`r roi_pct(backtest_naive_2b_win$roi)` here), so the gain is
relative to paper 2a, not a profit against the market. The ratio-threshold
sweep below shows the win ROI across selection thresholds.

```{r}
#| label: fig-roi-sweep-2b-win
#| fig-cap: "Depth-1 win ROI vs ratio threshold on the AW test set (ranking-fitted exploded model), 90% bootstrap CI shaded. Model-probability filter held at 0.13; dashed line at zero marks break-even."
ggplot2::ggplot(backtest_sweep_2b_win, ggplot2::aes(x = tau, y = roi)) +
  ggplot2::geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  ggplot2::geom_ribbon(ggplot2::aes(ymin = ci_lo, ymax = ci_hi), fill = "darkgreen", alpha = 0.18) +
  ggplot2::geom_line(linewidth = 0.8, colour = "darkgreen") +
  ggplot2::scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  ggplot2::labs(x = "Ratio threshold (model_prob / market_prob)", y = "ROI") +
  ggplot2::theme_minimal()
```
