# Evaluation on ranking metrics {#sec-evaluation}

This paper asks three questions, in order. **Q1** is the ranking question the
paper is built around: how well does the exploded model predict the top-3
finishing order, measured against the market? **Q2** is a byproduct:
because the model is fitted on richer (depth-3) ranking supervision rather
than depth-1 win supervision, how does it pick *winners* compared with paper
2a's win-fitted model? **Q3** is the applied question: turned into bets, does
that place-probability accuracy translate into betting value? The three are
answered separately, each against its own comparison.

## Ranking metrics

We evaluate ranking on two metrics, both computed out-of-sample on the
held-out test races.

**P1_rank** (ranking discrimination): the geometric mean, over test races,
of the Plackett–Luce probability assigned to the observed top-3 finishing
order. Computed as the exponentiated mean log-likelihood per race to avoid
numerical underflow:
$$
\text{P1}_{\text{rank}}
  = \exp\!\left( \frac{1}{n} \sum_{i} \log P(j_1^i, j_2^i, j_3^i) \right),
$$
where $P(j_1, j_2, j_3)$ is the depth-3 Plackett–Luce probability of the
observed order under the model or market. Higher is better. This is the
ranking analogue of Owen's P1. Working in log-space — summing
log-probabilities rather than multiplying probabilities directly — avoids
floating-point underflow when the product of many small probabilities would
otherwise round to zero; the geometric mean is recovered by exponentiating at
the end.

**Brier_place** (place calibration): the mean squared error between the
predicted place probability and the binary place outcome (top-3 or not),
averaged over all runner-race observations in the test set:
$$
\text{Brier}_{\text{place}}
  = \frac{1}{N} \sum_{i,j} \left( \text{place}_{ij} - p^{\text{place}}_{ij} \right)^2,
$$
where $\text{place}_{ij} = 1$ if horse $j$ finished in the top 3 in race
$i$, and $p^{\text{place}}_{ij}$ is the model- or market-implied
probability of a top-3 finish. Lower is better; a proper scoring rule,
analogous to Owen's P2.

The market's probabilities for both metrics use discounted Harville
($\alpha = 0.80$ for the second-place conditional, $0.65$ for the third;
@sec-data): the place probabilities for Brier_place and the depth-3 order
probability for P1_rank are both built from the same discounted starting-price
win probabilities. For each metric, the model's fitted win probabilities are
substituted into the same Plackett–Luce formula (with $\alpha = 1$, no
discounting) to produce its predicted ordering and place probabilities.

```{r}
#| label: p2b-eval-setup
#| include: false
roi_pct <- function(x) scales::percent(x, accuracy = 0.1)
pp      <- function(x) sprintf("%+.1f pp", 100 * x)   # ROI difference in points

p1  <- ranking_metrics_2b |> dplyr::filter(metric == "P1_rank")
brp <- ranking_metrics_2b |> dplyr::filter(metric == "Brier_place")

q1_races <- dplyr::n_distinct(ranking_eval_runners_2b$race_id)
q2_races <- dplyr::n_distinct(model_market_ratio_2b_win$race_id)
rd <- roi_difference_2b_vs_2a
```

## Q1 — Ranking performance against the market

The model is scored on `r format(q1_races, big.mark = ",")` test
races[^nfilter] against the market's discounted-Harville ordering
probabilities.

[^nfilter]: Q1 scores `r format(q1_races, big.mark = ",")` races and Q2
below scores `r format(q2_races, big.mark = ",")`; the difference is that
the ranking metrics require a clean, unambiguous top-3 (positions 1–3 each
recorded exactly once) and a complete market win-probability vector, a
filter the depth-1 win backtest in Q2 does not need.

```{r}
#| label: tbl-ranking-metrics
#| code-fold: true
ranking_metrics_2b |>
  dplyr::transmute(
    Metric    = metric,
    Model     = signif(model, 4),
    Market    = signif(market, 4),
    Better    = c("higher", "lower"),
    Winner    = dplyr::if_else(
      (metric == "P1_rank"     & model > market) |
      (metric == "Brier_place" & model < market), "model", "market")
  ) |>
  knitr::kable(caption = sprintf(
    "Ranking metrics on the %s scorable test races: exploded model vs the market's discounted-Harville probabilities.",
    format(q1_races, big.mark = ",")))
```

The model predicts the top-3 order with clear skill above chance. Random
guessing scores about 0.003 on P1_rank — roughly the
reciprocal of the number of possible top-3 orderings in a typical field —
and about 0.23 on Brier_place, the score from predicting every horse's place
chance at the base rate. The model scores `r signif(p1$model, 3)` on P1_rank
(higher is better) and `r signif(brp$model, 3)` on Brier_place (lower is
better), well inside the skilled range on both. Measured against the market's
discounted-Harville probabilities, the market scores higher on each —
`r signif(p1$market, 3)` on P1_rank and `r signif(brp$market, 3)` on
Brier_place. The model has genuine ranking skill; the market, here as in
papers 1 and 2a, sets the higher mark, consistent with the starting price
aggregating information the pre-race feature set does not contain.

## Q2 — Win performance as a byproduct

The same fully-specified three-course exploded model can be judged on the
*win* objective, by reading off its depth-1 win probabilities and running
paper 2a's exact betting backtest. The question is not whether it beats the
market — neither model does — but whether ranking supervision makes it a
better or worse winner-picker than paper 2a's win-fitted model.

```{r}
#| label: tbl-backtest-2b-win
backtest_naive_2b_win |>
  dplyr::transmute(
    `Prob threshold`  = prob_threshold,
    `Ratio threshold` = ratio_threshold,
    `Bets`            = n_bets,
    `Wins`            = n_wins,
    `Gross return`    = sprintf("%.2f", gross_return),
    `Profit`          = sprintf("%.2f", profit),
    `ROI`             = roi_pct(roi)
  ) |>
  knitr::kable(caption = "Exploded (ranking-fitted) model on the depth-1 win backtest, AW test races at Owen's exact thresholds.")
```

The ranking-fitted model returns a naive win ROI of
**`r roi_pct(backtest_naive_2b_win$roi)`**, against paper 2a's win-fitted
headline of **`r roi_pct(backtest_naive_w_final$roi)`** — both struck on the
identical `r format(q2_races, big.mark = ",")`-race test set. Since both
models are evaluated on the same test races, we compare them race by race —
computing the difference in per-race profit between the two models — and
bootstrap the mean of those differences (B = 2000, seed 42) to get a
confidence interval on the average gain: a point difference of
**`r pp(rd$diff_point)`** in the ranking model's favour, with a 90%
interval of **[`r pp(rd$ci_lo)`, `r pp(rd$ci_hi)`]**.

```{r}
#| label: tbl-roi-diff-2b
#| code-fold: true
roi_difference_2b_vs_2a |>
  dplyr::transmute(
    Contrast       = contrast,
    `ROI diff.`    = pp(diff_point),
    `90% CI`       = paste0("[", pp(ci_lo), ", ", pp(ci_hi), "]"),
    `Common races` = format(n_common, big.mark = ",")
  ) |>
  knitr::kable(caption = "Paired race-level ROI-difference bootstrap (2b ranking-fitted minus 2a win-fitted), restricted to the common test races.")
```

The 90% interval lies above zero, so on these test races the ranking-fitted
model's win ROI is distinguishable from the win-fitted model's. The reading
is that depth-3 ranking supervision yields a somewhat better winner-picker
than depth-1 win supervision. Two things keep it in proportion: the lower
bound of the interval (`r pp(rd$ci_lo)`) sits close to zero, so this is a
distinguishable difference rather than a wide one; and both models return a
negative ROI (`r roi_pct(backtest_naive_2b_win$roi)` here), so the gain is
relative to paper 2a, not a profit against the market. A threshold sweep
(varying $\tau$ from 0.9 to 2.0, as in papers 1 and 2a) shows the win ROI
falling further below zero as selection tightens. Restricting to one bet per
race changes nothing here: single-bet win returns
`r roi_pct(backtest_single_win_2b$roi)` (2b) against
`r roi_pct(backtest_single_win_2a$roi)` for paper 2a's win-fitted model — no
distinguishable improvement over backing all qualifiers, since the single-bet
effect is confined to the place and each-way markets (Q3).

## Q3 — Does place-probability accuracy translate into betting value?

Q1 and Q2 read the model against the market on scoring rules. The applied
question is whether that accuracy translates into betting value: priced and
selected as bets, do the model's probabilities return more than the market's?
We ask it of two bet types — a horse to finish in the top three (**place**)
and the same horse staked win-and-place at standard each-way terms
(**each-way**) — each priced and selected entirely from win starting prices,
with no separate place or each-way market prices used or needed.

```{r}
#| label: p2b-q3-setup
#| include: false
vb    <- value_bet_baselines_2b
vget  <- function(type, col) vb[[col]][vb$bet_type == type]
comma <- function(x) format(x, big.mark = ",")
orpc  <- function(x) scales::percent(x - 1, accuracy = 0.1)   # book sum -> over-round %
# 90% bootstrap-CI string for a bet's ROI at the naive threshold, from its sweep
ci_at <- function(sweep, tau0 = 1.3) {
  r <- sweep[abs(sweep$tau - tau0) < 1e-9, ]
  sprintf("[%s, %s]", roi_pct(r$ci_lo), roi_pct(r$ci_hi))
}
# single-bet each-way CI upper bound at tau = 1.3 (grazes zero)
ew_hi <- roi_pct(
  dplyr::filter(backtest_single_eachway_sweep_2b, abs(tau - 1.3) < 1e-9)$ci_hi
)
```

Owen's naive rule backs every horse clearing both thresholds (win probability
above 0.15 and model/market ratio above 1.3), which can mean several bets in
one race. A more realistic strategy places one bet per race: among the
qualifiers, back the single horse with the highest model win probability. We
take this **single-bet** rule as primary and report the **multi-bet** rule
(back all qualifiers) alongside for continuity with papers 1 and 2a — Owen
does not specify which he used. Both are priced at the real industry-SP book
(observed average over-round `r orpc(vget("place", "over_round"))`), the same
basis as the Q2 win backtest: the place price is derived from the raw
SP-implied win probabilities via Harville, and the each-way win leg pays the
raw SP win odds with the place leg at one fifth of those odds on a top-3
finish (standard terms). Both are swept over the ratio threshold $\tau$.

```{r}
#| label: tbl-value-bets
#| code-fold: true
tibble::tibble(
  `Bet type`       = c("Place", "Each-way"),
  `Bets`           = comma(c(backtest_single_place_2b$n_bets, backtest_single_eachway_2b$n_bets)),
  `Single-bet ROI` = roi_pct(c(backtest_single_place_2b$roi, backtest_single_eachway_2b$roi)),
  `90% CI`         = c(ci_at(backtest_single_place_sweep_2b), ci_at(backtest_single_eachway_sweep_2b)),
  `Multi-bet ROI`  = roi_pct(c(backtest_place_2b$roi, backtest_eachway_2b$roi))
) |>
  knitr::kable(
    align   = c("l", "r", "r", "r", "r"),
    caption = sprintf("Place and each-way betting at Owen's naive thresholds (ratio > 1.3), priced on the real industry-SP book (observed average over-round %s). Single-bet (one bet per race, the highest-model-win-probability qualifier) is the primary rule; multi-bet (all qualifiers) is shown for continuity with papers 1/2a. 90%% CI from a race-level bootstrap (B = 2000, seed 42).", orpc(vget("place", "over_round"))))
```

**Single bet per race.** With one bet per race, place returns
`r roi_pct(backtest_single_place_2b$roi)` on
`r comma(backtest_single_place_2b$n_bets)` bets and each-way
`r roi_pct(backtest_single_eachway_2b$roi)` on
`r comma(backtest_single_eachway_2b$n_bets)` bets — far smaller losses than the
multi-bet equivalents, and for place roughly half the ~16% structural SP
margin (against a zero-margin fair book, place selections return
`r roi_pct(vget("place", "model_fair"))`). @fig-single-vs-multi-sweep shows
these hold across the threshold range: the single-bet curve sits consistently
above the multi-bet curve for both bet types, the gap widening as $\tau$
tightens. The place result is robust to the tail — dropping the single-bet
strategy's largest winning place bet leaves
`r roi_pct(backtest_single_place_2b$roi_drop_top1)` (the single-bet recompute,
not the multi-bet figure). The each-way result is the more marginal: at
$\tau = 1.3$ its 90% interval runs to `r ew_hi` at the upper end, grazing zero,
so while it clearly improves on multi-bet we do not read it as comfortably
below zero.

```{r}
#| label: fig-single-vs-multi-sweep
#| fig-cap: "Single-bet (coloured) vs multi-bet (grey) ROI across the ratio threshold, place and each-way, 90% race-level bootstrap CI shaded; dashed line at break-even. Single-bet sits consistently above multi-bet, the gap widening as selection tightens. The win market (not shown) shows no distinguishable difference between the two."
dplyr::bind_rows(
  backtest_single_place_sweep_2b   |> dplyr::transmute(tau, roi, ci_lo, ci_hi, bet = "Place",    Strategy = "single-bet"),
  backtest_single_eachway_sweep_2b |> dplyr::transmute(tau, roi, ci_lo, ci_hi, bet = "Each-way", Strategy = "single-bet"),
  backtest_sweep_place_2b          |> dplyr::transmute(tau, roi, ci_lo, ci_hi, bet = "Place",    Strategy = "multi-bet"),
  backtest_sweep_eachway_2b        |> dplyr::transmute(tau, roi, ci_lo, ci_hi, bet = "Each-way", Strategy = "multi-bet")
) |>
  dplyr::mutate(bet = factor(bet, levels = c("Place", "Each-way"))) |>
  ggplot2::ggplot(ggplot2::aes(x = tau, y = roi, colour = Strategy, fill = Strategy)) +
  ggplot2::geom_hline(yintercept = 0, linetype = "dashed", colour = "grey60") +
  ggplot2::geom_ribbon(ggplot2::aes(ymin = ci_lo, ymax = ci_hi), alpha = 0.15, colour = NA) +
  ggplot2::geom_line(linewidth = 0.8) +
  ggplot2::facet_wrap(~ bet) +
  ggplot2::scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  ggplot2::scale_colour_manual(values = c("single-bet" = "firebrick", "multi-bet" = "grey45")) +
  ggplot2::scale_fill_manual(values   = c("single-bet" = "firebrick", "multi-bet" = "grey45")) +
  ggplot2::labs(x = "Ratio threshold (discounted model_prob / market_prob)", y = "ROI") +
  ggplot2::theme_minimal()
```

**Backing all qualifiers.** For continuity with papers 1 and 2a, the multi-bet
rule — backing every qualifier — returns `r roi_pct(backtest_place_2b$roi)` on
place and `r roi_pct(backtest_eachway_2b$roi)` on each-way, both worse than
single-bet because the additional qualifiers are lower-conviction picks.
Each-way is worse still. Backing every horse each-way at these terms loses
`r roi_pct(backtest_betall_eachway$roi)` before any selection — the 1/5 place
terms pay less than the fair place price in large fields — and the model's
weaker place selections, exposed twice (once in choosing which horses, once in
the place leg's payout), push multi-bet each-way below that floor.
