# What was searched {#sec-searched}

A betting rule on this data has three independent parts, and the first two
rounds of this work searched only two of them.

**Which horse.** Given a race worth betting, a *picker* chooses the runner.
Five are used throughout, and only the fifth reads anything about the filter:

```{r}
#| label: tbl-pickers
p6_tbl(tibble::tibble(
  Picker = c("P1", "P2", "P3", "P4", "P5"),
  Definition = c(
    "Highest model win probability",
    "Market favourite: shortest starting price",
    "Highest model probability $\\times$ starting price (expected return)",
    "Largest model probability minus market probability",
    "Highest model probability among runners passing the filter's cuts"),
  Reads = c("the field", "the field", "the field", "the field",
            "the filter's cuts")
), caption = "The five pickers. P5 is Owen's own picker; on the no-filter row every runner passes its cuts, so P5 is P1 there by construction.")
```

**Which races: the model's own output.** A *threshold* filter cuts on the
model probability $\hat P$ and on the ratio $\hat P / P^{\text{mkt}}$. Owen's
rule [@owen2019] is $\hat P > 0.15$ and ratio $> 1.3$ with picker P5, and the
grid searched here is $\hat P$ from 0.00 to 0.35 in steps of 0.01 crossed with
the ratio from 0.80 to 2.50 in steps of 0.05, plus a no-filter row that bets
every race — `r p6_fn(nrow(p6_explore_grid))` filters.

**Which races: the race itself.** A *race-characteristic* filter conditions on
what is known the night before the race and is independent of the model's
output: field size, class, course, distance, going. This family had never been
searched.

That omission is the reason this paper exists. Rounds 1 and 2 concluded that
nothing beat the rule paper 1 inherited from @owen2019, and read as a claim
about betting rules in general. It was a claim about one family. Worse, that
family is largely a single dial in disguise: across the
`r p6_fn(p6_price_band_fit$n_rules[p6_price_band_fit$bet == "win"])`
threshold-family rules that place at least 300 validation bets, win ROI
correlates
`r p6_fcor(p6_price_band_fit$cor_roi_price[p6_price_band_fit$bet == "win"])`
with the mean starting price the rule backs, and a quadratic in log mean price
explains
`r p6_fpct(p6_price_band_fit$r_squared[p6_price_band_fit$bet == "win"], 1)`
of the variation between rules (@sec-priceband). The search had been dense in
one dimension and absent everywhere else.

## The race-characteristic family, enumerated

Every contiguous field-size band over 4 to 16 runners; each class 2 to 5
singly and each contiguous pair; each of the four courses singly and each
pair; the distance terciles of the search set
(`r sprintf("%.0f", p6_rc_terciles[1])` and
`r sprintf("%.0f", p6_rc_terciles[2])` yards) and the single cuts at each
boundary; and the going ordinal split at each interior cut point. A
specification is data rather than a closure, so the identical filter is
evaluated against both splits, and the distance boundaries are fixed on the
search set and carried to the test split unchanged — a re-derived boundary
would make the two splits' rules different rules.

A filter is kept only if it admits at least 300 search-set races and at least
150 in each half of the search window, since a rule that cannot be ranked on
both halves cannot pass the stability guard in @sec-found.

```{r}
#| label: tbl-rc-specs
p6_tbl(
  p6_rc_pruned |>
    dplyr::group_by(Family = family) |>
    dplyr::summarise(
      Enumerated = dplyr::n(),
      Kept = sum(kept),
      `Struck for size` = sum(!kept),
      `Largest struck` = max(c(0L, n_races[!kept])),
      .groups = "drop"),
  caption = "The race-characteristic family. A filter is struck if it admits fewer than 300 search-set races or fewer than 150 in either half."
)
```

Two of the five sub-families are effectively empty before any scoring happens,
and both facts are about UK All-Weather racing rather than about the method.
**Going does not vary**:
`r p6_fn(sum(p6_rc_pruned$n_races[p6_rc_pruned$family == "going" & grepl("<= 4", p6_rc_pruned$label)]))`
of the `r p6_fn(p6_universe$n_val_races)` search-set races are reported
"Standard", so every split of the going scale puts almost everything on one
side; the two filters that survive pruning are near-complete sets of the data
and cannot distinguish anything. **Class is too concentrated**: classes 2 and
3 carry
`r p6_fn(sum(p6_rc_pruned$n_races[p6_rc_pruned$label == "class 2-3"]))`
races between them, below the 300-race floor even pooled, so only the class 4
and class 5 ends of the scale are searchable.

Field size is the sub-family with a prior behind it. Paper 4 found the
overnight forecast-price over-round widening steadily with the size of the
field, and the settled starting price on this data does the same thing:

```{r}
#| label: tbl-overround
p6_tbl(
  p6_overround_val |>
    dplyr::filter(n_races >= 20) |>
    dplyr::transmute(
      `Field size` = field_size, Races = n_races,
      `Mean book` = sprintf("%.3f", mean_book),
      `Margin` = p6_fpct(mean_margin_pct / 100, 2)),
  caption = "The starting-price over-round by field size, on the search set. The margin is what a flat backer pays away before any skill enters."
)
```

The margin a backer pays runs from
`r p6_fpct(p6_overround_val$mean_margin_pct[p6_overround_val$field_size == 4] / 100, 1)`
in a four-runner race to
`r p6_fpct(p6_overround_val$mean_margin_pct[p6_overround_val$field_size == 14] / 100, 1)`
in a fourteen-runner race. The prior is therefore real but it points *against*
big fields: whatever a rule gains by betting them, it starts further behind.

## Provenance, and the gates

Nothing here refits anything. The model is paper 5's rung-3b encoder, held
fixed, scored on the search set by its fitting-partition fit and on the test
split by the full-training-split refit paper 5 published.

```{r}
#| label: tbl-provenance
p6_tbl(
  p6_fit_provenance_tbl |>
    dplyr::transmute(Fit = fit, Scores = scores,
                     `Training races` = p6_fn(n_train_races),
                     Config = config, Epochs = epochs,
                     `Validation PL loss` = ifelse(is.na(val_pl_loss), "---",
                                                   sprintf("%.6f", val_pl_loss))),
  caption = "The two fits. They are two fits of one architecture and not the same model, which is why every threshold in this paper is checked on both splits rather than assumed to transfer."
)
```

The search set is `r p6_fn(p6_universe$n_val_races)` races and
`r p6_fn(p6_universe$n_val_rows)` runners; the test split is
`r p6_fn(p6_universe$n_test_races)` races and
`r p6_fn(p6_universe$n_test_rows)` runners. Race-id overlap between them is
`r p6_search_set_check$n_test_overlap`, and overlap with the
`r p6_fn(p6_search_set_check$n_races_fit)` races paper 5's fitting partition
saw is `r p6_search_set_check$n_fit_overlap`; both are asserted, not checked
by eye.

Three gates run before anything in this paper is read. The betting ledger
reproduces paper 5's published test backtest to the printed digit, and — added
for this paper — the search set's own ledger is rebuilt from the source
database table in code that calls none of the paper's own functions, matching
on the bet set as well as on the arithmetic. The sweep used for the searches
is a second, faster implementation of the same selection and settlement
arithmetic, and is asserted equal to the ledger on every rule expressible in
both. The round-3 scoring path is asserted to reproduce round 2's numbers on
all `r p6_fn(p6_round3_gate$rows_compared)` shared rows, with a maximum
absolute ROI difference of
`r sprintf("%.1e", p6_round3_gate$max_abs_roi_diff)`.
