# What was searched {#sec-searched}

A betting rule has three parts: which horse to back, which races to bet on the
model's output, and which races to bet on the race itself. Rounds 1 and 2
searched the first two and found nothing that beat the rule paper 1 took from
@owen2019. That was a claim about two families. This paper adds the third.

```{r}
#| label: tbl-pickers
p6_tbl(tibble::tibble(
  Picker = c("P1", "P2", "P3", "P4", "P5"),
  Definition = c(
    "Highest model win probability",
    "Market favourite: shortest starting price",
    "Highest model probability $\\times$ starting price (expected return)",
    "Largest model probability minus market probability",
    "Highest model probability among runners passing the filter's cuts"),
  Reads = c("the field", "the field", "the field", "the field",
            "the filter's cuts")
), caption = "The five pickers. P5 is Owen's; with no filter every runner passes its cuts, so P5 is then identical to P1.")
```

**Threshold filters** cut on the model probability $\hat P$ and the ratio
$\hat P / P^{\text{mkt}}$. Owen's rule is $\hat P > 0.15$, ratio $> 1.3$,
picker P5. The grid is $\hat P$ from 0.00 to 0.35 by 0.01 crossed with the
ratio from 0.80 to 2.50 by 0.05, plus a no-filter row:
`r p6_fn(nrow(p6_explore_grid))` filters.

**Race-characteristic filters** use conditions known the night before. Every
contiguous field-size band over 4 to 16 runners; each class 2 to 5 singly and
in contiguous pairs; each course singly and in pairs; the search set's
distance terciles (`r sprintf("%.0f", p6_rc_terciles[1])` and
`r sprintf("%.0f", p6_rc_terciles[2])` yards) and the cuts at each boundary;
the going ordinal at each interior cut. A filter is kept only if it admits 300
search-set races and 150 in each half of the window.

```{r}
#| label: tbl-rc-specs
p6_tbl(
  p6_rc_pruned |>
    dplyr::group_by(Family = family) |>
    dplyr::summarise(
      Enumerated = dplyr::n(),
      Kept = sum(kept),
      `Struck for size` = sum(!kept),
      `Largest struck` = max(c(0L, n_races[!kept])),
      .groups = "drop"),
  caption = "The race-characteristic family, before scoring."
)
```

Two sub-families are empty before scoring, and both facts are about the scope
of this series rather than the method. Going does not vary:
`r p6_fn(sum(p6_rc_pruned$n_races[p6_rc_pruned$family == "going" & grepl("<= 4", p6_rc_pruned$label)]))`
of `r p6_fn(p6_universe$n_val_races)` search races are "Standard". Class is
too concentrated: classes 2 and 3 hold
`r p6_fn(sum(p6_rc_pruned$n_races[p6_rc_pruned$label == "class 2-3"]))`
races between them, under the floor even pooled.

Field size is the one sub-family with a prior. Paper 4 found the overnight
forecast-price over-round widening with field size, and the settled starting
price does the same.

```{r}
#| label: tbl-overround
p6_tbl(
  p6_overround_val |>
    dplyr::filter(n_races >= 20) |>
    dplyr::transmute(
      `Field size` = field_size, Races = n_races,
      `Mean book` = sprintf("%.3f", mean_book),
      `Margin` = p6_fpct(mean_margin_pct / 100, 2)),
  caption = "The starting-price over-round by field size, search set. The margin is what a flat backer pays before any skill enters."
)
```

The prior is real, and it argues against big fields: a rule that bets them
starts further behind.

## Provenance and gates

Nothing here refits anything. The model is paper 5's rung-3b encoder, scored
on the search set by its fitting-partition fit and on the test split by the
full-training-split refit paper 5 published. These are two fits of one
architecture, so no threshold is assumed to transfer.

```{r}
#| label: tbl-provenance
p6_tbl(
  p6_fit_provenance_tbl |>
    dplyr::transmute(Fit = fit, Scores = scores,
                     `Training races` = p6_fn(n_train_races),
                     Config = config, Epochs = epochs,
                     `Validation PL loss` = ifelse(is.na(val_pl_loss), "---",
                                                   sprintf("%.6f", val_pl_loss))),
  caption = "The two fits."
)
```

Search set `r p6_fn(p6_universe$n_val_races)` races,
`r p6_fn(p6_universe$n_val_rows)` runners; test split
`r p6_fn(p6_universe$n_test_races)` races,
`r p6_fn(p6_universe$n_test_rows)` runners. Race-id overlap between them,
`r p6_search_set_check$n_test_overlap`; overlap with the
`r p6_fn(p6_search_set_check$n_races_fit)` fitting races,
`r p6_search_set_check$n_fit_overlap`. Both asserted in code.

Three gates run first. The ledger reproduces paper 5's published test backtest
to the printed digit, and the search set's ledger is rebuilt from the source
database in code calling none of this paper's functions, matching on the bet
set as well as the arithmetic. The faster sweep used for searching is asserted
equal to the ledger on every rule expressible in both. The round-3 scoring
path reproduces round 2 on all `r p6_fn(p6_round3_gate$rows_compared)` shared
rows, maximum absolute ROI difference
`r sprintf("%.1e", p6_round3_gate$max_abs_roi_diff)`.
