# Why most of the surface is price band, not picking {#sec-priceband}

The searches in @sec-found produce leaderboards, and a leaderboard invites the
reading that its top row has found something. On this surface most of what
separates the rows is which price band the rule bets in.

## The regression

Across every threshold-family rule placing at least 300 search-set bets, ROI
is regressed on the mean starting price the rule backs.

```{r}
#| label: tbl-priceband
p6_tbl(
  p6_price_band_fit |>
    dplyr::transmute(
      Market = unname(BET_LAB[bet]), Rules = n_rules,
      `cor(ROI, price)` = p6_fcor(cor_roi_price),
      `cor(ROI, log price)` = p6_fcor(cor_roi_logprice),
      `$R^2$, quadratic in log price` = p6_fpct(r_squared, 1),
      `Residual SD` = p6_fpct(resid_sd, 1),
      `Fitted peak at price` = p6_fprice(price_at_fitted_peak),
      `Median rule ROI` = p6_fpct(median_roi, 1, TRUE)),
  caption = "ROI against the mean price the rule backs, across the threshold family on the search set. Two thirds to four fifths of the variation between rules is a price effect."
)
```

On the win market the correlation is
`r p6_fcor(p6_price_band_fit$cor_roi_price[p6_price_band_fit$bet == "win"])`
and a quadratic in log mean price explains
`r p6_fpct(p6_price_band_fit$r_squared[p6_price_band_fit$bet == "win"], 1)`
of the between-rule variation; on corrected each-way terms it explains
`r p6_fpct(p6_price_band_fit$r_squared[p6_price_band_fit$bet == "eachway_corrected"], 1)`.
The surface is, to first order, a map of the favourite–longshot bias read at
different prices, with an interior optimum where the tilt toward longer prices
stops paying for the extra variance and the extra margin.

That is the substance behind two things that would otherwise look like
findings. It is why the threshold family's leaderboards are mostly ordered by
price. And it is why P4 in fields of 11 to 12 topped every search-set
leaderboard in @sec-found: P4 maximises the gap between the model's
probability and the market's, which in a field large enough to hold a long
price selects the longest price the model does not actively dislike. Its mean
backed price is
`r p6_fprice(p6_row(p6_rc_sweep_full, "win", "P4 / field size 11-12")$mean_sp)`
against `r p6_fprice(p6_price_context$mean_sp_top_rated)` for the model's
top-rated horse — and against
`r p6_fprice(p6_price_context$mean_sp_all_11_12)` for the average runner in an
eleven- or twelve-horse field. The second comparison is the informative one:
P4 in a big field backs a horse priced almost exactly like a runner drawn at
random from it, and only
`r p6_fpct(p6_row(p6_rc_sweep_full, "win", "P4 / field size 11-12")$share_favourite, 1)`
of its picks are the favourite. It is less selecting a price band than
declining to select at all, in the races where declining costs most.

## What is left after the price band

Owen's rule is a useful case because it is not the price band alone.

```{r}
#| label: tbl-owen-resid
p6_tbl(
  p6_price_band_fit |>
    dplyr::transmute(
      Market = unname(BET_LAB[bet]),
      `Owen ROI` = p6_fpct(owen_roi, 1, TRUE),
      `Mean price` = p6_fprice(owen_price),
      `Fitted from price alone` = p6_fpct(owen_fitted, 1, TRUE),
      Residual = p6_fpct(owen_resid, 1, TRUE),
      `Residual, in SDs` = sprintf("%+.2f", owen_resid_sd_units),
      `Rules ranked` = n_rules,
      `Owen's rank` = owen_rank),
  caption = "Owen's rule against the price-band fit on the search set. The residual is the part of its return the price band does not explain."
)
```

On the win market Owen's rule returns
`r p6_fpct(p6_price_band_fit$owen_roi[p6_price_band_fit$bet == "win"], 1, TRUE)`
where the price band alone predicts
`r p6_fpct(p6_price_band_fit$owen_fitted[p6_price_band_fit$bet == "win"], 1, TRUE)`,
a residual of
`r p6_fpct(p6_price_band_fit$owen_resid[p6_price_band_fit$bet == "win"], 1, TRUE)`
or
`r sprintf("%.2f", p6_price_band_fit$owen_resid_sd_units[p6_price_band_fit$bet == "win"])`
residual standard deviations. Two standard deviations of a residual is not a
finding on its own, and the next subsection is why.

## The halved window

The rules that top a search-set leaderboard are not stable inside the search
set. Splitting the `r p6_fn(p6_universe$n_val_races)` search races at
2012-01-15 into `r p6_fn(p6_round3_gate$n_races_h1)` and
`r p6_fn(p6_round3_gate$n_races_h2)`, Owen's cell moves from
`r p6_fpct(p6_owen_guard$roi_h1[p6_owen_guard$bet == "win"], 1, TRUE)` on the
first half to
`r p6_fpct(p6_owen_guard$roi_h2[p6_owen_guard$bet == "win"], 1, TRUE)` on the
second, and from rank
`r p6_owen_guard$rank_h1[p6_owen_guard$bet == "win"]` to rank
`r p6_owen_guard$rank_h2[p6_owen_guard$bet == "win"]` of
`r p6_fn(p6_owen_guard$n_eligible[p6_owen_guard$bet == "win"])` eligible
threshold-family rules. An earlier diagnostic sweep of Owen's own two
thresholds, without the picker dimension, put the same cell first of 617 on
the first half and 157th of 579 on the second. On that earlier surface Owen's
cell also sat second of 597 rules placing at least 300 bets — the 99.8th
percentile — which is exactly the shape a peak takes when a two-year window
happens to reward one price band: the 6-to-9 band in that window returned
about ten points more against its implied probability than the same band does
on the test split.

The reading, and it is the paper's central methodological point: **a
search-set leaderboard on 1,505 races mostly ranks price bands, and the
part that is not price band is not stable across halves of the same window.**
The stability guard in @sec-found was built to catch exactly that and it
struck Owen's rule on three of four markets. It did not save the search,
because the rule it passed —
`r p6_fpct(p6_row(p6_round3_ranks$table, "win", "P4 / field size 11-12")$val_roi_h1, 1, TRUE)`
on the first half,
`r p6_fpct(p6_row(p6_round3_ranks$table, "win", "P4 / field size 11-12")$val_roi_h2, 1, TRUE)`
on the second — was stable within the window and still collapsed to
`r p6_fpct(p6_row(p6_round3_ranks$table, "win", "P4 / field size 11-12")$test_roi, 1, TRUE)`
outside it. Within-window stability is not the same property as
across-window transfer, and this data has the first without the second.
