# Why most of the surface is price band {#sec-priceband}

Two thirds to four fifths of the difference in return between rules is
explained by the mean price the rule backs. A leaderboard on this data mostly
ranks price bands.

```{r}
#| label: tbl-priceband
p6_tbl(
  p6_price_band_fit |>
    dplyr::transmute(
      Market = unname(BET_LAB[bet]), Rules = n_rules,
      `cor(ROI, price)` = p6_fcor(cor_roi_price),
      `cor(ROI, log price)` = p6_fcor(cor_roi_logprice),
      `$R^2$, quadratic in log price` = p6_fpct(r_squared, 1),
      `Residual SD` = p6_fpct(resid_sd, 1),
      `Fitted peak at price` = p6_fprice(price_at_fitted_peak),
      `Median rule ROI` = p6_fpct(median_roi, 1, TRUE)),
  caption = "ROI against mean backed price, across the threshold family on the search set."
)
```

The surface is a map of the favourite-longshot bias read at different prices,
with an interior optimum where the tilt to longer prices stops paying for the
extra variance and margin.

That is why P4 in fields of 11 to 12 topped every search-set table. Its mean
backed price is
`r p6_fprice(p6_row(p6_rc_sweep_full, "win", "P4 / field size 11-12")$mean_sp)`,
against `r p6_fprice(p6_price_context$mean_sp_top_rated)` for the model's
top-rated horse and `r p6_fprice(p6_price_context$mean_sp_all_11_12)` for the
average runner in an 11-or-12-horse field. The second comparison is the
informative one. P4 in a big field backs a horse priced like a runner drawn at
random from it, and only
`r p6_fpct(p6_row(p6_rc_sweep_full, "win", "P4 / field size 11-12")$share_favourite, 1)`
of its picks are the favourite. It is not selecting a price band. It is
declining to select, in the races where declining costs most.

```{r}
#| label: tbl-owen-resid
p6_tbl(
  p6_price_band_fit |>
    dplyr::transmute(
      Market = unname(BET_LAB[bet]),
      `Owen ROI` = p6_fpct(owen_roi, 1, TRUE),
      `Mean price` = p6_fprice(owen_price),
      `Fitted from price alone` = p6_fpct(owen_fitted, 1, TRUE),
      Residual = p6_fpct(owen_resid, 1, TRUE),
      `Residual in SDs` = sprintf("%+.2f", owen_resid_sd_units),
      `Rules ranked` = n_rules,
      `Owen's rank` = owen_rank),
  caption = "Owen's rule against the price-band fit. The residual is the part of its return price does not explain."
)
```

Owen's win-market residual is
`r sprintf("%.2f", p6_price_band_fit$owen_resid_sd_units[p6_price_band_fit$bet == "win"])`
residual standard deviations. That is not a finding on its own, because the
residual does not survive halving the window.

Split at 2012-01-15, Owen's cell moves from
`r p6_fpct(p6_owen_guard$roi_h1[p6_owen_guard$bet == "win"], 1, TRUE)` to
`r p6_fpct(p6_owen_guard$roi_h2[p6_owen_guard$bet == "win"], 1, TRUE)`, and
from rank `r p6_owen_guard$rank_h1[p6_owen_guard$bet == "win"]` to rank
`r p6_owen_guard$rank_h2[p6_owen_guard$bet == "win"]` of
`r p6_fn(p6_owen_guard$n_eligible[p6_owen_guard$bet == "win"])`. An earlier
sweep of Owen's two thresholds alone put the same cell first of 617 on the
first half and 157th of 579 on the second, while also placing it second of 597
rules over the full window. That is what a peak looks like when a two-year
window rewards one price band: the 6-to-9 band in that window returned about
ten points more against its implied probability than the same band does on
test.

So a leaderboard on 1,505 races mostly ranks price bands, and the part that is
not price band does not hold across halves of one window.

The guard in @sec-found was built for that, and it struck Owen's rule on three
of four markets. It still did not save the search: the rule it passed returned
`r p6_fpct(p6_row(p6_round3_ranks$table, "win", "P4 / field size 11-12")$val_roi_h1, 1, TRUE)`
and
`r p6_fpct(p6_row(p6_round3_ranks$table, "win", "P4 / field size 11-12")$val_roi_h2, 1, TRUE)`
on the two halves and
`r p6_fpct(p6_row(p6_round3_ranks$table, "win", "P4 / field size 11-12")$test_roi, 1, TRUE)`
outside the window. Within-window stability is a different property from
across-window transfer. This data has the first and not the second.
