# Staking {#sec-staking}

Every search in this paper runs at a flat one-unit stake, and staking is
varied separately on a fixed selection. That separation is not tidiness. In
the first attempt at this paper the non-flat staking rules were searched
alongside the selection rules, and because both of them stake zero wherever
the model's probability does not exceed the price, they changed the *bet set*
as well as the stake. The search then preferred them for a reason that had
nothing to do with staking.

Three rules: flat at one unit; edge-proportional, $(\hat P - P^{\text{mkt}})$
divided by a five-point edge and capped at three units; and quarter-Kelly at
$\hat P$ against the starting price, on the same cap.

```{r}
#| label: tbl-staking-test
p6_tbl(
  p6_staking_test |>
    dplyr::transmute(
      Market = unname(BET_LAB[bet]), Rule = rule, Staking = staking,
      `Races selected` = n_races_selected, `Races staked` = n_bets,
      `Zero-stake share` = p6_fpct(share_zero_stake, 1),
      `Total staked` = sprintf("%.1f", total_stake),
      ROI = p6_fpct(roi, 1, TRUE),
      `Fair book` = p6_fpct(roi_fair, 1, TRUE)),
  caption = "Staking on the test split, selection held fixed. The zero-stake share is the fraction of the races the selection picked where the staking rule puts nothing on."
)
```

```{r}
#| label: tbl-staking-val
p6_tbl(
  p6_staking_val |>
    dplyr::transmute(
      Market = unname(BET_LAB[bet]), Rule = rule, Staking = staking,
      `Races selected` = n_races_selected, `Races staked` = n_bets,
      `Zero-stake share` = p6_fpct(share_zero_stake, 1),
      `Total staked` = sprintf("%.1f", total_stake),
      ROI = p6_fpct(roi, 1, TRUE),
      `Fair book` = p6_fpct(roi_fair, 1, TRUE)),
  caption = "The same on the search set, for reference."
)
```

```{r}
#| label: staking-counts
#| include: false
nf <- dplyr::bind_rows(
  dplyr::mutate(p6_staking_test, split = "test"),
  dplyr::mutate(p6_staking_val, split = "search set"))
k0 <- nf |> dplyr::filter(staking == "K0") |>
  dplyr::select(split, bet, rule, roi_k0 = roi)
cmp <- nf |> dplyr::filter(staking != "K0", !is.na(roi)) |>
  dplyr::left_join(k0, by = c("split", "bet", "rule")) |>
  dplyr::mutate(worse = roi < roi_k0,
                collapsed = !is.na(share_zero_stake) & share_zero_stake > 0.5)
n_nonflat <- nrow(cmp); n_worse <- sum(cmp$worse)
thin <- cmp |> dplyr::filter(collapsed)
```

**Neither non-flat rule helps.** Of the `r n_nonflat` non-flat arms scored
across the two splits, `r n_worse` return less than their own flat-stake
counterpart. Two mechanisms produce that, and both are visible in the tables.

Where the selection already requires the model's probability to exceed the
price — any rule with a ratio cut above 1 — the zero-stake share is 0.0% and
the staking rules simply lever up an identical bet set, so a negative return
becomes more negative.

Where the selection does not require it, the stake collapses instead.

```{r}
#| label: tbl-staking-thin
p6_tbl(
  thin |>
    dplyr::transmute(Split = split, Market = unname(BET_LAB[bet]),
                     Rule = rule, Staking = staking,
                     `Races selected` = n_races_selected,
                     `Races staked` = n_bets,
                     `Zero-stake share` = p6_fpct(share_zero_stake, 1),
                     `Total staked` = sprintf("%.1f", total_stake),
                     ROI = p6_fpct(roi, 1, TRUE)),
  caption = "The arms where the staking rule discards most of the bet set. An ROI computed on this many units is not a staking result."
)
```

The market-favourite rules are the case: the model usually rates the
favourite *below* the market, so both non-flat rules stake nothing on most of
the races the selection picked. In the worst case the rule stakes
`r p6_fn(min(thin$n_bets))` of `r p6_fn(max(thin$n_races_selected))` selected
races. Reporting only the ROI of such an arm would hide the fact that it is a
different, and much smaller, bet than the one it claims to be.

The conclusion carried forward from round 2 stands unchanged, and the reason
to report the zero-stake share rather than the ROI alone is that in the first
attempt at this paper the declared win rule staked 33 units across 2,183 test
races and the interval around its return was 54 points wide. Nothing about
that was visible in the ROI.
