# Staking {#sec-staking}

```{r}
#| label: staking-counts
#| include: false
nf <- dplyr::bind_rows(
  dplyr::mutate(p6_staking_test, split = "test"),
  dplyr::mutate(p6_staking_val, split = "search set"))
k0 <- nf |> dplyr::filter(staking == "K0") |>
  dplyr::select(split, bet, rule, roi_k0 = roi)
cmp <- nf |> dplyr::filter(staking != "K0", !is.na(roi)) |>
  dplyr::left_join(k0, by = c("split", "bet", "rule")) |>
  dplyr::mutate(worse = roi < roi_k0,
                collapsed = !is.na(share_zero_stake) & share_zero_stake > 0.5)
n_nonflat <- nrow(cmp); n_worse <- sum(cmp$worse)
thin <- cmp |> dplyr::filter(collapsed)
```

Neither non-flat staking rule helps. Of the `r n_nonflat` non-flat arms scored
across the two splits, `r n_worse` return less than their own flat-stake
counterpart.

Every search in this paper runs at a flat one-unit stake, and staking is
varied separately on a fixed selection. In the first attempt at this paper the
non-flat rules were searched alongside the selection rules. Because they stake
zero wherever the model's probability does not exceed the price, they changed
the bet set as well as the stake, and the search preferred them for a reason
that had nothing to do with staking.

Three rules: flat at one unit; edge-proportional,
$(\hat P - P^{\text{mkt}})$ over a five-point edge, capped at three units; and
quarter-Kelly at $\hat P$ against the starting price, on the same cap.

```{r}
#| label: tbl-staking-test
p6_tbl(
  p6_staking_test |>
    dplyr::transmute(
      Market = unname(BET_LAB[bet]), Rule = rule, Staking = staking,
      `Races selected` = n_races_selected, `Races staked` = n_bets,
      `Zero-stake share` = p6_fpct(share_zero_stake, 1),
      `Total staked` = sprintf("%.1f", total_stake),
      ROI = p6_fpct(roi, 1, TRUE),
      `Fair book` = p6_fpct(roi_fair, 1, TRUE)),
  caption = "Staking on the test split, selection held fixed. The zero-stake share is the fraction of selected races where the staking rule puts nothing on."
)
```

```{r}
#| label: tbl-staking-val
p6_tbl(
  p6_staking_val |>
    dplyr::transmute(
      Market = unname(BET_LAB[bet]), Rule = rule, Staking = staking,
      `Races selected` = n_races_selected, `Races staked` = n_bets,
      `Zero-stake share` = p6_fpct(share_zero_stake, 1),
      `Total staked` = sprintf("%.1f", total_stake),
      ROI = p6_fpct(roi, 1, TRUE),
      `Fair book` = p6_fpct(roi_fair, 1, TRUE)),
  caption = "The same on the search set."
)
```

Two mechanisms. Where the selection already requires the model's probability
to exceed the price, the zero-stake share is 0.0% and the staking rules lever
up an identical bet set, so a negative return becomes more negative. Where it
does not, the stake collapses.

```{r}
#| label: tbl-staking-thin
p6_tbl(
  thin |>
    dplyr::transmute(Split = split, Market = unname(BET_LAB[bet]),
                     Rule = rule, Staking = staking,
                     `Races selected` = n_races_selected,
                     `Races staked` = n_bets,
                     `Zero-stake share` = p6_fpct(share_zero_stake, 1),
                     `Total staked` = sprintf("%.1f", total_stake),
                     ROI = p6_fpct(roi, 1, TRUE)),
  caption = "The arms where the staking rule discards most of the bet set. An ROI on this many units is not a staking result."
)
```

That is the market-favourite rules: the model usually rates the favourite
below the market, so both non-flat rules stake nothing on most selected races.
At worst the rule stakes `r p6_fn(min(thin$n_bets))` of
`r p6_fn(max(thin$n_races_selected))`. Report the zero-stake share, not the
ROI alone. In the first attempt at this paper the declared win rule staked 33
units across 2,183 test races with an interval 54 points wide, and none of
that was visible in its ROI.
