# The result {#sec-result}

```{r}
#| label: result-numbers
#| include: false

inc_win  <- p6_res(p6_test_results, "incumbent", "win")
mkt_win  <- p6_res(p6_test_results, "market-only", "win")
dec_win  <- p6_res(p6_test_results, "^declared", "win")
inc_plc  <- p6_res(p6_test_results, "incumbent", "place")
mkt_plc  <- p6_res(p6_test_results, "market-only", "place")
dec_plc  <- p6_res(p6_test_results, "^declared", "place")
inc_ewc  <- p6_res(p6_test_results, "incumbent", "eachway_corrected")
mkt_ewc  <- p6_res(p6_test_results, "market-only", "eachway_corrected")
dec_ewc  <- p6_res(p6_test_results, "^declared", "eachway_corrected")
inc_ew5  <- p6_res(p6_test_results, "incumbent", "eachway")

con_win_s1 <- p6_con(p6_test_contrasts, "win", "incumbent")
con_win_s4 <- p6_con(p6_test_contrasts, "win", "market-only")
con_plc_s4 <- p6_con(p6_test_contrasts, "place", "market-only")

sel_win <- p6_declared_selection_sizes |> dplyr::filter(bet == "win")
```

Nothing wins.

Nine selection rules and three staking rules were searched on the
`r p6_comma(n_train_races)` training races. One rule per bet type was
declared by a mechanical rule, frozen in a committed file, and scored once on
the `r p6_comma(n_test_races)` test races. All three declared rules lose
money, all three lose more than the rule they were meant to beat, and every
paired interval against that rule contains zero. On the win market the
declared rule returns `r p6_fmt_pct(dec_win$roi)` on `r dec_win$n_bets` bets
against the incumbent's `r p6_fmt_pct(inc_win$roi)`, a difference of
`r p6_fmt_pts(con_win_s1$diff_point)` points with a 90% interval of
`r p6_fmt_pts_ci(con_win_s1$ci_lo, con_win_s1$ci_hi)`.

The result that matters is the control. S4 backs the market favourite in
every race at a flat stake and uses no model output of any kind. On the win
market it returns `r p6_fmt_pct(mkt_win$roi)` over all
`r p6_comma(mkt_win$n_bets)` test races, against
`r p6_fmt_pct(inc_win$roi)` for the model rule, and its own 90% interval,
`r p6_fmt_pct_ci(mkt_win$ci_lo, mkt_win$ci_hi)`, excludes zero. The paired
contrast of the declared rule against it is
`r p6_fmt_pts(con_win_s4$diff_point)` points,
`r p6_fmt_pts_ci(con_win_s4$ci_lo, con_win_s4$ci_hi)`. **A rule with no model
in it does at least as well as every rule that has one.** On the win market
it does better.

That does not generalise across bet types. On the place market the ordering
reverses: the model rule returns `r p6_fmt_pct(inc_plc$roi)` against the
favourite rule's `r p6_fmt_pct(mkt_plc$roi)`, and the declared rule beats the
favourite by `r p6_fmt_pts(con_plc_s4$diff_point)` points,
`r p6_fmt_pts_ci(con_plc_s4$ci_lo, con_plc_s4$ci_hi)` — the one interval in
the paper that excludes zero in the model's favour. Backing favourites to
place is an expensive way to lose money, because the place market prices
short-priced horses hardest; that is a statement about the place market's
shape, not about the model.

```{r}
#| label: tbl-headline
#| tbl-cap: "The test split, scored once. Each-way is reported under both paper 5's flat terms and the corrected industry terms; see @sec-terms."
p6_results_table(p6_test_results) |>
  kable(align = "llrrrrrrrl")
```

Two further things the numbers say, both structural rather than about the
model.

**The whole of the incumbent's loss is the over-round.** Priced at a
zero-margin book — the same starting prices, renormalised within race to sum
to one — the incumbent's win ROI moves from `r p6_fmt_pct(inc_win$roi)` to
`r p6_fmt_pct(inc_win$roi_fair)`, and the favourite rule's from
`r p6_fmt_pct(mkt_win$roi)` to `r p6_fmt_pct(mkt_win$roi_fair)`. The margin
paid is `r p6_fmt_pct(inc_win$roi_fair - inc_win$roi)` and
`r p6_fmt_pct(mkt_win$roi_fair - mkt_win$roi)` respectively. Both rules are
close to break-even or better against a fair book and lose against a real
one. That is the whole story of the series' betting results in one line, and
it is not a modelling result: it is what a bookmaker's margin costs.

**Staking rules select.** Both non-flat staking rules stake nothing when the
model's probability does not exceed the price, so they change the bet set as
well as the stake. In sample this costs little; out of sample it is severe.
The declared win rule selects `r p6_comma(sel_win$n_races_selected)` test races but places a
positive stake in only `r dec_win$n_bets` of them, for
`r sprintf("%.0f", dec_win$total_stake)` units of total stake across a
`r p6_comma(n_test_races)`-race split. Its interval is
`r p6_fmt_pct_ci(dec_win$ci_lo, dec_win$ci_hi)` wide. There is not enough
betting there to learn anything, and the declaration rule had no way to know
that from training numbers alone.

The rest of the paper is how those numbers were arrived at, and what is wrong
with the exercise.
