# Data and scope {#sec-data}

## Race selection and window

The data scope is unchanged from paper 1: UK All-Weather (AW) Flat
handicaps in Classes 2–5, run at the four AW courses (Kempton,
Lingfield, Southwell and Wolverhampton) between 2006-01-01 and
2015-10-14, with declared fields of 4–16 and exactly one winner per
race. We refer the reader to paper 1 for the full selection criteria and
their motivation (the `race_type` filter, the 2006 class-system cutoff,
the cross-surface career history, and the dead-heat and non-runner
handling), none of which we revisit here. The chronological 70/30
train/test split is also carried over verbatim: every race on or before
2012-12-30 is training, everything after is test, so no within-race
information crosses the boundary.

```{r}
#| label: tbl-data-counts
# Training set only for fitting; counts here describe the full split.
data_counts <- runners_model_ready |>
  dplyr::group_by(split) |>
  dplyr::summarise(
    Races   = dplyr::n_distinct(race_id),
    Runners = dplyr::n(),
    .groups = "drop"
  ) |>
  dplyr::mutate(split = dplyr::recode(split, train = "Training (≤ 2012-12-30)",
                                              test  = "Test (> 2012-12-30)")) |>
  dplyr::arrange(dplyr::desc(Races))

data_counts |>
  dplyr::mutate(dplyr::across(c(Races, Runners), scales::comma)) |>
  dplyr::rename(Split = split) |>
  knitr::kable(caption = "Race and runner counts by split, on the qualifying AW Class 2–5 handicap population.")
```

## Feature set

Paper 2 works from a two-tier predictor set: the Owen baseline,
re-estimated on the AW data, and a set of new candidates drawn from the
racing literature and the All-Weather context. @tbl-feature-set lists
them, grouped by tier and by whether each is a horse-level feature or a
race-level feature entered through an interaction.

```{r}
#| label: tbl-feature-set
tibble::tribble(
  ~Group,                 ~Feature,                ~Description,                                              ~Type,                         ~Source,
  "Owen baseline",        "age_diff",              "Symmetric distance from peak age (3 on AW)",              "horse-level",                 "Owen (2019)",
  "Owen baseline",        "daysLTO (log)",         "Log days since last run",                                 "horse-level",                 "Owen (2019)",
  "Owen baseline",        "trainerSR / sireSR",    "Cumulative pre-race strike rate",                         "horse-level",                 "Owen (2019)",
  "Owen baseline",        "position lags 1–3",     "Prior finishing positions (Model S encoding)",            "horse-level",                 "Owen (2019)",
  "Owen baseline",        "entire / gelding",      "Sex indicators",                                          "horse-level",                 "Owen (2019)",
  "Owen baseline",        "cheekpieces",           "Tack indicator",                                          "horse-level",                 "Owen (2019)",
  "New horse-level",      "jockeySR",              "Cumulative pre-race jockey strike rate",                  "horse-level",                 "Juola (2018)",
  "New horse-level",      "or_relative",           "Official rating minus race-mean rating",                  "horse-level",                 "Betwise blog",
  "New horse-level",      "or_missing",            "Indicator: official rating was NULL",                     "horse-level",                 "Betwise blog",
  "New horse-level",      "rel_weight",            "Weight carried minus race-mean weight",                   "horse-level",                 "Benter (1994)",
  "New horse-level",      "trainer_aw_premium",    "Trainer AW strike rate minus overall strike rate",        "horse-level",                 "Juola (2018)",
  "New horse-level",      "has_wins",              "Indicator: any prior career win",                         "horse-level",                 "Betwise blog",
  "Race-level × horse",   "rel_weight × distance", "Weight deviation interacted with race distance",          "race-level × horse-level",    "Benter (1994)",
  "Race-level × horse",   "draw × course",         "Normalised draw interacted with course",                  "race-level × horse-level",    "Betwise blog"
) |>
  knitr::kable(caption = "Paper-2 feature set: Owen baseline (re-estimated on AW), new horse-level candidates, and race-level interactions.")
```

## Data decisions

Three decisions in this paper warrant a note; each is taken up in detail
in the section where it is used.

**Official rating imputation.** The relative official rating `or_relative`
is NULL for about 10.5% of runners. The missingness is not at random —
unrated horses are typically unexposed (too few runs to be rated) or
returning from a long absence — so we do not drop them, which would
discard roughly a quarter of training races and skew the early window.
We impute `or_relative` $= 0$ and add a binary companion `or_missing`
flagging the imputation; this is the standard missing-indicator approach,
and it lets the model estimate the missing-rating group effect separately
rather than forcing those horses onto the zero point of the rating scale
(@sec-eda).

**Position encoding.** Paper 1 entered the three prior-finish lags as
twelve factor coefficients. We adopt a more parsimonious *Model S*
encoding — six coefficients, two per lag (a zero indicator plus a single
linear slope across placings 1–4) — which a likelihood-ratio test finds
loses nothing against the full factor grid. The encoding and the test
are set out in @sec-eda.

**Draw × course construction.** Normalised draw (`stall_normalised`)
showed near-zero standalone signal in the univariate screen, but draw
bias on AW is known to be track-specific. We enter it as four
course-specific slopes in the mixed logit (@sec-mixed-interactions); the
Lingfield term is individually non-significant and is dropped after a
one-degree-of-freedom reduction test (LR $=$
`r round(final_reduction_lr$lr_stat, 3)`, df $=$ `r final_reduction_lr$df`,
$p =$ `r round(final_reduction_lr$p_value, 3)`), leaving course-specific
draw effects for Kempton, Southwell and Wolverhampton.
