# Feature evaluation {#sec-eda}

All exploratory analysis in this section uses the **training split
only** (70/30 chronological, cutoff 2012-12-30). Win-rate summaries,
tables, and plots are computed after filtering to `split == "train"`,
so nothing here is informed by the held-out test races. (Paper 1
computed its exploratory summaries on the full joined tibble; that
paper is complete and is not being revisited — see the exploratory
analysis convention in the project notes.)

```{r}
#| label: eda-setup
#| include: false
# Training set only — see CLAUDE.md exploratory analysis convention
source(file.path(
  rprojroot::find_root(rprojroot::has_file("awracing.Rproj")),
  "papers", "02_extended_features", "_helpers.R"
))

# Single training-only working tibble used by every chunk below.
train <- runners_model_ready |> dplyr::filter(split == "train")

# Bin breaks reused across the win-rate plots.
sr_breaks      <- c(0, 0.05, 0.10, 0.15, 0.20, 0.30, 0.50, 1.0)
days_breaks    <- c(0, 7, 14, 30, 60, 90, 180, 365, Inf)
or_breaks      <- c(-Inf, -20, -10, -5, -2, 2, 5, 10, 20, Inf)
weight_breaks  <- c(-Inf, -8, -5, -3, -1, 1, 3, 5, 8, Inf)
premium_breaks <- c(-Inf, -0.10, -0.02, 0.02, 0.10, 0.25, Inf)
```

## Headline counts (training split)

```{r}
#| label: tbl-headline
# Training set only — see CLAUDE.md exploratory analysis convention
field_sizes <- train |> dplyr::count(race_id, name = "field_size")

tibble::tibble(
  `Training races`   = scales::comma(dplyr::n_distinct(train$race_id)),
  `Training runners` = scales::comma(nrow(train)),
  `Date range`       = paste(min(train$race_date), "to", max(train$race_date)),
  `Field size (min / median / max)` = paste(
    min(field_sizes$field_size),
    stats::median(field_sizes$field_size),
    max(field_sizes$field_size),
    sep = " / "
  ),
  `Win rate` = scales::percent(mean(train$won), accuracy = 0.01)
) |>
  tidyr::pivot_longer(dplyr::everything(), names_to = "Quantity", values_to = "Value") |>
  knitr::kable(caption = "Training-split headline counts (70/30 chronological, cutoff 2012-12-30).")
```

With field sizes between 4 and 16 and exactly one winner per race, the
headline win rate per runner-row is the reciprocal of the mean field
size; it is a property of the choice-set geometry, not of any feature.

## Automated feature evaluation

We rank every candidate feature by a univariate screen. For each
feature we fit a standalone no-intercept conditional logit (race =
choice set) on the training races and score it by the depth-1
Plackett–Luce $R^2$,
$$
R^2_{\text{PL}} = 1 - \frac{\log L_{\text{model}}}{\log L_{\text{null}}},
$$
where $\log L_{\text{null}}$ assigns every runner the equal probability
$1/\text{field size}$. This is the univariate analogue of the exploded
Plackett–Luce fit used later, following @juola2018 (§4.3.6). Each
feature is scored on its own complete-choice-set subset, so `n_races`
varies with the feature's coverage. The three prior-finish lags are
screened together as one conceptual feature (`position_lags`).

```{r}
#| label: tbl-feature-screen
# Training set only — see CLAUDE.md exploratory analysis convention
# (feature_screen is built from runners_augmented restricted to races_train)
feature_screen |>
  dplyr::transmute(
    Feature  = feature,
    `PL-R²`  = round(pl_r2, 5),
    `n races` = n_races
  ) |>
  knitr::kable(
    caption = "Univariate PL-R² screen, all 19 candidate features, ranked. Training races only."
  )
```

Two caveats temper the ranking. First, each score depends on the
assumed functional form: a feature entered linearly may screen weakly
yet carry signal under a better transformation (or as an interaction).
Second, standalone univariate signal is not the same as marginal
contribution — a feature can rank highly alone yet add little once
correlated features are already in the model. The screen is a
first-pass filter, not the final word; the conditional-logit fits in
@sec-extended-win are the proper multivariate test.

## Feature drop decisions

We drop the weakest candidates before model fitting, on the evidence
of the screen above.

**`weight_delta_lbs` (dropped).** Its NA mechanism is not
missing-at-random: a missing value means the horse has *no prior run*,
not that its weight was unchanged. Imputing 0 would assert "no change
since last time" for a horse that has no last time — a substantively
false statement. The univariate signal is weak (PL-R² 0.008), and
`rel_weight` already captures the within-race weight level without any
NA problem. Dropped.

**Weak-signal features (dropped).** Each of the following has a
univariate PL-R² at or near zero and is dropped before model fitting:

- `class_delta` (PL-R² 0.00004) — the handicapper already prices class
  moves into the allotted weight, so the residual signal is negligible.
- `sire_aw_premium` (0.0008) — no standalone All-Weather edge over the
  sire's overall rate.
- `jockey_aw_premium` (0.0007) — likewise for jockeys.
- `stall_normalised` (0.0003) — draw effect is near-zero pooled across
  courses and distances (a candidate for a course interaction in a
  later paper, not a main effect here).
- `first_time_aw` (0.0001) — AW debut carries essentially no marginal
  win signal once form is accounted for.

**Deferred decisions.** `gelding`, `cheekpieces`, `has_wins`, and
`trainer_aw_premium` are *not* dropped here. They enter the full
conditional-logit model in @sec-extended-win and are removed only if
their coefficients fail the significance test ($p > 0.05$) in the
reduced model — the same reduction rule paper 1 applied to Owen's
specification.

## Position encoding: from factors to Model S

The three prior-finish lags record a horse's finishing position in its
most recent, second-most-recent and third-most-recent runs. Each takes a
value in $\{0, 1, 2, 3, 4\}$: $1$–$4$ for a top-four finish, and $0$ for
"no qualifying recent finish" — either no prior run, or a finish worse
than fourth.

Paper 1 entered each lag as a *factor*, spending one coefficient on each
placing $1$–$4$ (with $0$ as the reference): four coefficients per lag,
twelve in all. That is maximally flexible but not parsimonious. We test
two cheaper encodings against it on the training races, position terms
only. **Model P** is the aggressive one — a single position score plus a
lag-decay term, two coefficients for all three lags combined. **Model S**
is the middle ground: per lag it gives the $0$ category its own term and
forces only placings $1$–$4$ onto a straight line, via a binary indicator
`pos_lagN_zero` (is the value $0$?) and a numeric slope `pos_lagN_nonzero`
(the placing $1$–$4$, else $0$) — two coefficients per lag, six in all.
@tbl-position-encoding shows how one lag's value enters the linear
predictor under the factor grid and under Model S.

| Last finish | Model F (factor) | Model S (adopted) |
|:---|:---:|:---:|
| 0 (no top-4) | reference | $\gamma_{\text{zero}}$ |
| 1st | $\beta_1$ | $\gamma_{\text{slope}}\times 1$ |
| 2nd | $\beta_2$ | $\gamma_{\text{slope}}\times 2$ |
| 3rd | $\beta_3$ | $\gamma_{\text{slope}}\times 3$ |
| 4th | $\beta_4$ | $\gamma_{\text{slope}}\times 4$ |
| **coefficients per lag** | **4** | **2** |

: One prior-finish lag's contribution to the linear predictor. Model F gives every placing a free effect; Model S separates the "no recent finish" category and constrains placings 1–4 to a straight line (three lags: twelve coefficients under F, six under S). {#tbl-position-encoding}

```{r}
#| label: model-s-lr
pt <- position_parsimony_test
```

Model P is decisively rejected against the factor grid (LR $=$
`r round(pt$lr_stat, 0)`, df $=$ `r pt$df`, $p < 0.001$): collapsing
everything to one slope forces the $0$ category onto the same numeric
scale as a win, which is substantively wrong. Model S, by contrast, is
**not** rejected (LR $=$ `r round(pt$lr_stat_s, 2)`, df $=$
`r pt$lr_df_s`, $p =$ `r round(pt$p_value_s, 3)`), and AIC prefers it to
the twelve-coefficient grid by about 7 points. So once the $0$ category
has its own term, placings $1$–$4$ are effectively linear, and six
coefficients capture what twelve did. We adopt Model S for every paper-2
fit; the empirical win-rate-by-position curves that motivate the
linearity are plotted later in this section.

## `or_relative` — coverage and imputation

`or_relative` (a runner's official rating minus the race-mean official
rating) is the strongest new feature in the screen, but it is also the
one with a real coverage cost.

```{r}
#| label: tbl-or-coverage
# Training set only — see CLAUDE.md exploratory analysis convention
or_runner_pct <- mean(train$or_missing == 1)
or_race_n     <- train |>
  dplyr::group_by(race_id) |>
  dplyr::summarise(any_missing = any(or_missing == 1), .groups = "drop") |>
  dplyr::summarise(n = sum(any_missing)) |>
  dplyr::pull(n)
or_race_total <- dplyr::n_distinct(train$race_id)

tibble::tibble(
  `Runners with NULL official rating` = scales::percent(or_runner_pct, accuracy = 0.1),
  `Training races affected`           = paste0(
    scales::comma(or_race_n), " of ", scales::comma(or_race_total),
    " (", scales::percent(or_race_n / or_race_total, accuracy = 1), ")"
  )
) |>
  tidyr::pivot_longer(dplyr::everything(), names_to = "Quantity", values_to = "Value") |>
  knitr::kable(caption = "Official-rating coverage on the training split.")
```

The missingness is **not at random**: a horse with no official rating
is typically unexposed (too few qualifying runs to be rated) or
returning from a long absence — a distinct population, not a random
sample of the field. Dropping these runners complete-case would discard
roughly a quarter of training races, and (because the gap is
concentrated in the early part of the window) would skew the training
period.

Our decision is therefore to **impute `or_relative` = 0** for unrated
runners and add a binary companion **`or_missing`** (1 where the rating
was NULL). The companion lets the model estimate the missing-OR group
effect separately, rather than forcing those horses onto the zero point
of the `or_relative` scale (which would otherwise read as "exactly
average"). This recovers full coverage at no cost to race count, and is
applied in the `runners_model_ready` target that every subsequent
modelling step consumes.

## Win-rate plots for retained features

The plots below follow paper 1's visual style: empirical win rate per
bin with per-bin standard errors, a dashed reference line at the
overall (training) win rate, and the bin count above each bar. All are
computed on the training split only. Dropped features are not plotted.

### Official rating (relative) and its missingness indicator

The continuous `or_relative` curve is shown on rated runners only
(`or_missing == 0`), so the binning reflects genuine rating spread
rather than the imputed zeros; `or_missing` is then shown on its own.

```{r}
#| label: tbl-or-relative
# Training set only — see CLAUDE.md exploratory analysis convention
or_summary <- train |>
  dplyr::filter(or_missing == 0) |>
  summarise_win_rate("or_relative", bins = or_breaks)
or_summary |>
  knitr::kable(digits = 4, caption = "Win rate by relative official rating (rated runners).")
```

```{r}
#| label: fig-or-relative
#| fig-cap: "Win rate by relative official rating (rated training runners)."
# Training set only — see CLAUDE.md exploratory analysis convention
or_summary |> plot_win_rate()
```

```{r}
#| label: fig-or-missing
#| fig-cap: "Win rate by or_missing (1 = no official rating)."
# Training set only — see CLAUDE.md exploratory analysis convention
train |> summarise_win_rate("or_missing") |> plot_win_rate()
```

### Relative weight

```{r}
#| label: fig-rel-weight
#| fig-cap: "Win rate by relative weight (lb above/below the race mean)."
# Training set only — see CLAUDE.md exploratory analysis convention
train |> summarise_win_rate("rel_weight", bins = weight_breaks) |> plot_win_rate()
```

### Strike rates: jockey, trainer, sire

`jockeySR` is the new jockey analogue of Owen's `trainerSR` / `sireSR`;
all three are cumulative pre-race win rates, uncapped, binned with finer
resolution at the low end where the mass sits.

```{r}
#| label: fig-strike-rates
#| fig-cap:
#|   - "Win rate by jockey strike rate."
#|   - "Win rate by trainer strike rate (AW training data)."
#|   - "Win rate by sire strike rate (AW training data)."
# Training set only — see CLAUDE.md exploratory analysis convention
for (feat in c("jockeySR", "trainerSR", "sireSR")) {
  print(plot_win_rate(summarise_win_rate(train, feat, bins = sr_breaks)))
}
```

### Trainer All-Weather premium

Carried into the full model; the drop decision is deferred to the
reduced-model coefficient test.

```{r}
#| label: fig-trainer-aw-premium
#| fig-cap: "Win rate by trainer AW premium (AW rate minus overall rate)."
# Training set only — see CLAUDE.md exploratory analysis convention
train |> summarise_win_rate("trainer_aw_premium", bins = premium_breaks) |> plot_win_rate()
```

### Owen baseline features carried forward

`age_diff` (symmetric distance from age 3) and `daysLTO` (binned on
log-spaced rest intervals) are replotted here on the AW training split.

```{r}
#| label: fig-age-diff
#| fig-cap: "Win rate by age_diff = min(|age - 3|, 5)."
# Training set only — see CLAUDE.md exploratory analysis convention
train |> summarise_win_rate("age_diff") |> plot_win_rate()
```

```{r}
#| label: fig-days-lto
#| fig-cap: "Win rate by days since last race, on log-spaced rest bins."
# Training set only — see CLAUDE.md exploratory analysis convention
train |> summarise_win_rate("days_LTO", bins = days_breaks) |> plot_win_rate()
```

### Binary sex / tack / form indicators

`entire`, `gelding`, and `cheekpieces` carry forward from Owen's
specification; `has_wins` (any prior career win) is a new binary
candidate carried into the full model.

```{r}
#| label: fig-binaries
#| fig-cap:
#|   - "Win rate by entire"
#|   - "Win rate by gelding"
#|   - "Win rate by cheekpieces"
#|   - "Win rate by has_wins"
# Training set only — see CLAUDE.md exploratory analysis convention
for (feat in c("entire", "gelding", "cheekpieces", "has_wins")) {
  print(plot_win_rate(summarise_win_rate(train, feat)))
}
```

### Prior finishing positions

The three prior-finish lags are screened as one feature but plotted
separately, most-recent first. Each is an integer in {0, 1, 2, 3, 4},
where 0 means "no prior run, or finished worse than 4th".

```{r}
#| label: fig-positions
#| fig-cap:
#|   - "Win rate by position1 (most recent prior finish)."
#|   - "Win rate by position2 (second-most-recent prior finish)."
#|   - "Win rate by position3 (third-most-recent prior finish)."
# Training set only — see CLAUDE.md exploratory analysis convention
for (feat in c("position1", "position2", "position3")) {
  print(plot_win_rate(summarise_win_rate(train, feat)))
}
```
