# Exploded conditional logit {#sec-exploded-logit}

The win model in @sec-extended-win uses only one bit of each race's
result: who won. The finishing order carries more. The **exploded
conditional logit** [@henery1981] uses it by treating the
observed order as a sequence of conditional win problems: the winner is
chosen from the full field, then the runner-up is chosen from the field
minus the winner, then third from the field minus the first two, and so
on. Each step is an ordinary conditional-logit choice on a shrinking
choice set, and pooling them and fitting a single conditional logit
across the pool is exactly maximising the Plackett–Luce likelihood. The
algebra — and the equivalence between the exploded representation and
the Plackett–Luce likelihood — is derived in Appendix A (lifted from
`notes/paper2_seed_plackett_luce.md`).

We truncate at depth $k = 3$: only positions 1, 2 and 3 are modelled.
The ordering among the tail of the field is largely noise [@henery1981],
so the fictitious sub-races beyond third are dropped from the
likelihood. The specification is otherwise **identical** to the reduced
win model — same 17 terms, same Model-S position encoding, same imputed
`or_relative`/`or_missing` — so any difference is the explosion, not the
features. Fitting is training-only; evaluation is depth-1 (win) on the
test set, directly comparable to @sec-extended-win and paper 1.

```{r}
#| label: p2exp-setup
#| include: false
# Fits loaded from targets (not fitted here) -> load mlogit's namespace so
# summary.mlogit / coef.mlogit dispatch (CLAUDE.md note).
loadNamespace("mlogit")

win_ct <- summary(model_p2_reduced)$CoefTable
exp_ct <- summary(model_p2_exploded)$CoefTable
terms  <- rownames(win_ct)

win_coef <- stats::setNames(win_ct[, "Estimate"], terms)
exp_coef <- stats::setNames(exp_ct[, "Estimate"], terms)
win_z    <- stats::setNames(win_ct[, "z-value"],  terms)
exp_z    <- stats::setNames(exp_ct[, "z-value"],  terms)

exp_sets   <- dplyr::n_distinct(exploded_train_data$exploded_race_id)
exp_races  <- dplyr::n_distinct(exploded_train_data$race_id)
exp_rows   <- nrow(exploded_train_data)

pl_r2_win <- model_p2_diagnostics$pl_r2[model_p2_diagnostics$model == "reduced"]
pl_r2_exp <- model_p2_exploded_diagnostics$pl_r2
```

## From win to ranking: the exploded construction

`prepare_exploded_data()` builds the pooled choice sets. For each race
it forms $k = 3$ fictitious sub-races: choice set $s$ is the horse that
finished $s$th together with every horse not yet ranked — all finishers
placed $s$-or-worse, plus any non-finishers, which never get ranked and
so stay in the pool at every depth. The depth-1 sub-race is therefore
exactly the win model's choice set (the full field, winner = 1st), which
is what makes the two models comparable at depth 1. Each sub-race gets a
composite id `"<race_id>_<s>"` as the choice index and a fresh per-set
`horse_ref`; a `stopifnot` confirms exactly one winner per set.

Two race filters apply. Races with any NA in the 19 modelling columns are
dropped, the same rule and race set as the win model. Races whose top-3
finishing positions are not each recorded exactly once — a dead-heat for
2nd or 3rd, or a gap — cannot be unambiguously exploded and are dropped
too (finishing position uses `coalesce(amended_position, finish_position)`
throughout).

```{r}
#| label: tbl-exploded-size
# exploded_train_data is a dfidx; use plain vector access (its special
# `idx` column breaks dplyr mutate/rename — see CLAUDE.md mlogit notes).
depth_rows <- as.integer(table(exploded_train_data$depth))
tibble::tibble(
  `Sub-race`    = paste0("depth ", seq_along(depth_rows), " (",
                         c("full field", "field − 1st", "field − 1st,2nd"), ")"),
  `Runner-rows` = depth_rows
) |>
  knitr::kable(caption = sprintf(
    "Exploded training data: %s valid races become %s choice sets (3 each) over %s runner-rows. The 42-race gap versus the win model's 5,065 training races is races with a tie or gap in the top 3 finishers.",
    format(exp_races, big.mark = ","),
    format(exp_sets, big.mark = ","),
    format(exp_rows, big.mark = ",")
  ))
```

## Coefficients: win model vs exploded

```{r}
#| label: tbl-coef-compare
#| code-fold: true

block_of <- function(term) {
  if (grepl("^pos_lag", term))                                    return("1. Position (Model S)")
  if (term %in% c("days_LTO_log", "trainerSR", "sireSR"))         return("2. Form / history")
  if (term %in% c("age_diff", "entire", "gelding", "cheekpieces")) return("3. Horse characteristics")
  "4. New features"
}
pretty_term <- function(t) dplyr::recode(t, days_LTO_log = "daysLTO (log)")

tibble::tibble(
  Block            = vapply(terms, block_of, character(1)),
  Term             = pretty_term(terms),
  `Win est.`       = round(win_coef[terms], 4),
  `Win z`          = round(win_z[terms], 2),
  `Exploded est.`  = round(exp_coef[terms], 4),
  `Exploded z`     = round(exp_z[terms], 2)
) |>
  dplyr::arrange(Block) |>
  knitr::kable(
    align   = c("l", "l", "r", "r", "r", "r"),
    caption = "Reduced win model (depth-1 win only) vs the exploded fit
(depth-3 Plackett–Luce), same 17-term specification. Estimates and Wald
z-values side by side."
  )
```

Two patterns stand out, and they pull in opposite directions.

**Estimates shrink.** Almost every coefficient is smaller in magnitude
under the explosion: `jockeySR` falls from `r sprintf("%+.2f", win_coef[["jockeySR"]])`
to `r sprintf("%+.2f", exp_coef[["jockeySR"]])`, `or_relative` from
`r sprintf("%+.3f", win_coef[["or_relative"]])` to
`r sprintf("%+.3f", exp_coef[["or_relative"]])`, and `rel_weight`
collapses from `r sprintf("%+.3f", win_coef[["rel_weight"]])` to
`r sprintf("%+.4f", exp_coef[["rel_weight"]])`. The reason is what each
model is asked to explain. The win model loads a feature with all of its
ability to separate the *winner* from the field; the exploded model must
make the same coefficient explain the 1-vs-rest, 2-vs-rest and
3-vs-rest contrasts at once. Features that matter more for *winning*
than for separating mid-pack places — weight carried being the clearest
case — are pulled toward their weaker average-over-depths effect.

**Precision sharpens.** Despite the smaller estimates, most z-values
*grow*, because depth-3 explosion roughly triples the number of choice
problems the same races contribute: `days_LTO_log` sharpens from
`r sprintf("%.1f", win_z[["days_LTO_log"]])` to
`r sprintf("%.1f", exp_z[["days_LTO_log"]])`, `pos_lag1_zero` from
`r sprintf("%.1f", win_z[["pos_lag1_zero"]])` to
`r sprintf("%.1f", exp_z[["pos_lag1_zero"]])`, `trainerSR` from
`r sprintf("%.1f", win_z[["trainerSR"]])` to
`r sprintf("%.1f", exp_z[["trainerSR"]])`. The form and position features
— the ones that order the whole field, not just crown the winner — gain
the most, which is the hypothesis the explosion was meant to test:
recent-form signal is estimated more precisely once the full finishing
order, not just the win, informs the fit. The exception is `rel_weight`,
whose z *falls* — confirming, from the other direction, that weight is a
win-specific rather than a rank-general signal.

## Out-of-sample diagnostics

We evaluate the exploded model exactly as the win model: depth-1 win
probabilities on the test races, against the over-round-adjusted market.
The calibration plot, P1/P2 scoring rules and betting backtest are the
same suite as @sec-extended-win and so are directly comparable.

```{r}
#| label: p2exp-eval-setup
#| include: false
# Test set only. Reuse the helper functions defined in @sec-extended-win
# (calibration_summary / scoring_summary / bin_breaks / thresholds — all
# in scope from the earlier included section).
eval_data_ex <- test_predictions_p2_exploded |> dplyr::rename(model_prob = predicted_prob)
```

```{r}
#| label: fig-calibration-exploded
#| fig-cap: "Calibration on the AW test set, exploded model (Owen Figure 7 equivalent): observed win proportion vs predicted / implied win probability, 5-point bins over [0, 0.50]. Circles = exploded model (depth-1, out-of-sample); plus signs = market; dashed line y = x."
calib_summary_ex <- dplyr::bind_rows(
  calibration_summary(eval_data_ex, "model_prob",  "Model (exploded)"),
  calibration_summary(eval_data_ex, "market_prob", "Market")
)
ggplot2::ggplot(calib_summary_ex,
                ggplot2::aes(x = bin_mid, y = observed, shape = source, colour = source)) +
  ggplot2::geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey50") +
  ggplot2::geom_point(size = 3) +
  ggplot2::scale_shape_manual(values = c("Model (exploded)" = 1L, Market = 3L)) +
  ggplot2::scale_colour_manual(values = c("Model (exploded)" = "darkgreen", Market = "tomato")) +
  ggplot2::scale_x_continuous(labels = scales::percent_format(accuracy = 1),
                              limits = c(0, 0.50), breaks = bin_breaks) +
  ggplot2::scale_y_continuous(labels = scales::percent_format(accuracy = 1), limits = c(0, NA)) +
  ggplot2::labs(x = "Predicted / implied win probability",
                y = "Observed win proportion", shape = NULL, colour = NULL) +
  ggplot2::theme_minimal()
```

```{r}
#| label: fig-p1-p2-exploded
#| fig-cap:
#|   - "P1 vs probability threshold, exploded model (Owen Figure 8a equivalent), AW test set."
#|   - "P2 vs probability threshold, exploded model (Owen Figure 8b equivalent), AW test set."
score_long_ex <- dplyr::bind_rows(
  scoring_summary(eval_data_ex, "model_prob",  "Model (exploded)", thresholds),
  scoring_summary(eval_data_ex, "market_prob", "Market",           thresholds)
)
p_score_ex <- function(yvar, ylab) {
  ggplot2::ggplot(score_long_ex,
                  ggplot2::aes(x = threshold, y = .data[[yvar]],
                               linetype = source, colour = source)) +
    ggplot2::geom_line(linewidth = 0.8) +
    ggplot2::scale_linetype_manual(values = c("Model (exploded)" = "solid", Market = "dashed")) +
    ggplot2::scale_colour_manual(values = c("Model (exploded)" = "darkgreen", Market = "tomato")) +
    ggplot2::scale_x_continuous(labels = scales::percent_format(accuracy = 1), limits = c(0, 0.50)) +
    ggplot2::labs(x = "Probability threshold", y = ylab, linetype = NULL, colour = NULL) +
    ggplot2::theme_minimal()
}
print(p_score_ex("P1", "P1 (geom. mean prob. on winners)"))
print(p_score_ex("P2", "P2 (geom. mean Bernoulli likelihood)"))
```

```{r}
#| label: tbl-backtest-naive-exploded
backtest_naive_p2_exploded |>
  dplyr::transmute(
    `Prob threshold`  = prob_threshold,
    `Ratio threshold` = ratio_threshold,
    `Bets`            = n_bets,
    `Wins`            = n_wins,
    `Gross return`    = sprintf("%.2f", gross_return),
    `Profit`          = sprintf("%.2f", profit),
    `ROI`             = scales::percent(roi, accuracy = 0.1)
  ) |>
  knitr::kable(caption = "Exploded model: betting backtest on the AW test races at Owen's exact thresholds.")
```

```{r}
#| label: fig-roi-sweep-exploded
#| fig-cap: "ROI vs ratio threshold on the AW test set (exploded model), 90% bootstrap CI shaded. Model-probability filter held at 0.13; dashed line at zero marks break-even."
ggplot2::ggplot(backtest_sweep_p2_exploded, ggplot2::aes(x = tau, y = roi)) +
  ggplot2::geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  ggplot2::geom_ribbon(ggplot2::aes(ymin = ci_lo, ymax = ci_hi), fill = "darkgreen", alpha = 0.18) +
  ggplot2::geom_line(linewidth = 0.8, colour = "darkgreen") +
  ggplot2::scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  ggplot2::labs(x = "Ratio threshold (model_prob / market_prob)", y = "ROI") +
  ggplot2::theme_minimal()
```

### Each model on both objectives

```{r}
#| label: tbl-depth-tradeoff
#| code-fold: true
# Win-model coefficients scored on the same exploded choice sets give the
# win model's depth-3 PL logLik (training races, in-sample). exploded_train_data
# is a dfidx -> pull columns as plain vectors.
beta_win  <- stats::coef(model_p2_reduced)
terms_win <- names(beta_win)
won_vec   <- as.integer(exploded_train_data$won)
Xw        <- vapply(terms_win, function(t) as.numeric(exploded_train_data[[t]]),
                    numeric(length(won_vec)))
win_depth3_ll <- tibble::tibble(
  set = exploded_train_data$exploded_race_id,
  won = won_vec,
  z   = as.vector(Xw %*% beta_win)
) |>
  dplyr::group_by(set) |>
  dplyr::mutate(p = exp(z - max(z)) / sum(exp(z - max(z)))) |>
  dplyr::summarise(ll = log(p[won == 1L]), .groups = "drop") |>
  dplyr::summarise(total = sum(ll)) |>
  dplyr::pull(total)

tibble::tibble(
  Model = c("Win model (depth-1 fit)", "Exploded model (depth-3 fit)"),
  `Depth-1 PL-R² (win)`      = round(c(pl_r2_win, pl_r2_exp), 4),
  `Depth-3 PL logLik (rank)` = round(c(win_depth3_ll,
                                       as.numeric(stats::logLik(model_p2_exploded))), 0)
) |>
  knitr::kable(
    caption = "Each model scored on both objectives, on the training races:
depth-1 win discrimination (PL-R²) and depth-3 Plackett–Luce ranking fit
(log-likelihood over the 15,069 exploded choice sets)."
  )
```

Each model is the maximum-likelihood fit of its own objective — the win
model leads at depth 1, the exploded model at depth 3 — and the two gaps
(≈0.006 in PL-R², ≈176 in ranking log-likelihood) quantify the cost of
optimising for one at the expense of the other.

## Does the exploded fit improve discrimination?

Not on the win, and that is the headline. The exploded model's depth-1
Plackett–Luce R² on the training races is
`r round(pl_r2_exp, 3)`, slightly **below** the win model's
`r round(pl_r2_win, 3)`. This is expected once stated carefully: the
exploded fit maximises the depth-3 ranking likelihood, not the win
likelihood, so it trades a little win-specific fit for fit across the
whole top-3 order. Judged purely on predicting the winner, the
win-optimised model is — by construction — at least as good. What the
explosion buys is **precision**: the form and position coefficients are
estimated far more sharply (the z-values above), because the full
finishing order supplies roughly three times the choice problems from
the same races.

The market comparison is unchanged in character. On calibration and the
P1/P2 rules the exploded model sits with the win model — closer to the
market than paper 1, still short of it. The betting backtest is the one
place the exploded fit edges ahead: a naive ROI of
**`r scales::percent(backtest_naive_p2_exploded$roi, accuracy = 0.1)`**,
against the win model's
`r scales::percent(backtest_naive_p2$roi, accuracy = 0.1)` and paper 1's
−28.2%. It is the least negative result in the series, but it is still a
loss, and the ratio-sweep CI continues to straddle and sit below zero —
so it is an improvement in degree, not a reversal.

The reading consistent with paper 1's market-efficiency framing is that
using the full finishing order sharpens *how well we can estimate* the
effect of each feature, but does not add a feature the starting price
does not already see. Discrimination against the market is bounded by the
information set, not by the precision of the fit, and on that axis the
exploded logit and the extended win model land in the same place. The
remaining lever is to widen the information set itself — race-level
conditions entered through horse-level interactions — which is
@sec-mixed-interactions.
