# Mixed race-level interactions {#sec-mixed-interactions}

The features so far are all horse-level. Race-level features — going,
distance, course — cannot enter the conditional logit directly: they are
constant within a race and cancel in the softmax. They enter only as
**interactions** with horse-level features, whose products vary across
runners within a race (the mixed discrete-choice mechanism derived in
notes/paper2_seed_mixed_choice.md, to be lifted into Appendix A). This
section tests two such interactions on top of the exploded model from
@sec-exploded-logit, which is the baseline throughout (Model E).

```{r}
#| label: p2int-setup
#| include: false
loadNamespace("mlogit")

ed_ct    <- summary(model_p2_ed)$CoefTable     # 4-course block fit (for Lingfield's p)
fin_ct   <- summary(model_p2_final)$CoefTable   # reduced final model (3 courses)
fin_coef <- stats::setNames(fin_ct[, "Estimate"], rownames(fin_ct))
roi_pct  <- function(x) scales::percent(x, accuracy = 0.1)
```

## Two interactions

**Weight × distance** (`rel_weight_x_dist = rel_weight × distance_furlongs`).
`rel_weight` is horse-level, `distance_furlongs` race-level; their product
varies across runners. The motivation is @benter1994: the burden of extra
weight should compound over longer trips, so a horse carrying weight above
the field mean ought to be penalised more in a staying race than a sprint.

**Draw × course.** Draw bias on All-Weather is known to depend on the
track — Kempton, Lingfield, Southwell and Wolverhampton have different
configurations, bend tightness and run-ins. `stall_normalised` had almost
no standalone signal (@sec-eda), but that pooled estimate could be hiding
course-specific effects that cancel. We split it into four course-masked
columns (`stall_normalised` where the race is at that course, else 0),
one coefficient per course.

```{r}
#| label: tbl-interaction-features
# Training set only — see CLAUDE.md exploratory analysis convention
ri_train <- runners_interactions |> dplyr::filter(split == "train")
tibble::tibble(
  Quantity = c("Distance range (furlongs)",
               "rel_weight_x_dist SD",
               "Runners with a Kempton draw term",
               "Runners with a Lingfield draw term",
               "Runners with a Southwell draw term",
               "Runners with a Wolverhampton draw term"),
  Value = c(
    paste(round(min(ri_train$distance_furlongs), 1), "–",
          round(max(ri_train$distance_furlongs), 1)),
    sprintf("%.1f", stats::sd(ri_train$rel_weight_x_dist, na.rm = TRUE)),
    scales::comma(sum(ri_train$stall_x_kempton       > 0, na.rm = TRUE)),
    scales::comma(sum(ri_train$stall_x_lingfield     > 0, na.rm = TRUE)),
    scales::comma(sum(ri_train$stall_x_southwell     > 0, na.rm = TRUE)),
    scales::comma(sum(ri_train$stall_x_wolverhampton > 0, na.rm = TRUE))
  )
) |>
  knitr::kable(caption = "Interaction features on the training split.")
```

## Likelihood-ratio tests

Four models, all fitted on one common exploded sample (k = 3) so the
tests are nested and valid: **E** (baseline), **EW** (+ weight×distance),
**ED** (+ draw×course), **EWD** (+ both). One training race with
draw-less runners is dropped from all four, so Model E here is the
exploded baseline refitted on that shared 5,022-race sample (it differs
from @sec-exploded-logit's fit by that single race; coefficients are
unchanged to three decimals).

```{r}
#| label: tbl-interaction-lr
#| code-fold: true
desc <- c(
  "EW|E"   = "Weight×distance vs baseline",
  "ED|E"   = "Draw×course vs baseline",
  "EWD|E"  = "Both vs baseline",
  "EWD|EW" = "Draw×course | weight×distance",
  "EWD|ED" = "Weight×distance | draw×course"
)
interaction_lr_tests |>
  dplyr::mutate(
    Test       = paste(model_a, "vs", model_b),
    Comparison = desc[paste(model_a, model_b, sep = "|")],
    `LR stat`  = round(lr_stat, 2),
    df         = df,
    `p-value`  = signif(p_value, 3),
    Verdict    = dplyr::if_else(decision == "significant",
                                "adds signal", "no improvement")
  ) |>
  dplyr::select(Test, Comparison, `LR stat`, df, `p-value`, Verdict) |>
  knitr::kable(caption = "Likelihood-ratio tests among the four interaction models.")
```

The verdict is clean. **Draw×course adds signal** (ED vs E: $p \approx 3
\times 10^{-5}$; and it survives conditional on weight, EWD vs EW same
$p$). **Weight×distance does not** — it is indistinguishable from zero
both on its own (EW vs E: $p = 0.88$) and added to draw (EWD vs ED:
$p = 0.91$). The most parsimonious model not rejected at $p < 0.05$ is
therefore **ED** (exploded + draw×course).

Within that four-course draw block, however, the Lingfield term is
individually non-significant (Wald $p = 0.91$ in the ED fit) — Lingfield
shows no draw bias. Since the reduced final model should carry only
individually significant terms, we drop it, and a 1-df likelihood-ratio
test confirms the reduction costs nothing.

```{r}
#| label: tbl-final-reduction
final_reduction_lr |>
  dplyr::transmute(
    Test       = paste(model_a, "vs", model_b),
    `LR stat`  = round(lr_stat, 3),
    df         = df,
    `p-value`  = signif(p_value, 3),
    Verdict    = dplyr::if_else(decision == "significant",
                                "Lingfield retained", "Lingfield dropped (no loss)")
  ) |>
  knitr::kable(caption = "Dropping the non-significant Lingfield draw term: 1-df LR test of the four-course block (ED) against the three-course reduced model.")
```

The **final paper-2 model** is therefore the exploded fit plus
course-specific draw terms for Kempton, Southwell and Wolverhampton —
20 coefficients in all.

## The final model vs the baseline

```{r}
#| label: tbl-ed-coef
#| code-fold: true
base_terms <- names(stats::coef(model_p2_e))
shift <- tibble::tibble(
  Term         = base_terms,
  `E est.`     = round(stats::coef(model_p2_e)[base_terms], 4),
  `Final est.` = round(stats::coef(model_p2_final)[base_terms], 4)
)
knitr::kable(shift, caption = "Base (17-term) coefficients: exploded baseline (E) vs the final model. The draw terms leave the base estimates essentially unchanged.")
```

```{r}
#| label: tbl-draw-course
#| code-fold: true
draw_rows <- grepl("^stall_x_", rownames(fin_ct))
tibble::tibble(
  Course       = c("Kempton", "Southwell", "Wolverhampton"),
  Estimate     = round(fin_ct[draw_rows, "Estimate"], 4),
  `Std. error` = round(fin_ct[draw_rows, "Std. Error"], 4),
  `z`          = round(fin_ct[draw_rows, "z-value"], 2),
  `p-value`    = signif(fin_ct[draw_rows, "Pr(>|z|)"], 3),
  Sig          = dplyr::case_when(
    fin_ct[draw_rows, "Pr(>|z|)"] < 0.001 ~ "***",
    fin_ct[draw_rows, "Pr(>|z|)"] < 0.01  ~ "**",
    fin_ct[draw_rows, "Pr(>|z|)"] < 0.05  ~ "*",
    TRUE                                  ~ ""
  )
) |>
  knitr::kable(caption = "Draw×course coefficients in the final model (Lingfield dropped). The slope is on stall_normalised (draw position / field size), per course.")
```

Adding the interaction barely moves the base coefficients — draw is
roughly orthogonal to form and rating — so the final model inherits the
whole extended-feature story and adds the draw effect on top.

**Which courses show draw bias, and the sign flips that hid it.** The
three retained courses do not agree on direction. Kempton
(`r sprintf("%+.3f", fin_coef[["stall_x_kempton"]])`, $p < 0.01$) and
Southwell (`r sprintf("%+.3f", fin_coef[["stall_x_southwell"]])`,
$p < 0.05$) favour **higher** draws, while Wolverhampton
(`r sprintf("%+.3f", fin_coef[["stall_x_wolverhampton"]])`, $p < 0.001$,
the strongest) favours **lower** draws; Lingfield, dropped above, shows
none. This is exactly why `stall_normalised` looked dead in the
univariate screen: the course effects point in opposite directions and
average out to almost nothing pooled. The interaction recovers a real,
track-specific signal that a single draw coefficient cannot represent —
the clearest vindication in paper 2 of the mixed-logit construction.

**Weight × distance: Benter's compounding not detected.** The
weight×distance term is statistically indistinguishable from zero on this
data. Benter's intuition — that weight bites harder over a longer trip —
may simply be too weak to register in Class 2–5 AW handicaps, where the
handicapper has already set weights to equalise the field and the
distance range is moderate, or it may be that `rel_weight` (a within-race
deviation) already absorbs the level effect with little left for a
distance modulation. Either way, it does not earn a place in the model.

## Out-of-sample diagnostics

We evaluate the final model exactly as before: depth-1 win probabilities
on the test races against the over-round-adjusted market, the same suite
as @sec-extended-win and @sec-exploded-logit.

```{r}
#| label: p2int-eval-setup
#| include: false
# Test set only. Reuse the helper functions from @sec-extended-win.
eval_data_fin <- test_predictions_p2_final |> dplyr::rename(model_prob = predicted_prob)
```

```{r}
#| label: fig-calibration-final
#| fig-cap: "Calibration on the AW test set, final model ED (Owen Figure 7 equivalent): observed win proportion vs predicted / implied win probability, 5-point bins over [0, 0.50]. Circles = final model (depth-1, out-of-sample); plus signs = market; dashed line y = x."
calib_summary_fin <- dplyr::bind_rows(
  calibration_summary(eval_data_fin, "model_prob",  "Model (final, ED)"),
  calibration_summary(eval_data_fin, "market_prob", "Market")
)
ggplot2::ggplot(calib_summary_fin,
                ggplot2::aes(x = bin_mid, y = observed, shape = source, colour = source)) +
  ggplot2::geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey50") +
  ggplot2::geom_point(size = 3) +
  ggplot2::scale_shape_manual(values = c("Model (final, ED)" = 1L, Market = 3L)) +
  ggplot2::scale_colour_manual(values = c("Model (final, ED)" = "purple", Market = "tomato")) +
  ggplot2::scale_x_continuous(labels = scales::percent_format(accuracy = 1),
                              limits = c(0, 0.50), breaks = bin_breaks) +
  ggplot2::scale_y_continuous(labels = scales::percent_format(accuracy = 1), limits = c(0, NA)) +
  ggplot2::labs(x = "Predicted / implied win probability",
                y = "Observed win proportion", shape = NULL, colour = NULL) +
  ggplot2::theme_minimal()
```

```{r}
#| label: fig-p1-p2-final
#| fig-cap:
#|   - "P1 vs probability threshold, final model (Owen Figure 8a equivalent), AW test set."
#|   - "P2 vs probability threshold, final model (Owen Figure 8b equivalent), AW test set."
score_long_fin <- dplyr::bind_rows(
  scoring_summary(eval_data_fin, "model_prob",  "Model (final, ED)", thresholds),
  scoring_summary(eval_data_fin, "market_prob", "Market",            thresholds)
)
p_score_fin <- function(yvar, ylab) {
  ggplot2::ggplot(score_long_fin,
                  ggplot2::aes(x = threshold, y = .data[[yvar]],
                               linetype = source, colour = source)) +
    ggplot2::geom_line(linewidth = 0.8) +
    ggplot2::scale_linetype_manual(values = c("Model (final, ED)" = "solid", Market = "dashed")) +
    ggplot2::scale_colour_manual(values = c("Model (final, ED)" = "purple", Market = "tomato")) +
    ggplot2::scale_x_continuous(labels = scales::percent_format(accuracy = 1), limits = c(0, 0.50)) +
    ggplot2::labs(x = "Probability threshold", y = ylab, linetype = NULL, colour = NULL) +
    ggplot2::theme_minimal()
}
print(p_score_fin("P1", "P1 (geom. mean prob. on winners)"))
print(p_score_fin("P2", "P2 (geom. mean Bernoulli likelihood)"))
```

```{r}
#| label: tbl-backtest-final
backtest_naive_p2_final |>
  dplyr::transmute(
    `Prob threshold`  = prob_threshold,
    `Ratio threshold` = ratio_threshold,
    `Bets`            = n_bets,
    `Wins`            = n_wins,
    `Gross return`    = sprintf("%.2f", gross_return),
    `Profit`          = sprintf("%.2f", profit),
    `ROI`             = roi_pct(roi)
  ) |>
  knitr::kable(caption = "Final model (ED): betting backtest on the AW test races at Owen's exact thresholds.")
```

```{r}
#| label: fig-roi-sweep-final
#| fig-cap: "ROI vs ratio threshold on the AW test set (final model ED), 90% bootstrap CI shaded. Model-probability filter held at 0.13; dashed line at zero marks break-even."
ggplot2::ggplot(backtest_sweep_p2_final, ggplot2::aes(x = tau, y = roi)) +
  ggplot2::geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  ggplot2::geom_ribbon(ggplot2::aes(ymin = ci_lo, ymax = ci_hi), fill = "purple", alpha = 0.15) +
  ggplot2::geom_line(linewidth = 0.8, colour = "purple") +
  ggplot2::scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  ggplot2::labs(x = "Ratio threshold (model_prob / market_prob)", y = "ROI") +
  ggplot2::theme_minimal()
```

## Do the interactions add anything?

Draw×course does; weight×distance does not. The final model's naive
backtest ROI is
**`r roi_pct(backtest_naive_p2_final$roi)`**, against the exploded
baseline's `r roi_pct(backtest_naive_p2_exploded$roi)`, the extended win
model's `r roi_pct(backtest_naive_p2$roi)`, and paper 1's −28.2%. The
series improves monotonically, and the single largest step is this one:
the course-specific draw effect is the most valuable thing paper 2 adds
to Owen's feature set. The reason is intuitive in the market-efficiency
frame — draw bias is a genuine, track-specific edge that a 13-feature
win model simply could not see, and the starting price, built partly on
punters who *do* know each course's draw, was pricing it while the model
was not.

And yet the result is still a loss. −17% ROI is the best in the series
and the ratio-sweep CI narrows, but it does not cross zero: the market
still prices the field more accurately than the model overall. Paper 2's
arc is therefore a clear *narrowing* of the gap to the starting price —
each addition (extended features, the exploded likelihood, course-draw
interactions) recovering a little more of the SP's information set — but
not a closing of it. What remains uncaptured is the part of the market's
edge that lives in information our pre-race feature set still does not
contain: late money, going on the day, and the non-linear, interacting
structure that a linear-predictor conditional logit cannot represent at
all. That last limitation is the explicit subject of paper 3.
