# Discussion {#sec-discussion}

```{r}
#| label: discussion-setup
#| include: false
loadNamespace("mlogit")

# Paper-1 reduced model depth-1 PL-R² (McFadden) from its own fitted
# object — padded probability slots are zero, real ones positive.
paper1_plr2 <- {
  pf  <- fitted_final$probabilities
  nrr <- rowSums(pf > 0)
  1 - as.numeric(stats::logLik(fitted_final)) / (-sum(log(nrr)))
}

fin_ct <- summary(model_p2_final)$CoefTable
roi_pct <- function(x) scales::percent(x, accuracy = 0.1)

# Mean bookmaker over-round across the test races the backtest bets on:
# raw 1/SP sums to > 1 per race; the excess is the bookmaker margin.
overround_mean <- test_predictions_p2_final |>
  dplyr::filter(!is.na(starting_price_decimal), starting_price_decimal > 1) |>
  dplyr::group_by(race_id) |>
  dplyr::summarise(book = sum(1 / starting_price_decimal), .groups = "drop") |>
  dplyr::summarise(m = mean(book) - 1) |>
  dplyr::pull(m)
```

## Model progression

Paper 2 makes three additions to paper 1's model, and each earns its
place. @tbl-progression lines up the four fits on a common footing: the
depth-1 (win) Plackett–Luce $R^2$ on the training races, and the naive
betting ROI and race count on the held-out test set.

```{r}
#| label: tbl-progression
#| code-fold: true
tibble::tibble(
  Model = c("Paper 1 (replication)",
            "Extended win model",
            "Exploded logit",
            "Final (exploded + draw×course)"),
  `PL-R² (depth-1)` = c(
    paper1_plr2,
    model_p2_diagnostics$pl_r2[model_p2_diagnostics$model == "reduced"],
    model_p2_exploded_diagnostics$pl_r2,
    model_p2_final_diagnostics$pl_r2
  ),
  `Naive ROI` = c(backtest_naive$roi, backtest_naive_p2$roi,
                  backtest_naive_p2_exploded$roi, backtest_naive_p2_final$roi),
  `Test races` = c(
    dplyr::n_distinct(test_predictions$race_id),
    dplyr::n_distinct(test_predictions_p2$race_id),
    dplyr::n_distinct(test_predictions_p2_exploded$race_id),
    dplyr::n_distinct(test_predictions_p2_final$race_id)
  )
) |>
  dplyr::mutate(
    `PL-R² (depth-1)` = round(`PL-R² (depth-1)`, 4),
    `Naive ROI`       = roi_pct(`Naive ROI`),
    `Test races`      = scales::comma(`Test races`)
  ) |>
  knitr::kable(caption = "Paper-2 model progression: depth-1 PL-R² on training, naive ROI and test-race count on the held-out set.")
```

The picture is one of steady, genuine improvement. The extended feature
set lifts depth-1 discrimination well above paper 1, the single largest
contributors being `or_relative` — the handicapper's own assessment of
ability, the strongest new predictor in the model — and, for the
backtest, the course-specific draw effect. The exploded fit is the one
step that does *not* raise depth-1 $R^2$: it optimises the full
finishing order rather than the win, so it sharpens the coefficient
estimates (their standard errors fall as the same races contribute three
times the choice problems) without improving win discrimination
specifically. The naive ROI nonetheless climbs at every step, from
−28.2% in paper 1 to `r roi_pct(backtest_naive_p2_final$roi)` in the
final model — a steady narrowing of the gap to the market.

## Why the market still wins

For all that, the final model still loses. Its naive ROI is
`r roi_pct(backtest_naive_p2_final$roi)`, and the ratio-threshold sweep
in @sec-mixed-interactions has a 90% bootstrap interval that does not
cross zero at any threshold. The reading is the one paper 1 reached, only
on a narrower margin: the starting price aggregates information our
pre-race feature set cannot observe. `or_relative` encodes the
handicapper's *historical* assessment of a horse, but not its current
readiness — paddock condition, work on the morning, stable confidence —
nor the late market moves of informed money that the SP absorbs in the
minutes before the off. The favourite–longshot bias documented in paper 1
persists here, and a model built only on form, ratings and draw cannot
price the part of the field's chances that lives in same-day, private
information.

A second qualification concerns the benchmark. The ROI is measured at the
**industry starting price**, which carries a substantial bookmaker
over-round — about `r scales::percent(overround_mean, accuracy = 0.1)` on
these test races — and is therefore the harshest price a bettor could
take. We remove that over-round only where we compare the model to the
market's *probabilities* (the calibration plot and the P1/P2 rules
renormalise 1/SP to sum to one); it is necessarily kept in the ROI,
because a winning bet returns the actual SP, margin and all. The same
selections struck on a betting exchange — where Betfair SP charges only
commission, of order 2–5%, rather than that margin — would return
materially more, though whether enough to reach break-even cannot be
settled without exchange prices, which this dataset does not carry. This
is the benchmark that matters for the live-trading direction noted under
limitations below.

## Draw bias by course

The draw×course result is the clearest illustration in the paper of why
the mixed discrete-choice construction matters. The three retained
courses disagree on direction: Kempton
(`r sprintf("%+.3f", fin_ct["stall_x_kempton", "Estimate"])`,
$p =$ `r signif(fin_ct["stall_x_kempton", "Pr(>|z|)"], 2)`) and Southwell
(`r sprintf("%+.3f", fin_ct["stall_x_southwell", "Estimate"])`,
$p =$ `r signif(fin_ct["stall_x_southwell", "Pr(>|z|)"], 2)`) favour
higher draws, while Wolverhampton
(`r sprintf("%+.3f", fin_ct["stall_x_wolverhampton", "Estimate"])`,
$p =$ `r signif(fin_ct["stall_x_wolverhampton", "Pr(>|z|)"], 2)`) favours
lower draws; Lingfield shows no significant bias. Because the course
effects point in opposite directions, a single pooled draw coefficient
averages them to almost nothing — which is exactly why
`stall_normalised` looked useless in the univariate screen. A feature
that appears dead in isolation carries strong, track-specific signal once
the race-level structure is respected through the interaction. This is
the mixed discrete-choice mechanism (@sec-appx-derivations) doing real
work, not decoration.

## Limitations

- **No live-data test.** The model is fitted and evaluated on the same
  2006–2015 window as paper 1; nothing here tests it on current racing,
  and the AW programme has changed since (new tracks, fixture inflation).
- **Industry-SP benchmark.** Profitability is evaluated at the bookmaker
  starting price, with its full over-round; a betting exchange (Betfair
  SP, commission only) is a more favourable — and, for the live-trading
  direction, more relevant — benchmark. The Smartform data carries only
  industry SP (the morning forecast price and Tote dividends aside), so a
  Betfair-SP backtest is left to future work.
- **Going deferred.** A horse-specific going-affinity feature
  (`going_ordinal × horse_going_sr`) would require a going-filtered
  career-history sub-query, and as a race-level feature it only enters
  the conditional logit through an interaction. It is the natural next
  feature for paper 3, where a tree-based model can take going as a raw
  predictor without the softmax-cancellation constraint that forces the
  interaction here.
- **`or_relative` missing-not-at-random.** The zero-imputation with a
  missing indicator is defensible, but unrated horses remain a distinct
  population rather than average ones; the fitted `or_missing` coefficient
  is negative and significant
  (`r sprintf("%+.3f", fin_ct["or_missing", "Estimate"])`,
  $p < 0.001$), confirming that they underperform even the imputed zero
  and that the indicator is carrying a real effect rather than a nuisance.

## Paper 3 direction

Paper 3 changes the model *class* while holding everything else fixed. It
replaces the conditional-logit linear predictor with a tree-based
ensemble fitted under a Plackett–Luce loss, keeping the same data scope,
the same chronological train/test split, and the same diagnostic suite
(calibration, the P1/P2 scoring rules, and the §3.4-style backtest). The
point of holding all of that constant is to isolate one thing: how much
of the remaining gap to the market is recoverable from model-class
flexibility alone — non-linearities, automatic interactions, and raw
features such as going — as distinct from the feature-set and
likelihood improvements that paper 2 has already banked.
