# Data and race selection {#sec-data}

## Source: the Smartform database

All race data come from the Smartform MySQL database. The two tables
we draw on are `historic_races`, which holds one row per race, and
`historic_runners`, which holds one row per declared runner per race.
A horse is identified by its `runner_id`, which is stable across
races and surfaces, so the same horse's runs at AW and turf courses
can be linked. The full extraction is driven from
`sql/qualifying_races.sql`, `sql/runners_for_races.sql` and
`sql/horse_full_history.sql`; pull-and-clean functions in `R/` lift
the resulting tibbles[^tibble] into the `targets` pipeline.

[^tibble]: A *tibble* is the tidyverse's tidy data frame — strict
    about column types and printing, otherwise behaves like a
    `data.frame`. See @sec-appx-software for the full software stack.

## The British class system, and why we begin in 2006

British flat racing is organised along two independent axes. The
first is a quality tier: Group races, then Listed races (together
making up Class 1), then Classes 2--7 in decreasing order of prize
money and minimum ability. The second is the method by which weights
are assigned: a *handicap* allocates each horse a weight set by the
BHA Handicapper from its Official Rating, while a *conditions* race
uses the race's own eligibility rules to level the field (most
commonly weight-for-age in maidens and novices).

The class scheme used today, with Classes 1--7, was introduced on
1 January 2006. Prior to that, the data carry a two-digit class code
(11, 42, 53, 64, …) whose meaning is not directly comparable. Rather
than recode pre-2006 entries, we simply restrict the window to
1 January 2006 onwards. The upper end of our window, 14 October
2015, is the most recent date for which the database extract we
work with is current.

Within Classes 2--7, the choice of which tier of race to model and
whether to restrict to handicaps or conditions races has a real
bearing on what the model can learn. Class 1 is qualitatively
different — its races are conditions-only and run between the best
horses in training, with the assigned-weights mechanism that
Classes 2--7 share entirely absent. The conditions-versus-handicap
distinction within Classes 2--7 also matters: handicap weights are
set by the same BHA Handicapper, against the same Official Rating,
across all four AW courses and all classes from 2--7, so the
mechanism by which a horse's *ability* gets translated into an
*assigned weight* is consistent throughout the scope we model.

We restrict to Class 2--5 handicaps on two grounds. First, Class
2--5 handicaps account for the overwhelming majority of races at
the four AW courses we model — Classes 6--7 are sparse on the
all-weather, and Class 1 races on these courses are negligible.
Second, restricting to handicaps means every race in our
population shares the same weight-assignment mechanism.

## All-Weather course scope

There are six AW courses in the UK and Ireland. We use four:
Kempton, Lingfield, Southwell and Wolverhampton. The other two are
excluded for date or jurisdiction reasons rather than from any
principled distinction:

- **Chelmsford City** opened in January 2015. Inside our window it
  contributes only nine months of racing — not enough for the
  career-history feature build to behave consistently with the four
  longer-established courses.
- **Newcastle's** AW track came on stream in May 2016, after our
  window ends. There is no Newcastle AW history in scope.
- **Dundalk** is an Irish course, outside the UK scope.

Within the four included courses there is real variation in surface,
configuration and racing style. Lingfield and Kempton race on
Polytrack; Wolverhampton and Southwell both race on Tapeta —
Southwell having switched to Tapeta from Fibresand at the end of
2021, after our window closes, so all Southwell races we use are
Fibresand-era. Lingfield is the fastest of the four — a short,
slightly downhill circuit with a notoriously short run-in.
Wolverhampton is a tight oval well suited to speed and an early
position. Kempton is a right-handed track (the others are
left-handed) with an unusual inner / outer loop arrangement and a
relatively long home straight. Southwell, on Fibresand, played
slower than the other three and traditionally rewarded stayers and
front-runners.

We do *not* control for surface or course in the headline model
specification in @sec-results, because Owen does not control for
them either: in his Turf Flat scope, almost all races share a single
surface family. The fact that our four courses span two surfaces
(Polytrack and Fibresand) is one of several differences from Owen
that we flag and absorb into the residual variation rather than
attempt to model explicitly. Adding a surface or course interaction
is one of the candidate features for paper 2, which extends the
feature set and refits with the exploded conditional logit.

::: {.callout-note title="The AW data scope, in one place"}
- **Source**: Smartform `historic_races` ⋈ `historic_runners`.
- **Race type**: All Weather Flat, filtered on `race_type` rather
  than the unreliable `all_weather` flag (the flag drops ~2,500
  legitimate AW races at our four courses).
- **Courses**: Kempton, Lingfield, Southwell, Wolverhampton.
- **Class**: 2--5.
- **Race kind**: handicaps only (`handicap = 1`); non-maiden.
- **Date window**: 2006-01-01 to 2015-10-14.
- **Field size**: 4--16 declared runners; re-checked as 4+ *actual*
  starters after Non-Runners are removed.
- **Winner**: `coalesce(amended_position, finish_position) == 1`,
  to pick up the seven post-disqualification promotions in scope.
- **Dead heats**: 35 races with two `won = 1` rows are dropped, as
  {mlogit} cannot fit a choice set with two chosen alternatives.
:::

After these filters, the SQL produces 7,507 candidate races. The
R-level cuts (post-Non-Runner field-size and dead-heat removal) take
this down to **`r format(nrow(qualifying_races), big.mark = ",")`
qualifying races** containing **`r format(nrow(qualifying_runners),
big.mark = ",")` runner-race observations**, drawn from
**`r format(length(unique(qualifying_runners$runner_id)),
big.mark = ",")` distinct horses**.

```{r}
#| label: data-counts-table
tibble::tribble(
  ~Quantity,                        ~Value,
  "Qualifying races",               nrow(qualifying_races),
  "Runner-race observations",       nrow(qualifying_runners),
  "Distinct horses",                length(unique(qualifying_runners$runner_id)),
  "Date range (first race)",        as.numeric(min(qualifying_races$meeting_date)),
  "Date range (last race)",         as.numeric(max(qualifying_races$meeting_date))
) |>
  dplyr::mutate(
    Value = dplyr::if_else(
      Quantity |> stringr::str_detect("Date range"),
      format(as.Date(Value, origin = "1970-01-01")),
      format(as.integer(Value), big.mark = ",")
    )
  ) |>
  knitr::kable(
    caption = "Headline counts for the qualifying AW population."
  )
```

The class distribution of the qualifying races is consistent with the
broader picture that AW racing is concentrated in mid-tier handicaps:

```{r}
#| label: class-distribution
qualifying_races |>
  dplyr::count(class, name = "Races") |>
  dplyr::rename(Class = class) |>
  knitr::kable(
    caption = "Number of qualifying races by Class."
  )
```

## Train / test split

For the modelling in @sec-results we split the qualifying races into
a training set and a held-out test set chronologically, by race
date, in a 70 / 30 ratio. The cutoff date is **2012-12-30**,
identified by sorting races by `meeting_date` and locating the 70th
percentile; we freeze the date so the split is reproducible against
an exact calendar boundary irrespective of upstream changes. The
split is on race date, so every runner in a race stays together in
one set — no within-race information leaks across the boundary.

```{r}
#| label: split-counts
tibble::tribble(
  ~Set,        ~Races,                  ~`Runner-rows`,
  "Training",  nrow(races_train),       sum(qualifying_runners$race_id %in% races_train$race_id),
  "Test",      nrow(races_test),        sum(qualifying_runners$race_id %in% races_test$race_id)
) |>
  knitr::kable(
    format.args = list(big.mark = ","),
    caption = "Chronological 70/30 train/test split. Cutoff: 2012-12-30 inclusive in training."
  )
```

## Why we use full cross-surface career history

A choice we make early is whether to compute history-based features
(days since last run, prior finishing positions, sire and trainer
strike rates) from each horse's AW races alone or from its full
career across all surfaces and courses. We do the latter; this
subsection sets out the empirical motivation. It is a digression
from the main thread but worth dwelling on because it affects every
history-based feature in @sec-eda and @sec-results.

### How much of each horse's career is spent at AW courses?

The four AW venues account for a minority of most horses' career
starts.

```{r}
#| label: fig-aw-share
#| fig-cap: "What share of career races were at AW courses?
#|   Histogram over the qualifying horse population. The dashed line is the median."

aw_share <- full_history |>
  dplyr::filter(runner_id %in% unique(qualifying_runners$runner_id)) |>
  dplyr::group_by(runner_id) |>
  dplyr::summarise(
    total_runs = dplyr::n(),
    aw_runs    = sum(course %in% aw_courses),
    pct_aw     = 100 * aw_runs / total_runs,
    .groups    = "drop"
  )

median_pct <- median(aw_share$pct_aw, na.rm = TRUE)

ggplot2::ggplot(aw_share, ggplot2::aes(x = pct_aw)) +
  ggplot2::geom_histogram(binwidth = 5, fill = "steelblue", colour = "white") +
  ggplot2::geom_vline(xintercept = median_pct, linetype = "dashed", colour = "firebrick") +
  ggplot2::annotate(
    "text",
    x     = median_pct + 1,
    y     = Inf,
    label = paste0("Median: ", round(median_pct, 1), "%"),
    hjust = 0, vjust = 1.5, size = 3.5, colour = "firebrick"
  ) +
  ggplot2::labs(
    x = "% of career runs at AW courses",
    y = "Number of horses"
  ) +
  ggplot2::theme_minimal()
```

The median qualifying horse runs only `r round(median_pct, 1)`% of
its career at AW courses. The distribution is broad — some horses
are pure AW specialists (100%) while others run at AW only as
occasional turf-career interludes (the left tail below 10%). A
meaningful fraction of our population has barely any AW history at
all at the moment we observe them in an AW race.

### Where else do AW horses run?

The four AW courses — and Wolverhampton in particular — dominate
the picture by volume, which is the mechanical consequence of
selecting the population on AW participation in the first place.
Beyond the AW set, runs are spread across more than fifty other
courses nationwide; the population has run more or less everywhere
in Britain.

```{r}
#| label: fig-course-map
#| fig-cap: "Total runs at each course across the qualifying horse population.
#|   AW courses highlighted in blue."

course_counts <- full_history |>
  dplyr::count(course, name = "runs") |>
  dplyr::mutate(
    aw_course = course %in% aw_courses,
    course    = forcats::fct_reorder(course, runs)
  )

ggplot2::ggplot(course_counts, ggplot2::aes(x = runs, y = course, fill = aw_course)) +
  ggplot2::geom_col() +
  ggplot2::scale_fill_manual(
    values = c("TRUE" = "steelblue", "FALSE" = "grey70"),
    labels = c("TRUE" = "All Weather", "FALSE" = "Other"),
    name   = "Surface"
  ) +
  ggplot2::labs(
    x = "Number of runs",
    y = "Course"
  ) +
  ggplot2::theme_minimal() +
  ggplot2::theme(axis.text.y = ggplot2::element_text(size = 7))
```

### Would AW-only daysLTO mislead us?

The single most testable consequence of restricting to AW history is
that `daysLTO` — days since the horse's last race — would be
computed against a thinner stream of prior runs. For each qualifying
runner-race we computed `daysLTO` two ways: using only prior AW
races, and using all prior races regardless of surface.

```{r}
#| label: fig-daysLTO-distortion
#| fig-cap: "Difference in days since last run when AW-only history is replaced by full cross-surface history,
#|   per qualifying runner-race. Negative values mean cross-surface history finds a more recent run that AW-only misses."

prior_all <- full_history |>
  dplyr::arrange(runner_id, meeting_date, race_id) |>
  dplyr::group_by(runner_id) |>
  dplyr::mutate(prev_date_all = dplyr::lag(meeting_date)) |>
  dplyr::ungroup() |>
  dplyr::select(runner_id, race_id, prev_date_all)

prior_aw <- full_history |>
  dplyr::filter(course %in% aw_courses) |>
  dplyr::arrange(runner_id, meeting_date, race_id) |>
  dplyr::group_by(runner_id) |>
  dplyr::mutate(prev_date_aw = dplyr::lag(meeting_date)) |>
  dplyr::ungroup() |>
  dplyr::select(runner_id, race_id, prev_date_aw)

daysLTO_distortion <- qualifying_runners |>
  dplyr::inner_join(
    full_history |> dplyr::distinct(runner_id, race_id, meeting_date),
    by = c("runner_id", "race_id")
  ) |>
  dplyr::left_join(prior_all, by = c("runner_id", "race_id")) |>
  dplyr::left_join(prior_aw,  by = c("runner_id", "race_id")) |>
  dplyr::mutate(
    days_lto_all = as.integer(meeting_date - prev_date_all),
    days_lto_aw  = as.integer(meeting_date - prev_date_aw),
    difference   = days_lto_all - days_lto_aw
  )

n_total   <- nrow(daysLTO_distortion)
n_na      <- sum(is.na(daysLTO_distortion$difference))
pct_na    <- round(100 * n_na / n_total, 1)

distortion <- daysLTO_distortion |> dplyr::filter(!is.na(difference))
distortion <- distortion |> dplyr::filter(difference >= -365, difference <= 0)

ggplot2::ggplot(distortion, ggplot2::aes(x = difference)) +
  ggplot2::geom_histogram(binwidth = 7, fill = "steelblue", colour = "white") +
  ggplot2::geom_vline(xintercept = 0, linetype = "dashed", colour = "firebrick") +
  ggplot2::labs(
    subtitle = glue::glue("{pct_na}% of runner-rows have no prior AW history at all and would be NA under AW-only."),
    x = "Difference in days (all surfaces − AW only)",
    y = "Number of runner-race observations"
  ) +
  ggplot2::theme_minimal()
```

The spike at zero is large — for most observations the most recent
prior run was an AW race and the two methods agree. But the left
tail is material, and `r pct_na`% of runner-rows have no prior AW
race at all, so an AW-only `daysLTO` cannot be defined for them.

We therefore build every history-based feature in this paper from
each horse's full cross-surface career history. Restricting to AW
history would discard real form information for most of the
population and would produce missing values at a rate that would
distort the model fit.
