# Exploratory analysis {#sec-eda}

```{r}
#| label: eda-setup
#| include: false
source(file.path(
  rprojroot::find_root(rprojroot::has_file("awracing.Rproj")),
  "papers", "01_replication", "_helpers.R"
))
```

This section reproduces, on the AW population, the kind of
feature-by-feature exploration that motivates Owen's choice of
predictors. For each candidate predictor we plot empirical win
rate against the feature, with bins chosen to give enough mass per
bar that the per-bin standard error remains informative. The
dashed reference line on each plot is the headline win rate over
the data shown.

The headline win rate over the full qualifying population is the
mechanical consequence of average field size: with field sizes
between 4 and 16 runners and exactly one winner per race, the
average win rate per runner-row is the reciprocal of the mean field
size.

```{r}
#| label: headline-win-rate
tibble::tibble(
  `Runner-rows`     = nrow(features),
  `Wins`            = sum(features$won),
  `Mean field size` = sum(features$won) / nrow(features) |> (\(p) 1 / p)(),
  `Win rate`        = scales::percent(mean(features$won), accuracy = 0.01)
) |>
  knitr::kable(digits = 2,
               caption = "Headline counts on the joined features tibble (training + test).")
```

We work off the joined `features` tibble produced by the pipeline.
It carries 13 modelling columns: `age_diff` (Owen's symmetric-
distance age encoding around 3), `sireSR` and `trainerSR` (cumulative
strike rates with strict no-leakage), `days_LTO_log = log(days_LTO +
1)`, the prior-finish lags `position1` / `position2` / `position3`
(each encoded as 0--4 with 0 meaning "no prior run, or finished
worse than 4th"), and the six binary tack / sex indicators
`entire`, `gelding`, `blinkers`, `cheekpieces`, `visor`,
`tonguetie`.

## Age

```{r}
#| label: tbl-age
age_summary <- features |> summarise_win_rate("age")
age_summary |>
  knitr::kable(
    digits  = 4,
    caption = "Win rate by raw age (years), with bin-level standard errors."
  )
```

```{r}
#| label: fig-age
#| fig-cap: "Win rate by raw age (years)."
age_summary |> plot_win_rate()
```

The win-rate-by-age curve on AW data peaks early and declines as
the horses get older. Owen [@owen2019] captures the age effect with
a symmetric-distance encoding, `age_diff = |age - 4.5|`, taking 4.5
as the peak from published evidence on thoroughbred ageing curves.

For AW we depart in two small ways. First, the empirical AW peak
sits noticeably younger than 4.5 — closer to age 3 — reflecting
that AW campaigns are disproportionately run by faster-maturing
horses; we therefore take the peak as 3. Second, the AW window
carries very few starters above age 8, so a raw absolute distance
extrapolates the effect into thinly-supported old-age bins; we cap
the unsigned offset at 5, holding `age_diff` flat above age 8.
Concretely the regression in @sec-results uses
`age_diff = pmin(|age - 3|, 5)`.

## Days since last race

We bin `days_LTO` into intervals matching the rest-effect ranges
Owen discusses. NA values become a `(missing)` bin, which here
captures first-time runners with no prior race to count days from.

```{r}
#| label: tbl-days-lto
days_breaks  <- c(0, 7, 14, 30, 60, 90, 180, 365, Inf)
days_summary <- features |> summarise_win_rate("days_LTO", bins = days_breaks)
days_summary |>
  knitr::kable(
    digits  = 4,
    caption = "Win rate by days since last race, with rest-effect bins."
  )
```

```{r}
#| label: fig-days-lto
#| fig-cap: "Win rate by days since last race, on rest-effect bins."
days_summary |> plot_win_rate()
```

Win rate softens as days since last race grows, with the long-tail
bins (180+ days) sitting noticeably below the headline. Owen reports
a small negative coefficient on raw `daysLTO` (~ −0.004). The
log-transformed version we use (`days_LTO_log = log(days_LTO + 1)`)
re-shapes the same effect on a more even scale; the empirical
relationship visible here is monotonically (and mildly) decreasing.

## Binary tack and sex indicators

The six binary features are a sex pair (`entire`, `gelding`) and
four tack indicators (`blinkers`, `cheekpieces`, `visor`,
`tonguetie`). They are aliased on each other in the obvious way —
fillies and mares are neither `entire` nor `gelding`, and a runner
without a piece of tack has the corresponding indicator 0.

```{r}
#| label: tbl-binaries
binaries         <- c("entire", "gelding", "blinkers", "cheekpieces", "visor", "tonguetie")
binary_summaries <- binaries |>
  purrr::set_names() |>
  purrr::map(\(feat) summarise_win_rate(features, feat))

binary_summaries |>
  dplyr::bind_rows() |>
  knitr::kable(
    digits  = 4,
    caption = "Win rate by binary feature: 0 / 1 split per indicator."
  )
```

```{r}
#| label: fig-binaries
#| fig-cap:
#|   - "Win rate by entire"
#|   - "Win rate by gelding"
#|   - "Win rate by blinkers"
#|   - "Win rate by cheekpieces"
#|   - "Win rate by visor"
#|   - "Win rate by tonguetie"
for (feat in binaries) {
  print(plot_win_rate(binary_summaries[[feat]]))
}
```

The sex pair carries visible gaps between the 0 and 1 bars in both
cases; Owen fits both as positive (+0.49 for `entire`, +0.55 for
`gelding`), reflecting the conditioning effect of sex on weight
allocation in handicaps. Among the four tack indicators, only
`cheekpieces` shows a clear deficit in the indicator-on bar;
`blinkers`, `visor` and `tonguetie` sit close to the headline rate.
Owen reports `cheekpieces` with a negative coefficient (−0.50) and
drops the other three as non-significant; the fit in @sec-results
will say whether the same pattern carries over to AW.

## Prior finishing positions

Owen's `position1`, `position2`, `position3` encode the horse's
finishing position in its most recent, second-most-recent and
third-most-recent prior races. Each is an integer in {0, 1, 2, 3,
4}, where the level corresponds to the actual finishing position
and 0 means "no prior run available, or finished worse than 4th".

```{r}
#| label: tbl-positions
positions          <- c("position1", "position2", "position3")
position_summaries <- positions |>
  purrr::set_names() |>
  purrr::map(\(feat) summarise_win_rate(features, feat))

position_summaries |>
  dplyr::bind_rows() |>
  knitr::kable(
    digits  = 4,
    caption = "Win rate by prior-finish lag (0--4 levels)."
  )
```

```{r}
#| label: fig-positions
#| fig-cap:
#|   - "Win rate by position1 (most recent prior finish)"
#|   - "Win rate by position2 (second-most-recent prior finish)"
#|   - "Win rate by position3 (third-most-recent prior finish)"
for (feat in positions) {
  print(plot_win_rate(position_summaries[[feat]]))
}
```

The pattern Owen reports is that the win-rate effect is largest for
`position1` and weakens as we look further back: `position1` levels
1--4 are positive and declining; `position2` levels 1--4 are smaller
and also declining; `position3` levels are dropped as
non-significant. The AW plots are consistent with this picture —
the lift from being a recent winner is larger than the lift from
being a winner-before-last, and the third-from-last race's
positional bin gaps are small relative to their standard errors.

## Sire and trainer strike rates

`sireSR` and `trainerSR` are cumulative win rates over all races
strictly before the runner's race date. They are continuous in
$[0, 1]$ and uncapped; we bin them with breaks that give finer
resolution at the low end — where most of the mass sits — and a
long tail bin above 0.50.

```{r}
#| label: tbl-sireSR
sr_breaks    <- c(0, 0.05, 0.10, 0.15, 0.20, 0.30, 0.50, 1.0)
sire_summary <- features |> summarise_win_rate("sireSR", bins = sr_breaks)
sire_summary |>
  knitr::kable(
    digits  = 4,
    caption = "Win rate by sire strike rate, with finer bins at the low end."
  )
```

```{r}
#| label: fig-sireSR
#| fig-cap: "Win rate by sire strike rate."
sire_summary |> plot_win_rate()
```

```{r}
#| label: tbl-trainerSR
trainer_summary <- features |> summarise_win_rate("trainerSR", bins = sr_breaks)
trainer_summary |>
  knitr::kable(
    digits  = 4,
    caption = "Win rate by trainer strike rate."
  )
```

```{r}
#| label: fig-trainerSR
#| fig-cap: "Win rate by trainer strike rate."
trainer_summary |> plot_win_rate()
```

Both curves are monotonically increasing in strike rate, which is
the picture Owen reports and on which his positive coefficients
(both ~ +0.05) rest. The two curves run roughly in parallel, with
trainer-rate-derived lift slightly larger than sire-rate-derived
lift in the higher bins — also consistent with Owen's coefficient
magnitudes.

We take these patterns into @sec-results as motivation for the
predictor set rather than as quantitative claims. The conditional-
logit fit is the proper test of which features carry signal once
the others are controlled for.
