# Model results {#sec-results}

This section fits Owen's conditional-logit win model on the AW
training data, compares the resulting coefficients to those Owen
reports in his Table 3 [@owen2019], and then evaluates the model's
calibration and proper scoring on the held-out test races.

The recipe is deliberately the same as Owen's. We fit the full
13-feature specification on the training subset, then — paralleling
the editing step that yields Owen's Table 3 — drop every term whose
$p$-value exceeds 0.05 in the full fit and refit on the same data.
The reduced fit is our headline; we treat it as the AW analogue of
Owen's Table 3, since Owen's Table 3 is itself a reduced model.

The conditional-logit machinery — softmax over the runners in a race,
no intercept, no race-specific covariates — is derived from first
principles in @sec-appx-derivation. In R, the fit is produced by
{mlogit}; the data is reshaped to {mlogit}'s long-form choice format
with `chid.var = "race_id"` and `alt.var = "horse_ref"` (a per-race
runner index in $\{1, \ldots, n_{\text{runners}}\}$) inside the
pipeline target `mlogit_train_data`.

## Specification

The full Owen specification fitted on the AW training races is

\begin{align*}
P(\text{horse } j \text{ wins race } i)
  = \frac{\exp(z_{ij})}{\sum_{m} \exp(z_{im})},
\end{align*}

with

\begin{align*}
z_{ij} =
\beta_1 \,\text{age\_diff}_{ij}
&+ \beta_2 \,\text{sireSR}_{ij}
+ \beta_3 \,\text{trainerSR}_{ij}
+ \beta_4 \,\text{days\_LTO\_log}_{ij} \\
&+ \beta_5 \,\text{position1}_{ij}
+ \beta_6 \,\text{position2}_{ij}
+ \beta_7 \,\text{position3}_{ij} \\
&+ \beta_8 \,\text{entire}_{ij}
+ \beta_9 \,\text{gelding}_{ij}
+ \beta_{10}\,\text{cheekpieces}_{ij} \\
&+ \beta_{11}\,\text{blinkers}_{ij}
+ \beta_{12}\,\text{visor}_{ij}
+ \beta_{13}\,\text{tonguetie}_{ij}.
\end{align*}

The three lag-position terms enter as factor variables with five
levels (0, 1, 2, 3, 4), so each contributes four estimated dummy
coefficients to the fit and twelve in total.

There are two material divergences from Owen's specification.
First, we use $\log(1 + \text{daysLTO})$ rather than raw `daysLTO`;
the substitution is a matter of better-conditioned numerical
scaling across the larger dynamic range produced by horses with
very long gaps between AW runs. Coefficient magnitudes on
`days_LTO_log` and `daysLTO` are therefore not directly comparable,
but signs and significance are. Second, we encode age as
`pmin(|age - 3|, 5)` rather than Owen's `|age - 4.5|`; the AW
empirical peak sits closer to 3 than 4.5, and the cap at distance
5 prevents the effect from extrapolating into the thinly-supported
old-age bins (@sec-eda).

## The full fit and term-by-term significance

```{r}
#| label: tbl-full-fit
#| code-fold: true

full_coefs <- model_diagnostics$coefficients |>
  dplyr::mutate(
    signif = dplyr::case_when(
      is.na(p.value)     ~ "?",
      p.value < 0.001    ~ "***",
      p.value < 0.01     ~ "**",
      p.value < 0.05     ~ "*",
      TRUE               ~ "ns"
    )
  )

full_coefs |>
  dplyr::select(parameter, estimate, std.error, p.value, signif) |>
  knitr::kable(
    digits  = 4,
    caption = "Full Owen specification fitted on the AW training data."
  )
```

```{r}
#| label: tbl-full-fitstats
tibble::tibble(
  Statistic = c("McFadden R²", "Null deviance", "Residual deviance",
                "Runner-rows (training)", "Races (training)"),
  Value     = c(
    sprintf("%.4f", model_diagnostics$fit_stats["McFadden"]),
    sprintf("%.1f", model_diagnostics$fit_stats["null.deviance"]),
    sprintf("%.1f", model_diagnostics$fit_stats["residual.deviance"]),
    format(model_diagnostics$fit_stats["nobs"],   big.mark = ","),
    format(model_diagnostics$fit_stats["nraces"], big.mark = ",")
  )
) |>
  knitr::kable(caption = "Aggregate fit statistics for the full specification.")
```

The runner-row and race counts here are slightly below the
training-split counts reported in @sec-data. We do not impute, and
we do not drop individual runners; instead we drop the **whole
race** if any of its runners has an NA in a modelling column. The
race-level drop is the right granularity here because {mlogit}
requires exactly one chosen alternative per choice set, and
removing a single runner could leave a race with zero or two
recorded winners. The NA cases are almost entirely a
first-time-starter sire (no prior strike rate defined) or a debut
trainer; the small reduction (1.13% of training races, 1.24% of
runner-rows) is the residual cost of the strict-no-leakage rule on
sire and trainer strike rates.

We follow Owen and drop any term whose every level has $p > 0.05$.
For factor terms (the three `position` lags), a term is retained if
*any* of its four dummy levels reaches significance.

```{r}
#| label: drop-terms

term_of <- function(coef_name) {
  if (grepl("^position1[1-4]$", coef_name)) return("position1")
  if (grepl("^position2[1-4]$", coef_name)) return("position2")
  if (grepl("^position3[1-4]$", coef_name)) return("position3")
  coef_name
}

orig_coefs_with_term <- full_coefs |>
  dplyr::mutate(term = vapply(parameter, term_of, character(1)))

term_significance <- orig_coefs_with_term |>
  dplyr::group_by(term) |>
  dplyr::summarise(
    n_levels = dplyr::n(),
    min_p    = suppressWarnings(min(p.value, na.rm = TRUE)),
    any_sig  = any(!is.na(p.value) & p.value < 0.05),
    .groups  = "drop"
  )

kept_terms    <- term_significance$term[ term_significance$any_sig]
dropped_terms <- term_significance$term[!term_significance$any_sig]
```

**Dropped terms** (no level with $p < 0.05$ in the full fit):
`r if (length(dropped_terms) == 0L) "none" else paste(dropped_terms, collapse = ", ")`.

## Reduced model (headline)

```{r}
#| label: fit-reduced

final_formula <- stats::as.formula(
  paste("won ~", paste(kept_terms, collapse = " + "), "| 0 | 0")
)

fitted_final <- mlogit::mlogit(final_formula, data = mlogit_train_data)

final_coef_table <- summary(fitted_final)$CoefTable
final_coefs <- tibble::tibble(
  parameter        = rownames(final_coef_table),
  final_estimate   = final_coef_table[, "Estimate"],
  final_std.error  = final_coef_table[, "Std. Error"],
  final_p.value    = final_coef_table[, "Pr(>|z|)"]
)
```

The reduced model is fitted on the same training data with the
formula automatically constructed from the retained terms. This is
the AW analogue of Owen's Table 3; @tbl-owen-comparison sets it
side-by-side with the values Owen reports.

```{r}
#| label: tbl-owen-comparison
#| code-fold: true

# Owen (2019) Table 3 exact values — the post-pruning (reduced) fit
# on UK Turf Flat 2014–2016. Every entry is significant by
# construction (Table 3 is the reduced specification), so `owen_signif`
# is derived from the sign of `owen_estimate`. Owen's `daysLTO` is on
# the raw scale; we map it onto our `days_LTO_log` row for the join
# and flag the scale difference in the Note column.

owen_table3 <- tibble::tribble(
  ~parameter,         ~owen_estimate, ~owen_se,  ~owen_p,
  "age_diff",         -0.152,         0.0313,    "<0.001",
  "sireSR",            0.048,         0.0093,    "<0.001",
  "trainerSR",         0.051,         0.0093,    "<0.001",
  "days_LTO_log",     -0.004,         0.0018,    "0.021",
  "position11",        0.607,         0.0918,    "<0.001",
  "position12",        0.329,         0.1003,    "<0.001",
  "position13",        0.313,         0.1026,    "<0.001",
  "position14",        0.157,         0.1081,    "<0.001",
  "position21",        0.377,         0.0970,    "<0.001",
  "position22",        0.373,         0.0975,    "<0.001",
  "position23",        0.075,         0.1070,    "<0.001",
  "position24",        0.220,         0.1048,    "<0.001",
  "entire",            0.489,         0.1294,    "<0.001",
  "gelding",           0.548,         0.0946,    "<0.001",
  "cheekpieces",      -0.503,         0.1454,    "0.001"
) |>
  dplyr::mutate(
    owen_signif = dplyr::if_else(owen_estimate > 0, "+", "-")
  )

aw_coefs <- final_coefs |>
  dplyr::mutate(
    aw_signif = dplyr::case_when(
      is.na(final_p.value) ~ "?",
      final_p.value < 0.001 ~ "***",
      final_p.value < 0.01  ~ "**",
      final_p.value < 0.05  ~ "*",
      TRUE                  ~ "ns"
    ),
    aw_direction = dplyr::case_when(
      aw_signif == "ns"     ~ "ns",
      final_estimate > 0    ~ "+",
      final_estimate < 0    ~ "-",
      TRUE                  ~ "0"
    )
  )

comparison <- owen_table3 |>
  dplyr::full_join(aw_coefs, by = "parameter") |>
  dplyr::mutate(
    difference = final_estimate - owen_estimate,
    note = dplyr::case_when(
      is.na(owen_estimate) & (is.na(final_estimate) | aw_signif == "ns") ~ "Both: NS / dropped",
      is.na(owen_estimate) & aw_signif != "ns"                            ~ "AW retains; Owen dropped",
      is.na(final_estimate)                                                ~ "AW dropped; Owen retains",
      owen_signif == aw_direction                                          ~ "Same direction",
      TRUE                                                                 ~ "Direction differs"
    ),
    note = dplyr::case_when(
      parameter == "days_LTO_log" ~
        paste0(note, " (AW: log(1 + daysLTO); Owen: raw daysLTO; scales differ)"),
      TRUE ~ note
    )
  ) |>
  dplyr::select(
    Parameter        = parameter,
    `Owen est.`      = owen_estimate,
    `Owen SE`        = owen_se,
    `Owen p`         = owen_p,
    `AW est.`        = final_estimate,
    `AW SE`          = final_std.error,
    `AW sig.`        = aw_signif,
    `AW − Owen`      = difference,
    Note             = note
  ) |>
  dplyr::arrange(
    factor(
      Parameter,
      levels = c(
        "age_diff",
        "sireSR", "trainerSR", "days_LTO_log",
        "position11", "position12", "position13", "position14",
        "position21", "position22", "position23", "position24",
        "position31", "position32", "position33", "position34",
        "entire", "gelding",
        "cheekpieces", "blinkers", "visor", "tonguetie"
      )
    )
  )

knitr::kable(
  comparison,
  digits  = 4,
  align   = c("l", "r", "r", "c", "r", "r", "c", "r", "l"),
  caption = "AW reduced model (training fit) vs Owen (2019) Table 3 (UK Turf Flat 2014–2016).
Two rows are not directly comparable on absolute magnitude: `days_LTO_log` (AW uses log; Owen uses raw `daysLTO`), and `sireSR` / `trainerSR` (Owen appears to report on a percentage scale, our coefficients are on a proportion scale — divide AW by 100 for a like-for-like comparison; see text)."
)
```

The two strike-rate rows in @tbl-owen-comparison deserve a moment.
Both `sireSR` and `trainerSR` are *defined* identically in our
pipeline and in Owen's — wins divided by races at the latest
meeting strictly before the runner's race date, returned as a
proportion in $[0, 1]$ (median around 0.10 in both our distribution
and his). The fitted AW coefficients (+2.355 for sireSR, +5.094 for
trainerSR) are roughly 50× and 100× larger than Owen's (+0.048,
+0.051). A direct read of the table would suggest a wildly stronger
strike-rate signal on AW than on Turf. The more likely
interpretation is a reporting-scale mismatch: Owen's coefficients
behave numerically as though his rates were on the percent scale
(0--100), while ours are on the proportion scale (0--1). Dividing
our coefficients by 100 puts them on the same footing:

- `trainerSR` — AW 0.0509 vs Owen 0.0510. A near-exact match.
- `sireSR` — AW 0.0235 vs Owen 0.0480. AW is about half Owen's.

The trainerSR agreement is striking. The sireSR halving is
substantively interesting: AW handicap horses are older and more
mature than the predominantly 3-year-old Turf flat population Owen
fits, and the inherited speed-and-stamina signal a sire carries is
plausibly weaker once horses have had more time to express their
own ability. Both observations would survive a refit with rescaled
predictors; we present them here as scale-corrected magnitudes
rather than refitting, since the qualitative direction of every
other comparison in the table is unaffected.

```{r}
#| label: tbl-fitstats-comparison

summarise_fit <- function(fitted, label) {
  # Null log-likelihood from choice-set sizes (robust to mlogit's
  # `summary()$lratio` not always being populated). Matches the
  # formula in `extract_model_diagnostics()` for consistency: padded
  # slots in `fitted$probabilities` are zero, real slots strictly
  # positive, so `> 0` selects actual alternatives.
  probs_mat          <- fitted$probabilities
  n_runners_per_race <- rowSums(probs_mat > 0)
  ll      <- as.numeric(stats::logLik(fitted))
  ll_null <- -sum(log(n_runners_per_race))
  tibble::tibble(
    Model           = label,
    `n params`      = length(stats::coef(fitted)),
    `logLik`        = ll,
    `McFadden R²`   = 1 - ll / ll_null,
    `Null dev.`     = -2 * ll_null,
    `Resid. dev.`   = -2 * ll
  )
}

dplyr::bind_rows(
  summarise_fit(fitted_mlogit, "Full"),
  summarise_fit(fitted_final,  "Reduced")
) |>
  knitr::kable(
    digits  = 4,
    caption = "Full vs reduced specification: aggregate fit statistics on the training data."
  )
```

The qualitative picture is largely consistent with Owen's: `age_diff`
attracts a significant negative coefficient (peak-ability encoding,
declining symmetrically away from peak); both strike-rate measures
sit positive; the `position1` and `position2` factors retain
significance on at least one level apiece, while `position3` is
dropped in both fits; `entire` and `gelding` come in positive and
`cheekpieces` negative; the remaining tack indicators are
non-significant. Magnitude differences between the two fits show on
the comparison table above; specific patterns of agreement and
drift are worth reading line-by-line rather than summarising
prematurely, given that the AW and Turf populations differ in
surface, window length and race-type scope.

## Out-of-sample evaluation: calibration

Owen's Figure 7 [@owen2019, §3.3] plots observed win proportion
against fitted (or implied) win probability, binned in 5-point
intervals across $[0, 0.50]$. Perfect calibration sits on the
$y = x$ line; points above indicate under-prediction, below
indicate over-prediction.

We compute model probabilities on the **test races** using the
reduced fit — i.e. these are out-of-sample forecasts, not in-sample
fits. Market-implied probabilities are constructed from the
runner's starting price and renormalised per race so the
probabilities in each race sum to one.

```{r}
#| label: build-probability-table

probs_mat_test <- stats::predict(fitted_final, newdata = mlogit_test_data)

model_probs_long <- probs_mat_test |>
  as.data.frame() |>
  tibble::rownames_to_column("race_id") |>
  tidyr::pivot_longer(
    cols      = -race_id,
    names_to  = "horse_ref",
    values_to = "model_prob"
  ) |>
  dplyr::filter(!is.na(model_prob), model_prob > 0) |>
  dplyr::mutate(
    race_id   = as.integer(race_id),
    horse_ref = as.integer(horse_ref)
  )

mldat_df <- as.data.frame(mlogit_test_data)
class(mldat_df) <- "data.frame"
attr(mldat_df, "index")    <- NULL
attr(mldat_df, "clseries") <- NULL

calib_base <- mldat_df |>
  dplyr::select(race_id, runner_id, horse_ref, won) |>
  dplyr::mutate(
    race_id   = as.integer(race_id),
    horse_ref = as.integer(horse_ref),
    won       = as.integer(won)
  ) |>
  dplyr::inner_join(model_probs_long, by = c("race_id", "horse_ref"))

market_probs <- qualifying_runners |>
  dplyr::select(race_id, runner_id, starting_price_decimal) |>
  dplyr::filter(!is.na(starting_price_decimal),
                starting_price_decimal > 1) |>
  dplyr::mutate(implied_raw = 1 / starting_price_decimal) |>
  dplyr::group_by(race_id) |>
  dplyr::mutate(market_prob = implied_raw / sum(implied_raw)) |>
  dplyr::ungroup() |>
  dplyr::select(race_id, runner_id, market_prob)

calib_data <- calib_base |>
  dplyr::left_join(market_probs, by = c("race_id", "runner_id"))
```

```{r}
#| label: fig-calibration
#| fig-cap: "Figure 7 (Owen 2019, §3.3, replicated on AW test data): observed
#|   win proportion vs predicted / implied win probability, binned in 5-point
#|   intervals over [0, 0.50]. Circles = reduced model out-of-sample; plus signs
#|   = market. Reference line y = x is perfect calibration."

bin_breaks <- seq(0, 0.50, by = 0.05)
bin_mids   <- head(bin_breaks, -1) + 0.025

calibration_summary <- function(df, prob_col, label) {
  df |>
    dplyr::filter(!is.na(.data[[prob_col]]),
                  .data[[prob_col]] >= 0,
                  .data[[prob_col]] <= 0.50) |>
    dplyr::mutate(
      bin = cut(.data[[prob_col]],
                breaks = bin_breaks,
                include.lowest = TRUE,
                right = FALSE)
    ) |>
    dplyr::group_by(bin) |>
    dplyr::summarise(
      n            = dplyr::n(),
      observed     = mean(won),
      predicted_mu = mean(.data[[prob_col]]),
      .groups      = "drop"
    ) |>
    dplyr::mutate(
      bin_mid = bin_mids[as.integer(bin)],
      source  = label
    )
}

calib_summary <- dplyr::bind_rows(
  calibration_summary(calib_data, "model_prob",  "Model (reduced)"),
  calibration_summary(calib_data, "market_prob", "Market")
)

ggplot2::ggplot(
  calib_summary,
  ggplot2::aes(x = bin_mid, y = observed, shape = source, colour = source)
) +
  ggplot2::geom_abline(slope = 1, intercept = 0,
                       linetype = "dashed", colour = "grey50") +
  ggplot2::geom_point(size = 3) +
  ggplot2::scale_shape_manual(values = c("Model (reduced)" = 1L, Market = 3L)) +
  ggplot2::scale_colour_manual(values = c("Model (reduced)" = "steelblue", Market = "tomato")) +
  ggplot2::scale_x_continuous(labels = scales::percent_format(accuracy = 1),
                              limits = c(0, 0.50), breaks = bin_breaks) +
  ggplot2::scale_y_continuous(labels = scales::percent_format(accuracy = 1),
                              limits = c(0, NA)) +
  ggplot2::labs(
    x      = "Predicted / implied win probability",
    y      = "Observed win proportion",
    shape  = NULL, colour = NULL
  ) +
  ggplot2::theme_minimal()
```

## Out-of-sample evaluation: P1 and P2 scoring rules

Owen's Figure 8 plots two scoring rules against a probability
threshold. One caveat to flag up front before any comparison
with Owen's headline numbers: Owen reports P1 and P2 *in-sample*
on the training set as a goodness-of-fit diagnostic, while the
values we compute below are *out-of-sample* on the held-out test
races. Out-of-sample scores will, on average, trail in-sample
scores for any non-trivial fit, so reading directly from his
quoted P1 / P2 magnitudes to ours overstates any apparent gap.
We return to this in @sec-future.

For each threshold $t$, we take the subset of runner-rows whose
respective probability exceeds $t$, then compute

- **P1** (geometric mean of probability assigned to winners):
  $$P_1(t) = \exp\!\left(\frac{1}{N_w}\sum_{i \in \text{winners}} \log p_{ij}\right).$$
- **P2** (geometric mean of per-row Bernoulli likelihood):
  $$P_2(t) = \exp\!\left(\frac{1}{N}\!\left[\sum_{\text{winners}}\log p + \sum_{\text{losers}} \log(1 - p)\right]\right).$$

P1 rewards high probability on actual winners; P2 also penalises
high probability on losers (proper scoring rule). Higher is better
for both.

```{r}
#| label: scoring-functions

add_loglik <- function(df, prob_col) {
  df |>
    dplyr::filter(!is.na(.data[[prob_col]]),
                  .data[[prob_col]] > 0,
                  .data[[prob_col]] < 1) |>
    dplyr::mutate(
      ll_winner = ifelse(won == 1L, log(.data[[prob_col]]),         NA_real_),
      ll_full   = ifelse(won == 1L, log(.data[[prob_col]]),
                                    log(1 - .data[[prob_col]]))
    )
}

score_at_threshold <- function(df, prob_col, threshold) {
  sub <- df[df[[prob_col]] > threshold, , drop = FALSE]
  if (nrow(sub) == 0L) {
    return(tibble::tibble(threshold = threshold, n = 0L, n_winners = 0L,
                          P1 = NA_real_, P2 = NA_real_))
  }
  winners <- sub[sub$won == 1L, , drop = FALSE]
  p1 <- if (nrow(winners) > 0L) exp(mean(winners$ll_winner, na.rm = TRUE)) else NA_real_
  p2 <- exp(mean(sub$ll_full, na.rm = TRUE))
  tibble::tibble(threshold = threshold, n = nrow(sub), n_winners = nrow(winners),
                 P1 = p1, P2 = p2)
}

scoring_summary <- function(df, prob_col, label, thresholds) {
  scored_df <- add_loglik(df, prob_col)
  purrr::map_dfr(thresholds, \(t) score_at_threshold(scored_df, prob_col, t)) |>
    dplyr::mutate(source = label)
}

thresholds <- seq(0, 0.50, by = 0.025)
score_long <- dplyr::bind_rows(
  scoring_summary(calib_data, "model_prob",  "Model (reduced)", thresholds),
  scoring_summary(calib_data, "market_prob", "Market",          thresholds)
)
```

```{r}
#| label: fig-P1
#| fig-cap: "Figure 8(a) (Owen 2019, §3.3, replicated on AW test data): P1 against
#|   probability threshold for the reduced model (solid) and the market (dashed).
#|   Higher = more probability concentrated on eventual winners."

ggplot2::ggplot(
  score_long,
  ggplot2::aes(x = threshold, y = P1, linetype = source, colour = source)
) +
  ggplot2::geom_line(linewidth = 0.8) +
  ggplot2::scale_linetype_manual(values = c("Model (reduced)" = "solid", Market = "dashed")) +
  ggplot2::scale_colour_manual(values = c("Model (reduced)" = "steelblue", Market = "tomato")) +
  ggplot2::scale_x_continuous(labels = scales::percent_format(accuracy = 1), limits = c(0, 0.50)) +
  ggplot2::labs(
    x        = "Probability threshold",
    y        = "P1 (geometric mean prob. on winners)",
    linetype = NULL, colour = NULL
  ) +
  ggplot2::theme_minimal()
```

```{r}
#| label: fig-P2
#| fig-cap: "Figure 8(b) (Owen 2019, §3.3, replicated on AW test data): P2 against
#|   probability threshold. Higher = better calibration plus better discrimination
#|   (proper scoring rule)."

ggplot2::ggplot(
  score_long,
  ggplot2::aes(x = threshold, y = P2, linetype = source, colour = source)
) +
  ggplot2::geom_line(linewidth = 0.8) +
  ggplot2::scale_linetype_manual(values = c("Model (reduced)" = "solid", Market = "dashed")) +
  ggplot2::scale_colour_manual(values = c("Model (reduced)" = "steelblue", Market = "tomato")) +
  ggplot2::scale_x_continuous(labels = scales::percent_format(accuracy = 1), limits = c(0, 0.50)) +
  ggplot2::labs(
    x        = "Probability threshold",
    y        = "P2 (geometric mean Bernoulli likelihood)",
    linetype = NULL, colour = NULL
  ) +
  ggplot2::theme_minimal()
```

```{r}
#| label: tbl-threshold-counts
#| code-fold: true
score_long |>
  dplyr::select(threshold, source, n, n_winners, P1, P2) |>
  tidyr::pivot_wider(names_from = source, values_from = c(n, n_winners, P1, P2)) |>
  dplyr::arrange(threshold) |>
  knitr::kable(
    digits  = 4,
    caption = "Subset sizes and scoring values at each threshold for the reduced model and the market.
As the threshold rises, fewer rows qualify; P1 and P2 estimates become noisier."
  )
```

## Discussion

The coefficient picture lands close to where Owen's recipe predicts.
Signs agree across the great majority of retained features, and the
features Owen drops as non-significant — `position3` and three of
the four tack indicators — drop out at p < 0.05 in his fit. On AW
some of those terms (`blinkers`, `tonguetie`, the `position3`
factor) do reach significance in our reduced fit; on roughly ten
times as many training races, smaller true effects become
statistically detectable, so this is a sample-size phenomenon rather
than a substantive AW / Turf difference. Two more substantive
magnitude differences are worth singling out from
@tbl-owen-comparison. First, the `gelding` coefficient is roughly
half Owen's (+0.28 vs +0.55). The AW Class 2--5 population is 65%
geldings — more than the broader Turf flat — so the contrast
between geldings and the non-gelding reference is mechanically
compressed on AW, and the coefficient with it. Second, the
strike-rate finding flagged in @sec-results: after the suspected
percent-vs-proportion scale correction, AW trainerSR matches Owen's
almost exactly while AW sireSR is about half. The trainer effect
appears to be a stable feature of British handicap racing
irrespective of surface; the sire effect is plausibly weaker on AW
because horses are older and more of their ability is visible in
their own record rather than inherited.

The probabilistic diagnostics are where this replication is
visibly less encouraging than Owen's. On Turf, Owen reports a
model that tracks the market quite tightly on calibration and runs
close to the market on both scoring rules across the threshold
range, with the market modestly sharper. On AW, the reduced model
underperforms the market on P1 and P2 across most of the threshold
range visible in @fig-P1 and @fig-P2, with the gap widening as the
threshold rises (i.e. as we restrict to higher-confidence picks).
Calibration is in the right ballpark but visibly noisier than
Owen's, with bin-level deviations from the $y = x$ line that
exceed sampling error in places. The forward-looking backtest in
@sec-future is consistent with this picture: Owen's bet-selection
rule ($\hat{P} > 0.15$, ratio $> 1.3$), which he reports yielding
+20.5% ROI on his Turf test set, does not produce a positive ROI
on the AW test set, and the ROI-vs-ratio sweep (@fig-roi-sweep)
has a 90% bootstrap CI that straddles zero across the full
threshold range.

The most plausible reading is that the conditional-logit
specification — Owen's, ours, or any thirteen-feature variant — is
asked to compete against a starting price that already aggregates
much richer information. The SP encodes weight carried, draw,
going, distance suitability, jockey, recent workout form, course
form, and the collective judgment of informed money. Our thirteen
features capture none of that directly. The surprise is not that
the model trails the market; it would be remarkable if it did not.
What is genuinely interesting in Owen's paper is that on a small
test set (264 bets, 512 races) his model nonetheless generated a
positive ROI of the order of 20%. With a sample that size, the
sampling distribution of ROI is wide enough that the 90% CI
plausibly includes zero, and Owen is careful to hedge accordingly.
Our headline ROI is materially negative across 1,713 bets and 2,224
races, with CIs comfortably below zero — a more confident *null* on
a much larger sample, rather than a contradiction of a confident
*positive*.

Paper 2 picks this up with three specific extensions to the
predictor set, all of which the database supplies but neither this
paper nor Owen's uses. **Draw** (stall number) is known to matter
on AW and varies with course and distance — it is a horse-race
pair feature and belongs in the next specification. **Weight
carried** (or each horse's deviation from the race's weight range)
is a direct readout of the handicapper's assessment of relative
ability within the race, structurally the most natural predictor
for a handicap and absent from Owen's feature list. **Going** and
**race class** become available as race-level features through the
interaction-with-horse-level trick saved for the paper 2 appendix.
The same paper also moves to the exploded conditional logit, so
the full finishing order contributes to the fit. Paper 3 then
leaves conditional logit behind for a Plackett--Luce R² tree.
