# Future predictive performance {#sec-future}

```{r}
#| label: future-setup
#| include: false
tar_load(c(
  test_predictions,
  model_market_ratio,
  backtest_naive,
  backtest_sweep
))
```

Owen's §3.4 deploys the Table 3 reduced model to the held-out test
set and evaluates a unit-stake betting strategy as a forward-looking
test of the model's edge against the market. We do the same on the
AW test races (the 30% chronological hold-out after 2012-12-30:
`r format(length(unique(test_predictions$race_id)), big.mark = ",")`
races,
`r format(nrow(test_predictions), big.mark = ",")` runner-rows).

Given the scoring picture established in @sec-results — the model
trails the market on both P1 and P2 across most of the threshold
range — we approach this analysis with low expectations. A model
whose fitted probabilities sit below the market's on actual winners
will not, in general, identify positive-EV bets against market odds.
The analysis below confirms this.

## Market probability construction

For each test race $k$, the over-round-adjusted market-implied win
probability of horse $j$ is constructed from the starting price by
renormalising the raw implied probabilities to sum to one over the
runners in that race:

$$
P^{\text{mkt}}_{jk} \;=\;
  \frac{1 / \text{SP}_{jk}}{\sum_{m \in \text{race}_k} 1 / \text{SP}_{m}},
$$

where $\text{SP}_{jk}$ is the runner's starting price in decimal
odds. Runners with missing SP are excluded from the bet-selection
step (their `market_prob` is NA).

The model/market probability ratio is then
$\text{ratio}_{jk} = \hat{P}_{jk} / P^{\text{mkt}}_{jk}$: ratios
above 1 are runners the model rates higher than the market; ratios
below 1 are runners the model rates lower.

## Backtest at Owen's thresholds

Owen's bet-selection rule places a unit stake on horse $j$ in race
$k$ when both of the following hold:

$$
\hat{P}_{jk} > 0.15 \qquad\text{and}\qquad
\text{ratio}_{jk} > 1.3.
$$

On Owen's Turf test set this rule places 264 bets and returns +54
units, an ROI of approximately +20.5%. Applied to our AW test set
with the reduced model fit, the same rule gives:

```{r}
#| label: tbl-backtest-naive
backtest_naive |>
  dplyr::transmute(
    `Prob threshold`  = prob_threshold,
    `Ratio threshold` = ratio_threshold,
    `Bets`            = n_bets,
    `Wins`            = n_wins,
    `Gross return`    = sprintf("%.2f", gross_return),
    `Profit`          = sprintf("%.2f", profit),
    `ROI`             = scales::percent(roi, accuracy = 0.1)
  ) |>
  knitr::kable(
    caption = "AW reduced model: betting backtest on the held-out test races at Owen's exact thresholds."
  )
```

The headline AW ROI is
**`r scales::percent(backtest_naive$roi, accuracy = 0.1)`** on
`r format(backtest_naive$n_bets, big.mark = ",")` bets, against
Owen's reported +20.5% on 264 bets. The gap is exactly the picture
the scoring-rule plots in @fig-P1 and @fig-P2 predict: a model that
assigns less probability to winners than the market does will not,
on average, identify market-beating bets at any sensible threshold.

## ROI across ratio thresholds

To check whether the headline result above is sensitive to the
exact 1.3 ratio threshold Owen uses, we sweep the ratio threshold
$\tau$ from 0.9 to 2.0 in steps of 0.05, hold the model-probability
threshold fixed at 0.13 (chosen on the AW scoring picture rather
than copied from Owen's 0.15), and compute point ROI plus a 90%
bootstrap confidence interval at each $\tau$. The bootstrap
resamples **races** (not runner-rows) with replacement, $B = 2000$
draws, seed fixed at 42 for reproducibility. Race-level
resampling preserves the one-winner-per-race structure of the
choice sets.

```{r}
#| label: fig-roi-sweep
#| fig-cap: "ROI vs ratio threshold on the AW test set, with 90% bootstrap CI (shaded). Model-probability filter held at 0.13. Dashed horizontal line at zero marks break-even."

ggplot2::ggplot(backtest_sweep, ggplot2::aes(x = tau, y = roi)) +
  ggplot2::geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  ggplot2::geom_ribbon(
    ggplot2::aes(ymin = ci_lo, ymax = ci_hi),
    fill  = "steelblue", alpha = 0.20
  ) +
  ggplot2::geom_line(linewidth = 0.8, colour = "steelblue") +
  ggplot2::scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  ggplot2::labs(
    x = "Ratio threshold (model_prob / market_prob)",
    y = "ROI"
  ) +
  ggplot2::theme_minimal()
```

```{r}
#| label: tbl-roi-sweep
backtest_sweep |>
  dplyr::transmute(
    `Ratio threshold` = tau,
    `Bets`            = n_bets,
    `Wins`            = n_wins,
    `ROI`             = scales::percent(roi, accuracy = 0.1),
    `90% CI low`      = scales::percent(ci_lo, accuracy = 0.1),
    `90% CI high`     = scales::percent(ci_hi, accuracy = 0.1)
  ) |>
  knitr::kable(
    caption = "ROI sweep over ratio thresholds with 90% bootstrap CIs (B = 2000, race-level resampling, seed 42)."
  )
```

The 90% bootstrap CI straddles zero across essentially the full
threshold range. There is no $\tau$ at which the lower bound
remains above zero, so we have no statistical evidence that any
filtered subset of the AW model's signal is positive-EV against
the market. This is the natural consequence of the P1 / P2 picture
in @sec-results.

## Owen's training-set vs our test-set scoring numbers

A caveat worth flagging on direct comparison with Owen's headline
numbers. Owen's reported P1 / P2 values (P1_model = 0.120,
P1_market = 0.136, P2_model = 0.860, P2_market = 0.831 in his
§3.3) are computed on the **training** set as goodness-of-fit
diagnostics. Our P1 / P2 in @sec-results are evaluated on the held-
out **test** set as out-of-sample forecasts. The two sets of
numbers are not directly comparable — out-of-sample scores will
typically trail in-sample for any non-trivial fit.

The §3.4 ROI comparison above, however, is like-for-like: Owen's
+20.5% and our AW ROI here are both evaluated out-of-sample on the
respective held-out test races.

## Closing remark

The AW backtest confirms the negative result the scoring rules
already implied. The interesting question is not *whether* the
model trails the market — given thirteen features against an SP
that aggregates weight, draw, jockey, going, distance form, recent
workouts and informed money, that direction is set in advance —
but *what is missing from the model* that lets the market lead.
Each of the candidates listed in @sec-results (draw, weight
carried, going, race class) is a property of horse-race pairs that
the SP already prices, and each is absent from Owen's specification
and ours. Paper 2 adds them, and reframes the comparison as a test
of how much of the market's edge is recoverable when the feature
set is widened to overlap meaningfully with the information set
that the SP itself sits on. That paper also moves to the exploded
conditional logit, so the full finishing order — not just the win —
contributes to the fit. Whether either change is enough to close
the gap is the question paper 2 will answer.
