# The test split, scored once {#sec-test}

```{r}
#| label: test-numbers
#| include: false

t_inc_win <- p6_res(p6_test_results, "incumbent", "win")
t_mkt_win <- p6_res(p6_test_results, "market-only", "win")
t_dec_win <- p6_res(p6_test_results, "^declared", "win")
t_inc_plc <- p6_res(p6_test_results, "incumbent", "place")
t_mkt_plc <- p6_res(p6_test_results, "market-only", "place")
t_dec_plc <- p6_res(p6_test_results, "^declared", "place")
t_inc_ewc <- p6_res(p6_test_results, "incumbent", "eachway_corrected")
t_mkt_ewc <- p6_res(p6_test_results, "market-only", "eachway_corrected")
t_dec_ewc <- p6_res(p6_test_results, "^declared", "eachway_corrected")
t_inc_ew5 <- p6_res(p6_test_results, "incumbent", "eachway")
t_mkt_ew5 <- p6_res(p6_test_results, "market-only", "eachway")

c_win_s1 <- p6_con(p6_test_contrasts, "win", "incumbent")
c_win_s4 <- p6_con(p6_test_contrasts, "win", "market-only")
c_plc_s1 <- p6_con(p6_test_contrasts, "place", "incumbent")
c_plc_s4 <- p6_con(p6_test_contrasts, "place", "market-only")
c_ewc_s1 <- p6_con(p6_test_contrasts, "eachway_corrected", "incumbent")
c_ewc_s4 <- p6_con(p6_test_contrasts, "eachway_corrected", "market-only")

sels <- p6_declared_selection_sizes
kelly_tr <- p6_kelly_edge_share$share_positive_edge[
  p6_kelly_edge_share$split == "train"]
kelly_te <- p6_kelly_edge_share$share_positive_edge[
  p6_kelly_edge_share$split == "test"]
drop_move <- abs(p6_test_results$roi_drop_top1 - p6_test_results$roi)
```

Five rules reached the test split and no others: the three declared, S1/K0
(paper 5's incumbent, the comparator for every headline number) and S4/K0
(the market-only control). Each was read on all four settlement columns. That
is the entire test contact.

## Results

```{r}
#| label: tbl-test
#| tbl-cap: "The test split. `ROI, fair book` prices the same bets at starting prices renormalised within race to sum to one; `ROI less top winner` drops the single largest return. Intervals are 90% race-level bootstrap, B = 2000, seed 42."
p6_results_table(p6_test_results) |> kable(align = "llrrrrrrrl")
```

Every arm loses money on every market, with one qualification: nothing here
is distinguishable from a small loss on the win market for the incumbent,
whose interval `r p6_fmt_pct_ci(t_inc_win$ci_lo, t_inc_win$ci_hi)` reaches
just above zero.

The declared rules are thin. The declared win rule, S3b/K2, selects
`r p6_comma(sels$n_races_selected[sels$bet == "win"])` of the
`r p6_comma(n_test_races)` test races but stakes a positive amount in only
`r t_dec_win$n_bets`, totalling `r sprintf("%.0f", t_dec_win$total_stake)`
units. The declared each-way rule, S4/K2, selects every race and stakes in
`r t_dec_ewc$n_bets`. This is the staking rules selecting: K1 and K2 both
stake zero where the model's probability does not exceed the price, and out
of sample it rarely does — the model's top-rated horse carries a positive
Kelly edge in `r p6_fmt_pct(kelly_tr, 1)` of training races but only
`r p6_fmt_pct(kelly_te, 1)` of test races, and inside the short-priced band
S3b selects, less again.

Whether the thinness comes from S3b or from K2 cannot be settled here. It
would take scoring S3b/K0 on the test split, and the design forbids a second
contact.

## The contrasts

```{r}
#| label: tbl-contrasts
#| tbl-cap: "Paired race-level bootstrap of the ROI difference, declared rule minus comparator, on the races both arms bet. B = 2000, seed 42. Differences are in percentage points."
p6_contrast_table(p6_test_contrasts) |> kable(align = "lllrrllr")
```

Seven of the eight intervals contain zero. The one that does not is the
declared place rule against the market-only control:
`r p6_fmt_pts(c_plc_s4$diff_point)` points,
`r p6_fmt_pts_ci(c_plc_s4$ci_lo, c_plc_s4$ci_hi)`, over
`r p6_comma(c_plc_s4$n_races)` common races. Against the incumbent the same
rule is `r p6_fmt_pts(c_plc_s1$diff_point)` points,
`r p6_fmt_pts_ci(c_plc_s1$ci_lo, c_plc_s1$ci_hi)` — indistinguishable, which
is what it should be, since the declared place rule is the incumbent's own
selection with a different stake.

**No declared rule beats the rule it was declared to beat.** All three point
differences against S1/K0 are negative. The search found nothing.

## The control {#sec-control}

S4/K0 backs the market favourite in every test race at a flat stake. It reads
no model output. It is the paper's central comparison, not a footnote: if a
rule containing no model does as well as the declared rule, the model adds
nothing to bet selection.

On the win market it does better than every model rule tested.

| | ROI | 90% interval | Bets |
|---|---|---|---|
| S4/K0, market favourite | `r p6_fmt_pct(t_mkt_win$roi)` | `r p6_fmt_pct_ci(t_mkt_win$ci_lo, t_mkt_win$ci_hi)` | `r p6_comma(t_mkt_win$n_bets)` |
| S1/K0, paper 5's incumbent | `r p6_fmt_pct(t_inc_win$roi)` | `r p6_fmt_pct_ci(t_inc_win$ci_lo, t_inc_win$ci_hi)` | `r p6_comma(t_inc_win$n_bets)` |
| declared, S3b/K2 | `r p6_fmt_pct(t_dec_win$roi)` | `r p6_fmt_pct_ci(t_dec_win$ci_lo, t_dec_win$ci_hi)` | `r p6_comma(t_dec_win$n_bets)` |

The favourite rule's interval excludes zero and the incumbent's does not, but
the two are not directly contrasted here — the declared rule is what the
design put against the control, and that contrast,
`r p6_fmt_pts(c_win_s4$diff_point)` points
`r p6_fmt_pts_ci(c_win_s4$ci_lo, c_win_s4$ci_hi)`, contains zero on
`r p6_comma(c_win_s4$n_races)` common races. What can be said is the weaker
and still uncomfortable statement: **on the win market, nothing the model
contributes to bet selection is visible against a rule that ignores it.**

On each-way at the corrected terms the same pattern holds, more starkly: the
favourite returns `r p6_fmt_pct(t_mkt_ewc$roi)` against the incumbent's
`r p6_fmt_pct(t_inc_ewc$roi)`.

The place market is the exception, and it goes the other way: the favourite
returns `r p6_fmt_pct(t_mkt_plc$roi)` against the incumbent's
`r p6_fmt_pct(t_inc_plc$roi)`. Backing favourites to place is a poor bet
because a synthetic place market built from the starting-price book prices
short horses hardest — the favourite's place probability is high, so its
place price is short, and the margin sits on top of a small number. That is a
property of the place market's shape, not evidence about the model.

## Margin paid, and horses picked {#sec-margin}

A rule can improve its ROI two ways: by picking better horses, or by betting
where the book takes less. The second is not skill, and separating them is
what the fair-book column is for.

```{r}
#| label: tbl-margin
#| tbl-cap: "The margin decomposition. `Margin paid` is the fair-book ROI less the real-SP ROI: the share of stake the over-round takes on the horses that arm actually backs."
p6_margin_table(p6_margin_decomposition) |> kable(align = "llrrrrrr")
```

At a zero-margin book the win-market picture inverts. The incumbent returns
`r p6_fmt_pct(t_inc_win$roi_fair)` and the favourite rule
`r p6_fmt_pct(t_mkt_win$roi_fair)` — both positive, the favourite rule by
more. A positive fair-book return on favourites is the favourite–longshot
bias, long documented and not a finding of this paper [@snowberg2010]:
favourites win more often than the renormalised book implies, and longshots
less. It is also not exploitable here, because the margin removed to produce
that number is the bookmaker's and there is no counterparty offering it.

**The declared win rule improves nothing but the margin it pays.** It pays
`r p6_fmt_pct(t_dec_win$roi_fair - t_dec_win$roi)`, against the incumbent's
`r p6_fmt_pct(t_inc_win$roi_fair - t_inc_win$roi)`, because a price cut
concentrates the bet in smaller fields and shorter prices where the book is
thinner. Its fair-book ROI, `r p6_fmt_pct(t_dec_win$roi_fair)`, is the worst
of the three win arms. Label it correctly: the rule reduced the margin it
paid and picked worse horses, and the two roughly cancel.

## Figures

```{r}
#| label: fig-cumulative
#| fig-cap: "Cumulative profit in stake units on the test split, in race-date order. Arms staking different amounts are not comparable in level; the shape is the point."
#| fig-height: 7
p6_fig_cumulative
```

The losses accumulate steadily. No arm's result rests on a small number of
outcomes: dropping the single largest winner moves every ROI in
@tbl-test by between `r sprintf("%.1f", 100 * min(drop_move))` and
`r sprintf("%.1f", 100 * max(drop_move))` points, the larger moves belonging
entirely to the arms with a few dozen bets.

```{r}
#| label: fig-train-vs-test
#| fig-cap: "The win market. Twenty-seven training-split ROIs as a rug, against the five arms that reached the test split with their 90% intervals. The training axis is in sample; see @sec-insample."
#| fig-height: 3.2
p6_fig_train_vs_test
```

@fig-train-vs-test is the honest summary of the search: a wide spread of
in-sample returns, none of which located anything that survived.
